### 2021-10-08 Output/Figures/E4/coverage_posteriormean.tex
E6 on dcc

% latex table generated in R 4.0.2 by xtable 1.8-4 package
% Wed Oct 06 16:48:58 2021
\begin{table}[ht]
\centering
\begin{tabular}{ccccc}
  \toprule
 & var & cover & mean\_overall & mean\_sd \\ 
  \midrule
1 & beta & 0.895 & 0.0025 & 0.000255 \\ 
  2 & gamma & 0.902 &    1 & 0.142 \\ 
  3 & R0 & 0.910 & 2.54 & 0.194 \\ 
   \bottomrule
\end{tabular}
\caption{Nomial coverage rate and distribution of posterior means among 2000 simulations. The true value of the parameters are (beta, gamma, R_0) = (0.0025, 1, 2.5)} 
\label{tab:cov}
\end{table}

### 2021-10-08 Experiment_output.R

@@ -27,7 +27,7 @@ for(i in 3 : 200) {
   load(paste0("Output/RDATA/E4_", i, ".RDATA"))
   results <- rbind(results, output_E4)
 }
-summary_4 <- experiment_4_output_analysis(results, path = "Output/Figures/E4")
+summary_E4 <- experiment_4_output_analysis(results, path = "Output/Figures/E4")
 
 load("Output/RDATA/E5.RDATA")
 summary_5 <- experiment_5_output(output_E5, path = "Output/Figures/E5")

### 2021-10-08 Experiments.R

@@ -4,8 +4,8 @@
 library(PDSIR)
 run_E1 <- FALSE
 run_E3 <- FALSE
-run_E4 <- TRUE
-run_E5 <- TRUE
+run_E5 <- FALSE
+run_E6 <- TRUE
 
 #
 # Experiments
@@ -32,21 +32,25 @@ if(run_E3){ # Rho
 }
 
 
-
-if(run_E4){ # coverage
-  set.seed(2)
-  output_E4 <- experiment_4_coverage(
-    N = 50e3, S0s = 1e3, R0s = 2, m = 200
-  )
-  save(output_E4, file = "Output/RDATA/E4.RDATA")
-  print(paste0("E4 is done - ", Sys.time()))
+if(run_E5){ # Ebola
+  set.seed(1)
+  output_E5 <- experiment_5_ebola(N = 1e6, rho = 0.1, thin = 10)
+  save(output_E5, file = "Output/RDATA/E5.RDATA")
+  print(paste0("E5 is done - ", Sys.time()))
 }
 
 
-if(run_E5){ # Ebola
-  set.seed(2)
-  output_E5 <- experiment_5_ebola(N = 1e6, rho = 0.1)
-  save(output_E5, file = "Output/RDATA/E5.RDATA")
-  print(paste0("E5 is done - ", Sys.time()))
+if(run_E6){ # Ebola
+
+  set.seed(1)
+  output_E6_single <- experiment_6_single_site_update(N = 1e6, save_fig = FALSE)
+  save(output_E6_single, file = "Output/RDATA/E6_single.RDATA")
+
+  set.seed(1)
+  output_E6_joint  <- experiment_1_proof_of_concept(N = 1e6, save_fig = FALSE)
+  save(output_E6_single, file = "Output/RDATA/E6_joint.RDATA" )
+
+  print(paste0("E6 is done - ", Sys.time()))
+
 }
 

### 2021-10-08 Experiments_array.R

@@ -4,8 +4,7 @@ run_E4 <- TRUE
 if(run_E4){ # coverage
   a <- as.integer(Sys.getenv("SLURM_ARRAY_TASK_ID"))
   output_E4 <- experiment_4_coverage(
-    N = 1e5, m = 10, thin = 1, a = a,
-    S0 = 1e3, theta = list(R0 = 2.5, gamma = 1), rho = 1/10
+    N = 1e5, m = 10, thin = 1, a = a
   )
   save(output_E4, file = paste0("Output/RDATA/E4_", a, ".RDATA"))
   print(paste0("E4 array ", a, " is done - ", Sys.time()))

### 2021-10-08 run_experiments.sh
fix .sh file for E6

@@ -1,8 +1,8 @@
 #!/bin/bash
-#SBATCH --output=DCC/experiment.out
-#SBATCH --error=DCC/experiment.err
-#SBATCH --mem=1G # GB RAM
-#SBATCH --partition=scavenger#
+#SBATCH -o slurm/experiment.out
+#SBATCH -e slurm/experiment.err
+#SBATCH --mem=2G
+#SBATCH -p scavenger
 hostname # print hostname
 module load R
 Export R_LIBS_USER = ~/R/x86_64-pc-linux-gnu-library/4.1

### 2021-10-08 Experiments.R
E6 thin

@@ -43,11 +43,11 @@ if(run_E5){ # Ebola
 if(run_E6){ # Ebola
 
   set.seed(1)
-  output_E6_single <- experiment_6_single_site_update(N = 1e6, save_fig = FALSE)
+  output_E6_single <- experiment_6_single_site_update(N = 1e6, save_fig = FALSE, thin = 10)
   save(output_E6_single, file = "Output/RDATA/E6_single.RDATA")
 
   set.seed(1)
-  output_E6_joint  <- experiment_1_proof_of_concept(N = 1e6, save_fig = FALSE)
+  output_E6_joint  <- experiment_1_proof_of_concept(N = 1e6, save_fig = FALSE, thin = 10)
   save(output_E6_single, file = "Output/RDATA/E6_joint.RDATA" )
 
   print(paste0("E6 is done - ", Sys.time()))

### 2021-10-08 Experiments.R
E6 rho

@@ -47,8 +47,8 @@ if(run_E6){ # Ebola
   save(output_E6_single, file = "Output/RDATA/E6_single.RDATA")
 
   set.seed(1)
-  output_E6_joint  <- experiment_1_proof_of_concept(N = 1e6, save_fig = FALSE, thin = 10)
-  save(output_E6_single, file = "Output/RDATA/E6_joint.RDATA" )
+  output_E6_joint  <- experiment_1_proof_of_concept(N = 1e6, save_fig = FALSE, thin = 10, rho = 1/10)
+  save(output_E6_joint, file = "Output/RDATA/E6_joint.RDATA" )
 
   print(paste0("E6 is done - ", Sys.time()))
 

### 2021-10-11 Manuscripts/Article/Article.tex
manuscript repo

%\documentclass{article}
%\usepackage[utf8]{inputenc}
%\usepackage[round,authoryear]{natbib}

\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{url,hyperref}
\usepackage{graphicx}
\usepackage{amsmath,amssymb,array,eucal, amsthm}
\linespread{1.5}
\setlength\parindent{35pt}
\input{macros}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{color}

\graphicspath{ {../Experiment Output/Figures/} }

\newcommand{\jx}[1]{{\color{blue}{ #1}}}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}{Proposition}[section]
\newtheorem{lemma}{Lemma}[section]

\begin{document}

	% Keywords command
	\providecommand{\keywords}[1]
	{
		\small	
		\textbf{\textit{Keywords---}} #1
	}

	\title{%Efficient Likelihood-Based Inference for Fitting Stochastic Epidemic Models to Incidence Data via Data Augmentation
	A Uniformly Ergodic Data-Augmented MCMC for Fitting Stochastic Epidemic Models to Incidence Data
	}
	\author{Rapha\"{e}l Morsomme$^{1}$\footnote{raphael.morsomme@duke.edu} \ and Jason Xu$^{1}$ \\
	\small $^{1}$Department of Statistical Science, Duke University \\
	}
	\date{\today}	
	\maketitle

	\begin{abstract}
		Stochastic epidemic models provide a realistic an interpretable description of the spread of a disease through a population. Yet, fitting these models in missing data settings is a notably difficult task; in particular, when the epidemic process is only partially observed, the likelihood of the model is intractable. To remedy this issue, this article introduces a data-augmented MCMC algorithm for fast and exact inference for the stochastic SIR model given discretely observed infection incidence counts. In a Metropolis-Hastings step, the algorithm jointly proposes event times of the augmented data according to a stochastic process whose dynamics closely resemble those of the SIR process, and from which we can efficiently generate an epidemic that is compatible with the observed data. Not only is the algorithm fast, but, since the augmented data are generated from a very faithful approximation of the target epidemic model, the algorithm can update a large portion of the augmented data per iteration while maintaining a relatively high acceptance rate, thereby exploring the high-dimensional latent space efficiently. Moreover, this MCMC algorithm is to be uniformly ergodic. While existing MCMC approaches become intractable for populations greater than a few thousand individuals, unless they rely on model simplifications or approximations, the proposed algorithm enables exact inference and scales to outbreaks in populations of hundred-thousands individuals
		+ even on a single laptop. We validate its performance via thorough simulation experiments, and a case study on the $2014$ Ebola outbreak in Western Africa. % in Gu\'eck\'edou, Guinea. %, in a population of $150,000$ individuals. 
	\end{abstract}
	\keywords{Stochastic epidemic model; incidence data; exact posterior inference; data-augmented MCMC; likelihood-based methods; uniform ergodicity}

	\section{Introduction}
	% Outline: SEM; intractable likelihood; existing methods: approximation (Cauchemez 2008, Fintzi 2020), particle filtering, ABC, data augmentation; poor mixing;

	The efficient control of a disease outbreak requires an understanding of the mechanisms underlying its spread. Mechanistic compartmental models, which describe the transition of individuals between various states, have a long mathematical modeling tradition in epidemiology \cite{Kermack.1927}. Due to their interpretability, they are commonly used to describe the dynamics of an outbreak and typically serve as the main source of information for predicting the course of an outbreak and identifying interventions that could be effective. Originally, deterministic versions of the models were employed by mathematicians and epidemiologists. These models are simple to analyze, but fail to capture the inherent randomness in the spread of a disease. For instance, they cannot be used to estimate the probability of a large-scale outbreak or its expected duration, and do not allow for uncertainty quantification when used within inferential procedures. Stochastic epidemic models (SEM), on the other hand, incorporate the random nature of infections and recoveries and therefore provide more realistic descriptions of the spread of a disease, and in turn more reliable inference from observed data.

	Conducting inference on SEMs is, however, a notably difficult task. The difficulty stems from the fact that the observed data typically provide incomplete information on a process that evolves continuously through time. This makes the likelihood of the model intractable; were the infection and recovery times of each individual of the population observed, conducting inference on SEMs would be straightforward, but this is rarely the case. In practice, one often has access to either incidence data such as weekly counts of new infections, or prevalence data such as the numbers of people infectious at discrete times. The marginal likelihood of these partially observed data becomes a bottleneck, requiring a large integration step that accounts for all possible configurations of the missing data. In particular, direct computation of this likelihood requires the transition probabilities between observation times for which no closed form are available. Numerical methods to obtain transition probabilities have only recently been developed in \cite{Ho.2018b, Ho.2018}, but the high computational costs of their algorithm makes the approach applicable to small populations only.

	In this article, we introduce a data-augmented MCMC algorithm to estimate the parameters of the stochastic SIR model, a commonly used SEM, from discretely observed infection incidence counts. Rather than marginalizing directly, our approach accounts for the latent space via sampling. The algorithm updates the event times of the augmented data by jointly proposing infection and removal times from a stochastic process whose dynamics closely resemble those of the SIR process, and from which we can efficiently generate an epidemic that is compatible with the observed data. This surrogate process serves as an efficient proposal scheme within a Metropolis-Hastings algorithm. Since the latent spaces are generated from a very faithful approximation of the target epidemic model, the algorithm can update a large portion of the event times per iteration while maintaining a relatively high acceptance rate, thereby exploring the high-dimensional latent space efficiently.

	The remainder of the article is structured as follows. Section \ref{sec:set} provides background information on the inference task. It introduces the stochastic SIR model and explains why conducting inference under partially observed data is a difficult task. Previous works addressing this problem are also presented. Section \ref{sec:pds} describes the data-augmented MCMC algorithm proposed in this article and assesses how closely it approximation the SIR process. Section \ref{sec:sim} presents the results of a simulation study examining the performance of the algorithm, while Section \ref{sec:ebo} describes an analysis of the $2014$ Ebola outbreak in Gu\'eck\'edou, Guinea. Finally, Section \ref{sec:dis} discusses the findings and concludes the article.

	\section{Background and Prior Work}
	\label{sec:set}
	%Outline: compartmental model; CTMC; rates; likelihood; inference (MLE, gamma conjugacy)

	\subsection{The Stochastic SIR Model}
	\label{sec:sir}
	%Outline: stochastic SIR; X; parameters; transition rate12

	Our point of departure is to consider the stochastic SIR model -- also referred to as the general epidemic model -- which offers a parsimonious and interpretable representation of the mechanistic, population-level dynamics of an epidemic. The stochastic SIR model is a compartmental model in which the individuals of a population transition through three types or compartments: susceptible (S), infectious (I) and removed (R). The only possible moves are from S to I (infections) and from I to R (removals). A susceptible individual becomes infected through contact with an infectious individual. Once infected, she is immediately infectious and remains so for some period of time after which she is removed from the process without the possibility of reinfection. In this formulation, demographics dynamics such as births and deaths of individuals are ignored since they usually occur at a much slower rate than infections and removals.

	Assuming a closed population with a fixed size $n$, the stochastic SIR model consists of a continuous-time vector-valued process
	\begin{equation}
		\label{eq:X}
		\X = \left\lbrace \X(t), t>0\right\rbrace \in \chi_{\X}
	\end{equation}
	with
	\begin{equation}
		\X(t) = (X_1(t), \dots, X_n(t)) \in \{S, I, R\}^n
	\end{equation}
	where the agent-level subprocess
	$$ X_i(t) = 
	\begin{cases}
		S, & t \in [0, \tau^T_i] \\
		I, & t \in (\tau^T_i, \tau^J_i] \\
		R, & t \in (\tau^J_i, \infty)
	\end{cases}
	,\quad \iton
	$$
	denotes the status (compartment) of individual $i$ at time $t$ with $\tau^T_i$ and $\tau^J_i$ respectively the infection and removal times of individual $i$. If individual $i$ never becomes infectious, then we set $\tau^T_i = \tau^J_i = \infty$ and $X_i(t) = S$ for $t \in [0, \infty)$. The set $\chi_{\X}$ consists of the trajectories compatible with the evolution of a disease, that is, the set of trajectories in which no infection occurs whenever the infectious compartment is depleted :
	\begin{equation}
		\label{eq:chi}
		\chi_{\X} = \{X:X(t) \in \{S,R\}^n \Rightarrow X(t+s) = X(t), \forall s>0 \}.
	\end{equation}

	The stochastic SIR model is specified by the rates at which individuals move from one compartment to another. If we assume a homogeneously mixing population where contacts between individuals occur independently at some rate constant $\beta$, then the contacts between two given individuals are said to follow a Poisson process with rate $\beta$. This parameter can be interpreted as the \textit{infection rate}: when a susceptible individual comes into contact with an infectious individual, she immediately becomes infectious.
	If we also make the common assumption that the infectious periods follow independent exponential distributions with rate $\gamma$, then the process \ref{eq:X} is a time–homogeneous continuous time Markov chain whose instantaneous transition rates 
	%from state $\x$ to $\x'$
	are given by the matrix $\Lambda = [\lambda_{\x, \x'}]$ with
	$$
	\lambda_{\x, \x'} = 
	\begin{cases}
		\beta I(t), & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = S \text{ and } x_j' = I, \\
		\gamma, & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = I \text{ and } x_j' = R, \\
		0, & \text{ otherwise.}
	\end{cases}
	$$
	where $I(t) = \sum_{i=1}^n \mathbf{1} \{x_i(t) = I\}$ is the total number of infectious individuals at time $t$. That is, the individual-level infection and removal rate at time $t$ are respectively $\beta I(t)$ and $\gamma$.
	%Equivalently \jx{The following is not an exact equality, but needs the $o(dt)$ term or similar; also uncapitalize X on the right}
%	$$
%	P(\X_t = \x, \X_{t+dt} = \x') =
%	\begin{cases}
%		\beta I(t) dt + o(dt), & \X \text{ and } \X' \text{ only differ at position } j \text{ with } x_j = S \text{ and } x_j' = I, \\
%		\gamma dt + o(dt), & \X \text{ and } \X' \text{ only differ at position } j \text{ with } x_j = I \text{ and } x_j' = R, \\
%		o(dt), & \text{otherwise}
%	\end{cases}
%	$$
%	where $dt>0$ is small.

	\subsection{Inference with Complete Data}
	\label{sec:icd}
	%Outline: complete-data likelihood; MLE; gamma conjugacy

	Assuming that the Markov process \ref{eq:X} is completely observed until time $t_{end}$, we obtain the following complete data likelihood
	\begin{align}
		L(\theta; \X)
		& = \prod_{T} \beta I(\tau^T_j) \prod_{J} \gamma \exp\left\lbrace - \int_{0}^{t_{end}}\beta I(t)S(t) + \gamma I(t) dt \right\rbrace  \nonumber \\
		\label{eq:cdl}
		& = \beta^{n_T} \gamma^{n_J}\prod_{J} I(\tau^T_j) \exp\left\lbrace - \int_{0}^{t_{end}}\beta I(t)S(t) + \gamma I(t) dt \right\rbrace
	\end{align}
	where $\theta = (\beta, \gamma) \in \chi_{\theta}$  are the parameters of the model and $\chi_{\theta} = (0, \infty) \times (0, \infty)$ is the parameter space, 
	$T = \{i:\tau^T_i \le t_{end}\}$ and $J = \{i:\tau^J_i \le t_{end}\}$ are the respective index sets of individual infected and removed by time $t_{end}$, 
	$n_T = |T|$ and $n_J = |J|$ are the respective numbers of observed infections and removals,
	and	$S(t) = \sum_{i=1}^n \mathbf{1}\{X_i(t) = S\}$ is the number of susceptible individuals at time $t$.

	We see that in this fully observed scenario, it is straightforward to conduct inference. The likelihood \ref{eq:cdl} belongs to the exponential family and the maximum likelihood estimates can be expressed in terms of the sufficient statistics defined above as
	$$\hat{\beta}_{MLE} = \dfrac{n_T}{ \int_{0}^{t_{end}} I(t)S(t)dt}, \quad \hat{\gamma}_{MLE} = \dfrac{n_J}{\int_{0}^{t_{end}} I(t)dt}.$$
	Since the functions $I$ and $S$ are constant between event times, the two integrals correspond to finite sums which are straightforward to compute.
	Furthermore, in a Bayesian context, inference is facilitated by the conjugacy of the gamma distribution: if we put independent gamma prior distributions on the parameters
		\begin{equation}
		\label{eq:pri}
		\beta \sim Ga(a_{\beta}, b_{\beta}), \qquad \gamma \sim Ga(a_{\gamma}, b_{\gamma})
	\end{equation}
	where $Ga(a,b)$ denotes the gamma distribution with mean $a/b$ and variance $a/b^2$, then the posterior distributions of the parameters are independent gamma distributions
	\begin{equation}
		\label{eq:pdb}
		\beta | \X \sim Ga\left( a_{\beta} + n_T, b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt\right)
	\end{equation}
	and
	\begin{equation}
		\label{eq:pdg}
		\gamma | \X \sim Ga\left( a_{\gamma} + n_J, b_{\gamma} + \int_{0}^{t_{end}} I(t) dt\right)
	\end{equation}
	from which we can easily generate independent values with a Monte Carlo sampler to explore the posterior distribution $\pi(\theta|\X)$.

	\subsection{Inference with Incomplete Data}
	\label{sec:iid}
	%Outline: observed data; intractable posterior; DAMCMC; parameters update; latent space update

	(New in this section: I am motivating the type of observed data  $T_{1:K}$ that we consider.)

	In practice, inference is complicated by the fact that the Markov process \ref{eq:X} is only partially observed. That is, we do not observe all epidemic events, and do not have access to the sufficient statistics needed to evaluate the likelihood \ref{eq:cdl}. Various types of partially observed data have been considered in the literature such as observing only the removal times \cite{Gibson.1998, ONeill.1999} and knowing the number of infectious individuals in the population at discrete points in time \cite{Fintzi.2017}. The former type of data for instance arise in animal experiment in which an animal that tests positive for the disease is immediately isolated from the rest of the population and therefore no longer contributes to the spread of the disease. In such case, the removal times are exactly known.

	In this article, we consider another type of partially observed data: incidence counts for the infections at discrete points in time. Such data arise when an infectious individual may continue to infect others even after being tested positive for the disease. This was for instance the case during the $2013$–$2016$ outbreak of Ebola virus. This pandemic was unprecedented in scale, with more than $10,000$ reported deaths, mainly in three Western African countries (Guinea, Liberia and Sierra Leone). This outbreak was characterized by the large number of infections that occurred after individuals were tested positive for the virus. Numerous infections happened at the hospitals where infected individuals received a treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased person during funerals is a tradition. Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.}. 
	It therefore seems appropriate to model a positive test for the virus as an indication that the individual became infectious some time prior to the test rather than as a removal time.

	Consider an observation schedule $t_{0:K}$ ($K \ge 1$) with $0 = t_0 < t_1 < \dots < t_K = t_{end}$. The observed data consist of the $K$-dimensional vector $\Y = T_{1:K}$ where $T_k = \sum_{i=1}^n \mathbf{1}\{\tau^T_i \in (t_{k-1}, t_k]\}$ is the number of infections during the $k^{\text{th}}$ time interval.
	For simplicity, we further assume that the initial configuration of the population $\X(0)$ is known.

	If we adopt a Bayesian approach, the posterior distribution of the parameters given the observed data is
	\begin{align}
		\label{eq:pdl}
		\pi(\theta|\Y) 
		& \propto L(\theta; \Y)\pi(\theta) \nonumber \\
		& = \int_{\chi^*_{\X}} L(\theta; \x) \pi(\theta) d\x
	\end{align}
	where $\pi(\theta)$ is the prior distribution on the parameters $\theta$ and
	$\chi^*_{\X} \subset \chi_{\X}$ is the set of trajectories of the process \ref{eq:X} that are compatible with the observed data $\Y$.
	The partial data likelihood $L(\theta; \Y)$ therefore consists of a high-dimensional integral over all epidemic paths compatible with the observed data and has no closed form solution available, even for a population of moderate size.

	\subsection{Prior Work}
	\label{sec:pre}
	Researchers have explored several approaches for conducting inference on partially observed stochastic epidemic processes. Early approaches include the use of martingale-based equations \cite{Becker.1977, Watson.1981, Sudbury.1985}. However, such methods are difficult to apply to dynamic models with partially observed data and are therefore not suitable to the stochastic SIR model with incomplete data.
	More recently, researchers have based their computation on a simpler process that approximates the model's dynamics and whose likelihood in the presence of partially observed data is tractable. These approximation models include binomial chains models \cite{Greenwood.1931, Abbey.1952}, diffusion processes \cite{Cauchemez.2008, Fintzi.2020} and Gaussian processes \cite{Jandarov.2014}. While these approximation-based approaches bypass the intractability of the partial data likelihood \ref{eq:pdl}, the assumptions on which they are based are questionable when the population is small, and, as a result, the dynamic of the stochastic process differs from its asymptotic behavior.

	Since the popularization of Markov chain Monte Carlo methods (MCMC) in the field of statistics \cite{Tanner.1987, Gelfand.1990, Tierney.1994}, researchers have developed sampling-based methods to directly work with the SIR likelihood instead of an approximation thereof. MCMC algorithms fall into two categories: model-based forward simulation and data augmented MCMC (DA-MCMC).
	Particle filtering \cite{King.2015} is an example of the former category that is extremely popular among practitioners. Its plug-and-play feature makes it applicable to a wide variety of models. However, model-based forward simulation suffers from two drawbacks: simulating data from a model as complex as the SEM that is compatible with the observed data is prohibitively slow, and these methods can fail to converge when the model does not fit the data well. Approximate Bayesian Computation (ABC) \cite{McKinley.2018} offers a solution to the latter problem, but its inference is based on an approximation of the model's likelihood. As a result, the inference is inevitably biased, and it is difficult to quantify how closely the approximate posterior resembles the original target distribution.

	The second family of MCMC-based inferential methods treats the unobserved event times as nuisance parameters; that is, the observed data are augmented with latent data that consist of the times and types of unobserved epidemic events. Researchers have developed algorithms to explore this latent space efficiently. Early attempts employed reversible-jump MCMC \cite{Green.1995} to explore models with different numbers of unobserved events \cite{Gibson.1998, ONeill.1999}. Their algorithm augments the observed data which consist of the recovery times with the unobserved infection times and explores the latent space by inserting, deleting or uniformly moving an infection time in each iteration of the MCMC algorithm. 
	More recently, Fintzi et al. \cite{Fintzi.2017} proposed a MCMC algorithm to conduct inference with discretely observed infection counts prevalence counts. They constructed latent data consisting of the infection and recovery times of each individual. The algorithm explores this latent space by updating the event times of an individual in each iteration of the algorithm according to the exact dynamics of the stochastic process. These DA-MCMC methods suffer from poor mixing in the presence of large epidemics. Since they update a single element of the latent data per iteration, the resulting Markov chains fail to explore the latent space efficiently. Although the update of multiple event times per iteration in \cite{Pooley.2015} and the non-centered parameterization in \cite{Neal.2005} improves mixing, the gains are modest and the Markov chains still suffer from a high level of auto-correlation. %Furthermore, generating event times from the exact dynamics of the SEM in \cite{Fintzi.2017} is computationally costly.

	%An advantage of DA-MCMC methods is the possibility to conduct inference in more complex models. For instance, Streftaris and Renshaw \cite{Streftaris.2002} generalized the approach of \cite{Gibson.1998} to a non-Markovian stochastic SIR model where the amount of time spent infectious is not necessarily exponentially distributed but follows an arbitrary distribution, and Bu et al. \cite{Bu.2020} analyzed models in which the population does not mix homogeneously but is characterized by a dynamic social network.

	\subsection{Proposed Data-Augmented Approach}
	\label{sec:con}

	TODO: Add short intro to not intimidating reader with notation.

	As shown in Section \ref{sec:icd}, the complete data likelihood \ref{eq:cdl} is amenable to computation. This suggests augmenting the observed data $\Y$ with latent data $\Z$ such that the likelihood $L(\theta; (\Y, \Z))$ has the closed form \ref{eq:cdl} and constructing a Markov chain on the space $\chi = \chi_{\theta} \times \chi_{\Z}$ whose stationary distribution is the joint posterior distribution $\pi(\theta, \X|\Y)$ \cite{Gibson.1998, ONeill.1999, Fintzi.2017}. We consider the latent data $\Z = \left\lbrace (z^T_i, z^J_i)\right\rbrace_{i=1}^{n} \in \chi_{\Z}$ which consist of the times and types of unobserved epidemic events, where the infection time of individual $i$ is
	$$z^T_i \begin{cases}
		\in [0, t_{end}) & \text{if individual } i \text{ becomes infectious before } t_{end} \\
		= \infty & \text{if individual } i \text{ does not become infectious before } t_{end} \\
	\end{cases}$$
	and the removal time of individual $i$ is
	$$z^J_i \begin{cases}
		\in (z_{T_i}, t_{end}] & \text{if individual } i \text{ is removed before } t_{end} \\
		= \infty & \text{if } z_{T_i} = \infty \text{ or individual } i \text{ is removed after } t_{end}. \\
	\end{cases}$$

    The set $\chi_{\Z} \subset (\bar{\R}^+ \times \bar{\R}^+)^n$ consists of the latent data $\Z$ compatible with the observed data $\Y$ and the progression of an epidemic (see Equation \ref{eq:chi}). Since $\Z$ contains all the information present in $\Y$, we can write
	$$L(\theta; (\Y, \Z)) = L(\theta; \Z).$$

	The data-augmented MCMC algorithm that we propose alternates between updates of the parameters $\theta$ given the current state of the latent data $\Z$, and updates of the latent data given the current state of the parameters and the observed data $\Y$. A one-step transition from $\x_0 = (\theta_0, \z_0)$ to $\x_1 = (\theta_1, \z_1)$ therefore looks like
	$$\x_0 = (\theta_0, \z_0) \rightarrow (\theta_1, \z_0) \rightarrow (\theta_1, \z_1) = \x_1.$$

	The transition kernel $P$ of the Markov chain $\{\x_i\}_i$ is thus a composition of two kernels $P_{\theta}$ and $P_{\z}$ which respectively update $\theta$ while keeping the $\z$ fixed, and update $\z$ while keeping $\theta$ fixed. Due to the gamma conjugacy of the complete-data likelihood mentioned in Section \ref{sec:icd}, sampling new parameter values poses no challenge. We can simply employ a Gibbs sampler to directly draw new values for $\theta = (\beta, \gamma)$ from the two independent full conditional posterior distributions given by Equations \ref{eq:pdb} and \ref{eq:pdg}. The kernel 
	$$
	P_{\theta}((\theta_0, \z), (\theta_1, \z)) = Ga\left( \beta_1; a_{\beta} + n_T, b_{\beta} + \int_0^{t_{end}} S(t)I(t) dt\right) Ga\left( \gamma_1; a_{\gamma} + n_J, b_{\gamma} + \int_0^{t_{end}} I(t) dt\right) d\beta_1 d\gamma_1
	$$
	therefore corresponds to a Gibbs kernel where $n_T, n_J, \int S(t)I(t) dt, \int I(t) dt$ are sufficient statistics from the latent data $\z$. The kernel
	$$
	P_{\theta}((\theta, \z_0), (\theta, \z_1)) = q(\z_1|\theta) \alpha((\theta, \z_0), (\theta, \z_1)) dz_1
	$$
	corresponds to a Metropolis-Hastings kernel where the proposal density $q$ will be defined in the following section and
	$$
	\alpha\left( \left( \theta^{(m)}, \Z^{(m-1)}\right) , \left( \theta^{(m)}, \Z^{(m)}\right) \right) = \min\left\lbrace 1, \dfrac{L\left( \theta^{(m)}; \Z^{(m)}\right) q\left( \Z^{(m - 1)}|\theta^{(m)}\right) }{L\left( \theta^{(m)};\Z^{(m - 1)}\right) q\left( \Z^{(m)}|\theta^{(m)}\right) }\right\rbrace
	$$
	corresponds to the Metropolis-Hasting acceptance ratio \cite{Tierney.1994}.	
	Algorithm \ref{alg:DA-MCMC} provides the details of the procedure.

		\begin{algorithm}
		\caption{Data-Augmented MCMC}
		\label{alg:DA-MCMC}
		\begin{algorithmic}
			\REQUIRE $\theta^{(1)}$
			\STATE $\Z^{(1)} \leftarrow draw(\Z|\theta^{(1)})$ (generate initial latent space)
			\FOR {$j = 2, \dots, N$}			
			\STATE $\beta^{(j)}|Z^{(j - 1)} \sim Ga\left( a_{\beta} + n_T^{(j - 1)}, b_{\beta} + \int_{0}^{t_{end}} I^{(j - 1)}(t)S^{(j - 1)}(t)dt\right)$ (Gibbs update)
			\STATE $\gamma^{(j)}|Z^{(j-1)} \sim Ga\left( a_{\gamma} + n_J^{(j - 1)}, b_{\gamma} + \int_{0}^{t_{end}} I^{(j - 1)}(t) dt\right)$ (Gibbs update)
			\STATE $\theta^{(j)} \leftarrow (\beta^{(j)}, \gamma^{(j)})$
			\STATE $\Z^* \leftarrow draw(\Z|\theta^{(j)})$ (generate latent data from the PD-SIR process)
			\STATE $\alpha = \min\left\lbrace 1, \dfrac{L(\theta^{(j)}; \Z^*)q(\Z^{(j - 1)}|\theta^{(j)})}{L(\theta; \Z^{(j - 1)})q(\Z^*|\theta^{(j)})}\right\rbrace  $
			\STATE $u \sim U(0,1)$
			\IF{$u<\alpha$}
			\STATE $\Z^{(j)} \leftarrow \Z^*$					
			\ELSE
			\STATE $\Z^{(j)} \leftarrow \Z^{(j - 1)}$				
			\ENDIF
			\ENDFOR
		\end{algorithmic}
	\end{algorithm}

	\section{Efficient Proposal Process for Latent Data}
	Simulating trajectories from discrete state Markov processes, such as the stochastic SIR process, that are compatible with partially observed data is, in general, a notoriously difficult task \cite{Hobolth.2009}. Instead of performing this conditional simulation from the SIR model directly, we will conditionally simulate from a surrogate process and then correct the approximation via a Metropolis-Hastings step.

	\subsection{The Piecewise-decoupled SIR Process}
	\label{sec:pds}
	% Outline: define PD-SIR; generate PD-SIR
	To generate the latent data, we consider a stochastic process whose dynamics closely resemble those of the SIR process and from which we can efficiently generate an epidemic that is compatible with the discretely observed data $\Y = T_{1:K}$. We refer to this surrogate process as the \textit{piecewise decoupled SIR} process (PD-SIR).
	Similarly to the SIR process, the PD-SIR process corresponds to a compartmental model in which individuals move from $S$ to $I$ and from $I$ to $R$.
	The removal dynamics are the same under both processes: infection periods follow independent exponential distributions with rate $\gamma$, resulting in a constant individual-level removal rate $\gamma$. The infection dynamics, however, vary slightly.
	In the SIR process, the population-level infection rate at time $t$ is $\mu_T(t) = \beta S(t)I(t)$ and, form the perspective of a susceptible individual, the individual-level infection rate is $\mu(t) = \frac{\mu_T(t)}{S(t)} = \beta I(t)$. Note that $\mu$ varies after each event since the variable $I$ changes after each infection and removal.
	In contrast, in the PD-SIR process, $\mu$ is kept constant over small time intervals.

	Consider a resetting schedule $r_{0:L}$ ($L\ge1$) with $0 = r_0 < r_1 < \dots < r_L$. Under the PD-SIR process the individual-level infection rate $\mu$ is set to $\beta I(r_{k - 1})$ for $t \in [r_{k - 1}, r_k)$ where $I(r_{k-1})$ denotes the number of infectious individuals at the beginning of the $k^{\text{th}}$ time interval; that is, the individual infection rate is kept constant over each interval of the resetting schedule. The infection rate $\mu$ and the evolution of the variable $I$ are therefore decoupled during each interval, and the infection rate is reset at their beginning. Over a single observation interval, the PD-SIR process is equivalent to the two-type branching process proposed in \cite{Ho.2018} to approximate SIR dynamics.

	If we let the resetting schedule coincide with the observation schedule $t_{0:K}$,
	$$L = K, \qquad r_k = t_k, \quad k=1,\dots,K,$$
	then we can straightforwardly generate a realization from the PD-SIR process that is compatible with the observed epidemic data $T_{1:K}$. Under the PD-SIR model, the infections occurring during the $k^{\text{th}}$ interval follow a linear death process with rate $\mu_k = \beta I(t_{k-1})$. Generating event times from a linear death process conditionally on the number of events happening by a certain time can be done extremely efficiently as shown by the following theorem whose proof is presented in Appendix \ref{app:ldp}.

	\begin{theorem}
		\label{theo:ldp}
		Consider a linear death process with death rate $\mu$ and let $d_{1:T} \in (t_l, t_u]$ be the times of the $T$ deaths occurring between times $t_l$ and $t_u$. Then 
		$$d_i \,{\buildrel d \over =}\, X_{(i)}, \quad i = 1, \dots, T$$
		where $X_{(i)}$ is the $i^{\text{th}}$ order statistics of $T$ iid random variables following a truncated exponential distribution with rate $\mu$, lower bound $t_l$ and upper bound $t_u$.
	\end{theorem}

	We generate latent data  $\z = \left\lbrace (z^T_i, z^J_i)\right\rbrace_{i=1}^{n}$ compatible with $\Y = T_{1:K}$ from the PD-SIR process as follows. For each time interval $[t_{k-1}, t_k)$, we compute $\mu_k = \beta I(t_{k-1})$ and $U_k = I(0) + \sum_{l=1}^{k-1} T_l$, the cumulative count of infections at the beginning of interval $k$. Note that $I(t_{k-1})$ only depends on past events and can therefore be computed at each given the PD-SIR process up to time $t_{k-1}$. Algorithm \ref{alg:PD-SIR} provides a simple recursive to compute this variable (referred to as $I_k$). For $i = U_k + 1, \dots, U_k + T_k$, following Theorem \ref{theo:ldp}, we generate the infection times $z^T_i$ as the order statistics of $T_k$ iid random variables following a truncated exponential distribution with rate $\mu_k$, lower bound $t_{k-1}$ and upper bound $t_k$. For the same indices $i$, we then generate the removal times $z^J_i$ from the mixture distribution
	$$(1 - p_i) \delta_{\infty} + p_i \TruncExp(\gamma, z^T_i, t_{end}) $$
	where $\delta_{\infty}$ corresponds to the Dirac distribution with mass $1$ on the element $\infty$,
	$$p_i = 1 - \exp\{-\gamma (t_{end} - z^T_i)\}$$
	is the cumulative distribution function of an exponential distribution with rate $\gamma$ and corresponds to the probability that particle $i$ is removed before $t_{end}$ given that it was infected at time $z^T_i$, and $\TruncExp(\gamma, z^T_i, t_{end})$ denotes a truncated exponential distribution with rate $\gamma$, lower bound $z^T_i$ and upper bounde $t_{end}$. By construction, this scheme generates a latent space from the PD-SIR process that is compatible with the observed data. Algorithm \ref{alg:PD-SIR} gives an algorithmic representation of the procedure.

	\begin{algorithm}
		\caption{Generating a PD-SIR process conditionally on discretely observed infection incidence counts $T_{1:K}$}
		\label{alg:PD-SIR}
		\begin{algorithmic}
			\REQUIRE $T_{1:K}, \theta = (\beta, \gamma)$, $I(0)$
			\FOR {$i = 1, \dots, I(0)$}
			\STATE $z^T_i \leftarrow 0$ (by the memoryless property of the exponential distribution).
			\STATE $p_i \leftarrow 1 - \exp\{-\gamma (t_{end} - 0)\}$
			\STATE $z^J_i \sim (1 - p_i) \delta_\infty + p_i \TruncExp(\gamma, 0, t_{end})$
			\ENDFOR
			\STATE $I_1 \leftarrow T_1$
			\FOR {$k = 1, \dots, K$}	
			\STATE $\mu_k \leftarrow \beta I_k$
			\STATE $X_{1 : T_k} \sim \TruncExp(\mu_k, t_{k-1}, t_k)$ independently
			\FOR {$i = U_k + 1, \dots, U_k + T_k$}
			\STATE $z^T_i \leftarrow X_{(i)}$ (infection times)
			\STATE $p_i \leftarrow 1 - \exp\{-\gamma(t_{end} - z^T_i)\}$
			\STATE $z^J_i \sim (1 - p_i) \delta_\infty + p_i \TruncExp(\gamma, z^T_i, t_{end})$ (removal times)
			\ENDFOR
			\STATE $J_k \leftarrow \sum_{i=1}^n \mathbf{1}\{\tau^J_i \in (t_{k-1}, t_k]\}$ (number of removals in the $k^{\text{th}}$ interval)
			\STATE $I_{k+1} \leftarrow I_k + T_k - J_k$
			\ENDFOR
		\end{algorithmic}
	\end{algorithm}

	The density $q$ of the PD-SIR process is then
	\begin{align*}
		q(\Z|\theta) =
		& \prod_{k=1}^K \prod_{j=U_k + 1}^{U_k + T_k} \TruncExp(\tau^T_j; \mu_{k}, t_{k-1}, t_{k}) \\
		& \times \prod_{i=1}^{n} \left( 1 - p_i \right)^{\mathbf{1}(z^J_i = \infty)} \left( p_i \TruncExp(z^J_i; \gamma, z^T_i, t_{end}) \right)^{\mathbf{1}(z^J_i \le t_{end})}
	\end{align*}
	where $\mu_k = \beta I(t_{k-1})$ and
	$$\TruncExp(x; \mu, l, u) = \dfrac{\mu \exp\{-\mu x\}}{\exp\{-\mu l\} - \exp\{-\mu u\}}, \quad x \in (l, u)$$
	denotes the density of a truncated exponential distribution with rate $\mu$, lower bound $l$ and upper bound $u$ at the value $x$.

	The following characteristics of our DA-MCMC algorithm are worth noting.
	First, initializing the Markov chain only requires setting $\theta^{(1)}$ since  $\Z^{(1)}$ can be generated from the PD-SIR process conditionally on $\theta^{(1)}$ and $\Y$ alone.
	Second, since the PD-SIR closely approximates the SIR model, our DA-MCMC algorithm can update a large portion of the latent data per iteration while keeping a relatively high acceptance rate. As a result, our algorithm makes larger jumps in the high-dimensional latent space and therefore explores it more efficiently. If, however, the acceptance rate of the latent data in the Metropolis-Hastings steps is too low -- which will be shown to occur when the size of the latent space grows -- then it is possible to update the event times of only a fraction $0<\rho\le 1$ of the individuals in the population in order to increase the acceptance rate and improve mixing.
	Third, if all event times are updated ($\rho = 1$), then the current and proposed latent spaces $\Z^{(k-1)}$ and $\Z^*$ are independent conditionally on the current values of the parameters: $\Z^{(k-1)} \perp \Z^* | \theta^{(k)}$. This contrasts with existing DA-MCMC algorithms that only update a small fraction of the latent data per iteration.
	Fourth, since the latent data are generated from a process that only resembles the SIR model, we ensure that the Markov chain converges to the correct distribution $\pi(\theta, \Z|\Y)$ by proposing and accepting the latent data according to a Metropolis-Hastings scheme. This results in exact Bayesian inference targeting the posterior distribution under the original SIR model.
	Fifth, not only is proposing latent data from the RD-SIR process extremely fast since all random variables can be generated via the inverse-CDF method, but ensuring that these data are compatible with the observed data can be done at no additional cost, making the proposal scalable to large outbreaks.
	Finally, our algorithm generates a Markov chain that is uniformly ergodic, which is the strongest form of ergodicity for a Markov chain \cite{Tierney.1994}. The proof of this result is provided in Appendix \ref{app:uni}.
	% our DA-MCMC does only requires I(0), not S(0).

	\subsection{Quality of Approximating Process}
	\label{sec:qua}
	% Outline: spaghetti plot, compare distribution of a particular infection time under SIR v approximation process (plot density and compute KS distance).

	Since we use a proposal for the latent data that is independent from that at of the previous iteration, the efficiency of the DA-MCMC algorithm to explore the latent space directly depends on how similar the surrogate process is to the target process. In particular, the Markov chain will have good mixing properties if the PD-SIR process is a close approximation of the SIR process. The PD-SIR process only differs from the SIR process in its infection dynamics, the removal dynamics being identical in the two processes. Figure \ref{fig:comparison} compares the trajectories of the compartments $S$, $I$ and $R$ of a SIR process of moderate size ($S(0) = 1,000$, $I_(0) = 10$, $R(0) = 0$, $(\beta, \gamma) = (0.003, 1)$ and $t_{end} = 6$) and those of four PD-SIR processes constrained to be compatible with the infection incidence data $T_{1:K}$ for $K \in (5, 10, 50, 1,000)$ observed in the SIR process. We see that the PD-SIR is qualitatively close to the SIR, even for small values of $K$. Moreover, the quality of the approximation improves as $K$ increases. 
	% In the limit as $K \rightarrow \infty$, the infection times are known.
	The fact that the PD-SIR is a faithful approximation of the SIR enables the algorithm to update a large portion of the augmented data at each iteration of the Markov chain while maintaining a relatively high acceptance rate in the Metropolis-Hastings step.
	% In contrast, existing DA-MCMC algorithm updates only a few event times per iteration. As a result, our MCMC algorithm explores the latent space much more efficiently. Moreover, even though the latent spaces are generated from an approximation of the SIR process, the Metropolis-Hastings step ensures that the Markov chain converges to the posterior distribution of the parameters under the SIR model; in other words, the inference is exact.

	\begin{figure}
		\centering
		\begin{subfigure}[b]{0.49\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E2_K5.jpg}
			\caption{PD-SIR process ($K = 5$)}
			\label{fig:comparison_RD_SIR_K5}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.49\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E2_K10.jpg}
			\caption{PD-SIR process ($K = 10$)}
			\label{fig:comparison_RD_SIR_K10}
		\end{subfigure}
		\\
		\begin{subfigure}[b]{0.49\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E2_K50.jpg}
			\caption{PD-SIR process ($K = 50$)}
			\label{fig:comparison_RD_SIR_K50}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.49\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E2_K1000.jpg}
			\caption{PD-SIR process ($K = 1000$)}
			\label{fig:comparison_RD_SIR_K1000}
		\end{subfigure}
		\caption{Trajectories of the compartments $S$, $I$ and $R$ in a SIR (in colors) and PD-SIR processes (in grey) with the same infection incidence data $T_{1:K}$.}
		\label{fig:comparison}
	\end{figure}

	\section{Performance on simulated and real epidemic data}
	\label{sec:per}

	\subsection{Simulation Study}
	\label{sec:sim}
	%Outline: setup; results: posterior mean, credible interval, frequentist coverage, run time, ACF, acceptance rate; table; figure: traceplot, ACF; tuning: update proportion v. acceptance rate; compare with others methods e.g. chain binomial, exact proposal.

	% 1. convergence in medium-size population
	% 2. rho
	% 3. nominal coverage should be 95%.
	% 4. distribution of estimate for PD-SIR (unbiased) vs. LNA (biased)
	% explain why not compare chain binomial, particle filtering (forward simulation).

	In this section, the convergence properties of the DA-MCMC algorithm are examined with a simulation study. Each data set in this study is simulated from the stochastic SIR model and the prior distributions on the parameters $\beta$ and $\gamma$ are independent weakly informative gamma distribution with scale and shape parameters set to $0.01$ (see Equation \ref{eq:pri}).

	First, the mixing property of the Markov chain is assessed in a medium-size population of $S(0)=1,000$ initially susceptible and $I(0)=10$ initially infectious individuals. The parameters are $(\beta, \gamma) = (0.003, 1)$ and the process is observed until time $t_{end} = 6$ when the pandemic has completed most of its course but is not over yet (see Figure \ref{fig:long}, where $I(t_{end}) = 55$). The numbers of infections in $K = 10$ intervals of equal length are observed and correspond to $T_{1:K} = (40, 111, 193, 259, 178, 93, 29, 19, 9, 6)$. The Markov chain is run for $100,000$ iterations from the initial values $(\beta^{(1)}, \gamma^{(1)}) = (0.0003, 0.1) = (\beta/10, \gamma/10)$ and the event times of $202$ ($\rho = 0.2$) individuals are updated in the augmented data in each Metropolis-Hastings step. The algorithm took less than $5$ minutes to complete on a personal laptop. Figure \ref{fig:traceplot} shows the traceplots of the parameters $\beta$ and $\gamma$, the log likelihood and the basic reproduction number $R_0 = S(0) \beta / \gamma$, which corresponds to the average number of secondary infections caused by an infectious individual at the beginning of the outbreak. We observe that the Markov chain quickly migrates from the low density region of the space where it started to a high density region where it mixes relatively well. The acceptance rate of the latent spaces proposed in the Metropolis-Hastings step is $0.11$.
	Excluding the first $10,000$ iterations of the Markov chain as a burn-in, the posterior means of $\beta$, $\gamma$ and $R_0$ are respectively $0.00304$, $0.995$ and $3.07$.

	% TODO: Results: convergence (R statistic, Geweke), credible intervals.
	% TODO: Tables: posterior means, credible intervals, effective sample size.

	% TODO: frequentist coverage for 3 different observation schedules

	% TODO: Geweke

	\begin{figure}
		\centering
		\includegraphics[scale = 0.3]{E1_trajectories.jpg}
		\caption{Trajectories of the $S$, $I$ and $R$ compartments of a SIR process for a medium-size population ($S_0 = 1,000$).}
		\label{fig:long}
	\end{figure}

	\begin{figure}
		\centering
		\begin{subfigure}[b]{0.41\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E1_no_burn_beta_tp.jpg}
			\caption{$\beta$}
			\label{fig:traceplot_beta}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.41\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E1_no_burn_gamma_tp.jpg}
			\caption{$\gamma$}
			\label{fig:traceplot_gamma}
		\end{subfigure}
		\\
		\begin{subfigure}[b]{0.41\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E1_no_burn_loglik_tp.jpg}
			\caption{Log likelihood}
			\label{fig:traceplot_loglik}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.41\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E1_no_burn_R0_tp.jpg}
			\caption{$R_0$}
			\label{fig:traceplot_R0}
		\end{subfigure}
		\caption{Traceplots of the Markov chain (including burn-in).}
		\label{fig:traceplot}
	\end{figure}

	Second, the impact of the tuning parameter $\rho$ on the performance of the MCMC algorithm is evaluated. In particular, we examine the acceptance rate of the Metropolis-Hastings step, the total run time of the algorithm and the auto-correlation function of the parameters.
	Three SIR processes with different population sizes $S(0) = (10^2, 10^3, 10^4)$ are simulated. We set $I(0) =10$, $\gamma=1$, $t_{end} = 3$ and $K = 50$ for all simulations and $\beta = (0.001, 0.01, 0.1)$ respectively for the three different population sizes. These values for $\beta$ ensure that the three processes have comparable dynamics. For each process, the DA-MCMC algorithm is run for $50,000$ iterations with different values of $\rho = (10^{-4}, 10^{-3}, 10^{-2}, 10^{-1}, 0.5, 1)$. The algorithm is run only if there is at least one event time updated per iteration, that is, if $\rho(S_0+I_0)>1$.
	Figure \ref{fig:accept} presents the acceptance rate across the range of $\rho$.  We observe that the acceptance rate decreases as the number of event times updated per iteration and the population size increase.
	Figure \ref{fig:run_time} shows the run time of the MCMC algorithm for these simulations. The run time increases with the population size and with $\rho$.
	Finally, Figure \ref{fig:acf} shows the effect of $\rho$ on the auto correlation function for the parameter $\beta$ for the medium-size population process ($S(0) = 1,000$) for three values of $\rho$. We observe that updating half of the augmented data per iteration yields the lowest auto-correlation function across the values of $\rho$ considered. This suggests that, in this setting, updating the entire latent space in the Metropolis-Hastings step results in too many rejections while updating a tenth or less prevents the chain from efficiently exploring the latent space.

	\begin{figure}
		\centering
		\includegraphics[scale = 0.4]{E3_1000_ESS_sec_R0.jpg}
		\caption{Acceptance rate of the proposed latent spaces in the Metropolis-Hastings step of the MCMC algorithm across different population sizes and values for $\rho$.}
		\label{fig:ESS_sec}
	\end{figure}

	\begin{figure}
		\centering	
\centering
\includegraphics[width=\textwidth]{E1_burn_R0_acf.jpg}
\caption{$\rho = 0.1$}
\label{fig:acf}
	\end{figure}

	%Figures: acceptance rate v. proportion of updates (see Neal, Roberts A case study in non-centering -- table 1 p.319).

	\subsection{Ebola Outbreak in Gu\'eck\'edou, Guinea}
	\label{sec:ebo}
	%Outline: background; setup; results; table; figure

	We now turn to a case study concerning the Ebola outbreak in Western Africa.
	Between the end of $2013$ and $2015$, Guinea, along with several neighboring countries, experienced the largest outbreak of the Ebola virus disease in history. The virus, which has a fatality rate of $70\%$, was responsible for the death of almost $2,000$ people in Guinea alone during the outbreak.
	The outbreak is believed to have originated in the Gu\'eck\'edou prefecture of Guinea at the end of November $2013$ \cite{Baize.2014}. Weekly infection incidence counts are available for each prefecture for the $73$ weeks between the end of December $2013$ and May $2015$.
	We fit the stochastic SIR model to these incidence counts for the Gu\'eck\'edou prefecture using the MCMC algorithm proposed in this article.
	For simplicity, we assume that the population of the prefecture forms a closed population, that the model's parameters remained constant throughout the outbreak and that the reported infections counts are exact. Applying our MCMC algorithm to stochastic epidemic models where these assumptions are relaxed is the subject of ongoing research.
	The \textit{effective} population size is set to $n = 150,000$, roughly half of the total population in $2014$. Note that as long as the number of infections is small compared to the total population size, the exact population size does not need to be known to estimate the parameters of the model. Furthermore, since the first documented infection occurred late November, one month before the first reported incidence count \cite{Baize.2014}, the number of infectious individuals at the beginning of the observation period is set to $I(0) = 10$, with the units of time corresponding to ``days" in the analysis.

	The Markov chain is run for $50,000$ iterations and the event times of $10\%$ of the individuals are updated each iteration. The initial values of the parameters are set to $(\beta^{(1)}, \gamma^{(1)}) = (10^{-7}, 0.05)$ and the conjugate distributions in Equation \ref{eq:pri} are used as priors with $a_{\beta} = b_{\beta} = a_{\gamma} = b_{\gamma} = 0.01$ to make them weakly informative. Even for such a large population, the total run time of the algorithm was less than $100$ minutes on a personal laptop.
	The Metropolis-Hastings step for the latent space proposals achieves a healthy $22.1\%$ acceptance rate, and the first $10,000$ iterations of the Markov chain are discarded as a burn-in.
	Figure \ref{fig:ebola} shows the joint posterior distributions of $(\beta, \gamma)$, whose posterior means are respectively $1.06 * 10^{-6}$ and $0.109$, as well as the marginal distribution of the basic reproduction number $R_0$. This value for $\gamma$ indicates that people remain infectious for around $9$ days on average, which is consistent with existing literature. As noted by numerous authors, conditionally on partially observed data, the parameters $\beta$ and $\gamma$ appear positively correlated. The posterior distribution of $R_0$ is unimodal, relatively symmetric, and centered around $1$. A basic reproduction number close to $1$ is consistent with an outbreak that lasted several months without infecting the majority of the population.

	\begin{figure}
		\centering
		\begin{subfigure}[b]{0.45\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E5_burn_HPD.jpeg}
			\caption{Joint posterior density of $\theta = (\beta, \gamma)$. The $50\%$ and $90\%$ confidence regions are shown.}
			\label{fig:density_joint}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.49\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E5_burn_R0_hist.jpg}
			\caption{Posterior density of $R_0$.}
			\label{fig:density_R0}
		\end{subfigure}
		\caption{Posterior densities of parameters of interest (excluding burn-in of $10,000$ iterations).}
		\label{fig:ebola}
	\end{figure}

	\section{Discussion and Conclusion}
	\label{sec:dis}
	The method proposed in this article enables the classical Metropolis-Hastings algorithm to be an efficient method to conduct full posterior inference of the stochastic SIR model given only incidence data. Existing attempts using Markov chain Monte Carlo either mix poorly or do not scale to populations of more than a few hundred individuals, while methods relying on forward simulation are limited to moderate-sized outbreaks and may suffer from degeneracy issues on a case-by-case basis depending on the amount of missing data. 	Moreover, the simplifying assumptions in \cite{Fintzi.2020} or the approximate-Bayesian-computation framework of \cite{McKinley.2018}, which compromise targeting the exact posterior for computational reasons, generate biased estimates that may lead to spurious conclusions. Instead, our data-augmented algorithm enables fast and exact inference, even for large outbreaks, and leverages a well-studied, transparent MCMC framework. %Most existing inferential methods for epidemic stochastic models are only applicable to prevalence data. To our knowledge, our algorithm is the first that permits exact Bayesian inference given incidence data.

	Central to the success of the DA-MCMC algorithm is an efficient proposal scheme that swiftly explores the latent space of epidemic paths that are compatible with the observed data. The PD-SIR process possesses three features that yield such an efficient algorithm. First, generating a PD-SIR process is extremely fast; it only requires the simulation of truncated exponential distributions, which can efficiently be realized via the inverse-CDF method. Moreover, the individual infection rates are only updated $K$ times, as opposed to after each event as in the SIR model. Second, generating a PD-SIR process that is constrained to be compatible with the observed incidence data can be done at no additional cost; in contrast, generating a SIR process compatible with the observed data would be prohibitively slow \cite{Hobolth.2009}. Third, the dynamics of the PD-SIR process closely resemble those of the SIR process: the removal dynamics are identical and, for short resetting intervals, the infection dynamics are also very similar in the two processes.

	The first two features make the algorithm extremely fast. While existing data-augmented MCMC algorithms have only be applied to populations of a few hundred of individuals, the analysis of the Ebola outbreak in Gu\'eck\'edou shows that our algorithm can be applied to populations of up to $150,000$ individuals, generating tens of thousands of posterior samples in a reasonable amount of time.	
	The latter feature of the PD-SIR process enables the algorithm to update a large portion of the augmented data in each iteration while maintaining a healthy acceptance rate. As a result, the Markov chain makes large jumps in the latent space and has very good mixing properties. In contrast, existing DA-MCMC algorithms keep most of the latent space fixed across iterations \cite{Gibson.1998, ONeill.1999, Fintzi.2017} and their Markov chains therefore mix much more slowly.

	The algorithm possesses a tuning parameter $\rho$ that determines the portion of the augmented data being updated each iteration. Larger values for $\rho$ enable the chain to make larger steps in the latent space but can result in a low acceptance rate, while smaller values of $\rho$ result in a higher acceptance rate but constrain the chain to make small jumps. Depending on the size of the population, different values of the tuning parameter are optimal. To find a value that optimizes the mixing properties of the chain, one can use several short runs of the algorithm with different values for $\rho$ and select the value that yields the lowest auto-correlation function (ACF). Since the run time of the algorithm increases with $\rho$, one could also look at the effective sample size per unit of time. For instance, for populations of a few hundred individuals, we observed that updating the entire augmented data ($\rho=1$) yields the lowest ACF. In this case, the current and proposed augmented data are independent conditionally on the current value of the parameters.

	The DA-MCMC algorithm proposed in this article is specific to the stochastic SIR model, which is arguably a simplistic representation of the spread of disease. The model relies on assumptions such as perfect reporting, constant infection rates, homogeneously mixing population and exponentially-distributed infectious periods. In future work, we will present applications of our DA-MCMC algorithm to processes where these assumptions are relaxed. In particular, we are considering extensions to epidemic models with under-reporting, a varying infection rate, stratified populations and non-Markovian dynamics, as well as the general stochastic SIR model of \cite{Severo.1969}.

	\appendix

	\section{Distribution of Death Times in Linear Death Process}
	\label{app:ldp}
	Theorem \ref{theo:ldp} was first proved by \cite{Neuts.1971}. Since this is a result that has rarely been mentioned in the literature, we provide a proof in this appendix from (cite Ross textbook).

	Consider a linear pure death process with $n$ particles and individual death rate $\mu$.
	Let $T_i$ be the time of the $i$th death. Then $W_1 = T_1 \sim \Exp(n\mu)$ and $W_i = T_i - T_{i-1} \sim \Exp((n-1)\mu)$ independently. Let $N$ be the number of deaths by time $t$. Then, 
	\begin{align*}
		& f(T_1 = t_1, \dots, T_N = t_N | N) \\
		& \propto f(T_1 = t_1, \dots, T_N = t_N, T_{N+1} > t) 1\{T_N < t\}\\
		& \propto f(W_1 = t_1, W_2 = t_2 - t_1, \dots, W_N = t_N - t_{N-1}, W_{N+1} > t - t_N) 1\{T_N < t\}\\
		& \propto \exp\{-n\mu t_1\} \exp\{-(n-1)\mu(t_2-t_1)\}\dots \exp\{-(n-N)\mu(t - t_N)\} 1\{T_N < t\}\\
		& \propto \exp\{-\mu t_1\}\exp\{-\mu t_2\} \dots \exp\{-\mu t_N\} 1\{T_N < t\}
	\end{align*}
which corresponds to the kernels of independent exponential distribution truncated above by $t$.

	\section{Uniform Ergodicity}
	\label{app:uni}
	We show that the state space $\chi = \chi_{\theta}\times\chi_{\z}$ of the Markov chain $\{(\theta, \z)^{(i)}\}_i$ is a small set for the transition kernel $P$, that is, there exists a probability measure $\nu$ on the $\sigma$-algebra $\sigma(\chi)$ such that
	$$\beta \nu(.) \le P^m(x,.), \forall x\in \chi.$$
	for some constants $m \ge 1$ and $\beta > 0$.
	By Proposition 2 in \cite{Tierney.1994}, this implies that the Markov chain $\{(\theta, \z)^{(i)}\}_i$ is uniformly ergodic and therefore satisfies
	$$
	\Vert P^n(x,.)-\pi(.)\Vert_{TV} \le M r^n
	$$
	for some finite $M$ and positive constant $r<1$, where $\Vert \mu \Vert_{TV}$ denotes the total variation norm of the measure $\mu$.

	The following three lemmas are used to show this result.

	% Gamma 1
	\begin{lemma}
		\label{pro:ga1}
		Let $Ga(x;a,b)$ denote the density of a gamma distribution with shape $a$ and rate $b$ evaluated at $x$. Then
		\begin{equation}
			\label{eq:ga1}
			\inf_{0\le \beta\le B} Ga(x;a,b+\beta) = 
			\begin{cases}
				Ga(x;a,b), & x<x_a^* \\ Ga(x;a,b+B), & x\ge x_a^*
			\end{cases}
		\end{equation}	
		where $x_a^*=\frac{a}{B}\log\left( 1+\frac{B}{b}\right) $. Moreover,
		\begin{equation}
			\label{eq:ga2}
			\inf_{0\le \alpha\le A} Ga(x;a+\alpha,b) = 
			\begin{cases}
				Ga(x;a,b), & x>x_b^* \\ Ga(x;a+A,b), & x\le x_b^*
			\end{cases}
		\end{equation}
		where $x_b^*=\frac{1}{b}\left[ \frac{\Gamma(a+A)}{\Gamma(a)}\right]^{1/A} $.
	\end{lemma}

	\begin{proof}
		Equation \ref{eq:ga1} is proven in Jones, Hobert (2004). For Equation \ref{eq:ga2}, note that $x_b^*$ is the only positive solution to $Ga(x;a,b) = Ga(x;a+A,b)$. Now, for all $0<x\le x_b^*$ and all $0\le\alpha\le A$, we have
		\begin{align*}
			\frac{Ga(x;a+A,b)}{Ga(x;a+\alpha,b)}
			& = b^{A-\alpha}x^{A-\alpha} \frac{\Gamma(a+\alpha)}{\Gamma(a+A)} \\
			& \le b^{A-\alpha}\left(\frac{1}{b}\left[ \frac{\Gamma(a+A)}{\Gamma(a)}\right]^{1/A} \right)^{A-\alpha} \frac{\Gamma(a+\alpha)}{\Gamma(a+A)} \\
			& = \left[ \frac{\Gamma(a+A)}{\Gamma(a)}\right]^{(A-\alpha)/A} \frac{\Gamma(a+\alpha)}{\Gamma(a+A)} \\
			& = \left( \frac{\Gamma_{a,a+A}}{\Gamma_{a+\alpha, a+A}}\right)^{A-\alpha} \\
			& \le 1,
		\end{align*}
		where $\Gamma_{d,e} = \left( \frac{\Gamma(e)}{\Gamma(d)} \right)^{\frac{1}{e-d}}$ is a geometric average,
		and where the last inequality holds because $\Gamma_{a,a+A}\le\Gamma_{a+\alpha, a+A}$.
		The case $x>x_b^*$ can be shown similarly.
	\end{proof}

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{gamma_minorization_1a.jpg}
		\caption{$\alpha \in [0, 1]$, $\beta=0$}
		\label{fig:gam1a}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.49\textwidth}
		\centering
		\includegraphics[width=\textwidth]{gamma_minorization_1b.jpg}
		\caption{$\alpha = 0$, $\beta \in [0, 1]$}
		\label{fig:gam1b}
	\end{subfigure}
	\caption{Example of minorization of the density of a gamma distribution $Ga(2+\alpha, 0.5+\beta)$ for $\alpha$ and $\beta$ separately.}
	\label{fig:gam1}
\end{figure}

% Gamma 2
Proposition \ref{pro:ga1} can be used to minorize a density of the form
\begin{equation}
	\label{eq:ga}
	Ga(x;a+\alpha,b+\beta) = \frac{(b+\beta)^{a+\alpha}}{\Gamma(a+\alpha)}x^{a+A-1}\exp\{-x(b+B)\}, \quad 0\le\alpha\le A, 0\le\beta\le B.
\end{equation}
over $\alpha$ and $\beta$ for each value of $x$. We can show the following.
\begin{lemma}	
	\label{pro:ga2}
	For a fix $x>0$, the density $Ga(x;a+\alpha,b+\beta)$ is minimized by $(\alpha, \beta) \in \{(A, 0), (0, B)\}$, the minimizing set of values depending on $x$. In particular,	
	$$\inf_{\begin{aligned}
			0\le \alpha\le A \\ 0\le \beta\le B
	\end{aligned}}Ga(x;a+\alpha,b+\beta) = 
	\begin{cases}
		Ga(x;a+A,b)                     , & x<x_a         \text{ or } x<x_{a+A}^* \vee x_b^*\\
		Ga(x;a,b+B)                     , & x_{a+A}^* < x \text{ or } x_a^* \wedge x_{b+B}^* < x\\
		\min\{Ga(x;a+A,b), Ga(x;a,b+B)\}, & x_a^* \wedge x_b^*\le x < x_{a+A}^* \vee x_{b+B}^*\\
	\end{cases}$$
\end{lemma}

\begin{proof}
	Given $x>0$, \ref{eq:ga1} shows that for a fixed $\alpha$, $\beta \in \{0, B\}$ is the minimizer of \ref{eq:ga}, and \ref{eq:ga2} shows that for a fixed $\beta$, $\alpha \in \{0, A\}$ is the minimizer of \ref{eq:ga}. This implies that for a fixed $x>0$, \ref{eq:ga} is minimized by $(\alpha, \beta) \in \{(0,0), (A, 0), (0, B), (A,B)\}$.

	A case-by-case analysis of the nine possibilities
	$$(x<x_a^*; x_a^*<x<x_{a+A}^*; x_{a+A}^* < x )\times(x<x_{b+B}^*; x_{b+B}^*<x<x_b^*; x_b^* < x)$$
	is presented in Table \ref{tab:ga9} and shows that it is sufficient to consider $(\alpha, \beta) \in \{(A, 0), (0, B)\}$.

	\begin{table}[H]
		\centering
		\begin{tabular}{l|c c c}
			& $x<x_a^*$  & $x_a^*\le x<x_{a+A}^*$ & $x_{a+A}^* < x$ \\ \hline
			$x<x_b^*$              & $(a+A, b)$ & $(a+A, b)$             &                 \\
			$x_b^*\le x<x_{b+B}^*$ & $(a+A, b)$ &  ad hoc                & $(a, b+B)$      \\
			$x_{b+B}^* < x$        &            & $(a, b+B)$             & $(a, b+B)$
		\end{tabular}
		\caption{Values of $(\alpha, \beta)$ that minimize $Ga(x;a+\alpha,b+\beta)$ for different values $x$.}
		\label{tab:ga9}
	\end{table}

	The two empty entries in Table \ref{tab:ga9} correspond to impossible configurations. Indeed, $x_b^* > x_a^*$ and $x_{a+A}^* > x_{b+B}^*$ for all $a, A, b, B$ since
	$$
	\frac{x_b^*}{x_a^*} = \dfrac{\left(\frac{\Gamma(a+A)}{\Gamma(a)}\right)^{1/A} b^{-1} }{a B^{-1} \log\left( 1+B/b\right)} = \dfrac{\left(\frac{\Gamma(a+A)}{\Gamma(a)}\right)^{1/A}}{a} \dfrac{B/b}{\log\left( 1+B/b\right)} = \dfrac{\Gamma_{a,a+A}}{a} \dfrac{B/b}{\log\left( 1+B/b\right)} \ge 1
	$$
	and
	$$
	\frac{x_{b+B}^*}{x_{a+A}^*} = \dfrac{\left(\frac{\Gamma(a+A)}{\Gamma(a)}\right)^{1/A} (b+B)^{-1} }{(a+A) B^{-1} \log\left( 1+B/b\right)} = \dfrac{\Gamma_{a,a+A}}{a+A} \dfrac{B}{b+B} \dfrac{1}{\log\left( 1+B/b\right)} \le 1
	$$
	where the inequalities hold since $a \le \Gamma_{a,a+A} \le a+A$ and $\dfrac{y}{\log\left( 1+y\right)}\le1$.

	If $x_a^* \wedge x_{b+B}^*\le x < x_{a+A}^* \vee x_{b}^*$, which corresponds to the middle entry of the center column in Table \ref{tab:ga9}, then one needs to directly check which set of values in $\{(0,0), (A, 0), (0, B), (A,B)\}$ minimizes Equation \ref{eq:ga}. In fact, it is sufficient to consider only $\{(A,0), (0, B)\}$ since
	$$Ga(x, a+A, b) \le \begin{cases}
		Ga(x;a,b), & x < x_b^* \\
		Ga(x;a+A,b+B), & x < x_{a+A}^* 
	\end{cases}$$
	and
	$$Ga(x, a, b+B) \le \begin{cases}
		Ga(x;a,b), & x > x_a^* \\
		Ga(x;a+A,b+B), & x > x_{b+B}^* 
	\end{cases}$$
\end{proof}

\begin{figure}	
	\centering
	\includegraphics[width=\textwidth]{gamma_minorization_2.jpg}
	\caption{Example of minorization of the density of a gamma distribution $Ga(2+\alpha, 0.5+\beta)$ for $\alpha$ and $\beta$ jointly.}
	\label{fig:gam2}
\end{figure}

% Truncated Exponential
\begin{lemma}
	\label{pro:tru}
	$$\min_{0 \le \mu \le M} \TruncExp(u; \mu, 0, u) = \TruncExp(u; M, 0, u) $$
\end{lemma}
\begin{proof}
	Remember that $\TruncExp(u; \mu, 0, u) = \frac{\mu \exp\{-u\mu\}}{1-\exp\{-u\mu\}}$ is the density of an exponential distribution bounded above by $u$.
	\begin{align*}
		\frac{d}{d\mu}\TruncExp(u; \mu, 0, u)
		& = \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \left[ 1-\mu u - \mu u \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \right]
	\end{align*}
	Now let $g(\mu) = 1-\mu u - \mu u \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right)$. By L'H\^{o}pital's rule, $g(0) = 0$. Moreover, for $\mu>0$, we have
	\begin{align*}
		g'(\mu)
		& = - \mu \left(1+\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \left[ 1-u\mu \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \right]  \\
		& = - \mu \left(1+\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \left[ 1- \frac{u\mu}{\exp\{u\mu\}-1} \right] \\
		& \le 0
	\end{align*}
	where the inequality follows from $\frac{u\mu}{\exp\{u\mu\}-1}\le1$.
	This implies that $g(\mu) \le 0$ for $\mu>0$. We therefore have $\frac{d}{d\mu}\TruncExp(u; \mu, 0, u) \le 0$ and the result of the proposition follows.
\end{proof}

% Uniform Ergodicity
\begin{proposition}
	\label{pro:uni}
	The transition kernel $P = P_{\theta}P_{\z}$ satisfies a minorization condition $M(1, \delta, \chi, \nu)$ and is therefore uniformly ergodic.
\end{proposition}

\begin{proof}
	First, note that the transition kernel $P$ of our Markov chain is a composition of two kernels
	$$ P = P_{\theta} P_{\z}$$
	where the kernel $ P_{\theta}$ updates the parameters $\theta$ of $\x$ while keeping the augmented data $\z$ fixed and $P_{\z}$ update $\z$ while keeping $\theta$ fixed. A one-step transition therefore corresponds to the following scheme 
	$$\x_1 = (\theta_1, \z_1) \rightarrow (\theta_2, \z_1) \rightarrow (\theta_2, \z_2) = \x_2.$$

	The kernel $P$ has a density $p$ with respect to the product measure $\mu = \lambda \mu_{\z}$ with $\lambda$ the Lebesgue measure on $\chi_{\theta}$, the space of $\theta$, and $\mu_{\z}$ a $\sigma$-finite measure on $\chi_{\z}$, the space of $\z$.
	\begin{align*}
		P(\x_1, d\x_2) 
		& = p(\x_1, \x_2) \mu(d\x_2) \\
		& = p((\theta_1, \z_1), (\theta_2, \z_2)) \lambda(d\theta_2)\mu_{\z}(d\z_2) \\
		& = p_{\theta}((\theta_1, \z_1), (\theta_2, \z_1)) p_{\z}((\theta_2, \z_1), (\theta_2, \z_2)) \mu_{\z}(d\z_2) \lambda(d\theta_2)
	\end{align*}
	where
	\begin{align*}
		p_{\theta}((\theta_1, \z_1), (\theta_2, \z_1))
		& = q_{\theta}((\theta_1, \z_1), (\theta_2, \z_1)) \alpha((\theta_1, \z_1), (\theta_2, \z_1)) \\
		& = \pi(\theta_2| \z_1)
	\end{align*}
	is the density of the transition kernel $P_{\theta}$ with respect to $\lambda$ and corresponds to the full conditional distribution $\pi(\theta_2| \z_1)$ since the kernel $P_{\theta}$ is a Gibbs sampler, and
	\begin{align*}
		p_{\z}((\theta_2, \z_1), (\theta_2, \z_2))
		& = q_{\z}((\theta_2, \z_1), (\theta_2, \z_2)) \alpha((\theta_2, \z_1), (\theta_2, \z_2)) \\
		& = g(\z_2; \theta_2) \min\left\lbrace 1, \dfrac{\pi(\theta_2, \z_2)g(\z_1; \theta_2)}{\pi(\theta_2, \z_1)g(\z_2; \theta_2)} \right\rbrace \\
		& = \min\left\lbrace g(\z_2; \theta_2), \dfrac{\pi(\theta_2, \z_2)g(\z_1; \theta_2)}{\pi(\theta_2, \z_1)} \right\rbrace
	\end{align*}
	is the density of the transition kernel $P_{\z}$ with respect to $\mu_{\z}$ and corresponds to one step of the Metropolis-Hasting algorithm where a new specification of the augmented data $\z_2$ is proposed conditionally on the current value of the parameters $\theta_2$, but independently of the current specification of the augmented data $\z_1$.

	To show that $\chi$ is a small state, it is sufficient to show that there exists a function $k$ such that
	$$p(\x_1, \x_2) \ge k(\x_2) > 0, \qquad \forall \x_1 \in \chi $$
	for all $\x_2 \in \chi$.
	Note that the density $p$ is a function of $\x_1 = (\theta_1, \z_1)$ only through the full conditional distribution $\pi(\theta_2| \z_1)$ and the proposal density $g(\z_1; \theta_2)$.
	\begin{align*}
		p((\theta_1, \z_1), (\theta_2, \z_2))
		& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2), \dfrac{\pi(\theta_2| \z_1) \pi(\theta_2, \z_2)}{\pi(\theta_2, \z_1)} g(\z_1; \theta_2) \right\rbrace \\
		& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2),
		\dfrac{
			\frac{f(\theta_2, \z_1, Y)}{f(\z_1, Y)}
			\frac{f(\theta_2, \z_2, Y)}{f(Y)}
		}{\frac{f(\theta_2, \z_1, Y)}{f(Y)}} 
		g(\z_1; \theta_2)\right\rbrace \\
		& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2),
		\dfrac{f(\theta_2, \z_2, Y)}{f(\z_1, Y)}
		g(\z_1; \theta_2)\right\rbrace \\
		& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2),
		\pi(\theta_2| \z_2)
		g(\z_1; \theta_2)\right\rbrace
	\end{align*}
It therefore suffices to show that there exist functions $k_1$ and $k_2$ such that
$$\pi(\theta| \z_1) \ge k_1(\theta) > 0 \qquad \text{and} \qquad  g(\z_1; \theta) \ge k_2(\theta) > 0.$$

	First, 
	\begin{align*}
		\pi(\theta| \z_1) 
		& = \pi(\beta| \z_1) \pi(\gamma| \z_1)  \\
		& = Ga(\beta; a_{\beta} + n_T, b_{\beta} + SI_1) Ga(\gamma; a_{\gamma} + {n_J}_1, b_{\gamma} + I_1) \\
		& \ge h_{\beta}(\beta) h_{\gamma}(\gamma) \\
		& = k_1(\theta) > 0
	\end{align*}
	where, by Proposition \ref{pro:ga1},
	$$h_{\beta}(\beta) = \begin{cases}
		Ga(\beta;a_{\beta}+n_T,b_{\beta}), & \beta<\frac{a_{\beta}+n_T}{t_{end}(n_T+I_0) n} \log \left( 1 + \frac{t_{end}(n_T+I_0) n}{b_{\beta}}\right) \\ Ga(x;a_{\beta}+n_T,b_{\beta}+n(n_T+I_0) t_{end}), & \text{else}
	\end{cases}$$
since $n_T = \sum_k T_k$ is known and 
$$SI_1 = \int_{0}^{t_{end}}S_1(t)I_1(t)dt \in [0, n (n_T+I_0) t_{end}];$$
and, by Proposition \ref{pro:ga2},
	$$h_{\gamma}(\gamma) = \min\{
		Ga(\gamma;a_{\gamma}+n_T+I_0,b_{\gamma}), Ga(x;a_{\gamma},b_{\gamma}+(n_T+I_0) t_{end})\}
		$$
since $0\le n_1^J \le n_T + I_0$ and
$$I_1 = \int_{0}^{t_{end}}I_1(t)dt \in [0, (n_T+I_0) t_{end}].$$

	Second, each factor in
	\begin{align*}
		& g(\z_1; \theta) = \prod_{i=1}^{n_T} \TruncExp(z^T_i; \beta I_k, t_{k(i)-1}, t_{k(i)}) (1-p_i)^{1\{\z^J_i = \infty\}} (p_i \TruncExp(z^J_i; \gamma, z^T_i, t_{end}))^{1\{\z^J_i \le t_{end}\}}
	\end{align*}
	can be minorized as follows. First,
	\begin{align*}
		\TruncExp(z^T_i; \beta I_k, t_{k(i)-1}, t_{k(i)}) 
		& \ge \TruncExp(t_{k(i)}; \beta I_k, t_{k(i)-1}, t_{k(i)}) \\
		& \ge \TruncExp(t_{k(i)}; \beta (n_T + I_0), t_{k(i)-1}, t_{k(i)}),
	\end{align*}
	with the last inequality following from proposition \ref{pro:tru}, second,
	$$1-p_i = \exp\{-\gamma (t_{end} - z^T_i)\} \ge \exp\{-\gamma (t_{end} - t_{k(i)-1})\},$$
	third,
	$$p_i \TruncExp(z^J_i; \gamma, z^T_i, t_{end}) = \Exp(z^J_i - z^T_i; \gamma)1(z^T_i < z^J_i \le t_{end}) \ge \Exp(t_{end} - t_{k(i) - 1}; \gamma)$$
	where the equality holds since $p_i = P(z^J_i \le t_{end}|z^J_i > z^T_i)$ is equal to the normalizing constant of the truncated exponential.

	Taken together, these inequalities give
	\begin{align*}
		g(\z_1; \theta)
		& \ge \prod_{i=1}^{n_T} \TruncExp(t_{k(i)}; \beta (I_0 + n_T), t_{k(i)-1}, t_{k(i)}) \min\{\exp\{-\gamma (t_{end} - t_{k(i) - 1})\}, \Exp(t_{end} - t_{k(i) - 1}; \gamma)\} \\
		& = k_2(\theta) > 0
	\end{align*}
	which completes the proof.
\end{proof}

	\bibliographystyle{plain}
	\bibliography{bibliography}

\end{document}

### 2021-10-11 Experiment_output.R
E5

@@ -30,4 +30,4 @@ for(i in 3 : 200) {
 summary_E4 <- experiment_4_output_analysis(results, path = "Output/Figures/E4")
 
 load("Output/RDATA/E5.RDATA")
-summary_5 <- experiment_5_output(output_E5, path = "Output/Figures/E5")
+summary_5 <- experiment_5_output(output_E5, path = "Output/Figures/E5", burnin = 50)

### 2021-10-11 Experiments.R

@@ -3,9 +3,10 @@
 # Setup
 library(PDSIR)
 run_E1 <- FALSE
+run_E2 <- FALSE
 run_E3 <- FALSE
-run_E5 <- FALSE
-run_E6 <- TRUE
+run_E5 <- TRUE
+run_E6 <- FALSE
 
 #
 # Experiments
@@ -22,6 +23,11 @@ if(run_E1){ # Proof of Concept
   ))
 }
 
+if(run_E2){
+  set.seed(1)
+  experiment_2_PDSIR_trajectories(path = "Output/Figures/E2")
+}
+
 if(run_E3){ # Rho
   set.seed(2)
   output_E3 <- experiment_3_acceptance_vs_rho(
@@ -34,7 +40,7 @@ if(run_E3){ # Rho
 
 if(run_E5){ # Ebola
   set.seed(1)
-  output_E5 <- experiment_5_ebola(N = 1e6, rho = 0.1, thin = 10)
+  output_E5 <- experiment_5_ebola(N = 1e3, rho = 0.1, thin = 10)
   save(output_E5, file = "Output/RDATA/E5.RDATA")
   print(paste0("E5 is done - ", Sys.time()))
 }

### 2021-10-11 Experiments.R
E5 long run

@@ -40,7 +40,7 @@ if(run_E3){ # Rho
 
 if(run_E5){ # Ebola
   set.seed(1)
-  output_E5 <- experiment_5_ebola(N = 1e3, rho = 0.1, thin = 10)
+  output_E5 <- experiment_5_ebola(N = 1e6, rho = 0.1, thin = 10)
   save(output_E5, file = "Output/RDATA/E5.RDATA")
   print(paste0("E5 is done - ", Sys.time()))
 }

### 2021-10-11 Experiment_output.R
E6 output analysis

@@ -31,3 +31,12 @@ summary_E4 <- experiment_4_output_analysis(results, path = "Output/Figures/E4")
 
 load("Output/RDATA/E5.RDATA")
 summary_5 <- experiment_5_output(output_E5, path = "Output/Figures/E5", burnin = 50)
+
+load("Output/RDATA/E6_joint.RDATA")
+load("Output/RDATA/E6_single.RDATA")
+summary_E6_joint <- experiment_1_output_analysis(
+  output_E6_joint, plot_id = "E6", path = "Output/Figures/E6/joint"
+)
+summary_E6_single <- experiment_1_output_analysis(
+  output_E6_single, plot_id = "E6", path = "Output/Figures/E6/single"
+)

### 2021-10-11 Experiments.R
cosmetic change

@@ -46,7 +46,7 @@ if(run_E5){ # Ebola
 }
 
 
-if(run_E6){ # Ebola
+if(run_E6){ # single-site updates
 
   set.seed(1)
   output_E6_single <- experiment_6_single_site_update(N = 1e6, save_fig = FALSE, thin = 10)

### 2021-10-12 Manuscripts/Article/Article.tex
set up .tex manuscript file

@@
 \setlength\parindent{35pt}
-\input{macros}
+\input{../macros}
 \usepackage{algorithm}
@@
 
-\graphicspath{ {../Experiment Output/Figures/} }
+\graphicspath{ {../Figures/} }
 
 \newcommand{\jx}[1]{{\color{blue}{ #1}}}
+\newcommand{\ram}[1]{{\color{green}{ #1}}}
 
@@
 	\title{%Efficient Likelihood-Based Inference for Fitting Stochastic Epidemic Models to Incidence Data via Data Augmentation
-	A Uniformly Ergodic Data-Augmented MCMC for Fitting Stochastic Epidemic Models to Incidence Data
-	}
+		Uniformly Ergodic Data-Augmented MCMC for Fitting Stochastic Epidemic Models to Incidence Data}
 	\author{Rapha\"{e}l Morsomme$^{1}$\footnote{raphael.morsomme@duke.edu} \ and Jason Xu$^{1}$ \\
@@
 	\begin{abstract}
-		Stochastic epidemic models provide a realistic an interpretable description of the spread of a disease through a population.
-Yet, fitting these models in missing data settings is a notably difficult task; in particular, when the epidemic process is only partially observed, the likelihood of the model is intractable.
-To remedy this issue, this article introduces a data-augmented MCMC algorithm for fast and exact inference for the stochastic SIR model given discretely observed infection incidence counts.
+		Stochastic epidemic models provide an interpretable probabilistic description of the spread of a disease through a population.
+Yet, fitting these models in missing data settings is a notably difficult task; in particular, when the epidemic process is only partially observed, the likelihood of many classic models becomes intractable.
+To remedy this issue, this article introduces a data-augmented MCMC algorithm for fast and exact inference under the stochastic SIR model given discretely observed incidence counts.
 In a Metropolis-Hastings step, the algorithm jointly proposes event times of the augmented data according to a stochastic process whose dynamics closely resemble those of the SIR process, and from which we can efficiently generate an epidemic that is compatible with the observed data.
-Not only is the algorithm fast, but, since the augmented data are generated from a very faithful approximation of the target epidemic model, the algorithm can update a large portion of the augmented data per iteration while maintaining a relatively high acceptance rate, thereby exploring the high-dimensional latent space efficiently.
-Moreover, this MCMC algorithm is to be uniformly ergodic.
-While existing MCMC approaches become intractable for populations greater than a few thousand individuals, unless they rely on model simplifications or approximations, the proposed algorithm enables exact inference and scales to outbreaks in populations of hundred-thousands individuals
-		+ even on a single laptop.
-We validate its performance via thorough simulation experiments, and a case study on the $2014$ Ebola outbreak in Western Africa.
+Not only is the algorithm fast, but since the augmented data are generated from a very faithful approximation of the target epidemic model, the sampler can update large portions of the augmented data per iteration without lowering the acceptance rate prohibitively.
+We find that the method explores the high-dimensional latent space efficiently, and show that the Markov chain underlying the algorithm is uniformly ergodic.
+% While existing exact MCMC approaches become intractable for populations greater than a few thousand individuals,
+		The proposed algorithm enables exact inference and scales to outbreaks in populations of hundreds of thousands, even on a single laptop.
+We validate its performance via thorough simulation experiments, consider and a case study on the $2014$ Ebola outbreak in Western Africa.
 % in Gu\'eck\'edou, Guinea.
 %, in a population of $150,000$ individuals.
+test.
 	\end{abstract}
-	\keywords{Stochastic epidemic model; incidence data; exact posterior inference; data-augmented MCMC; likelihood-based methods; uniform ergodicity}
+	\keywords{Compartmental model; incidence data; exact posterior inference; data-augmented MCMC; likelihood-based methods; uniform ergodicity}
 
@@
 	Conducting inference on SEMs is, however, a notably difficult task.
-The difficulty stems from the fact that the observed data typically provide incomplete information on a process that evolves continuously through time.
-This makes the likelihood of the model intractable; were the infection and recovery times of each individual of the population observed, conducting inference on SEMs would be straightforward, but this is rarely the case.
-In practice, one often has access to either incidence data such as weekly counts of new infections, or prevalence data such as the numbers of people infectious at discrete times.
-The marginal likelihood of these partially observed data becomes a bottleneck, requiring a large integration step that accounts for all possible configurations of the missing data.
-In particular, direct computation of this likelihood requires the transition probabilities between observation times for which no closed form are available.
-Numerical methods to obtain transition probabilities have only recently been developed in \cite{Ho.2018b, Ho.2018}, but the high computational costs of their algorithm makes the approach applicable to small populations only.
-
-	In this article, we introduce a data-augmented MCMC algorithm to estimate the parameters of the stochastic SIR model, a commonly used SEM, from discretely observed infection incidence counts.
+Challenges stem from the fact that the observed data typically provide incomplete information on a process that evolves continuously through time, making the likelihood of the model intractable.
+Were the infection and recovery times of each individual of the population observed, conducting inference on SEMs would be straightforward, but this is rarely the case.
+In practice, one often has access to either incidence data such as weekly counts of new infections, or prevalence data such as the numbers of people infectious at discrete reporting times.
+For instance, the $2013$–$2016$ outbreak of Ebola virus reported over $10,000$, mainly in three Western African countries (Guinea, Liberia and Sierra Leone).
+This outbreak was characterized by the large number of infections that occurred after individuals were tested positive for the virus, giving rise to noisy incidence data.
+Numerous infections happened at the hospitals where infected individuals received a treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased person during funerals is a tradition.
+Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.}.
+	It therefore seems appropriate to model a positive test for the virus as an indication that the individual became infectious some time prior to the test rather than as a removal time \jx{to do: stronger transition tying back to goal of partial observed inference}.
+
+	More specifically, the marginal likelihood of such partially observed data becomes a bottleneck, as it requires a large integration step that accounts for all possible configurations of the missing.
+In particular, direct computation of this likelihood requires the transition probabilities between observation times for which no closed form is available.
+\jx{to do: cite recent papers from Simon Spencer's group too that do so for discrete-time models, and mention classical matrix exponentiation being intractable}.
+Numerical methods to obtain transition probabilities have only recently been developed in \cite{Ho.2018b, Ho.2018}, but the high computational costs limit their algorithm to moderate sized outbreaks.
+
+	In this article, we make use of data augmentation to explore configurations of the missing data through latent variables.
+We show how a data-augmented MCMC algorithm can target the exact joint posterior of model parameters and these latent variables under the stochastic SIR model, a commonly used SEM, from discretely observed infection incidence counts.
 Rather than marginalizing directly, our approach accounts for the latent space via sampling.
 The algorithm updates the event times of the augmented data by jointly proposing infection and removal times from a stochastic process whose dynamics closely resemble those of the SIR process, and from which we can efficiently generate an epidemic that is compatible with the observed data.
-This surrogate process serves as an efficient proposal scheme within a Metropolis-Hastings algorithm.
+Its success lies in the design of this surrogate process serving as an efficient proposal scheme within a Metropolis-Hastings algorithm.
 Since the latent spaces are generated from a very faithful approximation of the target epidemic model, the algorithm can update a large portion of the event times per iteration while maintaining a relatively high acceptance rate, thereby exploring the high-dimensional latent space efficiently.
@@
 Once infected, she is immediately infectious and remains so for some period of time after which she is removed from the process without the possibility of reinfection.
-In this formulation, demographics dynamics such as births and deaths of individuals are ignored since they usually occur at a much slower rate than infections and removals.
+In this formulation, demographic dynamics such as births and deaths of individuals are ignored since they usually occur at a much slower rate than infections and removals.
 
@@
 	\begin{equation}
-		\X(t) = (X_1(t), \dots, X_n(t)) \in \{S, I, R\}^n
+		\X(t) = (X_1(t), \dots, X_n(t)) \in \{s, i, r\}^n
 	\end{equation}
 	where the agent-level subprocess
-	$$ X_i(t) = 
+	$$ X_j(t) = 
 	\begin{cases}
-		S, & t \in [0, \tau^T_i] \\
-		I, & t \in (\tau^T_i, \tau^J_i] \\
-		R, & t \in (\tau^J_i, \infty)
+		s, & t \in [0, \tau^I_j] \\
+		i, & t \in (\tau^I_j, \tau^R_j] \\
+		r, & t \in (\tau^R_j, \infty)
 	\end{cases}
@@
 	$$
-	denotes the status (compartment) of individual $i$ at time $t$ with $\tau^T_i$ and $\tau^J_i$ respectively the infection and removal times of individual $i$.
-If individual $i$ never becomes infectious, then we set $\tau^T_i = \tau^J_i = \infty$ and $X_i(t) = S$ for $t \in [0, \infty)$.
-The set $\chi_{\X}$ consists of the trajectories compatible with the evolution of a disease, that is, the set of trajectories in which no infection occurs whenever the infectious compartment is depleted :
+	denotes the status (compartment) of individual $j$ at time $t$ with $\tau^I_j$ and $\tau^R_j$ respectively the infection and removal times of individual $j$.
+If individual $j$ never becomes infectious, then we set $\tau^I_j = \tau^R_j = \infty$ and $X_j(t) = s$ for $t \in [0, \infty)$.
+Here $\chi_{\X}$ denotes the set of valid trajectories compatible with the evolution of a disease---that is, the set of trajectories in which no infection occurs whenever the infectious compartment is depleted :
 	\begin{equation}
 		\label{eq:chi}
-		\chi_{\X} = \{X:X(t) \in \{S,R\}^n \Rightarrow X(t+s) = X(t), \forall s>0 \}.
+		\chi_{\X} = \{X:X(t) \in \{s,r\}^n \Rightarrow X(t+u) = X(t), \forall u>0 \}.
 	\end{equation}
@@
 This parameter can be interpreted as the \textit{infection rate}: when a susceptible individual comes into contact with an infectious individual, she immediately becomes infectious.
-	If we also make the common assumption that the infectious periods follow independent exponential distributions with rate $\gamma$, then the process \ref{eq:X} is a time–homogeneous continuous time Markov chain whose instantaneous transition rates
+	If we also make the common assumption that the infectious periods follow independent exponential distributions with rate $\gamma$, then the process \eqref{eq:X} is a time–homogeneous continuous time Markov chain, whose instantaneous transition rates
 	%from state $\x$ to $\x'$
@@
 	\begin{cases}
-		\beta I(t), & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = S \text{ and } x_j' = I, \\
-		\gamma, & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = I \text{ and } x_j' = R, \\
+		\beta I(t), & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = s \text{ and } x_j' = i, \\
+		\gamma, & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = i \text{ and } x_j' = r, \\
 		0, & \text{ otherwise.}
@@
 	$$
-	where $I(t) = \sum_{i=1}^n \mathbf{1} \{x_i(t) = I\}$ is the total number of infectious individuals at time $t$.
-That is, the individual-level infection and removal rate at time $t$ are respectively $\beta I(t)$ and $\gamma$.
+	where $I(t) = \#\{x_j(t) = i\}$ is the total number of infectious individuals at time $t$.
+Thus, the individual-level infection and removal rates at time $t$ are respectively $\beta I(t)$ and $\gamma$.
 	%Equivalently \jx{The following is not an exact equality, but needs the $o(dt)$ term or similar; also uncapitalize X on the right}
@@
 
-	Assuming that the Markov process \ref{eq:X} is completely observed until time $t_{end}$, we obtain the following complete data likelihood
+	When the Markov process \eqref{eq:X} is completely observed until time $t_{end}$, we obtain the following likelihood
 	\begin{align}
 		L(\theta; \X)
-		& = \prod_{T} \beta I(\tau^T_j) \prod_{J} \gamma \exp\left\lbrace - \int_{0}^{t_{end}}\beta I(t)S(t) + \gamma I(t) dt \right\rbrace  \nonumber \\
+		& = \prod_{j \in \mathcal{I}} \beta I(\tau^I_j) \prod_{k \in \mathcal{R}} \gamma \exp\left\lbrace - \int_{0}^{t_{end}}\beta I(t)S(t) + \gamma I(t) dt \right\rbrace  \nonumber \\
 		\label{eq:cdl}
-		& = \beta^{n_T} \gamma^{n_J}\prod_{J} I(\tau^T_j) \exp\left\lbrace - \int_{0}^{t_{end}}\beta I(t)S(t) + \gamma I(t) dt \right\rbrace
+		& = \beta^{n_I} \gamma^{n_R}\prod_{j \in \mathcal{I}} I(\tau^I_j) \exp\left\lbrace - \int_{0}^{t_{end}}\beta I(t)S(t) + \gamma I(t) dt \right\rbrace
 	\end{align}
-	where $\theta = (\beta, \gamma) \in \chi_{\theta}$  are the parameters of the model and $\chi_{\theta} = (0, \infty) \times (0, \infty)$ is the parameter space, 
-	$T = \{i:\tau^T_i \le t_{end}\}$ and $J = \{i:\tau^J_i \le t_{end}\}$ are the respective index sets of individual infected and removed by time $t_{end}$, 
-	$n_T = |T|$ and $n_J = |J|$ are the respective numbers of observed infections and removals,
-	and	$S(t) = \sum_{i=1}^n \mathbf{1}\{X_i(t) = S\}$ is the number of susceptible individuals at time $t$.
-
-	We see that in this fully observed scenario, it is straightforward to conduct inference.
-The likelihood \ref{eq:cdl} belongs to the exponential family and the maximum likelihood estimates can be expressed in terms of the sufficient statistics defined above as
-	$$\hat{\beta}_{MLE} = \dfrac{n_T}{ \int_{0}^{t_{end}} I(t)S(t)dt}, \quad \hat{\gamma}_{MLE} = \dfrac{n_J}{\int_{0}^{t_{end}} I(t)dt}.$$
+	describing the complete data trajectory.
+	To establish notation,
+	$\theta = (\beta, \gamma) \in \chi_{\theta}$ are the model parameters 
+	and $\chi_{\theta} = (0, \infty) \times (0, \infty)$ denotes the parameter space, 
+	the index sets $\mathcal{I} = \{j\in (1, \cdots, n): \tau^I_j \in (0, t_{end}]\}$ and $\mathcal{R} = \{j\in (1, \cdots, n): \tau^R_j \in (0, t_{end}]\}$ denote the individuals that are respectively infected and removed during the observation interval $(0, t_{end}]$,
+	$n_I = |\mathcal{I}|$ and $n_R = |\mathcal{R}|$ are the numbers of observed infections and removals,
+	and	$S(t) = \#\{X_i(t) = S\}$ is the number of susceptible individuals at time $t$.
+
+	It is straightforward to conduct inference in this continuously observed scenario.
+In particular, the likelihood \eqref{eq:cdl} belongs to the exponential family, and maximum likelihood estimates of its parameters can be expressed in terms of the sufficient statistics defined above as
+	$$\hat{\beta}_{MLE} = \dfrac{n_I}{ \int_{0}^{t_{end}} I(t)S(t)dt}, \quad \hat{\gamma}_{MLE} = \dfrac{n_R}{\int_{0}^{t_{end}} I(t)dt}.$$
 	Since the functions $I$ and $S$ are constant between event times, the two integrals correspond to finite sums which are straightforward to compute.
-	Furthermore, in a Bayesian context, inference is facilitated by the conjugacy of the gamma distribution: if we put independent gamma prior distributions on the parameters
+	Furthermore, in a Bayesian context, inference is facilitated by the conjugacy with the Gamma distribution: if we let the parameters follow independent Gamma prior distributions
 	\begin{equation}
@@
 	\end{equation}
-	where $Ga(a,b)$ denotes the gamma distribution with mean $a/b$ and variance $a/b^2$, then the posterior distributions of the parameters are independent gamma distributions
+	where $Ga(a,b)$ denotes the parametrization with mean $a/b$ and variance $a/b^2$, then the posterior distributions of the parameters are independent Gamma distributions
 	\begin{equation}
 		\label{eq:pdb}
-		\beta | \X \sim Ga\left( a_{\beta} + n_T, b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt\right)
+		\beta | \X \sim Ga\left( a_{\beta} + n_I, b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt\right)
 	\end{equation}
@@
 		\label{eq:pdg}
-		\gamma | \X \sim Ga\left( a_{\gamma} + n_J, b_{\gamma} + \int_{0}^{t_{end}} I(t) dt\right)
+		\gamma | \X \sim Ga\left( a_{\gamma} + n_R, b_{\gamma} + \int_{0}^{t_{end}} I(t) dt\right)
 	\end{equation}
-	from which we can easily generate independent values with a Monte Carlo sampler to explore the posterior distribution $\pi(\theta|\X)$.
+	from which one can easily generate independent values with a Monte Carlo sampler to explore the posterior distribution of parameters $\pi(\theta|\X)$.
 
@@
 
-	(New in this section: I am motivating the type of observed data  $T_{1:K}$ that we consider.)
-
-	In practice, inference is complicated by the fact that the Markov process \ref{eq:X} is only partially observed.
-That is, we do not observe all epidemic events, and do not have access to the sufficient statistics needed to evaluate the likelihood \ref{eq:cdl}.
-Various types of partially observed data have been considered in the literature such as observing only the removal times \cite{Gibson.1998, ONeill.1999} and knowing the number of infectious individuals in the population at discrete points in time \cite{Fintzi.2017}.
-The former type of data for instance arise in animal experiment in which an animal that tests positive for the disease is immediately isolated from the rest of the population and therefore no longer contributes to the spread of the disease.
-In such case, the removal times are exactly known.
-
-	In this article, we consider another type of partially observed data: incidence counts for the infections at discrete points in time.
-Such data arise when an infectious individual may continue to infect others even after being tested positive for the disease.
-This was for instance the case during the $2013$–$2016$ outbreak of Ebola virus.
+	%	(New in this section: I am motivating the type of observed data  $T_{1:K}$ that we consider.)
+
+	In practice, inference is complicated by the fact that the process \eqref{eq:X} is only partially observed.
+When we do not observe all epidemic events, which is virtually always the case with real data, we do not have access to the sufficient statistics needed to evaluate the likelihood \eqref{eq:cdl}.
+Various types of partially observed data typify real data and have been considered in the literature, such as observing only the removal times \cite{Gibson.1998, ONeill.1999} or the number of infectious individuals in the population at discrete points in time \cite{Fintzi.2017}.
+For instance, the former arise in animal experiments in which positive cases are immediately isolated from the rest of the population and no longer contribute to disease spread.
+In such cases, exact times of removals are known, but such information is unavailable in typical observational studies.
+
+	In this article, we focus on partially observed \textit{incidence} counts of infections at discrete points in time.
+Such data arise when an infectious individual may be identified some time after onset, and may continue to infect others even after being tested positive for the disease.
+%This was for instance the case during the $2013$–$2016$ outbreak of Ebola virus.
 This pandemic was unprecedented in scale, with more than $10,000$ reported deaths, mainly in three Western African countries (Guinea, Liberia and Sierra Leone).
@@
 Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.}.
-	It therefore seems appropriate to model a positive test for the virus as an indication that the individual became infectious some time prior to the test rather than as a removal time.
-
+	%It therefore seems appropriate to model a positive test for the virus as an indication that the individual became infectious some time prior to the test rather than as a removal time.
 	Consider an observation schedule $t_{0:K}$ ($K \ge 1$) with $0 = t_0 < t_1 < \dots < t_K = t_{end}$.
-The observed data consist of the $K$-dimensional vector $\Y = T_{1:K}$ where $T_k = \sum_{i=1}^n \mathbf{1}\{\tau^T_i \in (t_{k-1}, t_k]\}$ is the number of infections during the $k^{\text{th}}$ time interval.
+The observed data consist of the $K$-dimensional vector $\Y = I_{1:K}$ where $I_k = \#\{\tau^I_j \in (t_{k-1}, t_k]\}$ is the number of infections during the $k^{\text{th}}$ time interval.
 	For simplicity, we further assume that the initial configuration of the population $\X(0)$ is known.
 
-	If we adopt a Bayesian approach, the posterior distribution of the parameters given the observed data is
-	\begin{align}
+	If we adopt a Bayesian approach, the posterior distribution of the parameters given the observed data is formally related to the joint posterior via integration:
+	\begin{equation}
 		\label{eq:pdl}
 		\pi(\theta|\Y) 
-		& \propto L(\theta; \Y)\pi(\theta) \nonumber \\
-		& = \int_{\chi^*_{\X}} L(\theta; \x) \pi(\theta) d\x
-	\end{align}
+		\, \, \propto \, \, L(\theta; \Y)\pi(\theta) \, \, = \, \, \int_{\chi^*_{\X}} L(\theta; \x) \pi(\theta) d\x
+	\end{equation}
 	where $\pi(\theta)$ is the prior distribution on the parameters $\theta$ and
-	$\chi^*_{\X} \subset \chi_{\X}$ is the set of trajectories of the process \ref{eq:X} that are compatible with the observed data $\Y$.
-	The partial data likelihood $L(\theta; \Y)$ therefore consists of a high-dimensional integral over all epidemic paths compatible with the observed data and has no closed form solution available, even for a population of moderate size.
+	$\chi^*_{\X} \subset \chi_{\X}$ is the set of trajectories of the process \eqref{eq:X} that are compatible with the observed data $\Y$.
+	The partial data likelihood $L(\theta; \Y)$ therefore consists of a high-dimensional integral over all epidemic paths compatible with the observed data.
+This marginalization step has no closed form solution, and presents computational challenges even for a population of moderate size.
 
@@
 	More recently, researchers have based their computation on a simpler process that approximates the model's dynamics and whose likelihood in the presence of partially observed data is tractable.
-These approximation models include binomial chains models \cite{Greenwood.1931, Abbey.1952}, diffusion processes \cite{Cauchemez.2008, Fintzi.2020} and Gaussian processes \cite{Jandarov.2014}.
+Popular approximations include chain binomial models \cite{Greenwood.1931, Abbey.1952}, diffusion processes \cite{Cauchemez.2008, Fintzi.2020} and Gaussian processes \cite{Jandarov.2014}.
 While these approximation-based approaches bypass the intractability of the partial data likelihood \ref{eq:pdl}, the assumptions on which they are based are questionable when the population is small, and, as a result, the dynamic of the stochastic process differs from its asymptotic behavior.
@@
 For instance, Streftaris and Renshaw \cite{Streftaris.2002} generalized the approach of \cite{Gibson.1998} to a non-Markovian stochastic SIR model where the amount of time spent infectious is not necessarily exponentially distributed but follows an arbitrary distribution, and Bu et al.
-\cite{Bu.2020} analyzed models in which the population does not mix homogeneously but is characterized by a dynamic social network.
-
-	\subsection{Proposed Data-Augmented Approach}
+\cite{Bu.2020} analyzed models in which the population does not mix homogeneously but is characterized by a dynamic social network, and Kypraios (2018, 2020) model the infection dynamics nonparameterically.
+
+	\section{Exact Inference with a  Data-Augmented Approach}
 	\label{sec:con}
 
-	TODO: Add short intro to not intimidating reader with notation.
-
-	As shown in Section \ref{sec:icd}, the complete data likelihood \ref{eq:cdl} is amenable to computation.
-This suggests augmenting the observed data $\Y$ with latent data $\Z$ such that the likelihood $L(\theta; (\Y, \Z))$ has the closed form \ref{eq:cdl} and constructing a Markov chain on the space $\chi = \chi_{\theta} \times \chi_{\Z}$ whose stationary distribution is the joint posterior distribution $\pi(\theta, \X|\Y)$ \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
-We consider the latent data $\Z = \left\lbrace (z^T_i, z^J_i)\right\rbrace_{i=1}^{n} \in \chi_{\Z}$ which consist of the times and types of unobserved epidemic events, where the infection time of individual $i$ is
-	$$z^T_i \begin{cases}
-		\in [0, t_{end}) & \text{if individual } i \text{ becomes infectious before } t_{end} \\
-		= \infty & \text{if individual } i \text{ does not become infectious before } t_{end} \\
+	We adopt a data-augmented MCMC approach that bridges the challenging partially observed setting to the tractable complete data likelihood by way of latent variables.
+In particular, our method hinges on the efficacy of a carefully designed proposal process for the latent variables that faithfully approximates the original process dynamics.
+	%By correcting with a Metropolis-Hastings step
+	The sampler targets the exact posterior of the model given partially observed incidence data and enjoys the efficiency of fast proposals in the high-dimensional latent space as well as fast computations involving the complete data likelihood.
+
+	As shown in Section \ref{sec:icd}, the complete data likelihood \eqref{eq:cdl} is amenable to computation.
+This suggests augmenting the observed data $\Y$ with latent data $\Z$ such that the likelihood $L(\theta; (\Y, \Z))$ has the closed form \eqref{eq:cdl} and constructing a Markov chain $\{(\theta^{(m)}, \Z^{(m)})\}_{m=0}^M$ with state space $\chi = \chi_{\theta} \times \chi_{\Z}$ whose stationary distribution is the joint posterior distribution $\pi(\theta, \X|\Y)$ \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
+We consider the latent data $\Z = \left\lbrace (z^I_j, z^R_j)\right\rbrace_{j=1}^{n} \in \chi_{\Z}$ which consist of the times and types of unobserved epidemic events, where the infection and removal times for individual $j$ are
+	$$z^I_j \begin{cases}
+		\in [0, t_{end}) & \text{if individual } j \text{ becomes infectious before } t_{end} \\
+		= \infty & \text{if individual } j \text{ does not become infectious before } t_{end} \\
 	\end{cases}$$
-	and the removal time of individual $i$ is
-	$$z^J_i \begin{cases}
-		\in (z_{T_i}, t_{end}] & \text{if individual } i \text{ is removed before } t_{end} \\
-		= \infty & \text{if } z_{T_i} = \infty \text{ or individual } i \text{ is removed after } t_{end}. \\
+	and
+	$$z^R_j \begin{cases}
+		\in (z^I_j, t_{end}] & \text{if individual } j \text{ is removed before } t_{end} \\
+		= \infty & \text{if } z^I_j = \infty \text{ or individual } j \text{ is removed after } t_{end}. \\
 	\end{cases}$$
@@
 
-	The data-augmented MCMC algorithm that we propose alternates between updates of the parameters $\theta$ given the current state of the latent data $\Z$, and updates of the latent data given the current state of the parameters and the observed data $\Y$.
+	%    \jx{Brief transition by saying that drawing from $\chi_Z$ would require conditional simulation, remind reader of the Hobolth stone reference/mention that Lam's method could in theory draw from these conditionals, but it's a hard problem computationally that becomes intractable and instead we will use Metropolis-Hastings with a fast proposal for Z.
+This was all mentioned in the intro but can be recapped in one/two sentences.}
+
+	The data-augmented MCMC algorithm that we propose alternates between updates of the parameters $\theta$ given the current state of the latent data $\Z$, and updates of the latent data given the current state of the parameters.
 A one-step transition from $\x_0 = (\theta_0, \z_0)$ to $\x_1 = (\theta_1, \z_1)$ therefore looks like
@@
 
-	The transition kernel $P$ of the Markov chain $\{\x_i\}_i$ is thus a composition of two kernels $P_{\theta}$ and $P_{\z}$ which respectively update $\theta$ while keeping the $\z$ fixed, and update $\z$ while keeping $\theta$ fixed.
-Due to the gamma conjugacy of the complete-data likelihood mentioned in Section \ref{sec:icd}, sampling new parameter values poses no challenge.
-We can simply employ a Gibbs sampler to directly draw new values for $\theta = (\beta, \gamma)$ from the two independent full conditional posterior distributions given by Equations \ref{eq:pdb} and \ref{eq:pdg}.
-The kernel
-	$$
-	P_{\theta}((\theta_0, \z), (\theta_1, \z)) = Ga\left( \beta_1; a_{\beta} + n_T, b_{\beta} + \int_0^{t_{end}} S(t)I(t) dt\right) Ga\left( \gamma_1; a_{\gamma} + n_J, b_{\gamma} + \int_0^{t_{end}} I(t) dt\right) d\beta_1 d\gamma_1
-	$$
-	therefore corresponds to a Gibbs kernel where $n_T, n_J, \int S(t)I(t) dt, \int I(t) dt$ are sufficient statistics from the latent data $\z$. The kernel
-	$$
-	P_{\theta}((\theta, \z_0), (\theta, \z_1)) = q(\z_1|\theta) \alpha((\theta, \z_0), (\theta, \z_1)) dz_1
-	$$
-	corresponds to a Metropolis-Hastings kernel where the proposal density $q$ will be defined in the following section and
+	The transition kernel $P$ of the Markov chain $\{\x^{(m)}\}_m$ is thus a composition of two kernels $P_{\theta}$ and $P_{\z}$ which alternatively updates $\theta$ while keeping $\z$ fixed, and then updates $\z$ while keeping $\theta$ fixed.
+Due to the Gamma conjugacy of the complete-data likelihood mentioned in Section \ref{sec:icd}, drawing new parameter values poses no challenge: we can simply employ a Gibbs sampler to directly draw new values for $\theta = (\beta, \gamma)$ from the two independent full conditional posterior distributions given by Equations \ref{eq:pdb} and \ref{eq:pdg}.
+In these expressions, $n_I, n_R, \int S(t)I(t) dt, \int I(t) dt$ are sufficient statistics from the latent data $\z$.
+The latent data is updated in a Metropolis-Hastings step whose kernel is
+	$$
+	P_{\z}((\theta, \z_0), (\theta, \z_1)) = q(\z_1|\theta) \alpha((\theta, \z_0), (\theta, \z_1)) dz_1
+	$$
+	where the proposal density $q$ will be defined in the following section and
 	$$
@@
 	corresponds to the Metropolis-Hasting acceptance ratio \cite{Tierney.1994}.	
-	Algorithm \ref{alg:DA-MCMC} provides the details of the procedure.
+	Algorithm \ref{alg:DA-MCMC} provides the details of the Data-Augmented MCMC.
 
@@
 		\begin{algorithmic}
-			\REQUIRE $\theta^{(1)}$
-			\STATE $\Z^{(1)} \leftarrow draw(\Z|\theta^{(1)})$ (generate initial latent space)
-			\FOR {$j = 2, \dots, N$}			
-			\STATE $\beta^{(j)}|Z^{(j - 1)} \sim Ga\left( a_{\beta} + n_T^{(j - 1)}, b_{\beta} + \int_{0}^{t_{end}} I^{(j - 1)}(t)S^{(j - 1)}(t)dt\right)$ (Gibbs update)
-			\STATE $\gamma^{(j)}|Z^{(j-1)} \sim Ga\left( a_{\gamma} + n_J^{(j - 1)}, b_{\gamma} + \int_{0}^{t_{end}} I^{(j - 1)}(t) dt\right)$ (Gibbs update)
+			\REQUIRE $\theta^{(0)}$
+			\STATE $\Z^{(0)} \sim q(\Z|\theta^{(0)})$ (generate the initial latent space)
+			\FOR {$j = 1, \dots, N$}
+			\STATE $\beta^{(j)}|Z^{(j - 1)} \sim Ga\left( a_{\beta} + n_I^{(j - 1)}, b_{\beta} + \int_{0}^{t_{end}} I^{(j - 1)}(t)S^{(j - 1)}(t)dt\right)$ (Gibbs update)
+			\STATE $\gamma^{(j)}|Z^{(j-1)} \sim Ga\left( a_{\gamma} + n_R^{(j - 1)}, b_{\gamma} + \int_{0}^{t_{end}} I^{(j - 1)}(t) dt\right)$ (Gibbs update)
 			\STATE $\theta^{(j)} \leftarrow (\beta^{(j)}, \gamma^{(j)})$
-			\STATE $\Z^* \leftarrow draw(\Z|\theta^{(j)})$ (generate latent data from the PD-SIR process)
-			\STATE $\alpha = \min\left\lbrace 1, \dfrac{L(\theta^{(j)}; \Z^*)q(\Z^{(j - 1)}|\theta^{(j)})}{L(\theta; \Z^{(j - 1)})q(\Z^*|\theta^{(j)})}\right\rbrace  $
+			\STATE $\Z^\star \sim q(\Z|\theta^{(j)})$ (generate latent data from the PD-SIR process)
+			\STATE $\alpha = \min\left\lbrace 1, \dfrac{L(\theta^{(j)}; \Z^\star)q(\Z^{(j - 1)}|\theta^{(j)})}{L(\theta^{(j)}; \Z^{(j - 1)})q(\Z^\star|\theta^{(j)})}\right\rbrace  $
 			\STATE $u \sim U(0,1)$
@@
 
-	\section{Efficient Proposal Process for Latent Data}
-	Simulating trajectories from discrete state Markov processes, such as the stochastic SIR process, that are compatible with partially observed data is, in general, a notoriously difficult task \cite{Hobolth.2009}.
-Instead of performing this conditional simulation from the SIR model directly, we will conditionally simulate from a surrogate process and then correct the approximation via a Metropolis-Hastings step.
-
-	\subsection{The Piecewise-decoupled SIR Process}
-	\label{sec:pds}
+	\subsection{Efficient Proposal Process for Latent Data}	\label{sec:pds}
+	Though forward simulation of the stochastic SIR process is straightforward, 
+	simulating trajectories \textit{conditionally} on the observed incidence data $\Y$ is a notoriously difficult task \cite{Hobolth.2009}.
+Instead, we will generate the latent variables conditionally on $\Y$ from a surrogate process, and accept or reject them in a Metropolis-Hastings step.
+
+	%\subsection{The Piecewise-decoupled SIR Process}
 	% Outline: define PD-SIR; generate PD-SIR
-	To generate the latent data, we consider a stochastic process whose dynamics closely resemble those of the SIR process and from which we can efficiently generate an epidemic that is compatible with the discretely observed data $\Y = T_{1:K}$.
+	To this end, we consider a stochastic process whose dynamics closely resemble those of the SIR process and which is designed for efficient simulation of epidemic trajectories compatible with the incidence data $\Y = T_{1:K}$.
 We refer to this surrogate process as the \textit{piecewise decoupled SIR} process (PD-SIR).
-	Similarly to the SIR process, the PD-SIR process corresponds to a compartmental model in which individuals move from $S$ to $I$ and from $I$ to $R$.
-	The removal dynamics are the same under both processes: infection periods follow independent exponential distributions with rate $\gamma$, resulting in a constant individual-level removal rate $\gamma$.
+	Similarly to the SIR process, the PD-SIR process corresponds to a compartmental model in which individuals move from the compartments $S$ to $I$ and from $I$ to $R$.
+	The removal dynamics are identical under both processes: infection periods follow independent exponential distributions with rate $\gamma$, resulting in a constant individual-level removal rate $\gamma$.
+
 	The infection dynamics, however, vary slightly.
-	In the SIR process, the population-level infection rate at time $t$ is $\mu_T(t) = \beta S(t)I(t)$ and, form the perspective of a susceptible individual, the individual-level infection rate is $\mu(t) = \frac{\mu_T(t)}{S(t)} = \beta I(t)$.
-Note that $\mu$ varies after each event since the variable $I$ changes after each infection and removal.
-	In contrast, in the PD-SIR process, $\mu$ is kept constant over small time intervals.
-
+	In the SIR process, the population-level infection rate at time $t$ is $\mu_T(t) = \beta S(t)I(t)$ and so, from the perspective of each susceptible individual, the individual-level infection rate is $\mu(t) = \frac{\mu_T(t)}{S(t)} = \beta I(t)$.
+Note that $\mu$ varies after every event since the value of $I$ changes after an infection or a removal.
+	In contrast, in the PD-SIR process, the infection rate is kept constant over small time intervals.
 Consider a resetting schedule $r_{0:L}$ ($L\ge1$) with $0 = r_0 < r_1 < \dots < r_L$.
-Under the PD-SIR process the individual-level infection rate $\mu$ is set to $\beta I(r_{k - 1})$ for $t \in [r_{k - 1}, r_k)$ where $I(r_{k-1})$ denotes the number of infectious individuals at the beginning of the $k^{\text{th}}$ time interval; that is, the individual infection rate is kept constant over each interval of the resetting schedule.
-The infection rate $\mu$ and the evolution of the variable $I$ are therefore decoupled during each interval, and the infection rate is reset at their beginning.
-Over a single observation interval, the PD-SIR process is equivalent to the two-type branching process proposed in \cite{Ho.2018} to approximate SIR dynamics.
-
-	If we let the resetting schedule coincide with the observation schedule $t_{0:K}$,
-	$$L = K, \qquad r_k = t_k, \quad k=1,\dots,K,$$
-	then we can straightforwardly generate a realization from the PD-SIR process that is compatible with the observed epidemic data $T_{1:K}$.
-Under the PD-SIR model, the infections occurring during the $k^{\text{th}}$ interval follow a linear death process with rate $\mu_k = \beta I(t_{k-1})$.
+The individual-level infection rate of the PD-SIR process is defined as
+	$$\tilde{\mu}(t) = \beta I(r_{k - 1}), \forall t \in [r_{k - 1}, r_k)$$
+	where $I(r_{k-1})$ denotes the number of infectious individuals at the beginning of the time interval.
+That is, the infection rate is kept constant over short periods of time.
+The infection rate $\mu$ and the evolution of the variable $I$ are therefore decoupled during each interval, and the infection rate is reset at the beginning %or left endpoint
+	of the intervals.
+Over a single interval, the PD-SIR process is equivalent to a two-type branching process approximation of the SIR dynamics \cite{Ho.2018}.
+\jx{I really like how this is written but I wonder if there's some simple figure we can make to illustrate the model and difference between SIR at a glance => line plot of mu over time with colored vertical line for every event.
+Done.}.
+
+	\begin{figure}
+		\centering
+		\includegraphics[scale = 0.6]{infection_rate_SIR_PDSIR.jpg}
+		\caption{Infection dynamics of the SIR (dashed line) and PD-SIR (solid line) processes in a small population ($(S(0), I(0)) = (10,2)$).
+			$(\beta, \gamma) = (0.2, 1)$ and the resetting schedule of the PD-SIR contains $L=5$ intervals of equal length.
+			The vertical dotted lines indicate the infection (red) and removal (blue) times.}
+		\label{fig:mu}
+	\end{figure}
+
+	Now, by letting the resetting schedule $r_{0:L}$ coincide with the observation schedule $t_{0:K}$,
+	$$L = K, \text{ and } r_k = t_k, \quad k=1,\dots,K,$$
+	we can straightforwardly simulate realizations from the PD-SIR process conditionally on the observations $I_{1:K}$.
+By construction, under the PD-SIR model the population of susceptible individuals $S(t)$ follows a linear death process during each time interval with death rate $\mu_k = \beta I(t_{k-1})$.
 Generating event times from a linear death process conditionally on the number of events happening by a certain time can be done extremely efficiently as shown by the following theorem whose proof is presented in Appendix \ref{app:ldp}.
@@
 		\label{theo:ldp}
-		Consider a linear death process with death rate $\mu$ and let $d_{1:T} \in (t_l, t_u]$ be the times of the $T$ deaths occurring between times $t_l$ and $t_u$. Then 
-		$$d_i \,{\buildrel d \over =}\, X_{(i)}, \quad i = 1, \dots, T$$
-		where $X_{(i)}$ is the $i^{\text{th}}$ order statistics of $T$ iid random variables following a truncated exponential distribution with rate $\mu$, lower bound $t_l$ and upper bound $t_u$.
+		Consider a linear death process with death rate $\mu$ and let $d_{1:T} \in (t_l, t_u]$ be the times of the $N$ deaths occurring between times $t_l$ and $t_u$. Then 
+		$$d_j \,{\buildrel d \over =}\, X_{(j)}, \quad j = 1, \dots, N$$
+		where $X_{(j)}$ is the $j^{\text{th}}$ order statistics of $N$ i.i.d.\ random variables following a truncated exponential distribution with rate $\mu$, lower bound $t_l$ and upper bound $t_u$.
 	\end{theorem}
 
-	We generate latent data  $\z = \left\lbrace (z^T_i, z^J_i)\right\rbrace_{i=1}^{n}$ compatible with $\Y = T_{1:K}$ from the PD-SIR process as follows.
-For each time interval $[t_{k-1}, t_k)$, we compute $\mu_k = \beta I(t_{k-1})$ and $U_k = I(0) + \sum_{l=1}^{k-1} T_l$, the cumulative count of infections at the beginning of interval $k$.
-Note that $I(t_{k-1})$ only depends on past events and can therefore be computed at each given the PD-SIR process up to time $t_{k-1}$.
-Algorithm \ref{alg:PD-SIR} provides a simple recursive to compute this variable (referred to as $I_k$).
-For $i = U_k + 1, \dots, U_k + T_k$, following Theorem \ref{theo:ldp}, we generate the infection times $z^T_i$ as the order statistics of $T_k$ iid random variables following a truncated exponential distribution with rate $\mu_k$, lower bound $t_{k-1}$ and upper bound $t_k$.
-For the same indices $i$, we then generate the removal times $z^J_i$ from the mixture distribution
-	$$(1 - p_i) \delta_{\infty} + p_i \TruncExp(\gamma, z^T_i, t_{end}) $$
+	We can use this result to generate latent data  $\z = \left\lbrace (z^I_i, z^R_i)\right\rbrace_{i=1}^{n}$ compatible with $\Y = I_{1:K}$ from the PD-SIR process as follows.
+For each time interval $[t_{k-1}, t_k)$, we compute $\mu_k = \beta I(t_{k-1})$ and $U_k = I(0) + \sum_{l=1}^{k-1} I_l$, the total number of infections that have happened before $t_{k-1}$.
+The index set $\mathcal{I}_k = (U_k + 1, \dots, U_k + I_k)$ therefore denotes the $I_k$ individuals infected during the $k$th interval.
+Note that $I(t_{k-1})$ only depends on past events and can therefore be computed given the PD-SIR process up to time $t_{k-1}$.
+Algorithm \ref{alg:PD-SIR} provides a simple recursion to compute this variable.
+For $j \in \mathcal{I}_k$, by following Theorem \ref{theo:ldp}, we draw the infection times $z^I_j$ as the order statistics of $I_k$ i.i.d.\ truncated exponential random variables with rate $\mu_k$, lower bound $t_{k-1}$ and upper bound $t_k$.
+For the same indices $j$, we then generate the removal times $z^R_j$ from the removal dynamics of the SIR.
+To accomplish this, we propose the removal time of individual $j$ from the mixed distribution
+	$$(1 - p_j) \delta_{\infty} + p_j \TruncExp(\gamma; z^I_j, t_{end})$$
+	placing point mass $(1 - p_j)$ at $\infty$ and continuous mass on $(z^I_j, t_{end}]$,
 	where $\delta_{\infty}$ corresponds to the Dirac distribution with mass $1$ on the element $\infty$,
-	$$p_i = 1 - \exp\{-\gamma (t_{end} - z^T_i)\}$$
-	is the cumulative distribution function of an exponential distribution with rate $\gamma$ and corresponds to the probability that particle $i$ is removed before $t_{end}$ given that it was infected at time $z^T_i$, and $\TruncExp(\gamma, z^T_i, t_{end})$ denotes a truncated exponential distribution with rate $\gamma$, lower bound $z^T_i$ and upper bounde $t_{end}$.
-By construction, this scheme generates a latent space from the PD-SIR process that is compatible with the observed data.
-Algorithm \ref{alg:PD-SIR} gives an algorithmic representation of the procedure.
+	$$p_j = 1 - \exp\{-\gamma (t_{end} - z^I_j)\} = P(z^R_j < t_end | z^I_j)$$
+	is the cumulative distribution function of an exponential distribution with rate $\gamma$, % and corresponds to the probability that particle $j$ is removed before $t_{end}$ given that it was infected at time $z^I_j$,
+	and $\TruncExp(\gamma; z^I_j, t_{end})$ denotes a truncated exponential distribution with rate $\gamma$ bounded between $z^I_j$ and  $t_{end}$.
+By construction, this scheme generates latent data from the PD-SIR process that are compatible with the observed data.
+Algorithm \ref{alg:PD-SIR} provides a summary of the procedure in pseudo-code.
 
 	\begin{algorithm}
-		\caption{Generating a PD-SIR process conditionally on discretely observed infection incidence counts $T_{1:K}$}
+		\caption{Generating a PD-SIR process conditionally on discretely observed infection incidence counts $I_{1:K}$}
 		\label{alg:PD-SIR}
 		\begin{algorithmic}
-			\REQUIRE $T_{1:K}, \theta = (\beta, \gamma)$, $I(0)$
-			\FOR {$i = 1, \dots, I(0)$}
-			\STATE $z^T_i \leftarrow 0$ (by the memoryless property of the exponential distribution).
-			\STATE $p_i \leftarrow 1 - \exp\{-\gamma (t_{end} - 0)\}$
-			\STATE $z^J_i \sim (1 - p_i) \delta_\infty + p_i \TruncExp(\gamma, 0, t_{end})$
+			\REQUIRE $I_{1:K}, \theta = (\beta, \gamma)$, $I(0)$
+			\FOR {$j = 1, \dots, I(0)$}
+			\STATE $z^I_j \leftarrow 0$ (by the memoryless property of the exponential distribution).
+			\STATE $p_j \leftarrow 1 - \exp\{-\gamma (t_{end} - 0)\}$
+			\STATE $z^R_j \sim (1 - p_j) \delta_\infty + p_j \TruncExp(\gamma, 0, t_{end})$
 			\ENDFOR
-			\STATE $I_1 \leftarrow T_1$
+
 			\FOR {$k = 1, \dots, K$}	
-			\STATE $\mu_k \leftarrow \beta I_k$
-			\STATE $X_{1 : T_k} \sim \TruncExp(\mu_k, t_{k-1}, t_k)$ independently
-			\FOR {$i = U_k + 1, \dots, U_k + T_k$}
-			\STATE $z^T_i \leftarrow X_{(i)}$ (infection times)
-			\STATE $p_i \leftarrow 1 - \exp\{-\gamma(t_{end} - z^T_i)\}$
-			\STATE $z^J_i \sim (1 - p_i) \delta_\infty + p_i \TruncExp(\gamma, z^T_i, t_{end})$ (removal times)
+			\STATE $\mu_k \leftarrow \beta I(t_{k-1})$
+			\STATE $X_{1 : I_k} \sim \TruncExp(\mu_k, t_{k-1}, t_k)$ i.i.d.\
+			\STATE $U_k \leftarrow I(0) + \sum_{l=1}^{k-1} I_l$
+			\STATE $\mathcal{I}_k \leftarrow (U_k + 1, \dots, U_k + I_k)$
+
+			\FOR {$j \in \mathcal{I}_k$}
+			\STATE $z^I_j \leftarrow X_{(j)}$ (infection times)
+			\STATE $p_j \leftarrow 1 - \exp\{-\gamma(t_{end} - z^I_j)\}$
+			\STATE $z^R_j \sim (1 - p_j) \delta_\infty + p_j \TruncExp(\gamma, z^I_j, t_{end})$ (removal times)
 			\ENDFOR
-			\STATE $J_k \leftarrow \sum_{i=1}^n \mathbf{1}\{\tau^J_i \in (t_{k-1}, t_k]\}$ (number of removals in the $k^{\text{th}}$ interval)
-			\STATE $I_{k+1} \leftarrow I_k + T_k - J_k$
+			\STATE $R_k \leftarrow \#\{i:\tau^R_i \in (t_{k-1}, t_k]\}$ (number of removals in the $k^{\text{th}}$ interval)
+			\STATE $I(t_k) \leftarrow I(t_{k-1}) + I_k - R_k$
 			\ENDFOR
@@
 
-	The density $q$ of the PD-SIR process is then
-	\begin{align*}
-		q(\Z|\theta) =
-		& \prod_{k=1}^K \prod_{j=U_k + 1}^{U_k + T_k} \TruncExp(\tau^T_j; \mu_{k}, t_{k-1}, t_{k}) \\
-		& \times \prod_{i=1}^{n} \left( 1 - p_i \right)^{\mathbf{1}(z^J_i = \infty)} \left( p_i \TruncExp(z^J_i; \gamma, z^T_i, t_{end}) \right)^{\mathbf{1}(z^J_i \le t_{end})}
-	\end{align*}
-	where $\mu_k = \beta I(t_{k-1})$ and
+	Based on this construction, we see that the density $q$ of the PD-SIR process is
+	\begin{align*}
+		q(\z|\theta) & = && \prod_{k=1}^K \prod_{j\in\mathcal{I}_k} \TruncExp(\tau^I_j; \mu_{k}, t_{k-1}, t_{k}) \\
+		& && \times \prod_{i=1}^{n} \left( 1 - p_i \right)^{\mathbf{1}(z^R_i = \infty)} \left( p_i \TruncExp(z^I_i; \gamma, z^I_i, t_{end}) \right)^{\mathbf{1}(z^R_i \le t_{end})} \\
+		& = && \prod_{k=1}^K \prod_{j\in\mathcal{I}_k} \TruncExp(\tau^I_j; \mu_{k}, t_{k-1}, t_{k}) \\
+		& && \times \prod_{i=1}^{n} \left( 1 - p_i \right)^{\mathbf{1}(z^R_i = \infty)} \left( p_i \TruncExp(z^I_i; \gamma, z^I_i, t_{end}) \right)^{\mathbf{1}(z^R_i \le t_{end})}
+	\end{align*}
+	where %$\mu_k = \beta I(t_{k-1})$ and
 	$$\TruncExp(x; \mu, l, u) = \dfrac{\mu \exp\{-\mu x\}}{\exp\{-\mu l\} - \exp\{-\mu u\}}, \quad x \in (l, u)$$
-	denotes the density of a truncated exponential distribution with rate $\mu$, lower bound $l$ and upper bound $u$ at the value $x$.
-
-	The following characteristics of our DA-MCMC algorithm are worth noting.
-	First, initializing the Markov chain only requires setting $\theta^{(1)}$ since  $\Z^{(1)}$ can be generated from the PD-SIR process conditionally on $\theta^{(1)}$ and $\Y$ alone.
-	Second, since the PD-SIR closely approximates the SIR model, our DA-MCMC algorithm can update a large portion of the latent data per iteration while keeping a relatively high acceptance rate.
-As a result, our algorithm makes larger jumps in the high-dimensional latent space and therefore explores it more efficiently.
-If, however, the acceptance rate of the latent data in the Metropolis-Hastings steps is too low -- which will be shown to occur when the size of the latent space grows -- then it is possible to update the event times of only a fraction $0<\rho\le 1$ of the individuals in the population in order to increase the acceptance rate and improve mixing.
-	Third, if all event times are updated ($\rho = 1$), then the current and proposed latent spaces $\Z^{(k-1)}$ and $\Z^*$ are independent conditionally on the current values of the parameters: $\Z^{(k-1)} \perp \Z^* | \theta^{(k)}$.
-This contrasts with existing DA-MCMC algorithms that only update a small fraction of the latent data per iteration.
-	Fourth, since the latent data are generated from a process that only resembles the SIR model, we ensure that the Markov chain converges to the correct distribution $\pi(\theta, \Z|\Y)$ by proposing and accepting the latent data according to a Metropolis-Hastings scheme.
-This results in exact Bayesian inference targeting the posterior distribution under the original SIR model.
-	Fifth, not only is proposing latent data from the RD-SIR process extremely fast since all random variables can be generated via the inverse-CDF method, but ensuring that these data are compatible with the observed data can be done at no additional cost, making the proposal scalable to large outbreaks.
-	Finally, our algorithm generates a Markov chain that is uniformly ergodic, which is the strongest form of ergodicity for a Markov chain \cite{Tierney.1994}.
-The proof of this result is provided in Appendix \ref{app:uni}.
-	% our DA-MCMC does only requires I(0), not S(0).
-
-	\subsection{Quality of Approximating Process}
+	denotes the density of a truncated exponential distribution with parameters as notated previously.
+
+	\subsection{Quality of Approximation}
 	\label{sec:qua}
@@
 
-	Since we use a proposal for the latent data that is independent from that at of the previous iteration, the efficiency of the DA-MCMC algorithm to explore the latent space directly depends on how similar the surrogate process is to the target process.
+	Since the proposal for the latent data is independent from the current configuration of the latent data, the efficiency of the DA-MCMC algorithm to explore the latent space directly depends on how similar the surrogate process is to the target process.
 In particular, the Markov chain will have good mixing properties if the PD-SIR process is a close approximation of the SIR process.
 The PD-SIR process only differs from the SIR process in its infection dynamics, the removal dynamics being identical in the two processes.
-Figure \ref{fig:comparison} compares the trajectories of the compartments $S$, $I$ and $R$ of a SIR process of moderate size ($S(0) = 1,000$, $I_(0) = 10$, $R(0) = 0$, $(\beta, \gamma) = (0.003, 1)$ and $t_{end} = 6$) and those of four PD-SIR processes constrained to be compatible with the infection incidence data $T_{1:K}$ for $K \in (5, 10, 50, 1,000)$ observed in the SIR process.
+Figure \ref{fig:comparison} compares the trajectories of the compartments $S$, $I$ and $R$ of a SIR process of moderate size ($(S(0) = 1000, I(0), R(0)) = (1000, 10, 0)$ with $(\beta, \gamma) = (0.003, 1)$ and $t_{end} = 6$) and those of four PD-SIR processes constrained to be compatible with the observed incidence data $I_{1:K}$ from the SIR process for $K \in (5, 10, 50, 1,000)$.
 We see that the PD-SIR is qualitatively close to the SIR, even for small values of $K$.
-Moreover, the quality of the approximation improves as $K$ increases.
+Unsurprisingly, the quality of the approximation improves as $K$ increases.
 	% In the limit as $K \rightarrow \infty$, the infection times are known.
-	The fact that the PD-SIR is a faithful approximation of the SIR enables the algorithm to update a large portion of the augmented data at each iteration of the Markov chain while maintaining a relatively high acceptance rate in the Metropolis-Hastings step.
-	% In contrast, existing DA-MCMC algorithm updates only a few event times per iteration.
-As a result, our MCMC algorithm explores the latent space much more efficiently.
-Moreover, even though the latent spaces are generated from an approximation of the SIR process, the Metropolis-Hastings step ensures that the Markov chain converges to the posterior distribution of the parameters under the SIR model; in other words, the inference is exact.
+	The fact that the PD-SIR is a faithful approximation of the SIR leads to a relatively high acceptance rate in the Metropolis-Hastings step for the latent data.
+As a result, our algorithm makes larger jumps in the high-dimensional latent space and therefore explores it more efficiently.
+	%Moreover, even though the latent spaces are generated from an approximation of the SIR process, the Metropolis-Hastings step ensures that the Markov chain converges to the posterior distribution of the parameters under the SIR model; in other words, the inference is exact.
 
@@
 			\centering
-			\includegraphics[width=\textwidth]{E2_K5.jpg}
-			\caption{PD-SIR process ($K = 5$)}
+			\includegraphics[width=\textwidth]{E2_K5}
+			\caption{$K = 5$}
 			\label{fig:comparison_RD_SIR_K5}
@@
 			\centering
-			\includegraphics[width=\textwidth]{E2_K10.jpg}
-			\caption{PD-SIR process ($K = 10$)}
+			\includegraphics[width=\textwidth]{E2_K10}
+			\caption{$K = 10$}
 			\label{fig:comparison_RD_SIR_K10}
@@
 			\centering
-			\includegraphics[width=\textwidth]{E2_K50.jpg}
-			\caption{PD-SIR process ($K = 50$)}
+			\includegraphics[width=\textwidth]{E2_K50}
+			\caption{$K = 50$}
 			\label{fig:comparison_RD_SIR_K50}
@@
 			\centering
-			\includegraphics[width=\textwidth]{E2_K1000.jpg}
-			\caption{PD-SIR process ($K = 1000$)}
+			\includegraphics[width=\textwidth]{E2_K1000}
+			\caption{$K = 1000$}
 			\label{fig:comparison_RD_SIR_K1000}
@@
 	\end{figure}
-
-	\section{Performance on simulated and real epidemic data}
-	\label{sec:per}
-
-	\subsection{Simulation Study}
-	\label{sec:sim}
-	%Outline: setup; results: posterior mean, credible interval, frequentist coverage, run time, ACF, acceptance rate; table; figure: traceplot, ACF; tuning: update proportion v.
-acceptance rate; compare with others methods e.g.
-chain binomial, exact proposal.
-
-	% 1. convergence in medium-size population
-	% 2. rho
-	% 3. nominal coverage should be 95%.
-	% 4. distribution of estimate for PD-SIR (unbiased) vs. LNA (biased)
-	% explain why not compare chain binomial, particle filtering (forward simulation).
-
-	In this section, the convergence properties of the DA-MCMC algorithm are examined with a simulation study.
-Each data set in this study is simulated from the stochastic SIR model and the prior distributions on the parameters $\beta$ and $\gamma$ are independent weakly informative gamma distribution with scale and shape parameters set to $0.01$ (see Equation \ref{eq:pri}).
-
-	First, the mixing property of the Markov chain is assessed in a medium-size population of $S(0)=1,000$ initially susceptible and $I(0)=10$ initially infectious individuals.
-The parameters are $(\beta, \gamma) = (0.003, 1)$ and the process is observed until time $t_{end} = 6$ when the pandemic has completed most of its course but is not over yet (see Figure \ref{fig:long}, where $I(t_{end}) = 55$).
-The numbers of infections in $K = 10$ intervals of equal length are observed and correspond to $T_{1:K} = (40, 111, 193, 259, 178, 93, 29, 19, 9, 6)$.
-The Markov chain is run for $100,000$ iterations from the initial values $(\beta^{(1)}, \gamma^{(1)}) = (0.0003, 0.1) = (\beta/10, \gamma/10)$ and the event times of $202$ ($\rho = 0.2$) individuals are updated in the augmented data in each Metropolis-Hastings step.
-The algorithm took less than $5$ minutes to complete on a personal laptop.
-Figure \ref{fig:traceplot} shows the traceplots of the parameters $\beta$ and $\gamma$, the log likelihood and the basic reproduction number $R_0 = S(0) \beta / \gamma$, which corresponds to the average number of secondary infections caused by an infectious individual at the beginning of the outbreak.
-We observe that the Markov chain quickly migrates from the low density region of the space where it started to a high density region where it mixes relatively well.
-The acceptance rate of the latent spaces proposed in the Metropolis-Hastings step is $0.11$.
-	Excluding the first $10,000$ iterations of the Markov chain as a burn-in, the posterior means of $\beta$, $\gamma$ and $R_0$ are respectively $0.00304$, $0.995$ and $3.07$.
-
-	% TODO: Results: convergence (R statistic, Geweke), credible intervals.
-	% TODO: Tables: posterior means, credible intervals, effective sample size.
-
-	% TODO: frequentist coverage for 3 different observation schedules
-
-	% TODO: Geweke
-
-	\begin{figure}
-		\centering
-		\includegraphics[scale = 0.3]{E1_trajectories.jpg}
-		\caption{Trajectories of the $S$, $I$ and $R$ compartments of a SIR process for a medium-size population ($S_0 = 1,000$).}
-		\label{fig:long}
-	\end{figure}
-
-	\begin{figure}
-		\centering
-		\begin{subfigure}[b]{0.41\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E1_no_burn_beta_tp.jpg}
-			\caption{$\beta$}
-			\label{fig:traceplot_beta}
-		\end{subfigure}
-		\hfill
-		\begin{subfigure}[b]{0.41\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E1_no_burn_gamma_tp.jpg}
-			\caption{$\gamma$}
-			\label{fig:traceplot_gamma}
-		\end{subfigure}
-		\\
-		\begin{subfigure}[b]{0.41\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E1_no_burn_loglik_tp.jpg}
-			\caption{Log likelihood}
-			\label{fig:traceplot_loglik}
-		\end{subfigure}
-		\hfill
-		\begin{subfigure}[b]{0.41\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E1_no_burn_R0_tp.jpg}
-			\caption{$R_0$}
-			\label{fig:traceplot_R0}
-		\end{subfigure}
-		\caption{Traceplots of the Markov chain (including burn-in).}
-		\label{fig:traceplot}
-	\end{figure}
-
-	Second, the impact of the tuning parameter $\rho$ on the performance of the MCMC algorithm is evaluated.
-In particular, we examine the acceptance rate of the Metropolis-Hastings step, the total run time of the algorithm and the auto-correlation function of the parameters.
-	Three SIR processes with different population sizes $S(0) = (10^2, 10^3, 10^4)$ are simulated.
-We set $I(0) =10$, $\gamma=1$, $t_{end} = 3$ and $K = 50$ for all simulations and $\beta = (0.001, 0.01, 0.1)$ respectively for the three different population sizes.
-These values for $\beta$ ensure that the three processes have comparable dynamics.
-For each process, the DA-MCMC algorithm is run for $50,000$ iterations with different values of $\rho = (10^{-4}, 10^{-3}, 10^{-2}, 10^{-1}, 0.5, 1)$.
-The algorithm is run only if there is at least one event time updated per iteration, that is, if $\rho(S_0+I_0)>1$.
-	Figure \ref{fig:accept} presents the acceptance rate across the range of $\rho$.
-We observe that the acceptance rate decreases as the number of event times updated per iteration and the population size increase.
-	Figure \ref{fig:run_time} shows the run time of the MCMC algorithm for these simulations. The run time increases with the population size and with $\rho$.
-	Finally, Figure \ref{fig:acf} shows the effect of $\rho$ on the auto correlation function for the parameter $\beta$ for the medium-size population process ($S(0) = 1,000$) for three values of $\rho$.
-We observe that updating half of the augmented data per iteration yields the lowest auto-correlation function across the values of $\rho$ considered.
-This suggests that, in this setting, updating the entire latent space in the Metropolis-Hastings step results in too many rejections while updating a tenth or less prevents the chain from efficiently exploring the latent space.
-
-	\begin{figure}
-		\centering
-		\includegraphics[scale = 0.4]{E3_1000_ESS_sec_R0.jpg}
-		\caption{Acceptance rate of the proposed latent spaces in the Metropolis-Hastings step of the MCMC algorithm across different population sizes and values for $\rho$.}
-		\label{fig:ESS_sec}
-	\end{figure}
-
-	\begin{figure}
-		\centering	
-\centering
-\includegraphics[width=\textwidth]{E1_burn_R0_acf.jpg}
-\caption{$\rho = 0.1$}
-\label{fig:acf}
-	\end{figure}
-
-	%Figures: acceptance rate v. proportion of updates (see Neal, Roberts A case study in non-centering -- table 1 p.319).
-
-	\subsection{Ebola Outbreak in Gu\'eck\'edou, Guinea}
-	\label{sec:ebo}
-	%Outline: background; setup; results; table; figure
-
-	We now turn to a case study concerning the Ebola outbreak in Western Africa.
-	Between the end of $2013$ and $2015$, Guinea, along with several neighboring countries, experienced the largest outbreak of the Ebola virus disease in history.
-The virus, which has a fatality rate of $70\%$, was responsible for the death of almost $2,000$ people in Guinea alone during the outbreak.
-	The outbreak is believed to have originated in the Gu\'eck\'edou prefecture of Guinea at the end of November $2013$ \cite{Baize.2014}.
-Weekly infection incidence counts are available for each prefecture for the $73$ weeks between the end of December $2013$ and May $2015$.
-	We fit the stochastic SIR model to these incidence counts for the Gu\'eck\'edou prefecture using the MCMC algorithm proposed in this article.
-	For simplicity, we assume that the population of the prefecture forms a closed population, that the model's parameters remained constant throughout the outbreak and that the reported infections counts are exact.
-Applying our MCMC algorithm to stochastic epidemic models where these assumptions are relaxed is the subject of ongoing research.
-	The \textit{effective} population size is set to $n = 150,000$, roughly half of the total population in $2014$.
-Note that as long as the number of infections is small compared to the total population size, the exact population size does not need to be known to estimate the parameters of the model.
-Furthermore, since the first documented infection occurred late November, one month before the first reported incidence count \cite{Baize.2014}, the number of infectious individuals at the beginning of the observation period is set to $I(0) = 10$, with the units of time corresponding to ``days" in the analysis.
-
-	The Markov chain is run for $50,000$ iterations and the event times of $10\%$ of the individuals are updated each iteration.
-The initial values of the parameters are set to $(\beta^{(1)}, \gamma^{(1)}) = (10^{-7}, 0.05)$ and the conjugate distributions in Equation \ref{eq:pri} are used as priors with $a_{\beta} = b_{\beta} = a_{\gamma} = b_{\gamma} = 0.01$ to make them weakly informative.
-Even for such a large population, the total run time of the algorithm was less than $100$ minutes on a personal laptop.
-	The Metropolis-Hastings step for the latent space proposals achieves a healthy $22.1\%$ acceptance rate, and the first $10,000$ iterations of the Markov chain are discarded as a burn-in.
-	Figure \ref{fig:ebola} shows the joint posterior distributions of $(\beta, \gamma)$, whose posterior means are respectively $1.06 * 10^{-6}$ and $0.109$, as well as the marginal distribution of the basic reproduction number $R_0$.
-This value for $\gamma$ indicates that people remain infectious for around $9$ days on average, which is consistent with existing literature.
-As noted by numerous authors, conditionally on partially observed data, the parameters $\beta$ and $\gamma$ appear positively correlated.
-The posterior distribution of $R_0$ is unimodal, relatively symmetric, and centered around $1$.
-A basic reproduction number close to $1$ is consistent with an outbreak that lasted several months without infecting the majority of the population.
-
-	\begin{figure}
-		\centering
-		\begin{subfigure}[b]{0.45\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E5_burn_HPD.jpeg}
-			\caption{Joint posterior density of $\theta = (\beta, \gamma)$. The $50\%$ and $90\%$ confidence regions are shown.}
-			\label{fig:density_joint}
-		\end{subfigure}
-		\hfill
-		\begin{subfigure}[b]{0.49\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E5_burn_R0_hist.jpg}
-			\caption{Posterior density of $R_0$.}
-			\label{fig:density_R0}
-		\end{subfigure}
-		\caption{Posterior densities of parameters of interest (excluding burn-in of $10,000$ iterations).}
-		\label{fig:ebola}
-	\end{figure}
-
-	\section{Discussion and Conclusion}
-	\label{sec:dis}
-	The method proposed in this article enables the classical Metropolis-Hastings algorithm to be an efficient method to conduct full posterior inference of the stochastic SIR model given only incidence data.
-Existing attempts using Markov chain Monte Carlo either mix poorly or do not scale to populations of more than a few hundred individuals, while methods relying on forward simulation are limited to moderate-sized outbreaks and may suffer from degeneracy issues on a case-by-case basis depending on the amount of missing data.
-	Moreover, the simplifying assumptions in \cite{Fintzi.2020} or the approximate-Bayesian-computation framework of \cite{McKinley.2018}, which compromise targeting the exact posterior for computational reasons, generate biased estimates that may lead to spurious conclusions.
-Instead, our data-augmented algorithm enables fast and exact inference, even for large outbreaks, and leverages a well-studied, transparent MCMC framework.
-%Most existing inferential methods for epidemic stochastic models are only applicable to prevalence data.
-To our knowledge, our algorithm is the first that permits exact Bayesian inference given incidence data.
-
-	Central to the success of the DA-MCMC algorithm is an efficient proposal scheme that swiftly explores the latent space of epidemic paths that are compatible with the observed data.
-The PD-SIR process possesses three features that yield such an efficient algorithm.
-First, generating a PD-SIR process is extremely fast; it only requires the simulation of truncated exponential distributions, which can efficiently be realized via the inverse-CDF method.
-Moreover, the individual infection rates are only updated $K$ times, as opposed to after each event as in the SIR model.
-Second, generating a PD-SIR process that is constrained to be compatible with the observed incidence data can be done at no additional cost; in contrast, generating a SIR process compatible with the observed data would be prohibitively slow \cite{Hobolth.2009}.
-Third, the dynamics of the PD-SIR process closely resemble those of the SIR process: the removal dynamics are identical and, for short resetting intervals, the infection dynamics are also very similar in the two processes.
-
-	The first two features make the algorithm extremely fast.
-While existing data-augmented MCMC algorithms have only be applied to populations of a few hundred of individuals, the analysis of the Ebola outbreak in Gu\'eck\'edou shows that our algorithm can be applied to populations of up to $150,000$ individuals, generating tens of thousands of posterior samples in a reasonable amount of time.
-	The latter feature of the PD-SIR process enables the algorithm to update a large portion of the augmented data in each iteration while maintaining a healthy acceptance rate.
-As a result, the Markov chain makes large jumps in the latent space and has very good mixing properties.
-In contrast, existing DA-MCMC algorithms keep most of the latent space fixed across iterations \cite{Gibson.1998, ONeill.1999, Fintzi.2017} and their Markov chains therefore mix much more slowly.
-
-	The algorithm possesses a tuning parameter $\rho$ that determines the portion of the augmented data being updated each iteration.
-Larger values for $\rho$ enable the chain to make larger steps in the latent space but can result in a low acceptance rate, while smaller values of $\rho$ result in a higher acceptance rate but constrain the chain to make small jumps.
-Depending on the size of the population, different values of the tuning parameter are optimal.
-To find a value that optimizes the mixing properties of the chain, one can use several short runs of the algorithm with different values for $\rho$ and select the value that yields the lowest auto-correlation function (ACF).
-Since the run time of the algorithm increases with $\rho$, one could also look at the effective sample size per unit of time.
-For instance, for populations of a few hundred individuals, we observed that updating the entire augmented data ($\rho=1$) yields the lowest ACF.
-In this case, the current and proposed augmented data are independent conditionally on the current value of the parameters.
-
-	The DA-MCMC algorithm proposed in this article is specific to the stochastic SIR model, which is arguably a simplistic representation of the spread of disease.
-The model relies on assumptions such as perfect reporting, constant infection rates, homogeneously mixing population and exponentially-distributed infectious periods.
-In future work, we will present applications of our DA-MCMC algorithm to processes where these assumptions are relaxed.
-In particular, we are considering extensions to epidemic models with under-reporting, a varying infection rate, stratified populations and non-Markovian dynamics, as well as the general stochastic SIR model of \cite{Severo.1969}.
-
-	\appendix
-
-	\section{Distribution of Death Times in Linear Death Process}
-	\label{app:ldp}
-	Theorem \ref{theo:ldp} was first proved by \cite{Neuts.1971}. Since this is a result that has rarely been mentioned in the literature, we provide a proof in this appendix from (cite Ross textbook).
-
-	Consider a linear pure death process with $n$ particles and individual death rate $\mu$.
-	Let $T_i$ be the time of the $i$th death. Then $W_1 = T_1 \sim \Exp(n\mu)$ and $W_i = T_i - T_{i-1} \sim \Exp((n-1)\mu)$ independently. Let $N$ be the number of deaths by time $t$. Then, 
-	\begin{align*}
-		& f(T_1 = t_1, \dots, T_N = t_N | N) \\
-		& \propto f(T_1 = t_1, \dots, T_N = t_N, T_{N+1} > t) 1\{T_N < t\}\\
-		& \propto f(W_1 = t_1, W_2 = t_2 - t_1, \dots, W_N = t_N - t_{N-1}, W_{N+1} > t - t_N) 1\{T_N < t\}\\
-		& \propto \exp\{-n\mu t_1\} \exp\{-(n-1)\mu(t_2-t_1)\}\dots \exp\{-(n-N)\mu(t - t_N)\} 1\{T_N < t\}\\
-		& \propto \exp\{-\mu t_1\}\exp\{-\mu t_2\} \dots \exp\{-\mu t_N\} 1\{T_N < t\}
-	\end{align*}
-which corresponds to the kernels of independent exponential distribution truncated above by $t$.
-
-	\section{Uniform Ergodicity}
-	\label{app:uni}
-	We show that the state space $\chi = \chi_{\theta}\times\chi_{\z}$ of the Markov chain $\{(\theta, \z)^{(i)}\}_i$ is a small set for the transition kernel $P$, that is, there exists a probability measure $\nu$ on the $\sigma$-algebra $\sigma(\chi)$ such that
+	\smallskip
+
+	The following characteristics of our DA-MCMC algorithm are worth noting.
+	First, initializing the Markov chain only requires setting $\theta^{(0)} = (\beta^{(0)}, \gamma^{(0)})$ since $\Z^{(0)}$ can be generated from the PD-SIR process conditionally on $\theta^{(0)}$ and $\Y$ alone.
+	Second, since the PD-SIR closely approximates the SIR model, the acceptance rate in the Metropolis-Hastings step for the latent data is relatively high.
+For large population, however, the acceptance rate may be too low, thereby hindering the mixing of the Markov chain.
+To address this issue, we propose to update the event times of only a fraction $0<\rho\le 1$ of the individuals in the population.
+Smaller values for $\rho$ result in smaller jumps in the latent space and a larger acceptance rate.
+	Third, if all event times are updated ($\rho = 1$), then the proposed and current latent data are independent conditionally on the current values of the parameters: $\Z^{(k-1)} \perp \Z^* | \theta^{(k)}$.
+This contrasts with existing DA-MCMC algorithms which only update a small fraction of the latent data per iteration.
+	%Fourth, since the latent data are generated from a process that only resembles the SIR model, we ensure that the Markov chain converges to the correct distribution $\pi(\theta, \Z|\Y)$ by proposing and accepting the latent data according to a Metropolis-Hastings scheme.
+This results in exact Bayesian inference targeting the posterior distribution under the original SIR model.
+	%Fifth, 
+	Fourth, not only is proposing latent data from the RD-SIR process extremely fast since all random variables can be generated via the inverse-CDF method, but ensuring that these data are compatible with the observed data can be done at no additional cost, making the proposal scalable to large outbreaks.
+	Finally, we show in the following section that the algorithm generates a Markov chain that is \textit{uniformly ergodic}, the strongest form of ergodicity for a Markov chain \cite{Tierney.1994}.
+%The proof of this result is provided in Appendix \ref{app:uni}.
+	% our DA-MCMC does only requires I(0), not S(0).
+
+	\subsection{Uniform Ergodicity}
+
+	It is straightforward to show that the Markov chain $\{(\theta^{(m)}, \z^{(m)})\}_m$ is harris ergodic and converges to the target posterior distribution $\pi(\theta, \z|\Y)$.
+For any $\pi$-integrable function $g$, the estimator
+	$$\bar{g}_m := \frac{1}{m} \sum_{i=0}^{m-1} g(\theta^{(m)}, \z^{(m)})$$
+	is therefore consistent for $E_\pi g$, regardless of the starting values of the chain.
+A Markov chain is uniformly ergodic if for some finite $M$ and positive constant $r<1$ the $n$-step transition kernel $P^n$ satisfies
+	$$
+	\Vert P^n(x,.)-\pi(.)\Vert_{TV} \le M r^n, \forall x \in \chi
+	$$
+	where $\Vert \mu \Vert_{TV}$ denotes the total variation norm of the measure $\mu$.
+	Uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$ (ref), that is,
+	$$\sqrt{m}(\bar{g}_m - E_\pi g) \Rightarrow N(0, \sigma^2)$$
+	for some positive, finite $\sigma^2$. Furthermore, uniform ergodicity implies that the batch means and spectral variance estimators of $\sigma^2$ are consistent (ref).
+
+	%First, the Markov chain  $\{(\theta^{(m)}, \z^{(m)})\}_m$ is Harris recurrent and converges to the target posterior distribution.
+Indeed, by construction, the joint posterior density $\pi(\theta, \z|\Y)$ is invariant for the Gibbs kernel $P_\theta$ and the Metropolis-Hastings kernel $P_\z$ and the density of the composite kernel $P = P_\theta P_\theta$ is strictly positive, implying that the chain is aperiodic and $\pi$-irreducible.
+Together with the existence of an invariant distribution, $\pi$-irreducibility implies that the chain is positive Harris recurrent.
+
+	Toward establishing uniform ergodicity of our DA-MCMC algorithm, we will show that the state space $\chi = \chi_{\theta}\times\chi_{\z}$ of the Markov chain $\{(\theta^{(m)}, \z^{(m)})\}_m$ is a \textit{small set} for the transition kernel $P$.
+That is, there exists a probability measure $\nu$ on the $\sigma$-algebra $\sigma(\chi)$ such that
 	$$\beta \nu(.) \le P^m(x,.), \forall x\in \chi.$$
 	for some constants $m \ge 1$ and $\beta > 0$.
-	By Proposition 2 in \cite{Tierney.1994}, this implies that the Markov chain $\{(\theta, \z)^{(i)}\}_i$ is uniformly ergodic and therefore satisfies
-	$$
-	\Vert P^n(x,.)-\pi(.)\Vert_{TV} \le M r^n
-	$$
-	for some finite $M$ and positive constant $r<1$, where $\Vert \mu \Vert_{TV}$ denotes the total variation norm of the measure $\mu$.
-
-	The following three lemmas are used to show this result.
+	By Proposition 2 in \cite{Tierney.1994}, this implies that the Markov is uniformly ergodic.
+
+	To derive this result, we will make use of three lemmas. The appendix contains a detailed a proof of the latter two.
+	\rm{Figures for lemma 3.1 and 3.2?}
 
@@
 		\label{pro:ga1}
-		Let $Ga(x;a,b)$ denote the density of a gamma distribution with shape $a$ and rate $b$ evaluated at $x$. Then
+		Let $Ga(x;a,b)$ denote the density of the gamma distribution with shape $a$ and rate $b$ evaluated at $x$. Then
 		\begin{equation}
@@
 	\end{lemma}
-
 	\begin{proof}
@@
 
-\begin{figure}
-	\centering
-	\begin{subfigure}[b]{0.45\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{gamma_minorization_1a.jpg}
-		\caption{$\alpha \in [0, 1]$, $\beta=0$}
-		\label{fig:gam1a}
-	\end{subfigure}
-	\hfill
-	\begin{subfigure}[b]{0.49\textwidth}
-		\centering
-		\includegraphics[width=\textwidth]{gamma_minorization_1b.jpg}
-		\caption{$\alpha = 0$, $\beta \in [0, 1]$}
-		\label{fig:gam1b}
-	\end{subfigure}
-	\caption{Example of minorization of the density of a gamma distribution $Ga(2+\alpha, 0.5+\beta)$ for $\alpha$ and $\beta$ separately.}
-	\label{fig:gam1}
-\end{figure}
-
-% Gamma 2
-Proposition \ref{pro:ga1} can be used to minorize a density of the form
+	Lemma \ref{pro:ga1} can be used to minorize a gamma density of the form
 	\begin{equation}
 		\label{eq:ga}
-	Ga(x;a+\alpha,b+\beta) = \frac{(b+\beta)^{a+\alpha}}{\Gamma(a+\alpha)}x^{a+A-1}\exp\{-x(b+B)\}, \quad 0\le\alpha\le A, 0\le\beta\le B.
+		Ga(x;a+\alpha,b+\beta) = \frac{(b+\beta)^{a+\alpha}}{\Gamma(a+\alpha)}x^{a+\alpha-1}\exp\{-x(b+\beta)\}, \quad 0\le\alpha\le A, 0\le\beta\le B.
 	\end{equation}
-over $\alpha$ and $\beta$ for each value of $x$. We can show the following.
+	over $\alpha$ and $\beta$ for each value of $x$. To this end, we establish the following technical lemma.
 	\begin{lemma}	
@@
 	\end{lemma}
+	The final lemma can be used to minorize truncated exponential densities.
+	\begin{lemma}
+		\label{pro:tru}
+		$$\min_{0 \le \mu \le M} \TruncExp(x; \mu, l, u) = \TruncExp(x; M, l, u) $$
+	\end{lemma}
+
+	%\noindent
+	We can now proceed with the proof of uniform ergodicity.
+	% Uniform Ergodicity
+	\begin{proposition}
+		\label{pro:uni}
+		The transition kernel $P = P_{\theta}P_{\z}$ of our DA-MCMC algorithm satisfies a minorization condition $M(1, \delta, \chi, \nu)$ and, as a result, is uniformly ergodic.
+	\end{proposition}
+
+	\begin{proof}
+		First, note that the transition kernel $P$ is a composition of two kernels
+		$$ P = P_{\theta} P_{\z}$$
+		where the kernel $ P_{\theta}$ updates the parameters $\theta$ while keeping the latent data $\z$ fixed and $P_{\z}$ update $\z$ while keeping $\theta$ fixed.
+A one-step transition therefore corresponds to the following scheme
+		$$\x_1 = (\theta_1, \z_1) \rightarrow (\theta_2, \z_1) \rightarrow (\theta_2, \z_2) = \x_2.$$
+
+		The kernel $P$ has a density $p$ with respect to the product measure $\mu = \lambda \mu_{\z}$ with $\lambda$ the Lebesgue measure on $\chi_{\theta}$, the space of $\theta$, and $\mu_{\z}$ a $\sigma$-finite measure on $\chi_{\z}$, the space of $\z$.
+		\begin{align*}
+			P(\x_1, d\x_2) 
+			& = p(\x_1, \x_2) \mu(d\x_2) \\
+			& = p((\theta_1, \z_1), (\theta_2, \z_2)) \lambda(d\theta_2)\mu_{\z}(d\z_2) \\
+			& = p_{\theta}((\theta_1, \z_1), (\theta_2, \z_1)) p_{\z}((\theta_2, \z_1), (\theta_2, \z_2)) \mu_{\z}(d\z_2) \lambda(d\theta_2)
+		\end{align*}
+		where
+		\begin{align*}
+			p_{\theta}((\theta_1, \z_1), (\theta_2, \z_1))
+			%	& = q_{\theta}((\theta_1, \z_1), (\theta_2, \z_1)) \alpha((\theta_1, \z_1), (\theta_2, \z_1)) \\
+			& = \pi(\theta_2| \z_1) \\
+			& = \pi(\beta_2|\z_1) \pi(\gamma_2|\z_1)
+		\end{align*}
+		is the density of the transition kernel $P_{\theta}$ with respect to $\lambda$ and corresponds to the product of the full conditional distributions of $\beta$ and $\gamma$ since the kernel $P_{\theta}$ is a Gibbs sampler, and
+		\begin{align*}
+			p_{\z}((\theta_2, \z_1), (\theta_2, \z_2))
+			& = q_{\z}((\theta_2, \z_1), (\theta_2, \z_2)) \alpha((\theta_2, \z_1), (\theta_2, \z_2)) \\
+			& = q(\z_2; \theta_2) \min\left\lbrace 1, \dfrac{\pi(\theta_2, \z_2)q(\z_1; \theta_2)}{\pi(\theta_2, \z_1)q(\z_2; \theta_2)} \right\rbrace \\
+			& = \min\left\lbrace q(\z_2; \theta_2), \dfrac{\pi(\theta_2, \z_2)q(\z_1; \theta_2)}{\pi(\theta_2, \z_1)} \right\rbrace
+		\end{align*}
+		is the density of the transition kernel $P_{\z}$ with respect to $\mu_{\z}$ and corresponds to one step of the Metropolis-Hasting algorithm where a new configuration of the latent data $\z_2$ is proposed conditionally on the current value of the parameters $\theta_2$, but independently of the current configuration $\z_1$.
+
+		To show that $\chi$ is a small state, it is sufficient to show that there exists a function $k$ such that
+		$$p(\x_1, \x_2) \ge k(\x_2) > 0, \qquad \forall \x_1 \in \chi $$
+		for all $\x_2 \in \chi$.
+		The density $p$ 
+		\begin{align*}
+			p((\theta_1, \z_1), (\theta_2, \z_2))
+			& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2), \dfrac{\pi(\theta_2| \z_1) \pi(\theta_2, \z_2)}{\pi(\theta_2, \z_1)} g(\z_1; \theta_2) \right\rbrace \\
+			& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2), \dfrac{\pi(\theta_2, \z_2)}{\pi(\z_1)} g(\z_1; \theta_2) \right\rbrace \\
+			%	& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2),		\dfrac{			\frac{f(\theta_2, \z_1, Y)}{f(\z_1, Y)}			\frac{f(\theta_2, \z_2, Y)}{f(Y)}		}{\frac{f(\theta_2, \z_1, Y)}{f(Y)}} 		g(\z_1; \theta_2)\right\rbrace \\
+			%	& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2),		\dfrac{f(\theta_2, \z_2, Y)}{f(\z_1, Y)}		g(\z_1; \theta_2)\right\rbrace \\
+			& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2),
+			\pi(\theta_2| \z_2)
+			g(\z_1; \theta_2)\right\rbrace
+		\end{align*}
+		depends on $x_1 = (\theta_1, \z_1)$ only through the full conditional $\pi(\theta_2| \z_1)$ and the proposal density $g(\z_1; \theta_2)$.
+
+		It therefore suffices to show that there exist functions $k_1$ and $k_2$ such that
+		$$\pi(\theta| \z_1) \ge k_1(\theta) > 0 \qquad \text{and} \qquad  g(\z_1; \theta) \ge k_2(\theta) > 0.$$
+		First, 
+		\begin{align*}
+			\pi(\theta| \z_1) 
+			& = \pi(\beta| \z_1) \pi(\gamma| \z_1)  \\
+			& = Ga(\beta; a_{\beta} + n_T, b_{\beta} + SI_1) Ga(\gamma; a_{\gamma} + {n_J}_1, b_{\gamma} + I_1) \\
+			& \ge h_{\beta}(\beta) h_{\gamma}(\gamma) \\
+			& = k_1(\theta) > 0
+		\end{align*}
+		where, by Proposition \ref{pro:ga1},
+		$$h_{\beta}(\beta) = \begin{cases}
+			Ga(\beta;a_{\beta}+n_T,b_{\beta}), & \beta<\frac{a_{\beta}+n_T}{t_{end}(n_T+I_0) n} \log \left( 1 + \frac{t_{end}(n_T+I_0) n}{b_{\beta}}\right) \\ Ga(x;a_{\beta}+n_T,b_{\beta}+n(n_T+I_0) t_{end}), & \text{else}
+		\end{cases}$$
+		since $n_T = \sum_k T_k$ is known and 
+		$$SI_1 = \int_{0}^{t_{end}}S_1(t)I_1(t)dt \in [0, n (n_T+I_0) t_{end}];$$
+		and, by Proposition \ref{pro:ga2},
+		$$h_{\gamma}(\gamma) = \min\{
+		Ga(\gamma;a_{\gamma}+n_T+I_0,b_{\gamma}), Ga(x;a_{\gamma},b_{\gamma}+(n_T+I_0) t_{end})\}
+		$$
+		since $0\le n_1^J \le n_T + I_0$ and
+		$$I_1 = \int_{0}^{t_{end}}I_1(t)dt \in [0, (n_T+I_0) t_{end}].$$
+
+		Second, each factor in
+		\begin{align*}
+			& g(\z_1; \theta) = \prod_{i=1}^{n_T} \TruncExp(z^T_i; \beta I_k, t_{k(i)-1}, t_{k(i)}) (1-p_i)^{1\{\z^J_i = \infty\}} (p_i \TruncExp(z^J_i; \gamma, z^T_i, t_{end}))^{1\{\z^J_i \le t_{end}\}}
+		\end{align*}
+		can be minorized as follows. First,
+		\begin{align*}
+			\TruncExp(z^T_i; \beta I_k, t_{k(i)-1}, t_{k(i)}) 
+			& \ge \TruncExp(t_{k(i)}; \beta I_k, t_{k(i)-1}, t_{k(i)}) \\
+			& \ge \TruncExp(t_{k(i)}; \beta (n_T + I_0), t_{k(i)-1}, t_{k(i)}),
+		\end{align*}
+		with the last inequality following from Proposition \ref{pro:tru}, second,
+		$$1-p_i = \exp\{-\gamma (t_{end} - z^T_i)\} \ge \exp\{-\gamma (t_{end} - t_{k(i)-1})\},$$
+		third,
+		$$p_i \TruncExp(z^J_i; \gamma, z^T_i, t_{end}) = \Exp(z^J_i - z^T_i; \gamma)1(z^T_i < z^J_i \le t_{end}) \ge \Exp(t_{end} - t_{k(i) - 1}; \gamma)$$
+		where the equality holds since $p_i = P(z^J_i \le t_{end}|z^J_i > z^T_i)$ is equal to the normalizing constant of the truncated exponential.
+
+		Taken together, these inequalities give
+		\begin{align*}
+			g(\z_1; \theta)
+			& \ge \prod_{i=1}^{n_T} \TruncExp(t_{k(i)}; \beta (I_0 + n_T), t_{k(i)-1}, t_{k(i)}) \min\{\exp\{-\gamma (t_{end} - t_{k(i) - 1})\}, \Exp(t_{end} - t_{k(i) - 1}; \gamma)\} \\
+			& = k_2(\theta) > 0
+		\end{align*}
+		which completes the proof.
+	\end{proof}
+
+	The kernel 
+	$$
+	P_{\theta}((\theta_0, \z), (\theta_1, \z)) = Ga\left( \beta_1; a_{\beta} + n_I, b_{\beta} + \int_0^{t_{end}} S(t)I(t) dt\right) Ga\left( \gamma_1; a_{\gamma} + n_R, b_{\gamma} + \int_0^{t_{end}} I(t) dt\right) d\beta_1 d\gamma_1
+	$$
+	therefore corresponds to a Gibbs kernel where $n_I, n_R, \int S(t)I(t) dt, \int I(t) dt$ are sufficient statistics from the latent data $\z$
+
+	\section{Performance on simulated and real epidemic data}
+	\label{sec:per}
+
+	\subsection{Simulation Study}
+	\label{sec:sim}
+	%Outline: setup; results: posterior mean, credible interval, frequentist coverage, run time, ACF, acceptance rate; table; figure: traceplot, ACF; tuning: update proportion v.
+acceptance rate; compare with others methods e.g.
+chain binomial, exact proposal.
+
+	% 1. convergence in medium-size population
+	% 2. rho
+	% 3. nominal coverage should be 95%: boxplot of posterior means, including coverage (one setting is enough for the moment)
+	% 4. distribution of estimate for PD-SIR (unbiased) vs. LNA (biased)
+	% explain why not compare chain binomial, particle filtering (forward simulation), LNA requires noisy data.
+
+	% TODO: Tables: posterior means, credible intervals, effective sample size.
+	% TODO: frequentist coverage for 3 different observation schedules
+
+	In this section, the convergence properties of the DA-MCMC algorithm are examined with a simulation study.
+Each data set in this study is simulated from the stochastic SIR process and the prior distributions on the parameters are the independent weakly informative distributions $\beta \sim Ga(0.1, 1)$ and $\gamma \sim Ga(1,1)$.
+
+	First, the mixing property of the Markov chain is assessed in a medium-size population whose initial configuration is $(S(0), I(0)) = (1000, 10)$.
+The parameters are $(\beta, \gamma) = (0.003, 1)$ and the process is observed until time $t_{end} = 6$ when the pandemic has completed most of its course but is not over yet (see Figure \ref{fig:long}, where $I(t_{end}) = 55$).
+The numbers of infections in $K = 10$ time intervals of equal length are observed and correspond to $T_{1:K} = (40, 111, 193, 259, 178, 93, 29, 19, 9, 6)$.
+	The Markov chain is run for $100,000$ iterations from the initial values $(\beta^{(1)}, \gamma^{(1)}) = (0.0003, 0.1) = (\beta/10, \gamma/10)$ and the event times of $202$ ($\rho = 0.2$) individuals are updated in each Metropolis-Hastings step for the latent data.
+The algorithm takes in $2.5$ minutes to complete on a personal laptop.
+	Figure \ref{fig:traceplot} shows the traceplots of the parameters $\beta$ and $\gamma$, the log likelihood and the basic reproduction number $R_0 = S(0) \beta / \gamma$, which corresponds to the average number of secondary infections generated by an infectious individual in a susceptible population.
+We observe that the Markov chain quickly migrates from the low density region of the space where $(\beta^{(1)}, \gamma^{(1)})$ is located to a high density region where it mixes well.
+The acceptance rate of the latent spaces proposed in the Metropolis-Hastings step is $0.11$.
+	Excluding the first $10,000$ iterations of the Markov chain as a burn-in, the posterior means of $\beta$, $\gamma$ and $R_0$ are respectively $0.00304$, $0.995$ and $3.07$.
+
+	% TODO: Results: convergence (R statistic, Geweke), credible intervals.
+
+	% TODO: Geweke
+
+	\begin{figure}
+		\centering
+		\includegraphics[scale = 0.3]{E1_trajectories}
+		\caption{Trajectories of the $S$, $I$ and $R$ compartments of a SIR process for a medium-size population ($S_0 = 1,000$).}
+		\label{fig:long}
+	\end{figure}
+
+	\begin{figure}
+		\centering
+		\begin{subfigure}[b]{0.41\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E1_no_burn_beta_tp}
+			\caption{$\beta$}
+			\label{fig:traceplot_beta}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.41\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E1_burn_beta_acf}
+			\caption{$\gamma$}
+			\label{fig:traceplot_gamma}
+		\end{subfigure}
+		\\
+		\begin{subfigure}[b]{0.41\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E1_no_burn_R0_hist}
+			\caption{Log likelihood}
+			\label{fig:traceplot_loglik}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.41\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E1_burn_R0_acf}
+			\caption{$R_0$}
+			\label{fig:traceplot_R0}
+		\end{subfigure}
+		\caption{Traceplots of the Markov chain (including burn-in).}
+		\label{fig:traceplot}
+	\end{figure}
+
+	Second, the impact of the tuning parameter $\rho$ on the performance of the MCMC algorithm is evaluated.
+In particular, we examine the acceptance rate of the Metropolis-Hastings step, the total run time of the algorithm and the auto-correlation function of the parameters.
+	Three SIR processes with different population sizes $S(0) = (10^2, 10^3, 10^4)$ are simulated.
+We set $I(0) =10$, $\gamma=1$, $t_{end} = 3$ and $K = 50$ for all simulations and $\beta = (0.001, 0.01, 0.1)$ respectively for the three different population sizes.
+These values for $\beta$ ensure that the three processes have comparable dynamics.
+For each process, the DA-MCMC algorithm is run for $50,000$ iterations with different values of $\rho = (10^{-4}, 10^{-3}, 10^{-2}, 10^{-1}, 0.5, 1)$.
+The algorithm is run only if there is at least one event time updated per iteration, that is, if $\rho(S_0+I_0)>1$.
+	Figure \ref{fig:accept} presents the acceptance rate across the range of $\rho$.
+We observe that the acceptance rate decreases as the number of event times updated per iteration and the population size increase.
+	Figure \ref{fig:run_time} shows the run time of the MCMC algorithm for these simulations. The run time increases with the population size and with $\rho$.
+	Finally, Figure \ref{fig:acf} shows the effect of $\rho$ on the auto correlation function for the parameter $\beta$ for the medium-size population process ($S(0) = 1,000$) for three values of $\rho$.
+We observe that updating half of the augmented data per iteration yields the lowest auto-correlation function across the values of $\rho$ considered.
+This suggests that, in this setting, updating the entire latent space in the Metropolis-Hastings step results in too many rejections while updating a tenth or less prevents the chain from efficiently exploring the latent space.
+
+	\begin{figure}
+		\centering
+		\includegraphics[scale = 0.4]{E3_facet_ESSsecR0}
+		\caption{ESS/sec for $R_0$
+			%Acceptance rate of the proposed latent spaces in the Metropolis-Hastings step of the MCMC algorithm across different population sizes and values for $\rho$.
+		}
+		\label{fig:ESS_sec}
+	\end{figure}
+
+	\begin{figure}
+		\centering	
+		\centering
+		\includegraphics[width=\textwidth]{E1_burn_R0_acf}
+		\caption{$\rho = 0.1$}
+		\label{fig:acf}
+	\end{figure}
+
+	%Figures: acceptance rate v. proportion of updates (see Neal, Roberts A case study in non-centering -- table 1 p.319).
+
+	\subsection{Ebola Outbreak in Gu\'eck\'edou, Guinea}
+	\label{sec:ebo}
+	%Outline: background; setup; results; table; figure
+
+	We now turn to a case study concerning the Ebola outbreak in Western Africa.
+	Between the end of $2013$ and $2015$, Guinea, along with several neighboring countries, experienced the largest outbreak of the Ebola virus disease in history.
+The virus, which has a fatality rate of $70\%$, was responsible for the death of almost $2,000$ people in Guinea alone during the outbreak.
+	The outbreak is believed to have originated in the Gu\'eck\'edou prefecture of Guinea at the end of November $2013$ \cite{Baize.2014}.
+Weekly infection incidence counts are available for each prefecture for the $73$ weeks between the end of December $2013$ and May $2015$.
+	We fit the stochastic SIR model to these incidence counts for the Gu\'eck\'edou prefecture using the MCMC algorithm proposed in this article.
+	For simplicity, we assume that the population of the prefecture forms a closed population, that the model's parameters remained constant throughout the outbreak and that the reported infections counts are exact.
+Applying our MCMC algorithm to stochastic epidemic models where these assumptions are relaxed is the subject of ongoing research.
+	The \textit{effective} population size is set to $n = 150,000$, roughly half of the total population in $2014$.
+Note that as long as the number of infections is small compared to the total population size, the exact population size does not need to be known to estimate the parameters of the model.
+Furthermore, since the first documented infection occurred late November, one month before the first reported incidence count \cite{Baize.2014}, the number of infectious individuals at the beginning of the observation period is set to $I(0) = 10$, with the units of time corresponding to ``days" in the analysis.
+
+	The Markov chain is run for $50,000$ iterations and the event times of $10\%$ of the individuals are updated each iteration.
+The initial values of the parameters are set to $(\beta^{(1)}, \gamma^{(1)}) = (10^{-7}, 0.05)$ and the conjugate distributions in Equation \ref{eq:pri} are used as priors with $a_{\beta} = b_{\beta} = a_{\gamma} = b_{\gamma} = 0.01$ to make them weakly informative.
+Even for such a large population, the total run time of the algorithm was less than $100$ minutes on a personal laptop.
+	The Metropolis-Hastings step for the latent space proposals achieves a healthy $22.1\%$ acceptance rate, and the first $10,000$ iterations of the Markov chain are discarded as a burn-in.
+	Figure \ref{fig:ebola} shows the joint posterior distributions of $(\beta, \gamma)$, whose posterior means are respectively $1.06 * 10^{-6}$ and $0.109$, as well as the marginal distribution of the basic reproduction number $R_0$.
+This value for $\gamma$ indicates that people remain infectious for around $9$ days on average, which is consistent with existing literature.
+As noted by numerous authors, conditionally on partially observed data, the parameters $\beta$ and $\gamma$ appear positively correlated.
+The posterior distribution of $R_0$ is unimodal, relatively symmetric, and centered around $1$.
+A basic reproduction number close to $1$ is consistent with an outbreak that lasted several months without infecting the majority of the population.
+
+	\begin{figure}
+		\centering
+		\begin{subfigure}[b]{0.45\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E5_expected_infection_length_tp}
+			\caption{Joint posterior density of $\theta = (\beta, \gamma)$. The $50\%$ and $90\%$ confidence regions are shown.}
+			\label{fig:density_joint}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.49\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E5_expected_infection_length_hist.jpg}
+			\caption{Posterior density of $R_0$.}
+			\label{fig:density_R0}
+		\end{subfigure}
+		\caption{Posterior densities of parameters of interest (excluding burn-in of $10,000$ iterations).}
+		\label{fig:ebola}
+	\end{figure}
+
+	\section{Discussion and Conclusion}
+	\label{sec:dis}
+	The method proposed in this article enables the classical Metropolis-Hastings algorithm to be an efficient method to conduct full posterior inference of the stochastic SIR model given only incidence data.
+Existing attempts using Markov chain Monte Carlo either mix poorly or do not scale to populations of more than a few hundred individuals, while methods relying on forward simulation are limited to moderate-sized outbreaks and may suffer from degeneracy issues on a case-by-case basis depending on the amount of missing data.
+	Moreover, the simplifying assumptions in \cite{Fintzi.2020} or the approximate-Bayesian-computation framework of \cite{McKinley.2018}, which compromise targeting the exact posterior for computational reasons, generate biased estimates that may lead to spurious conclusions.
+Instead, our data-augmented algorithm enables fast and exact inference, even for large outbreaks, and leverages a well-studied, transparent MCMC framework.
+%Most existing inferential methods for epidemic stochastic models are only applicable to prevalence data.
+To our knowledge, our algorithm is the first that permits exact Bayesian inference given incidence data.
+
+	Central to the success of the DA-MCMC algorithm is an efficient proposal scheme that swiftly explores the latent space of epidemic paths that are compatible with the observed data.
+The PD-SIR process possesses three features that yield such an efficient algorithm.
+First, generating a PD-SIR process is extremely fast; it only requires the simulation of truncated exponential distributions, which can efficiently be realized via the inverse-CDF method.
+Moreover, the individual infection rates are only updated $K$ times, as opposed to after each event as in the SIR model.
+Second, generating a PD-SIR process that is constrained to be compatible with the observed incidence data can be done at no additional cost; in contrast, generating a SIR process compatible with the observed data would be prohibitively slow \cite{Hobolth.2009}.
+Third, the dynamics of the PD-SIR process closely resemble those of the SIR process: the removal dynamics are identical and, for short resetting intervals, the infection dynamics are also very similar in the two processes.
+
+	The first two features make the algorithm extremely fast.
+While existing data-augmented MCMC algorithms have only be applied to populations of a few hundred of individuals, the analysis of the Ebola outbreak in Gu\'eck\'edou shows that our algorithm can be applied to populations of up to $150,000$ individuals, generating tens of thousands of posterior samples in a reasonable amount of time.
+	The latter feature of the PD-SIR process enables the algorithm to update a large portion of the augmented data in each iteration while maintaining a healthy acceptance rate.
+As a result, the Markov chain makes large jumps in the latent space and has very good mixing properties.
+In contrast, existing DA-MCMC algorithms keep most of the latent space fixed across iterations \cite{Gibson.1998, ONeill.1999, Fintzi.2017} and their Markov chains therefore mix much more slowly.
+
+	The algorithm possesses a tuning parameter $\rho$ that determines the portion of the augmented data being updated each iteration.
+Larger values for $\rho$ enable the chain to make larger steps in the latent space but can result in a low acceptance rate, while smaller values of $\rho$ result in a higher acceptance rate but constrain the chain to make small jumps.
+Depending on the size of the population, different values of the tuning parameter are optimal.
+To find a value that optimizes the mixing properties of the chain, one can use several short runs of the algorithm with different values for $\rho$ and select the value that yields the lowest auto-correlation function (ACF).
+Since the run time of the algorithm increases with $\rho$, one could also look at the effective sample size per unit of time.
+For instance, for populations of a few hundred individuals, we observed that updating the entire augmented data ($\rho=1$) yields the lowest ACF.
+In this case, the current and proposed augmented data are independent conditionally on the current value of the parameters.
+
+	The DA-MCMC algorithm proposed in this article is specific to the stochastic SIR model, which is arguably a simplistic representation of the spread of disease.
+The model relies on assumptions such as perfect reporting, constant infection rates, homogeneously mixing population and exponentially-distributed infectious periods.
+In future work, we will present applications of our DA-MCMC algorithm to processes where these assumptions are relaxed.
+In particular, we are considering extensions to epidemic models with under-reporting, a varying infection rate, stratified populations and non-Markovian dynamics, as well as the general stochastic SIR model of \cite{Severo.1969}.
+
+	\appendix
+
+	\section{Distribution of Death Times in Linear Death Process}
+	\label{app:ldp}
+	Theorem \ref{theo:ldp} was first proved by \cite{Neuts.1971}. Since this is a result that has rarely been mentioned in the literature, we provide a proof in this appendix from (cite Ross textbook).
+
+	Consider a linear pure death process with $n$ particles and individual death rate $\mu$.
+	Let $T_i$ be the time of the $i$th death. Then $W_1 = T_1 \sim \Exp(n\mu)$ and $W_i = T_i - T_{i-1} \sim \Exp((n-1)\mu)$ independently. Let $N$ be the number of deaths by time $t$. Then, 
+	\begin{align*}
+		& f(T_1 = t_1, \dots, T_N = t_N | N) \\
+		& \propto f(T_1 = t_1, \dots, T_N = t_N, T_{N+1} > t) 1\{T_N < t\}\\
+		& \propto f(W_1 = t_1, W_2 = t_2 - t_1, \dots, W_N = t_N - t_{N-1}, W_{N+1} > t - t_N) 1\{T_N < t\}\\
+		& \propto \exp\{-n\mu t_1\} \exp\{-(n-1)\mu(t_2-t_1)\}\dots \exp\{-(n-N)\mu(t - t_N)\} 1\{T_N < t\}\\
+		& \propto \exp\{-\mu t_1\}\exp\{-\mu t_2\} \dots \exp\{-\mu t_N\} 1\{T_N < t\}
+	\end{align*}
+	which corresponds to the kernels of independent exponential distribution truncated above by $t$.
+
+	\section{Proofs}
+	\label{app:uni}
+	% 	We show that the state space $\chi = \chi_{\theta}\times\chi_{\z}$ of the Markov chain $\{(\theta, \z)^{(i)}\}_i$ is a small set for the transition kernel $P$, that is, there exists a probability measure $\nu$ on the $\sigma$-algebra $\sigma(\chi)$ such that
+	% 	$$\beta \nu(.) \le P^m(x,.), \forall x\in \chi.$$
+	% 	for some constants $m \ge 1$ and $\beta > 0$.
+	% 	By Proposition 2 in \cite{Tierney.1994}, this implies that the Markov chain $\{(\theta, \z)^{(i)}\}_i$ is uniformly ergodic and therefore satisfies
+	% 	$$
+	% 	\Vert P^n(x,.)-\pi(.)\Vert_{TV} \le M r^n
+	% 	$$
+	% 	for some finite $M$ and positive constant $r<1$, where $\Vert \mu \Vert_{TV}$ denotes the total variation norm of the measure $\mu$.
+
+	% 	The following three lemmas are used to show this result.
+
+	% 	% Gamma 1
+	% 	\begin{lemma}
+	% 		\label{pro:ga1}
+	% 		Let $Ga(x;a,b)$ denote the density of a gamma distribution with shape $a$ and rate $b$ evaluated at $x$. Then
+	% 		\begin{equation}
+	% 			\label{eq:ga1}
+	% 			\inf_{0\le \beta\le B} Ga(x;a,b+\beta) = 
+	% 			\begin{cases}
+	% 				Ga(x;a,b), & x<x_a^* \\ Ga(x;a,b+B), & x\ge x_a^*
+	% 			\end{cases}
+	% 		\end{equation}	
+	% 		where $x_a^*=\frac{a}{B}\log\left( 1+\frac{B}{b}\right) $. Moreover,
+	% 		\begin{equation}
+	% 			\label{eq:ga2}
+	% 			\inf_{0\le \alpha\le A} Ga(x;a+\alpha,b) = 
+	% 			\begin{cases}
+	% 				Ga(x;a,b), & x>x_b^* \\ Ga(x;a+A,b), & x\le x_b^*
+	% 			\end{cases}
+	% 		\end{equation}
+	% 		where $x_b^*=\frac{1}{b}\left[ \frac{\Gamma(a+A)}{\Gamma(a)}\right]^{1/A} $.
+	% 	\end{lemma}
+
+	%Proof of Lemma \ref{pro:ga1}.
+
+	\begin{figure}
+		\centering
+		\begin{subfigure}[b]{0.45\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{gamma_minorization_1a.jpg}
+			\caption{$\alpha \in [0, 1]$, $\beta=0$}
+			\label{fig:gam1a}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.49\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{gamma_minorization_1b.jpg}
+			\caption{$\alpha = 0$, $\beta \in [0, 1]$}
+			\label{fig:gam1b}
+		\end{subfigure}
+		\caption{Example of minorization of the density of a gamma distribution $Ga(2+\alpha, 0.5+\beta)$ for $\alpha$ and $\beta$ separately.}
+		\label{fig:gam1}
+	\end{figure}
+
+	% Gamma 2
+	Proof of Lemma \ref{pro:ga2}: % can be used to minorize a density of the form
+	% \begin{equation}
+	% 	\label{eq:ga}
+	% 	Ga(x;a+\alpha,b+\beta) = \frac{(b+\beta)^{a+\alpha}}{\Gamma(a+\alpha)}x^{a+A-1}\exp\{-x(b+B)\}, \quad 0\le\alpha\le A, 0\le\beta\le B.
+	% \end{equation}
+	% over $\alpha$ and $\beta$ for each value of $x$. We can show the following.
+	% \begin{lemma}	
+	% 	\label{pro:ga2}
+	% 	For a fix $x>0$, the density $Ga(x;a+\alpha,b+\beta)$ is minimized by $(\alpha, \beta) \in \{(A, 0), (0, B)\}$, the minimizing set of values depending on $x$. In particular,	
+	% 	$$\inf_{\begin{aligned}
+	% 			0\le \alpha\le A \\ 0\le \beta\le B
+	% 	\end{aligned}}Ga(x;a+\alpha,b+\beta) = 
+	% 	\begin{cases}
+	% 		Ga(x;a+A,b)                     , & x<x_a         \text{ or } x<x_{a+A}^* \vee x_b^*\\
+	% 		Ga(x;a,b+B)                     , & x_{a+A}^* < x \text{ or } x_a^* \wedge x_{b+B}^* < x\\
+	% 		\min\{Ga(x;a+A,b), Ga(x;a,b+B)\}, & x_a^* \wedge x_b^*\le x < x_{a+A}^* \vee x_{b+B}^*\\
+	% 	\end{cases}$$
+	% \end{lemma}
 
@@
 	% Truncated Exponential
-\begin{lemma}
-	\label{pro:tru}
-	$$\min_{0 \le \mu \le M} \TruncExp(u; \mu, 0, u) = \TruncExp(u; M, 0, u) $$
-\end{lemma}
+	Proof of Lemma \ref{pro:tru}
+
 	\begin{proof}
@@
 
-% Uniform Ergodicity
-\begin{proposition}
-	\label{pro:uni}
-	The transition kernel $P = P_{\theta}P_{\z}$ satisfies a minorization condition $M(1, \delta, \chi, \nu)$ and is therefore uniformly ergodic.
-\end{proposition}
-
-\begin{proof}
-	First, note that the transition kernel $P$ of our Markov chain is a composition of two kernels
-	$$ P = P_{\theta} P_{\z}$$
-	where the kernel $ P_{\theta}$ updates the parameters $\theta$ of $\x$ while keeping the augmented data $\z$ fixed and $P_{\z}$ update $\z$ while keeping $\theta$ fixed.
-A one-step transition therefore corresponds to the following scheme
-	$$\x_1 = (\theta_1, \z_1) \rightarrow (\theta_2, \z_1) \rightarrow (\theta_2, \z_2) = \x_2.$$
-
-	The kernel $P$ has a density $p$ with respect to the product measure $\mu = \lambda \mu_{\z}$ with $\lambda$ the Lebesgue measure on $\chi_{\theta}$, the space of $\theta$, and $\mu_{\z}$ a $\sigma$-finite measure on $\chi_{\z}$, the space of $\z$.
-	\begin{align*}
-		P(\x_1, d\x_2) 
-		& = p(\x_1, \x_2) \mu(d\x_2) \\
-		& = p((\theta_1, \z_1), (\theta_2, \z_2)) \lambda(d\theta_2)\mu_{\z}(d\z_2) \\
-		& = p_{\theta}((\theta_1, \z_1), (\theta_2, \z_1)) p_{\z}((\theta_2, \z_1), (\theta_2, \z_2)) \mu_{\z}(d\z_2) \lambda(d\theta_2)
-	\end{align*}
-	where
-	\begin{align*}
-		p_{\theta}((\theta_1, \z_1), (\theta_2, \z_1))
-		& = q_{\theta}((\theta_1, \z_1), (\theta_2, \z_1)) \alpha((\theta_1, \z_1), (\theta_2, \z_1)) \\
-		& = \pi(\theta_2| \z_1)
-	\end{align*}
-	is the density of the transition kernel $P_{\theta}$ with respect to $\lambda$ and corresponds to the full conditional distribution $\pi(\theta_2| \z_1)$ since the kernel $P_{\theta}$ is a Gibbs sampler, and
-	\begin{align*}
-		p_{\z}((\theta_2, \z_1), (\theta_2, \z_2))
-		& = q_{\z}((\theta_2, \z_1), (\theta_2, \z_2)) \alpha((\theta_2, \z_1), (\theta_2, \z_2)) \\
-		& = g(\z_2; \theta_2) \min\left\lbrace 1, \dfrac{\pi(\theta_2, \z_2)g(\z_1; \theta_2)}{\pi(\theta_2, \z_1)g(\z_2; \theta_2)} \right\rbrace \\
-		& = \min\left\lbrace g(\z_2; \theta_2), \dfrac{\pi(\theta_2, \z_2)g(\z_1; \theta_2)}{\pi(\theta_2, \z_1)} \right\rbrace
-	\end{align*}
-	is the density of the transition kernel $P_{\z}$ with respect to $\mu_{\z}$ and corresponds to one step of the Metropolis-Hasting algorithm where a new specification of the augmented data $\z_2$ is proposed conditionally on the current value of the parameters $\theta_2$, but independently of the current specification of the augmented data $\z_1$.
-
-	To show that $\chi$ is a small state, it is sufficient to show that there exists a function $k$ such that
-	$$p(\x_1, \x_2) \ge k(\x_2) > 0, \qquad \forall \x_1 \in \chi $$
-	for all $\x_2 \in \chi$.
-	Note that the density $p$ is a function of $\x_1 = (\theta_1, \z_1)$ only through the full conditional distribution $\pi(\theta_2| \z_1)$ and the proposal density $g(\z_1; \theta_2)$.
-	\begin{align*}
-		p((\theta_1, \z_1), (\theta_2, \z_2))
-		& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2), \dfrac{\pi(\theta_2| \z_1) \pi(\theta_2, \z_2)}{\pi(\theta_2, \z_1)} g(\z_1; \theta_2) \right\rbrace \\
-		& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2),
-		\dfrac{
-			\frac{f(\theta_2, \z_1, Y)}{f(\z_1, Y)}
-			\frac{f(\theta_2, \z_2, Y)}{f(Y)}
-		}{\frac{f(\theta_2, \z_1, Y)}{f(Y)}} 
-		g(\z_1; \theta_2)\right\rbrace \\
-		& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2),
-		\dfrac{f(\theta_2, \z_2, Y)}{f(\z_1, Y)}
-		g(\z_1; \theta_2)\right\rbrace \\
-		& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2),
-		\pi(\theta_2| \z_2)
-		g(\z_1; \theta_2)\right\rbrace
-	\end{align*}
-It therefore suffices to show that there exist functions $k_1$ and $k_2$ such that
-$$\pi(\theta| \z_1) \ge k_1(\theta) > 0 \qquad \text{and} \qquad  g(\z_1; \theta) \ge k_2(\theta) > 0.$$
-
-	First, 
-	\begin{align*}
-		\pi(\theta| \z_1) 
-		& = \pi(\beta| \z_1) \pi(\gamma| \z_1)  \\
-		& = Ga(\beta; a_{\beta} + n_T, b_{\beta} + SI_1) Ga(\gamma; a_{\gamma} + {n_J}_1, b_{\gamma} + I_1) \\
-		& \ge h_{\beta}(\beta) h_{\gamma}(\gamma) \\
-		& = k_1(\theta) > 0
-	\end{align*}
-	where, by Proposition \ref{pro:ga1},
-	$$h_{\beta}(\beta) = \begin{cases}
-		Ga(\beta;a_{\beta}+n_T,b_{\beta}), & \beta<\frac{a_{\beta}+n_T}{t_{end}(n_T+I_0) n} \log \left( 1 + \frac{t_{end}(n_T+I_0) n}{b_{\beta}}\right) \\ Ga(x;a_{\beta}+n_T,b_{\beta}+n(n_T+I_0) t_{end}), & \text{else}
-	\end{cases}$$
-since $n_T = \sum_k T_k$ is known and 
-$$SI_1 = \int_{0}^{t_{end}}S_1(t)I_1(t)dt \in [0, n (n_T+I_0) t_{end}];$$
-and, by Proposition \ref{pro:ga2},
-	$$h_{\gamma}(\gamma) = \min\{
-		Ga(\gamma;a_{\gamma}+n_T+I_0,b_{\gamma}), Ga(x;a_{\gamma},b_{\gamma}+(n_T+I_0) t_{end})\}
-		$$
-since $0\le n_1^J \le n_T + I_0$ and
-$$I_1 = \int_{0}^{t_{end}}I_1(t)dt \in [0, (n_T+I_0) t_{end}].$$
-
-	Second, each factor in
-	\begin{align*}
-		& g(\z_1; \theta) = \prod_{i=1}^{n_T} \TruncExp(z^T_i; \beta I_k, t_{k(i)-1}, t_{k(i)}) (1-p_i)^{1\{\z^J_i = \infty\}} (p_i \TruncExp(z^J_i; \gamma, z^T_i, t_{end}))^{1\{\z^J_i \le t_{end}\}}
-	\end{align*}
-	can be minorized as follows. First,
-	\begin{align*}
-		\TruncExp(z^T_i; \beta I_k, t_{k(i)-1}, t_{k(i)}) 
-		& \ge \TruncExp(t_{k(i)}; \beta I_k, t_{k(i)-1}, t_{k(i)}) \\
-		& \ge \TruncExp(t_{k(i)}; \beta (n_T + I_0), t_{k(i)-1}, t_{k(i)}),
-	\end{align*}
-	with the last inequality following from proposition \ref{pro:tru}, second,
-	$$1-p_i = \exp\{-\gamma (t_{end} - z^T_i)\} \ge \exp\{-\gamma (t_{end} - t_{k(i)-1})\},$$
-	third,
-	$$p_i \TruncExp(z^J_i; \gamma, z^T_i, t_{end}) = \Exp(z^J_i - z^T_i; \gamma)1(z^T_i < z^J_i \le t_{end}) \ge \Exp(t_{end} - t_{k(i) - 1}; \gamma)$$
-	where the equality holds since $p_i = P(z^J_i \le t_{end}|z^J_i > z^T_i)$ is equal to the normalizing constant of the truncated exponential.
-
-	Taken together, these inequalities give
-	\begin{align*}
-		g(\z_1; \theta)
-		& \ge \prod_{i=1}^{n_T} \TruncExp(t_{k(i)}; \beta (I_0 + n_T), t_{k(i)-1}, t_{k(i)}) \min\{\exp\{-\gamma (t_{end} - t_{k(i) - 1})\}, \Exp(t_{end} - t_{k(i) - 1}; \gamma)\} \\
-		& = k_2(\theta) > 0
-	\end{align*}
-	which completes the proof.
-\end{proof}
-
 	\bibliographystyle{plain}
-	\bibliography{bibliography}
+	\bibliography{../bibliography}
 

### 2021-10-12 Experiments.R
E1 thinned

@@ -2,17 +2,17 @@
 #
 # Setup
 library(PDSIR)
-run_E1 <- FALSE
+run_E1 <- T
 run_E2 <- FALSE
 run_E3 <- FALSE
-run_E5 <- TRUE
+run_E5 <- F
 run_E6 <- FALSE
 
 #
 # Experiments
 if(run_E1){ # Proof of Concept
   set.seed(2)
-  output_E1 <- experiment_1_proof_of_concept(N = 1e6, save_fig = FALSE)
+  output_E1 <- experiment_1_proof_of_concept(N = 1e6, save_fig = FALSE, thin = 10)
   save(output_E1, file = "Output/RDATA/E1.RDATA")
 
   print(list(

### 2021-10-12 Experiment_output.R
E5 burnin

@@ -30,7 +30,7 @@ for(i in 3 : 200) {
 summary_E4 <- experiment_4_output_analysis(results, path = "Output/Figures/E4")
 
 load("Output/RDATA/E5.RDATA")
-summary_5 <- experiment_5_output(output_E5, path = "Output/Figures/E5", burnin = 50)
+summary_5 <- experiment_5_output(output_E5, path = "Output/Figures/E5", burnin = 1e4)
 
 load("Output/RDATA/E6_joint.RDATA")
 load("Output/RDATA/E6_single.RDATA")

### 2021-10-15 Experiments.R
.sh email test

@@ -2,7 +2,7 @@
 #
 # Setup
 library(PDSIR)
-run_E1 <- T
+run_E1 <- F
 run_E2 <- FALSE
 run_E3 <- FALSE
 run_E5 <- F
@@ -60,3 +60,7 @@ if(run_E6){ # single-site updates
 
 }
 
+
+x <- cummax(stats::rnorm(1e9))
+y <- x[seq(1, length(x), by = 1e3)]
+save(x, y, file = "test.RDATA")

### 2021-10-15 run_experiments.sh

@@ -3,6 +3,8 @@
 #SBATCH -e slurm/experiment.err
 #SBATCH --mem=2G
 #SBATCH -p scavenger
+#SBATCH --mail-type=ALL
+#SBATCH --mail-user=raphael.morsomme@duke.edu
 hostname # print hostname
 module load R
 Export R_LIBS_USER = ~/R/x86_64-pc-linux-gnu-library/4.1

### 2021-10-15 Experiments.R
.sh trial email 1e8

@@ -61,6 +61,6 @@ if(run_E6){ # single-site updates
 }
 
 
-x <- cummax(stats::rnorm(1e9))
+x <- cummax(stats::rnorm(1e8))
 y <- x[seq(1, length(x), by = 1e3)]
 save(x, y, file = "test.RDATA")

### 2021-10-15 Experiments.R
.sh email 1e7

@@ -61,6 +61,6 @@ if(run_E6){ # single-site updates
 }
 
 
-x <- cummax(stats::rnorm(1e8))
+x <- cummax(stats::rnorm(1e7))
 y <- x[seq(1, length(x), by = 1e3)]
 save(x, y, file = "test.RDATA")

### 2021-10-15 Manuscripts/Article/Article.tex
test overleaf
move .bib file
Updates from Overleaf

@@
 %The proof of this result is provided in Appendix \ref{app:uni}.
-	% our DA-MCMC does only requires I(0), not S(0).
+	% our DA-MCMC only requires I(0), not S(0).
 
@@
 
-	In this section, the convergence properties of the DA-MCMC algorithm are examined with a simulation study.
-Each data set in this study is simulated from the stochastic SIR process and the prior distributions on the parameters are the independent weakly informative distributions $\beta \sim Ga(0.1, 1)$ and $\gamma \sim Ga(1,1)$.
-
-	First, the mixing property of the Markov chain is assessed in a medium-size population whose initial configuration is $(S(0), I(0)) = (1000, 10)$.
-The parameters are $(\beta, \gamma) = (0.003, 1)$ and the process is observed until time $t_{end} = 6$ when the pandemic has completed most of its course but is not over yet (see Figure \ref{fig:long}, where $I(t_{end}) = 55$).
-The numbers of infections in $K = 10$ time intervals of equal length are observed and correspond to $T_{1:K} = (40, 111, 193, 259, 178, 93, 29, 19, 9, 6)$.
-	The Markov chain is run for $100,000$ iterations from the initial values $(\beta^{(1)}, \gamma^{(1)}) = (0.0003, 0.1) = (\beta/10, \gamma/10)$ and the event times of $202$ ($\rho = 0.2$) individuals are updated in each Metropolis-Hastings step for the latent data.
-The algorithm takes in $2.5$ minutes to complete on a personal laptop.
+	We examine the convergence properties of the proposed DA-MCMC algorithm with simulation studies.
+First, we assess the performance of the algorithm in a medium-sized population of with over $1000$ individuals.
+We simulate a pandemic with parameters $(\beta, \gamma) = (0.0025, 1)$, which gives $R_0 = 2.5$, starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$ when the pandemic has completed most of its course but is not over yet (see Figure \ref{fig:fig:E1_trajectories}, where $I(t_{end}) = 55$).
+The numbers of infections in $K = 10$ time intervals of equal length are observed, corresponding to $I_{1:K} = (40, 111, 193, 259, 178, 93, 29, 19, 9, 6)$.
+	We set the independent weakly informative prior distributions on the parameters $\beta \sim Ga(0.01, 1)$ and $\gamma \sim Ga(2,2)$.
+	The Markov chain is initialized in a low density region at $(\beta^{(0)}, \gamma^{(0)}) = (0.0003, 0.1) = (\beta/10, \gamma/10)$ and is run for $1$ million iterations.
+We keep every tenth draw for memory reason and set $\rho = 0.1$ so that the trajectories of $\rho (S(0)+I(0)) = 101$ individuals are updated in each Metropolis-Hastings step.
+The algorithm runs takes $25$ minutes to complete on a personal laptop.
+
 	Figure \ref{fig:traceplot} shows the traceplots of the parameters $\beta$ and $\gamma$, the log likelihood and the basic reproduction number $R_0 = S(0) \beta / \gamma$, which corresponds to the average number of secondary infections generated by an infectious individual in a susceptible population.
@@
 		\centering
-		\includegraphics[scale = 0.3]{E1_trajectories}
-		\caption{Trajectories of the $S$, $I$ and $R$ compartments of a SIR process for a medium-size population ($S_0 = 1,000$).}
-		\label{fig:long}
-	\end{figure}
-
-	\begin{figure}
-		\centering
-		\begin{subfigure}[b]{0.41\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E1_no_burn_beta_tp}
-			\caption{$\beta$}
-			\label{fig:traceplot_beta}
+		\begin{subfigure}[b]{0.42\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E1_trajectories}
+			\caption{Compartment trajectories}
+			\label{fig:E1_trajectories}
 		\end{subfigure}
@@
 			\centering
-			\includegraphics[width=\textwidth]{E1_burn_beta_acf}
+			\includegraphics[width=\textwidth]{E1_no_burn_gamma_tp}
 			\caption{$\gamma$}
@@
 			\centering
-			\includegraphics[width=\textwidth]{E1_no_burn_R0_hist}
+			\includegraphics[width=\textwidth]{E1_no_burn_gamma_hist}
 			\caption{Log likelihood}
@@
 			\centering
-			\includegraphics[width=\textwidth]{E1_burn_R0_acf}
+			\includegraphics[width=\textwidth]{E1_burn_gamma_acf}
 			\caption{$R_0$}
@@
 		\end{subfigure}
-		\caption{Traceplots of the Markov chain (including burn-in).}
-		\label{fig:traceplot}
+		\caption{Performance of the proposed DA-MCMC in a medium-sized population $(S(0), I(0)) = (1000, 10)$.}
+		\label{fig:E1}
 	\end{figure}
@@
 	\bibliographystyle{plain}
-	\bibliography{../bibliography}
+	\bibliography{bibliography}
 

### 2021-10-15 Experiment_output.R
E1 output for manuscript

@@ -4,14 +4,16 @@ library(PDSIR)
 load("Output/RDATA/E1.RDATA")
 
 summary_E1 <- experiment_1_output_analysis(
-  output_E1, plot_id = "E1", path = "Output/Figures/E1",
-  n_max = 1e6, thin = 10
-  )
+  output_E1, plot_id = "E1_short", path = "Output/Figures/E1", n_max = 1e4, burnin = 1e3
+)
+summary_E1 <- experiment_1_output_analysis(
+  output_E1, plot_id = "E1", path = "Output/Figures/E1"
+)
 
 print(list(
   summary_E1$summary_no_burn$ESS,
   summary_E1$summary_no_burn$ESS_sec,
-  summary_E1$summary_no_burn$mean_quant,
+  summary_E1$summary_burn$post_stat,
   output_E1$SIR$MLE
 ))
 

### 2021-10-15 Experiments.R

@@ -2,17 +2,20 @@
 #
 # Setup
 library(PDSIR)
-run_E1 <- F
+run_E1 <- T
 run_E2 <- FALSE
-run_E3 <- FALSE
+run_E3 <- T
 run_E5 <- F
 run_E6 <- FALSE
 
 #
 # Experiments
 if(run_E1){ # Proof of Concept
-  set.seed(2)
-  output_E1 <- experiment_1_proof_of_concept(N = 1e6, save_fig = FALSE, thin = 10)
+  set.seed(1)
+  output_E1 <- experiment_1_proof_of_concept(
+    N = 1e6, save_fig = FALSE, thin = 10,
+    theta_0_factor = 0.1
+    )
   save(output_E1, file = "Output/RDATA/E1.RDATA")
 
   print(list(

### 2021-10-15 Manuscripts/Article/Article.tex
Updates from Overleaf

@@
 	from which one can easily generate independent values with a Monte Carlo sampler to explore the posterior distribution of parameters $\pi(\theta|\X)$.
+
+	% the basic reproduction number $R_0 = S(0) \beta / \gamma$, which corresponds to the average number of secondary infections generated by an infectious individual in a susceptible population
 
@@
 
-	Figure \ref{fig:traceplot} shows the traceplots of the parameters $\beta$ and $\gamma$, the log likelihood and the basic reproduction number $R_0 = S(0) \beta / \gamma$, which corresponds to the average number of secondary infections generated by an infectious individual in a susceptible population.
-We observe that the Markov chain quickly migrates from the low density region of the space where $(\beta^{(1)}, \gamma^{(1)})$ is located to a high density region where it mixes well.
-The acceptance rate of the latent spaces proposed in the Metropolis-Hastings step is $0.11$.
+	Figure \ref{fig:E1_no_burn_gamma_tp} shows the traceplot of the parameter $\gamma$ among the first $10000$ iterations.
+We observe that the Markov chain quickly migrates from the low density region where it is initialized to the high density region where it mixes well.
+The acceptance rate of the latent spaces proposed in the Metropolis-Hastings step is $0.18$.
+
 	Excluding the first $10,000$ iterations of the Markov chain as a burn-in, the posterior means of $\beta$, $\gamma$ and $R_0$ are respectively $0.00304$, $0.995$ and $3.07$.
@@
 			\caption{$\gamma$}
-			\label{fig:traceplot_gamma}
+			\label{fig:E1_no_burn_gamma_tp}
 		\end{subfigure}
@@
 			\caption{Log likelihood}
-			\label{fig:traceplot_loglik}
+			\label{fig:E1_no_burn_gamma_hist}
 		\end{subfigure}
@@
 			\centering
-			\includegraphics[width=\textwidth]{E1_burn_gamma_acf}
+			\includegraphics[width=\textwidth]{E1_no_burn_gamma_acf}
 			\caption{$R_0$}
-			\label{fig:traceplot_R0}
+			\label{fig:\textbf{traceplot_R0}}
 		\end{subfigure}

### 2021-10-21 Manuscripts/Article/Article.tex

@@
 \newcommand{\ram}[1]{{\color{green}{ #1}}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
 
@@
 For large population, however, the acceptance rate may be too low, thereby hindering the mixing of the Markov chain.
-To address this issue, we propose to update the event times of only a fraction $0<\rho\le 1$ of the individuals in the population.
-Smaller values for $\rho$ result in smaller jumps in the latent space and a larger acceptance rate.
+To address this issue, we propose to update the event times of only a fraction $0<\rho\le 1$ of the individuals in the population; that is, in the Metropolis-Hastings step we only update the infection and removal times of $\lfloor\rho (S(0)+I(0))\rfloor$ individuals chosen randomly.
+	Smaller values for $\rho$ result in smaller jumps in the latent space and a larger acceptance rate, which may lead to better mixing in certain cases.
 	Third, if all event times are updated ($\rho = 1$), then the proposed and current latent data are independent conditionally on the current values of the parameters: $\Z^{(k-1)} \perp \Z^* | \theta^{(k)}$.
@@
 	\label{sec:sim}
-	%Outline: setup; results: posterior mean, credible interval, frequentist coverage, run time, ACF, acceptance rate; table; figure: traceplot, ACF; tuning: update proportion v.
-acceptance rate; compare with others methods e.g.
-chain binomial, exact proposal.
-
-	% 1. convergence in medium-size population
-	% 2. rho
-	% 3. nominal coverage should be 95%: boxplot of posterior means, including coverage (one setting is enough for the moment)
-	% 4. distribution of estimate for PD-SIR (unbiased) vs. LNA (biased)
+	%Outline: (1) proof of concept, (1bis) coverage, (2) rho (3) single-site versus joint proposal
+
 	% explain why not compare chain binomial, particle filtering (forward simulation), LNA requires noisy data.
 
-	% TODO: Tables: posterior means, credible intervals, effective sample size.
-	% TODO: frequentist coverage for 3 different observation schedules
-
 	We examine the convergence properties of the proposed DA-MCMC algorithm with simulation studies.
-First, we assess the performance of the algorithm in a medium-sized population of with over $1000$ individuals.
-We simulate a pandemic with parameters $(\beta, \gamma) = (0.0025, 1)$, which gives $R_0 = 2.5$, starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$ when the pandemic has completed most of its course but is not over yet (see Figure \ref{fig:fig:E1_trajectories}, where $I(t_{end}) = 55$).
-The numbers of infections in $K = 10$ time intervals of equal length are observed, corresponding to $I_{1:K} = (40, 111, 193, 259, 178, 93, 29, 19, 9, 6)$.
-	We set the independent weakly informative prior distributions on the parameters $\beta \sim Ga(0.01, 1)$ and $\gamma \sim Ga(2,2)$.
-	The Markov chain is initialized in a low density region at $(\beta^{(0)}, \gamma^{(0)}) = (0.0003, 0.1) = (\beta/10, \gamma/10)$ and is run for $1$ million iterations.
-We keep every tenth draw for memory reason and set $\rho = 0.1$ so that the trajectories of $\rho (S(0)+I(0)) = 101$ individuals are updated in each Metropolis-Hastings step.
-The algorithm runs takes $25$ minutes to complete on a personal laptop.
-
-	Figure \ref{fig:E1_no_burn_gamma_tp} shows the traceplot of the parameter $\gamma$ among the first $10000$ iterations.
-We observe that the Markov chain quickly migrates from the low density region where it is initialized to the high density region where it mixes well.
-The acceptance rate of the latent spaces proposed in the Metropolis-Hastings step is $0.18$.
-
-	Excluding the first $10,000$ iterations of the Markov chain as a burn-in, the posterior means of $\beta$, $\gamma$ and $R_0$ are respectively $0.00304$, $0.995$ and $3.07$.
-
-	% TODO: Results: convergence (R statistic, Geweke), credible intervals.
-
-	% TODO: Geweke
+First, we assess the performance of the DA-MCMC algorithm in a medium-sized population of over $1000$ individuals.
+We simulate a pandemic with parameters $(\beta, \gamma) = (0.0025, 1)$, which gives $R_0 = 2.5$, starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$ when the pandemic has completed most of its course but is not over yet (see Figure \ref{fig:E1_trajectories}, where $I(t_{end}) = 55$).
+The numbers of infections in $K = 10$ time intervals of equal length are observed, corresponding to $I_{1:K} = (9, 14, 21, 42, 56, 121, 190, 162, 107, 73)$.
+	We put the conjugate prior distributions on the parameters $\beta \sim Ga(0.001, 1)$ and $\gamma \sim Ga(1,1)$ independently, which are weakly informative.
+	The Markov chain is initialized in a low density region at $(\beta^{(0)}, \gamma^{(0)}) = (\beta/10, \gamma/10) = (0.00025, 0.1)$ and is run for $1$ million iterations.
+We keep every tenth draw for storage reason and set $\rho = 0.2$ so that the trajectories of $\lfloor\rho (S(0)+I(0))\rfloor = 202$ individuals are updated in each Metropolis-Hastings step.
+The algorithm takes less than $30$ minutes to complete on a personal laptop.
+
+	Figure \ref{fig:E1_short_no_burn_gamma_tp} shows the traceplot of the parameter $\gamma$ for the first $10000$ iterations.
+We observe that the Markov chain quickly migrates from the low density region where it is initialized to the mode of the target distribution.
+Figures  \ref{fig:E1_burn_gamma_tp} and \ref{fig:E1_burn_gamma_acf}, which show the traceplot of $\gamma$ after a burn-in period of $10000$ iterations and its auto-correlation function, indicate that once the chain reaches the high density region it mixes well.
+The acceptance rate of the Metropolis-Hastings step is $0.21$.
+	The posterior means of $\beta$, $\gamma$ and $R_0$ are respectively $0.00217$, $0.815$ and $2.75$, respectively based on $398$, $370$ and $438$ effective sample sizes.
+The $90\%$ credible intervals are $(0.00166, 0.00278)$, $(0.492, 1.19)$ and $(2.26, 3.45)$ and cover the true values of the parameters.
+To verify the coverage of the DA-MCMC algorithm, the experiment was repeated $2000$ times independently.
+Tables \ref{tab:coverage} provides the nominal coverage of the $90\%$ credible intervals for the parameters $\beta$, $\gamma$ and $R_0$ along with the average and variance of the posterior means.
+The algorithm has the appropriate coverage, suggesting that running the chain for $1$ million iterations is sufficient to approximate the posterior distribution of the parameters.
 
@@
 			\centering
-			\includegraphics[width=\textwidth]{E1_no_burn_gamma_tp}
-			\caption{$\gamma$}
-			\label{fig:E1_no_burn_gamma_tp}
+			\includegraphics[width=\textwidth]{E1_short_no_burn_gamma_tp}
+			\caption{Traceplot of $\gamma$ during the transient phase}
+			\label{fig:E1_short_no_burn_gamma_tp}
 		\end{subfigure}
@@
 			\centering
-			\includegraphics[width=\textwidth]{E1_no_burn_gamma_hist}
-			\caption{Log likelihood}
-			\label{fig:E1_no_burn_gamma_hist}
+			\includegraphics[width=\textwidth]{E1_burn_gamma_tp}
+			\caption{Traceplot of $\gamma$ after the transient phase}
+			\label{fig:E1_burn_gamma_tp}
 		\end{subfigure}
@@
 			\centering
-			\includegraphics[width=\textwidth]{E1_no_burn_gamma_acf}
-			\caption{$R_0$}
-			\label{fig:\textbf{traceplot_R0}}
-		\end{subfigure}
-		\caption{Performance of the proposed DA-MCMC in a medium-sized population $(S(0), I(0)) = (1000, 10)$.}
+			\includegraphics[width=\textwidth]{E1_burn_gamma_acf}
+			\caption{Auto-correlation function for $\gamma$}
+			\label{fig:E1_burn_gamma_acf}
+		\end{subfigure}
+		\caption{Performance of the proposed DA-MCMC in a medium-sized population $(S(0), I(0)) = (1000, 10)$ for the parameter $\gamma$.}
 		\label{fig:E1}
@@
 
-	Second, the impact of the tuning parameter $\rho$ on the performance of the MCMC algorithm is evaluated.
-In particular, we examine the acceptance rate of the Metropolis-Hastings step, the total run time of the algorithm and the auto-correlation function of the parameters.
-	Three SIR processes with different population sizes $S(0) = (10^2, 10^3, 10^4)$ are simulated.
-We set $I(0) =10$, $\gamma=1$, $t_{end} = 3$ and $K = 50$ for all simulations and $\beta = (0.001, 0.01, 0.1)$ respectively for the three different population sizes.
-These values for $\beta$ ensure that the three processes have comparable dynamics.
-For each process, the DA-MCMC algorithm is run for $50,000$ iterations with different values of $\rho = (10^{-4}, 10^{-3}, 10^{-2}, 10^{-1}, 0.5, 1)$.
-The algorithm is run only if there is at least one event time updated per iteration, that is, if $\rho(S_0+I_0)>1$.
-	Figure \ref{fig:accept} presents the acceptance rate across the range of $\rho$.
-We observe that the acceptance rate decreases as the number of event times updated per iteration and the population size increase.
-	Figure \ref{fig:run_time} shows the run time of the MCMC algorithm for these simulations. The run time increases with the population size and with $\rho$.
-	Finally, Figure \ref{fig:acf} shows the effect of $\rho$ on the auto correlation function for the parameter $\beta$ for the medium-size population process ($S(0) = 1,000$) for three values of $\rho$.
+    \begin{table}
+    \centering
+    \begin{tabular}{ |C{2cm}| *{3}{C{3cm}}|}
+        \hline
+        Parameter & Coverage rate of the $90\%$ credible intervals & Average of the posterior means & SD of the posterior means \\ 
+        \hline
+        $\beta$ & 0.895 & 0.0025 & 0.000255\\ 
+        $\gamma$ & 0.902 & 1.00 & 0.142 \\ 
+        $R_0$ & 0.910 & 2.54 & 0.194 \\
+        \hline
+    \end{tabular}
+    \caption{Nominal coverage of the DA-MCMC algorithm and distribution of the posterior means from $2000$ independent runs in a medium-sized population.
+The true values of the parameters are $(\beta.
+\gamma.
+R_0) = (0.0025, 1, 2.5)$.}
+    \label{tab:coverage}
+    \end{table}
+
+	Second, we evaluate the impact of the tuning parameter $\rho$ on the mixing properties of the algorithm across different population sizes.
+	We simulate three SIR processes with respectively $S(0) \in \{250, 500, 1000\}$.
+We set $I(0)=10$, $\gamma=1$, $t_{end} = 6$ and $K = 10$ for each simulation and use $R_0 \in \{3, 2.5, 2\}$ respectively for the different population sizes.
+These values for $R_0$ were chosen to give pandemics with comparable dynamics.
+We consider the values $\rho \in \{0.02, 0.05, 0.1, 0.25, 0.5, 1\}$.
+For each population size and each value of $\rho$, the DA-MCMC algorithm is run for $1$ million iterations.
+%Note that the algorithm is not run if the number of trajectories updated $\rho(S_0+I_0)$ is smaller than $1$.
+
+	Figure \ref{fig:E3_runtime} and \ref{fig:E3_accept} presents the run time of the algorithm and the acceptance rate in the Metropolis-Hastings step.
+We observe that the run time increases with the population size and with $\rho$ and that, as expected, the acceptance rate decreases as the number of trajectories updated per iteration $\rho(S(0)+I(0))$ increases.
+	Finally, Figure \ref{fig:E3_facet_ESSsecR0} shows the effect of $\rho$ on the auto correlation function for the parameter $\beta$ for the medium-size population process ($S(0) = 1,000$) for three values of $\rho$.
 We observe that updating half of the augmented data per iteration yields the lowest auto-correlation function across the values of $\rho$ considered.
@@
 		\centering
-		\includegraphics[scale = 0.4]{E3_facet_ESSsecR0}
+		\begin{subfigure}[b]{0.42\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E3_runtime}
+			\caption{Run time in minutes}
+			\label{fig:E3_runtime}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.41\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E3_accept}
+			\caption{Acceptance rate}
+			\label{fig:E3_accept}
+		\end{subfigure}
+		\\
+		\begin{subfigure}[b]{0.41\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E3_facet_ESSsecR0}
 			\caption{ESS/sec for $R_0$
@@
 			}
-		\label{fig:ESS_sec}
+			\label{fig:E3_facet_ESSsecR0}
+		\end{subfigure}
+		\caption{Impact of $\rho$ on the performance of the DA-MCMC across different population sizes.}
+		\label{fig:E3}
 	\end{figure}
 
+	Finally, we compare the performance of the proposed DA-MCMC algorithm in which event times are jointly proposed to that of a single-site-update DA-MCMC (SSU DA-MCMC) that is similar in spirit to \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
+The SSU DA-MCMC that we consider here is a special case of the proposed DA-MCMC in which $\rho = (S(0)+I(0))^{-1}$, so that the infection and removal times of a single individual are updated per iteration.
+The two algorithms are run on the same data for $1$ million iterations.
+Figure \ref{fig:E6} presents the traceplots for $\gamma$ as well as the auto-correlation function.
+We observe that the Markov chain mixes much better when event times are jointly proposed in the Metropolis-Hastings step.
+This is confirmed by Table \ref{tab:E6} which shows that the effective sample size per second obtained with the proposed DA-MCMC is between 7 and 20 times larger than that of the SSU DA-MCMC.
+
 	\begin{figure}
 		\centering
-		\centering
-		\includegraphics[width=\textwidth]{E1_burn_R0_acf}
-		\caption{$\rho = 0.1$}
-		\label{fig:acf}
+		\begin{subfigure}[b]{0.42\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E6_no_burn_beta_tp_joint.jpg}
+			\caption{DA-MCMC}
+			\label{fig:E6_no_burn_beta_tp_joint}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.41\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E6_no_burn_beta_tp_single.jpg}
+			\caption{SSU DA-MCMC}
+			\label{fig:E6_no_burn_beta_tp_single}
+		\end{subfigure}
+		\\
+		\begin{subfigure}[b]{0.42\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E6_burn_beta_acf_joint.jpg}
+			\caption{DA-MCMC}
+			\label{fig:E6_burn_beta_acf_joint}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.41\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E6_burn_beta_acf_single.jpg}
+			\caption{SSU DA-MCMC}
+			\label{fig:E6_burn_beta_acf_single}
+		\end{subfigure}
+		\caption{Traceplots and auto-correlation functions of the proposed DA-MCMC and the SSU DA-MCMC for $\beta$.}
+		\label{fig:E6}
 	\end{figure}
 
-	%Figures: acceptance rate v. proportion of updates (see Neal, Roberts A case study in non-centering -- table 1 p.319).
+	 \begin{table}
+        \centering
+        \begin{tabular}{ |C{2cm}| *{2}{C{3cm}}|}
+            \hline
+            Parameter & DA-MCMC & SSU DA-MCMC \\ 
+            \hline
+            $\beta$ & 0.20 & 0.01 \\ 
+            $\gamma$ & 0.19 & 0.01 \\ 
+            $R_0$ & 0.38 & 0.05 \\
+            \hline
+        \end{tabular}
+        \caption{Effective sample size per second for for the proposed DA-MCMC and the SSU DA-MCMC.}
+        \label{tab:E6}
+    \end{table}
 
@@
 	Between the end of $2013$ and $2015$, Guinea, along with several neighboring countries, experienced the largest outbreak of the Ebola virus disease in history.
-The virus, which has a fatality rate of $70\%$, was responsible for the death of almost $2,000$ people in Guinea alone during the outbreak.
+The virus, which has a fatality rate of $70\%$, was responsible for the death of almost $2,000$ people in Guinea alone.
 	The outbreak is believed to have originated in the Gu\'eck\'edou prefecture of Guinea at the end of November $2013$ \cite{Baize.2014}.
 Weekly infection incidence counts are available for each prefecture for the $73$ weeks between the end of December $2013$ and May $2015$.
+
 	We fit the stochastic SIR model to these incidence counts for the Gu\'eck\'edou prefecture using the MCMC algorithm proposed in this article.
 	For simplicity, we assume that the population of the prefecture forms a closed population, that the model's parameters remained constant throughout the outbreak and that the reported infections counts are exact.
-Applying our MCMC algorithm to stochastic epidemic models where these assumptions are relaxed is the subject of ongoing research.
-	The \textit{effective} population size is set to $n = 150,000$, roughly half of the total population in $2014$.
-Note that as long as the number of infections is small compared to the total population size, the exact population size does not need to be known to estimate the parameters of the model.
-Furthermore, since the first documented infection occurred late November, one month before the first reported incidence count \cite{Baize.2014}, the number of infectious individuals at the beginning of the observation period is set to $I(0) = 10$, with the units of time corresponding to ``days" in the analysis.
-
-	The Markov chain is run for $50,000$ iterations and the event times of $10\%$ of the individuals are updated each iteration.
-The initial values of the parameters are set to $(\beta^{(1)}, \gamma^{(1)}) = (10^{-7}, 0.05)$ and the conjugate distributions in Equation \ref{eq:pri} are used as priors with $a_{\beta} = b_{\beta} = a_{\gamma} = b_{\gamma} = 0.01$ to make them weakly informative.
+	\ram{Justify these assumptions.}
+	Applying the proposed DA-MCMC algorithm to stochastic epidemic models where these assumptions are relaxed is the subject of ongoing research.
+	The population size is set to $n = 292000$, the estimated population of the Gu\'eck\'edou prefecture in $2014$.
+	The units of time correspond to ``days" in the analysis and $t=0$ corresponds to Monday December 30, 2013, the first day of the observation period.
+	Since the first documented infection occurred late November \cite{Baize.2014}, one month before the first reported incidence count, we set $I(0) = 5$.
+
+	The Markov chain is run for $1$ million iterations and the event times of $\rho=10\%$ of the individuals are updated each iteration.
+The initial values of the parameters are set to $(\beta^{(0)}, \gamma^{(0)}) = (10^{-7}, 0.05)$ and the conjugate distributions in Equation \ref{eq:pri} are used as priors with $a_{\beta} = b_{\beta} = a_{\gamma} = b_{\gamma} = 1$ to make them weakly informative.
 Even for such a large population, the total run time of the algorithm was less than $100$ minutes on a personal laptop.
-	The Metropolis-Hastings step for the latent space proposals achieves a healthy $22.1\%$ acceptance rate, and the first $10,000$ iterations of the Markov chain are discarded as a burn-in.
-	Figure \ref{fig:ebola} shows the joint posterior distributions of $(\beta, \gamma)$, whose posterior means are respectively $1.06 * 10^{-6}$ and $0.109$, as well as the marginal distribution of the basic reproduction number $R_0$.
-This value for $\gamma$ indicates that people remain infectious for around $9$ days on average, which is consistent with existing literature.
-As noted by numerous authors, conditionally on partially observed data, the parameters $\beta$ and $\gamma$ appear positively correlated.
+	The Metropolis-Hastings step for the latent data proposals achieves a healthy $22.1\%$ acceptance rate, and the first $50,000$ iterations of the Markov chain are discarded as a burn-in.
+	Figure \ref{fig:ebola} shows the marginal posterior distributions $\gamma$, whose means is $0.109$.
+This indicates that people remained infectious for around $9$ days on average, which is consistent with existing literature.
+	%As noted by numerous authors, conditionally on partially observed data, the parameters $\beta$ and $\gamma$ appear positively correlated.
 	The posterior distribution of $R_0$ is unimodal, relatively symmetric, and centered around $1$.
@@
 			\centering
-			\includegraphics[width=\textwidth]{E5_expected_infection_length_tp}
-			\caption{Joint posterior density of $\theta = (\beta, \gamma)$. The $50\%$ and $90\%$ confidence regions are shown.}
-			\label{fig:density_joint}
+			\includegraphics[width=\textwidth]{E5_gamma_hist}
+			\caption{Histogram of $\gamma$}
+			\label{fig:E5_gamma_hist}
 		\end{subfigure}
@@
 			\centering
-			\includegraphics[width=\textwidth]{E5_expected_infection_length_hist.jpg}
+			\includegraphics[width=\textwidth]{E5_gamma_acf}
 			\caption{Posterior density of $R_0$.}
@@
 	\label{app:ldp}
-	Theorem \ref{theo:ldp} was first proved by \cite{Neuts.1971}. Since this is a result that has rarely been mentioned in the literature, we provide a proof in this appendix from (cite Ross textbook).
-
+
+	Theorem \ref{theo:ldp} was first proved by by \cite{Neuts.1971}. Ross provides a simpler proof which we now give.
+
+	\begin{proof}
 	Consider a linear pure death process with $n$ particles and individual death rate $\mu$.
@@
 	which corresponds to the kernels of independent exponential distribution truncated above by $t$.
+
+	\end{proof}
 

### 2021-11-02 Experiment_output.R
cosmetic changes

@@ -1,13 +1,14 @@
 
 library(PDSIR)
 
+# E1
 load("Output/RDATA/E1.RDATA")
 
 summary_E1 <- experiment_1_output_analysis(
   output_E1, plot_id = "E1_short", path = "Output/Figures/E1", n_max = 1e4, burnin = 1e3
 )
 summary_E1 <- experiment_1_output_analysis(
-  output_E1, plot_id = "E1", path = "Output/Figures/E1"
+  output_E1, plot_id = "E1", path = "Output/Figures/E1", burnin = 1e4
 )
 
 print(list(
@@ -17,7 +18,7 @@ print(list(
   output_E1$SIR$MLE
 ))
 
-
+# E3
 load("Output/RDATA/E3.RDATA")
 experiment_3_output_analysis(output_E3, path = "Output/Figures/E3")
 
@@ -32,7 +33,9 @@ for(i in 3 : 200) {
 summary_E4 <- experiment_4_output_analysis(results, path = "Output/Figures/E4")
 
 load("Output/RDATA/E5.RDATA")
-summary_5 <- experiment_5_output(output_E5, path = "Output/Figures/E5", burnin = 1e4)
+summary_5 <- experiment_5_output(
+  output_E5, path = "Output/Figures/E5", burnin = 5e4, thin = 10
+  )
 
 load("Output/RDATA/E6_joint.RDATA")
 load("Output/RDATA/E6_single.RDATA")

### 2021-11-02 Experiments.R

@@ -11,9 +11,9 @@ run_E6 <- FALSE
 #
 # Experiments
 if(run_E1){ # Proof of Concept
-  set.seed(1)
+  set.seed(0)
   output_E1 <- experiment_1_proof_of_concept(
-    N = 1e6, save_fig = FALSE, thin = 10,
+    N = 1e6, save_fig = FALSE, thin = 10, rho = 1/5,
     theta_0_factor = 0.1
     )
   save(output_E1, file = "Output/RDATA/E1.RDATA")

### 2021-11-02 Manuscripts/Article/Article.tex

@@
-%\documentclass{article}
 %\usepackage[utf8]{inputenc}
@@
 		Stochastic epidemic models provide an interpretable probabilistic description of the spread of a disease through a population.
-Yet, fitting these models in missing data settings is a notably difficult task; in particular, when the epidemic process is only partially observed, the likelihood of many classic models becomes intractable.
+Yet, fitting these models in missing data settings is a difficult task; in particular, when the epidemic process is only partially observed, the likelihood of many classic models becomes intractable.
 To remedy this issue, this article introduces a data-augmented MCMC algorithm for fast and exact inference under the stochastic SIR model given discretely observed incidence counts.
@@
 % While existing exact MCMC approaches become intractable for populations greater than a few thousand individuals,
-		The proposed algorithm enables exact inference and scales to outbreaks in populations of hundreds of thousands, even on a single laptop.
+		The proposed algorithm enables exact inference that scales to outbreaks in populations of hundreds of thousands, even on a single laptop.
 We validate its performance via thorough simulation experiments, consider and a case study on the $2014$ Ebola outbreak in Western Africa.
@@
 Challenges stem from the fact that the observed data typically provide incomplete information on a process that evolves continuously through time, making the likelihood of the model intractable.
-Were the infection and recovery times of each individual of the population observed, conducting inference on SEMs would be straightforward, but this is rarely the case.
+Were the infection and recovery times of each individual of the population observed, inference would be straightforward, but this is rarely the case.
 In practice, one often has access to either incidence data such as weekly counts of new infections, or prevalence data such as the numbers of people infectious at discrete reporting times.
@@
 
-	Our point of departure is to consider the stochastic SIR model -- also referred to as the general epidemic model -- which offers a parsimonious and interpretable representation of the mechanistic, population-level dynamics of an epidemic.
+	Our point of departure is to consider the stochastic SIR model -- also referred to as the general stochastic epidemic model (Bailey, 1975) -- which offers a parsimonious and interpretable representation of the mechanistic dynamics of an epidemic.
 The stochastic SIR model is a compartmental model in which the individuals of a population transition through three types or compartments: susceptible (S), infectious (I) and removed (R).
@@
 
-	Assuming a closed population with a fixed size $n$, the stochastic SIR model consists of a continuous-time vector-valued process
+	Assuming a closed population of size $n$, the stochastic SIR model consists of a continuous-time vector-valued process
 	\begin{equation}
@@
 	denotes the status (compartment) of individual $j$ at time $t$ with $\tau^I_j$ and $\tau^R_j$ respectively the infection and removal times of individual $j$.
-If individual $j$ never becomes infectious, then we set $\tau^I_j = \tau^R_j = \infty$ and $X_j(t) = s$ for $t \in [0, \infty)$.
-Here $\chi_{\X}$ denotes the set of valid trajectories compatible with the evolution of a disease---that is, the set of trajectories in which no infection occurs whenever the infectious compartment is depleted :
+If individual $j$ never becomes infectious, we set $\tau^I_j = \tau^R_j = \infty$ and $X_j(t) = s$ for $t \in [0, \infty)$.
+Here $\chi_{\X}$ denotes the set of trajectories compatible with the evolution of a disease---that is, the set of trajectories in which no infection occurs whenever the infectious compartment is depleted:
 	\begin{equation}
@@
 	The stochastic SIR model is specified by the rates at which individuals move from one compartment to another.
-If we assume a homogeneously mixing population where contacts between individuals occur independently at some rate constant $\beta$, then the contacts between two given individuals are said to follow a Poisson process with rate $\beta$.
+If we assume a homogeneously mixing population where contacts between individuals occur independently at some constant rate $\beta$, the contacts between two given individuals are said to follow a Poisson process with rate $\beta$.
 This parameter can be interpreted as the \textit{infection rate}: when a susceptible individual comes into contact with an infectious individual, she immediately becomes infectious.
 	If we also make the common assumption that the infectious periods follow independent exponential distributions with rate $\gamma$, then the process \eqref{eq:X} is a time–homogeneous continuous time Markov chain, whose instantaneous transition rates
-	%from state $\x$ to $\x'$
 	are given by the matrix $\Lambda = [\lambda_{\x, \x'}]$ with
@@
 	%Outline: complete-data likelihood; MLE; gamma conjugacy
-
-	When the Markov process \eqref{eq:X} is completely observed until time $t_{end}$, we obtain the following likelihood
+	\jx{Need some citations here for form of likelihood, conjugate/MLE forms, etc} 
+	When the Markov process \eqref{eq:X} is completely observed until time $t_{end}$, we obtain the following likelihood \ram{ref for likelihood}
 	\begin{align}
@@
 	and $\chi_{\theta} = (0, \infty) \times (0, \infty)$ denotes the parameter space, 
-	the index sets $\mathcal{I} = \{j\in (1, \cdots, n): \tau^I_j \in (0, t_{end}]\}$ and $\mathcal{R} = \{j\in (1, \cdots, n): \tau^R_j \in (0, t_{end}]\}$ denote the individuals that are respectively infected and removed during the observation interval $(0, t_{end}]$,
+	the index sets $\mathcal{I} = \{j: \tau^I_j \in (0, t_{end}]\}$ and $\mathcal{R} = \{j: \tau^R_j \in (0, t_{end}]\}$ denote the individuals that are respectively infected and removed during the observation interval $(0, t_{end}]$,
 	$n_I = |\mathcal{I}|$ and $n_R = |\mathcal{R}|$ are the numbers of observed infections and removals,
-	and	$S(t) = \#\{X_i(t) = S\}$ is the number of susceptible individuals at time $t$.
+	and	$S(t) = \#\{X_i(t) = s\}$ is the number of susceptible individuals at time $t$.
 
 	It is straightforward to conduct inference in this continuously observed scenario.
-In particular, the likelihood \eqref{eq:cdl} belongs to the exponential family, and maximum likelihood estimates of its parameters can be expressed in terms of the sufficient statistics defined above as
+The likelihood \eqref{eq:cdl} belongs to the exponential family, and maximum likelihood estimates of its parameters can be expressed in terms of the sufficient statistics defined above as \ram{ref for exponential family and MLE}
 	$$\hat{\beta}_{MLE} = \dfrac{n_I}{ \int_{0}^{t_{end}} I(t)S(t)dt}, \quad \hat{\gamma}_{MLE} = \dfrac{n_R}{\int_{0}^{t_{end}} I(t)dt}.$$
-	Since the functions $I$ and $S$ are constant between event times, the two integrals correspond to finite sums which are straightforward to compute.
-	Furthermore, in a Bayesian context, inference is facilitated by the conjugacy with the Gamma distribution: if we let the parameters follow independent Gamma prior distributions
+	Note that since the functions $I$ and $S$ are constant between event times, the two integrals correspond to finite sums which are straightforward to compute.
+	Furthermore, in a Bayesian context, inference is facilitated by the conjugacy with the Gamma distribution: if we use independent Gamma prior distributions
 	\begin{equation}
@@
 	\end{equation}
-	where $Ga(a,b)$ denotes the parametrization with mean $a/b$ and variance $a/b^2$, then the posterior distributions of the parameters are independent Gamma distributions
+	where $Ga(a,b)$ denotes the parametrization with mean $a/b$ and variance $a/b^2$, we obtain independent Gamma posterior distributions
 	\begin{equation}
-		\label{eq:pdb}
-		\beta | \X \sim Ga\left( a_{\beta} + n_I, b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt\right)
-	\end{equation}
-	and
-	\begin{equation}
-		\label{eq:pdg}
+		\label{eq:posterior_theta}
+		\beta | \X \sim Ga\left( a_{\beta} + n_I, b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt\right), \qquad 
 		\gamma | \X \sim Ga\left( a_{\gamma} + n_R, b_{\gamma} + \int_{0}^{t_{end}} I(t) dt\right)
 	\end{equation}
-	from which one can easily generate independent values with a Monte Carlo sampler to explore the posterior distribution of parameters $\pi(\theta|\X)$.
-
-	% the basic reproduction number $R_0 = S(0) \beta / \gamma$, which corresponds to the average number of secondary infections generated by an infectious individual in a susceptible population
+	from which one can easily generate independent values with a Monte Carlo sampler to explore the posterior distribution of parameters $\pi(\theta|\X)$ \cite{Tierney.1994}.
+
+	% the basic reproduction number $R_0 = S(0) \beta / \gamma$, which corresponds to the expected number of secondary infections caused by an infectious individual in a susceptible population.
+$\R_0 = S(0) \beta/\gamma$.
 
@@
 
-	In practice, inference is complicated by the fact that the process \eqref{eq:X} is only partially observed.
-When we do not observe all epidemic events, which is virtually always the case with real data, we do not have access to the sufficient statistics needed to evaluate the likelihood \eqref{eq:cdl}.
+	In practice, inference is complicated by the fact that the epidemic process \eqref{eq:X} is typically only partially observed.
+When we do not observe all event times, we do not have access to the sufficient statistics needed to evaluate the likelihood \eqref{eq:cdl}.
 Various types of partially observed data typify real data and have been considered in the literature, such as observing only the removal times \cite{Gibson.1998, ONeill.1999} or the number of infectious individuals in the population at discrete points in time \cite{Fintzi.2017}.
-For instance, the former arise in animal experiments in which positive cases are immediately isolated from the rest of the population and no longer contribute to disease spread.
+For instance, the former arises in animal experiments in which positive cases are immediately isolated from the rest of the population and no longer contribute to disease spread.
 In such cases, exact times of removals are known, but such information is unavailable in typical observational studies.
 
-	In this article, we focus on partially observed \textit{incidence} counts of infections at discrete points in time.
-Such data arise when an infectious individual may be identified some time after onset, and may continue to infect others even after being tested positive for the disease.
+	In this article, we consider partial data consisting of \textit{incidence} counts of infections during given time intervals, e.g.\ weekly infection counts.
+Such data arise when an infectious individual is identified some time after onset and may continue to infect others after being tested positive for the disease.
 %This was for instance the case during the $2013$–$2016$ outbreak of Ebola virus.
@@
 	%It therefore seems appropriate to model a positive test for the virus as an indication that the individual became infectious some time prior to the test rather than as a removal time.
-	Consider an observation schedule $t_{0:K}$ ($K \ge 1$) with $0 = t_0 < t_1 < \dots < t_K = t_{end}$.
-The observed data consist of the $K$-dimensional vector $\Y = I_{1:K}$ where $I_k = \#\{\tau^I_j \in (t_{k-1}, t_k]\}$ is the number of infections during the $k^{\text{th}}$ time interval.
-	For simplicity, we further assume that the initial configuration of the population $\X(0)$ is known.
-
-	If we adopt a Bayesian approach, the posterior distribution of the parameters given the observed data is formally related to the joint posterior via integration:
+	Given an observation schedule $t_{0:K}$ ($K \ge 1$) with $0 = t_0 < t_1 < \dots < t_K = t_{end}$, the observed data consist of the $K$-dimensional vector $\Y = I_{1:K}$ where $I_k = \#\{\tau^I_j \in (t_{k-1}, t_k]\}$ is the number of infections during the $k^{\text{th}}$ interval.
+	%For simplicity, we further assume that the initial configuration of the population $\X(0)$ is known.
+
+	Under a Bayesian approach, the posterior distribution of the parameters given the observed data is formally related to the joint posterior via integration:
 	\begin{equation}
@@
 		\pi(\theta|\Y) 
-		\, \, \propto \, \, L(\theta; \Y)\pi(\theta) \, \, = \, \, \int_{\chi^*_{\X}} L(\theta; \x) \pi(\theta) d\x
+		\, \, \propto \, \, L(\theta; \Y)\pi(\theta) \, \, = \, \, \int_{\chi^*_{\X}} L(\theta; \x) d\x \pi(\theta)
 	\end{equation}
-	where $\pi(\theta)$ is the prior distribution on the parameters $\theta$ and
+	where $\pi(\theta)$ is the prior distribution on $\theta$, and
 	$\chi^*_{\X} \subset \chi_{\X}$ is the set of trajectories of the process \eqref{eq:X} that are compatible with the observed data $\Y$.
 	The partial data likelihood $L(\theta; \Y)$ therefore consists of a high-dimensional integral over all epidemic paths compatible with the observed data.
-This marginalization step has no closed form solution, and presents computational challenges even for a population of moderate size.
+This marginalization step has no known closed form solution and presents computational challenges even for a population of moderate size.
 
@@
 Approximate Bayesian Computation (ABC) \cite{McKinley.2018} offers a solution to the latter problem, but its inference is based on an approximation of the model's likelihood.
-As a result, the inference is inevitably biased, and it is difficult to quantify how closely the approximate posterior resembles the original target distribution.
+As a result, the inference is inevitably biased, and it is difficult to evaluate the degree of approximation involved.
+%quantify how closely the approximate posterior resembles the original target distribution.
 
 	The second family of MCMC-based inferential methods treats the unobserved event times as nuisance parameters; that is, the observed data are augmented with latent data that consist of the times and types of unobserved epidemic events.
-Researchers have developed algorithms to explore this latent space efficiently.
+	Researchers have mostly focused on the development of better and more efficient proposal schemes to explore this high-dimensional latent space efficiently.
 Early attempts employed reversible-jump MCMC \cite{Green.1995} to explore models with different numbers of unobserved events \cite{Gibson.1998, ONeill.1999}.
-Their algorithm augments the observed data which consist of the recovery times with the unobserved infection times and explores the latent space by inserting, deleting or uniformly moving an infection time in each iteration of the MCMC algorithm.
+These authors augment the observed data which consist of the recovery times with the unobserved infection times and explore the latent space by uniformly inserting, deleting or moving an infection time in each iteration of the MCMC algorithm.
 	More recently, Fintzi et al.
-\cite{Fintzi.2017} proposed a MCMC algorithm to conduct inference with discretely observed infection counts prevalence counts.
+\cite{Fintzi.2017} proposed a MCMC algorithm to conduct inference with discretely observed infection prevalence counts.
 They constructed latent data consisting of the infection and recovery times of each individual.
-The algorithm explores this latent space by updating the event times of an individual in each iteration of the algorithm according to the exact dynamics of the stochastic process.
+The algorithm explores this latent space by updating the event times of an individual in each iteration of the algorithm.
+These event times are sampled from a distribution that approximate the exact dynamics of the stochastic process.
+	In the discrete-time setting, Touloupou et al. \cite{Touloupou.2020} used a Gibbs sampler to update the trajectory of an individual per iteration.
 	These DA-MCMC methods suffer from poor mixing in the presence of large epidemics.
-Since they update a single element of the latent data per iteration, the resulting Markov chains fail to explore the latent space efficiently.
-Although the update of multiple event times per iteration in \cite{Pooley.2015} and the non-centered parameterization in \cite{Neal.2005} improves mixing, the gains are modest and the Markov chains still suffer from a high level of auto-correlation.
-%Furthermore, generating event times from the exact dynamics of the SEM in \cite{Fintzi.2017} is computationally costly.
+Since they update a very small number of latent data per iteration, the resulting Markov chains is sticky: it makes very small jumps in the latent space and therefore requires a large number of iterations to explore it completely.
+	Although the update of a slightly larger number of event times per iteration in \cite{Pooley.2015} and the non-centered parameterization of \cite{Neal.2005} may improve the performance, the gains are modest; the Markov chains still suffer from a high level of auto-correlation and do not possess satisfactory mixing properties in populations over a few hundred individuals.
+	Even though these DA-MCMC algorithms may update more than one element per iteration, we will refer to them as \textit{single}-site DA-MCMC algorithms to contrast them to the proposed method in which the entire latent data are jointly updated.
 
@@
 For instance, Streftaris and Renshaw \cite{Streftaris.2002} generalized the approach of \cite{Gibson.1998} to a non-Markovian stochastic SIR model where the amount of time spent infectious is not necessarily exponentially distributed but follows an arbitrary distribution, and Bu et al.
-\cite{Bu.2020} analyzed models in which the population does not mix homogeneously but is characterized by a dynamic social network, and Kypraios (2018, 2020) model the infection dynamics nonparameterically.
+\cite{Bu.2020} analyzed models in which the population does not mix homogeneously but is characterized by a dynamic social network, and Kypraios  model the infection dynamics in a nonparameteric framework.
 
@@
 	%By correcting with a Metropolis-Hastings step
-	The sampler targets the exact posterior of the model given partially observed incidence data and enjoys the efficiency of fast proposals in the high-dimensional latent space as well as fast computations involving the complete data likelihood.
+	The Metropolis-Hastings sampler targets the exact posterior of the model given partially observed incidence data, and enjoys the efficiency of fast proposals in the high-dimensional latent space as well as fast computations involving the complete data likelihood.
 
 	As shown in Section \ref{sec:icd}, the complete data likelihood \eqref{eq:cdl} is amenable to computation.
-This suggests augmenting the observed data $\Y$ with latent data $\Z$ such that the likelihood $L(\theta; (\Y, \Z))$ has the closed form \eqref{eq:cdl} and constructing a Markov chain $\{(\theta^{(m)}, \Z^{(m)})\}_{m=0}^M$ with state space $\chi = \chi_{\theta} \times \chi_{\Z}$ whose stationary distribution is the joint posterior distribution $\pi(\theta, \X|\Y)$ \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
-We consider the latent data $\Z = \left\lbrace (z^I_j, z^R_j)\right\rbrace_{j=1}^{n} \in \chi_{\Z}$ which consist of the times and types of unobserved epidemic events, where the infection and removal times for individual $j$ are
+This suggests augmenting the observed data $\Y$ with latent data $\Z$ such that the likelihood $L(\theta; (\Y, \Z))$ has the closed form \eqref{eq:cdl}, and constructing a Markov chain $\{(\theta^{(m)}, \Z^{(m)})\}_{m=0}^M$ with state space $\chi = \chi_{\theta} \times \chi_{\Z}$ whose stationary distribution is the joint posterior distribution $\pi(\theta, \X|\Y)$ \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
+We consider the latent data $\Z = \left\lbrace (z^I_j, z^R_j)\right\rbrace_{j=1}^{n}$ which consist of the times and types of unobserved epidemic events, where the infection and removal times for individual $j$ are
 	$$z^I_j \begin{cases}
-		\in [0, t_{end}) & \text{if individual } j \text{ becomes infectious before } t_{end} \\
-		= \infty & \text{if individual } j \text{ does not become infectious before } t_{end} \\
+		\in [0, t_{end}) & \text{if individual } j \text{ is infected before } t_{end} \\
+		= \infty & \text{if individual } j \text{ is not infected before } t_{end} \\
 	\end{cases}$$
@@
 	\end{cases}$$
-
-	The set $\chi_{\Z} \subset (\bar{\R}^+ \times \bar{\R}^+)^n$ consists of the latent data $\Z$ compatible with the observed data $\Y$ and the progression of an epidemic (see Equation \ref{eq:chi}).
 	Since $\Z$ contains all the information present in $\Y$, we can write
@@
 
-	The data-augmented MCMC algorithm that we propose alternates between updates of the parameters $\theta$ given the current state of the latent data $\Z$, and updates of the latent data given the current state of the parameters.
-A one-step transition from $\x_0 = (\theta_0, \z_0)$ to $\x_1 = (\theta_1, \z_1)$ therefore looks like
-	$$\x_0 = (\theta_0, \z_0) \rightarrow (\theta_1, \z_0) \rightarrow (\theta_1, \z_1) = \x_1.$$
-
-	The transition kernel $P$ of the Markov chain $\{\x^{(m)}\}_m$ is thus a composition of two kernels $P_{\theta}$ and $P_{\z}$ which alternatively updates $\theta$ while keeping $\z$ fixed, and then updates $\z$ while keeping $\theta$ fixed.
-Due to the Gamma conjugacy of the complete-data likelihood mentioned in Section \ref{sec:icd}, drawing new parameter values poses no challenge: we can simply employ a Gibbs sampler to directly draw new values for $\theta = (\beta, \gamma)$ from the two independent full conditional posterior distributions given by Equations \ref{eq:pdb} and \ref{eq:pdg}.
-In these expressions, $n_I, n_R, \int S(t)I(t) dt, \int I(t) dt$ are sufficient statistics from the latent data $\z$.
-The latent data is updated in a Metropolis-Hastings step whose kernel is
-	$$
-	P_{\z}((\theta, \z_0), (\theta, \z_1)) = q(\z_1|\theta) \alpha((\theta, \z_0), (\theta, \z_1)) dz_1
-	$$
-	where the proposal density $q$ will be defined in the following section and
-	$$
-	\alpha\left( \left( \theta^{(m)}, \Z^{(m-1)}\right) , \left( \theta^{(m)}, \Z^{(m)}\right) \right) = \min\left\lbrace 1, \dfrac{L\left( \theta^{(m)}; \Z^{(m)}\right) q\left( \Z^{(m - 1)}|\theta^{(m)}\right) }{L\left( \theta^{(m)};\Z^{(m - 1)}\right) q\left( \Z^{(m)}|\theta^{(m)}\right) }\right\rbrace
-	$$
-	corresponds to the Metropolis-Hasting acceptance ratio \cite{Tierney.1994}.	
-	Algorithm \ref{alg:DA-MCMC} provides the details of the Data-Augmented MCMC.
+	We propose a data-augmented (DA) MCMC algorithm that alternates between updates of the parameters $\theta$ given the current configuration of the latent data $\Z$, and updates of the latent data given the current values of the parameters.
+A one-step transition from $\x_1 = (\theta_1, \z_1)$ to $\x_2 = (\theta_2, \z_2)$ therefore looks like
+	$$\x_1 = (\theta_1, \z_1) \rightarrow (\theta_2, \z_1) \rightarrow (\theta_2, \z_2) = \x_2$$
+
+	%The transition kernel $P$ of the Markov chain $\{\x^{(m)}\}_m = \{(\theta^{(m)}, \z^{(m)})\}_m$ is thus a composition of two kernels $P_{\theta}$ and $P_{\z}$ which alternatively updates $\theta$ while keeping $\z$ fixed, and then updates $\z$ while keeping $\theta$ fixed.
+	Due to the Gamma conjugacy of the complete-data likelihood (see Section \ref{sec:icd}), sampling new values for $\theta$ given $\z$ poses no challenge: we can simply employ a Gibbs sampler to directly draw from the two independent full conditional distributions \ref{eq:posterior_theta}.
+In these expressions, $n_I, n_R, \int S(t)I(t) dt, \int I(t) dt$ are sufficient statistics from the latent data $\z$ which are easy to compute.
+	The latent data is updated in a Metropolis-Hastings (MH) step.
+Given some current $\theta$ and $\z$, the MH step proceeds by simulating a candidate $\z^\star$ from a surrogate process $q(\z^\star|\theta)$.
+Then $\z^\star$ is accepted with probability
+	$$
+	\alpha\left( \left( \theta, \z\right) , \left( \theta, \z^\star\right) \right) =	\min\left\lbrace 1, \dfrac{L\left( \theta; \Z^\star\right) q\left( \Z|\theta\right)}{L\left( \theta;\Z\right) q\left( \Z^\star|\theta\right)}\right\rbrace
+	$$
+	and otherwise the current $\z$ is retained \cite{Tierney.1994}.
+	%whose kernel is
+	%$$
+	%P_{\z}((\theta, \z_0), (\theta, \z_1)) = q(\z_1|\theta) %\alpha((\theta, \z_0), (\theta, \z_1)) dz_1
+	%$$
+	%where the proposal density $q$ will be defined in the following section and
+	Algorithm \ref{alg:DA-MCMC} provides the details of the DA-MCMC algorithm.
 
@@
 			\REQUIRE $\theta^{(0)}$
+			\RETURN $\{(\Z^{(i)}, \theta^{(i)})\}_{i=0}^N$ 
 			\STATE $\Z^{(0)} \sim q(\Z|\theta^{(0)})$ (generate the initial latent space)
@@
 
-	\subsection{Efficient Proposal Process for Latent Data}	\label{sec:pds}
+	\subsection{Efficient Proposal Process for Latent Data}
+	\label{sec:pds}
+	% Outline: define PD-SIR; generate PD-SIR
 	Though forward simulation of the stochastic SIR process is straightforward, 
 	simulating trajectories \textit{conditionally} on the observed incidence data $\Y$ is a notoriously difficult task \cite{Hobolth.2009}.
-Instead, we will generate the latent variables conditionally on $\Y$ from a surrogate process, and accept or reject them in a Metropolis-Hastings step.
-
-	%\subsection{The Piecewise-decoupled SIR Process}
-	% Outline: define PD-SIR; generate PD-SIR
-	To this end, we consider a stochastic process whose dynamics closely resemble those of the SIR process and which is designed for efficient simulation of epidemic trajectories compatible with the incidence data $\Y = T_{1:K}$.
+Instead, we will generate the latent data conditionally on $\Y$ from a surrogate process, and accept or reject them in a MH step.
+
+	To this end, we consider a stochastic process whose dynamics closely resemble those of the SIR  and which is designed for efficient simulation of epidemic trajectories compatible with the incidence data $\Y = T_{1:K}$.
 We refer to this surrogate process as the \textit{piecewise decoupled SIR} process (PD-SIR).
-	Similarly to the SIR process, the PD-SIR process corresponds to a compartmental model in which individuals move from the compartments $S$ to $I$ and from $I$ to $R$.
+	Similarly to the SIR, the PD-SIR process corresponds to a compartmental model in which individuals move from the compartments $S$ to $I$ and from $I$ to $R$.
 	The removal dynamics are identical under both processes: infection periods follow independent exponential distributions with rate $\gamma$, resulting in a constant individual-level removal rate $\gamma$.
@@
 Note that $\mu$ varies after every event since the value of $I$ changes after an infection or a removal.
-	In contrast, in the PD-SIR process, the infection rate is kept constant over small time intervals.
+	In contrast, in the PD-SIR process, the infection rate is kept constant over small periods of time.
 Consider a resetting schedule $r_{0:L}$ ($L\ge1$) with $0 = r_0 < r_1 < \dots < r_L$.
 The individual-level infection rate of the PD-SIR process is defined as
-	$$\tilde{\mu}(t) = \beta I(r_{k - 1}), \forall t \in [r_{k - 1}, r_k)$$
+	$$\tilde{\mu}(t) = \beta I(r_{k - 1}), \quad \forall t \in [r_{k - 1}, r_k)$$
 	where $I(r_{k-1})$ denotes the number of infectious individuals at the beginning of the time interval.
-That is, the infection rate is kept constant over short periods of time.
-The infection rate $\mu$ and the evolution of the variable $I$ are therefore decoupled during each interval, and the infection rate is reset at the beginning %or left endpoint
-	of the intervals.
-Over a single interval, the PD-SIR process is equivalent to a two-type branching process approximation of the SIR dynamics \cite{Ho.2018}.
-\jx{I really like how this is written but I wonder if there's some simple figure we can make to illustrate the model and difference between SIR at a glance => line plot of mu over time with colored vertical line for every event.
-Done.}.
+As shown in \ref{fig:mu}, the infection rate $\tilde{\mu}$ and the evolution of the variable $I$ are therefore decoupled during each interval, and the infection rate is reset at the beginning %or left endpoint
+	of the intervals. Over a single interval, the PD-SIR process is equivalent to a two-type branching process approximation of the SIR dynamics \cite{Ho.2018}.
 
@@
 		\includegraphics[scale = 0.6]{infection_rate_SIR_PDSIR.jpg}
-		\caption{Infection dynamics of the SIR (dashed line) and PD-SIR (solid line) processes in a small population ($(S(0), I(0)) = (10,2)$).
-			$(\beta, \gamma) = (0.2, 1)$ and the resetting schedule of the PD-SIR contains $L=5$ intervals of equal length.
+		\caption{Infection dynamics of the SIR (dashed line) and PD-SIR (solid line) processes in a small population ($(S(0), I(0)) = (10,2)$) with $\beta=1$.
+The resetting schedule of the PD-SIR is $r_{0:3}=(0.0.2, 0.4, 0.6)$.
 		The vertical dotted lines indicate the infection (red) and removal (blue) times.}
@@
 	we can straightforwardly simulate realizations from the PD-SIR process conditionally on the observations $I_{1:K}$.
-By construction, under the PD-SIR model the population of susceptible individuals $S(t)$ follows a linear death process during each time interval with death rate $\mu_k = \beta I(t_{k-1})$.
+By construction, under the PD-SIR model the population of susceptible individuals $S(t)$ follows a linear death process during each interval with death rate $\mu_k = \beta I(t_{k-1})$.
 Generating event times from a linear death process conditionally on the number of events happening by a certain time can be done extremely efficiently as shown by the following theorem whose proof is presented in Appendix \ref{app:ldp}.
@@
 		\label{theo:ldp}
-		Consider a linear death process with death rate $\mu$ and let $d_{1:T} \in (t_l, t_u]$ be the times of the $N$ deaths occurring between times $t_l$ and $t_u$. Then 
+		Consider a linear death process with death rate $\mu$ and let $d_{1:N} \in (t_l, t_u]$ be the times of the $N$ deaths occurring between times $t_l$ and $t_u$. Then 
 		$$d_j \,{\buildrel d \over =}\, X_{(j)}, \quad j = 1, \dots, N$$
@@
 	We can use this result to generate latent data  $\z = \left\lbrace (z^I_i, z^R_i)\right\rbrace_{i=1}^{n}$ compatible with $\Y = I_{1:K}$ from the PD-SIR process as follows.
-For each time interval $[t_{k-1}, t_k)$, we compute $\mu_k = \beta I(t_{k-1})$ and $U_k = I(0) + \sum_{l=1}^{k-1} I_l$, the total number of infections that have happened before $t_{k-1}$.
-The index set $\mathcal{I}_k = (U_k + 1, \dots, U_k + I_k)$ therefore denotes the $I_k$ individuals infected during the $k$th interval.
+For each time interval $[t_{k-1}, t_k)$, we compute $\mu_k = \beta I(t_{k-1})$ and $U_k = I(0) + \sum_{l=1}^{k-1} I_l$, the cumulative number of infections that happened before $t_{k-1}$.
+The index set $\mathcal{I}_k = \{U_k + 1, \dots, U_k + I_k\}$ therefore denotes the $I_k$ individuals infected during the $k$th interval.
 Note that $I(t_{k-1})$ only depends on past events and can therefore be computed given the PD-SIR process up to time $t_{k-1}$.
-Algorithm \ref{alg:PD-SIR} provides a simple recursion to compute this variable.
-For $j \in \mathcal{I}_k$, by following Theorem \ref{theo:ldp}, we draw the infection times $z^I_j$ as the order statistics of $I_k$ i.i.d.\ truncated exponential random variables with rate $\mu_k$, lower bound $t_{k-1}$ and upper bound $t_k$.
+Algorithm \ref{alg:PD-SIR} provides a simple recursion to compute this variable efficiently.
+For $j \in \mathcal{I}_k$, following Theorem \ref{theo:ldp}, we sample the infection times $z^I_j$ from $\TruncExp(\mu_k; t_{k-1}, t_k)$, a truncated exponential distribution with rate $\mu_k$ bounded between $t_{k-1}$ and $t_k$ from which can efficiently be done with the inverse CDF method.
 For the same indices $j$, we then generate the removal times $z^R_j$ from the removal dynamics of the SIR.
-To accomplish this, we propose the removal time of individual $j$ from the mixed distribution
+To accomplish this, we sample the removal time of individual $j$ from the mixed distribution
 	$$(1 - p_j) \delta_{\infty} + p_j \TruncExp(\gamma; z^I_j, t_{end})$$
@@
 	where $\delta_{\infty}$ corresponds to the Dirac distribution with mass $1$ on the element $\infty$,
-	$$p_j = 1 - \exp\{-\gamma (t_{end} - z^I_j)\} = P(z^R_j < t_end | z^I_j)$$
+	$$p_j = 1 - \exp\{-\gamma (t_{end} - z^I_j)\} = P(z^R_j < t_{end} | z^I_j)$$
 	is the cumulative distribution function of an exponential distribution with rate $\gamma$, % and corresponds to the probability that particle $j$ is removed before $t_{end}$ given that it was infected at time $z^I_j$,
-	and $\TruncExp(\gamma; z^I_j, t_{end})$ denotes a truncated exponential distribution with rate $\gamma$ bounded between $z^I_j$ and  $t_{end}$.
-By construction, this scheme generates latent data from the PD-SIR process that are compatible with the observed data.
-Algorithm \ref{alg:PD-SIR} provides a summary of the procedure in pseudo-code.
+	. By construction, this scheme generates latent data from the PD-SIR process that are compatible with the observed data. Algorithm \ref{alg:PD-SIR} provides a summary of the procedure in pseudo-code.
 
@@
 
-	Based on this construction, we see that the density $q$ of the PD-SIR process is
-	\begin{align*}
-		q(\z|\theta) & = && \prod_{k=1}^K \prod_{j\in\mathcal{I}_k} \TruncExp(\tau^I_j; \mu_{k}, t_{k-1}, t_{k}) \\
-		& && \times \prod_{i=1}^{n} \left( 1 - p_i \right)^{\mathbf{1}(z^R_i = \infty)} \left( p_i \TruncExp(z^I_i; \gamma, z^I_i, t_{end}) \right)^{\mathbf{1}(z^R_i \le t_{end})} \\
-		& = && \prod_{k=1}^K \prod_{j\in\mathcal{I}_k} \TruncExp(\tau^I_j; \mu_{k}, t_{k-1}, t_{k}) \\
-		& && \times \prod_{i=1}^{n} \left( 1 - p_i \right)^{\mathbf{1}(z^R_i = \infty)} \left( p_i \TruncExp(z^I_i; \gamma, z^I_i, t_{end}) \right)^{\mathbf{1}(z^R_i \le t_{end})}
-	\end{align*}
+	The density $q$ of the PD-SIR process can therefore be written as
+	\begin{alignat*}{3}
+		q(\z|\theta) & = && \prod_{k=1}^K \prod_{j\in\mathcal{I}_k} \TruncExp(z^I_j; \mu_{k}, t_{k-1}, t_{k}) \\
+		& && \times \prod_{i=1}^{n} \left( 1 - p_i \right)^{\mathbf{1}(z^R_i = \infty)} \left( p_i \TruncExp(z^R_i; \gamma, z^I_i, t_{end}) \right)^{\mathbf{1}(z^R_i \le t_{end})} \\
+		& = && \prod_{k=1}^K \prod_{j\in\mathcal{I}_k} \TruncExp(z^I_j; \mu_{k}, t_{k-1}, t_{k}) \\
+		& && \times \prod_{i=1}^{n} \left( 1 - p_i \right)^{\mathbf{1}(z^R_i = \infty)} \left( p_i \TruncExp(z^R_i; \gamma, z^I_i, t_{end}) \right)^{\mathbf{1}(z^R_i \le t_{end})}
+	\end{alignat*}
 	where %$\mu_k = \beta I(t_{k-1})$ and
@@
 	Since the proposal for the latent data is independent from the current configuration of the latent data, the efficiency of the DA-MCMC algorithm to explore the latent space directly depends on how similar the surrogate process is to the target process.
-In particular, the Markov chain will have good mixing properties if the PD-SIR process is a close approximation of the SIR process.
+The Markov chain will have good mixing properties if the PD-SIR process is a close approximation of the SIR process.
 The PD-SIR process only differs from the SIR process in its infection dynamics, the removal dynamics being identical in the two processes.
-Figure \ref{fig:comparison} compares the trajectories of the compartments $S$, $I$ and $R$ of a SIR process of moderate size ($(S(0) = 1000, I(0), R(0)) = (1000, 10, 0)$ with $(\beta, \gamma) = (0.003, 1)$ and $t_{end} = 6$) and those of four PD-SIR processes constrained to be compatible with the observed incidence data $I_{1:K}$ from the SIR process for $K \in (5, 10, 50, 1,000)$.
+Figure \ref{fig:comparison} compares the trajectories of the compartments $S$, $I$ and $R$ of a SIR process of moderate size ($(S(0) = 1000, I(0), R(0)) = (1000, 10, 0)$ with $(\beta, \gamma) = (0.003, 1)$ and $t_{end} = 6$) and those of four PD-SIR processes constrained to be compatible with the observed incidence data $I_{1:K}$ from the SIR process for $K \in \{5, 10, 50, 1,000\}$.
 We see that the PD-SIR is qualitatively close to the SIR, even for small values of $K$.
 Unsurprisingly, the quality of the approximation improves as $K$ increases.
-	% In the limit as $K \rightarrow \infty$, the infection times are known.
+	% In the limit as $K \rightarrow \infty$, the infection times are effectively known.
 	The fact that the PD-SIR is a faithful approximation of the SIR leads to a relatively high acceptance rate in the Metropolis-Hastings step for the latent data.
-As a result, our algorithm makes larger jumps in the high-dimensional latent space and therefore explores it more efficiently.
+As a result, our algorithm frequently makes larger jumps in the high-dimensional latent space and therefore explores it more efficiently.
 	%Moreover, even though the latent spaces are generated from an approximation of the SIR process, the Metropolis-Hastings step ensures that the Markov chain converges to the posterior distribution of the parameters under the SIR model; in other words, the inference is exact.
+
+	\ram{Piece-wise linear trajectory of S compartment comes from the fact that it follows a linear death process with piece-wise constant rate.}
 
@@
 	% our DA-MCMC only requires I(0), not S(0).
+
+	\jx{We may insert a brief note explicitly contrasting with previous single-site RJMCMC work, giving a bit of intuition why we are able to avoid that change of space, and this could be a natural place to put it.
+Then, they also have a little more context for the experiment you've since added}
 
@@
 
-	% explain why not compare chain binomial, particle filtering (forward simulation), LNA requires noisy data.
-
-	We examine the convergence properties of the proposed DA-MCMC algorithm with simulation studies.
+	We validate the convergence properties of the proposed DA-MCMC algorithm empirically via a suite of simulation studies.
 First, we assess the performance of the DA-MCMC algorithm in a medium-sized population of over $1000$ individuals.
-We simulate a pandemic with parameters $(\beta, \gamma) = (0.0025, 1)$, which gives $R_0 = 2.5$, starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$ when the pandemic has completed most of its course but is not over yet (see Figure \ref{fig:E1_trajectories}, where $I(t_{end}) = 55$).
+We simulate an epidemic from true parameters $(\beta, \gamma) = (0.0025, 1)$, which gives $R_0 = 2.5$, starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$ when the outbreak has completed most of its course but is not over yet (see Figure \ref{fig:E1_trajectories}, where $I(t_{end}) = 55$).
 The numbers of infections in $K = 10$ time intervals of equal length are observed, corresponding to $I_{1:K} = (9, 14, 21, 42, 56, 121, 190, 162, 107, 73)$.
-	We put the conjugate prior distributions on the parameters $\beta \sim Ga(0.001, 1)$ and $\gamma \sim Ga(1,1)$ independently, which are weakly informative.
-	The Markov chain is initialized in a low density region at $(\beta^{(0)}, \gamma^{(0)}) = (\beta/10, \gamma/10) = (0.00025, 0.1)$ and is run for $1$ million iterations.
-We keep every tenth draw for storage reason and set $\rho = 0.2$ so that the trajectories of $\lfloor\rho (S(0)+I(0))\rfloor = 202$ individuals are updated in each Metropolis-Hastings step.
-The algorithm takes less than $30$ minutes to complete on a personal laptop.
+	We use the conjugate prior distributions \ref{eq:pri} $\beta \sim Ga(0.001, 1)$ and $\gamma \sim Ga(1,1)$, which are only weakly informative.
+	The Markov chain is initialized in a low density region at $(\beta^{(0)}, \gamma^{(0)}) = (\beta/10, \gamma/10) = (0.00025, 0.1)$, and we keep every tenth draw for storage reasons, and set $\rho = 0.2$ so that the trajectories of $\lfloor\rho (S(0)+I(0))\rfloor = 202$ individuals are updated in each Metropolis-Hastings step.
+The algorithm takes less than $30$ minutes to generate $1$ million samples on a personal laptop.
 
@@
 The $90\%$ credible intervals are $(0.00166, 0.00278)$, $(0.492, 1.19)$ and $(2.26, 3.45)$ and cover the true values of the parameters.
-To verify the coverage of the DA-MCMC algorithm, the experiment was repeated $2000$ times independently.
-Tables \ref{tab:coverage} provides the nominal coverage of the $90\%$ credible intervals for the parameters $\beta$, $\gamma$ and $R_0$ along with the average and variance of the posterior means.
+To validate the uncertainty quantification of the posteriors under our DA-MCMC algorithm, we examine frequentist coverage properties of posterior credible intervals under $2000$ independently repeats of the experiment.
+Tables \ref{tab:coverage} provides the empirical coverage of the $90\%$ credible intervals for the parameters $\beta$, $\gamma$ and $R_0$ along with the average and variance of the posterior means.
 The algorithm has the appropriate coverage, suggesting that running the chain for $1$ million iterations is sufficient to approximate the posterior distribution of the parameters.
@@
     \end{tabular}
-    \caption{Nominal coverage of the DA-MCMC algorithm and distribution of the posterior means from $2000$ independent runs in a medium-sized population.
+    \caption{Empirical coverage of  $90\%$ posterior credible intervals and distribution of the posterior means from $2000$ independent runs in a medium-sized population.
 The true values of the parameters are $(\beta.
@@
 
-	Second, we evaluate the impact of the tuning parameter $\rho$ on the mixing properties of the algorithm across different population sizes.
-	We simulate three SIR processes with respectively $S(0) \in \{250, 500, 1000\}$.
+	Second, we explore the impact of the tuning parameter $\rho$ on the mixing properties of the algorithm across different population sizes.
+	We simulate three SIR processes with $S(0) \in \{250, 500, 1000\}$.
 We set $I(0)=10$, $\gamma=1$, $t_{end} = 6$ and $K = 10$ for each simulation and use $R_0 \in \{3, 2.5, 2\}$ respectively for the different population sizes.
@@
 
-	Figure \ref{fig:E3_runtime} and \ref{fig:E3_accept} presents the run time of the algorithm and the acceptance rate in the Metropolis-Hastings step.
+	Figure \ref{fig:E3_runtime} and \ref{fig:E3_accept} present the run time of the algorithm and the acceptance rate in the Metropolis-Hastings step.
 We observe that the run time increases with the population size and with $\rho$ and that, as expected, the acceptance rate decreases as the number of trajectories updated per iteration $\rho(S(0)+I(0))$ increases.
@@
 We observe that updating half of the augmented data per iteration yields the lowest auto-correlation function across the values of $\rho$ considered.
-This suggests that, in this setting, updating the entire latent space in the Metropolis-Hastings step results in too many rejections while updating a tenth or less prevents the chain from efficiently exploring the latent space.
+This suggests that, in this setting, updating the entire latent space in the Metropolis-Hastings step results in too many rejections, while updating a tenth or less prevents the chain from efficiently exploring the latent space.
 
@@
 We observe that the Markov chain mixes much better when event times are jointly proposed in the Metropolis-Hastings step.
-This is confirmed by Table \ref{tab:E6} which shows that the effective sample size per second obtained with the proposed DA-MCMC is between 7 and 20 times larger than that of the SSU DA-MCMC.
+This is supported quantitatively in Table \ref{tab:E6} which shows that the effective sample size per second obtained with the proposed DA-MCMC is between 7 and 20 times larger than that of the SSU DA-MCMC.
+
+ \ram{explain why not compare chain binomial, particle filtering (forward simulation), LNA requires noisy data.}
 
@@
 Weekly infection incidence counts are available for each prefecture for the $73$ weeks between the end of December $2013$ and May $2015$.
+	The analysis serves to show that fast and exact Bayesian inference can be made regarding the Ebola pandemic with the proposed DA-MCMC algorithm instead of providing new insights into this outbreak or the Ebola virus in general.
 
@@
 Even for such a large population, the total run time of the algorithm was less than $100$ minutes on a personal laptop.
-	The Metropolis-Hastings step for the latent data proposals achieves a healthy $22.1\%$ acceptance rate, and the first $50,000$ iterations of the Markov chain are discarded as a burn-in.
+	The Metropolis-Hastings step for the latent data proposals achieves a healthy $20.1\%$ acceptance rate, and the first $50,000$ iterations of the Markov chain are discarded as a burn-in.
 	Figure \ref{fig:ebola} shows the marginal posterior distributions $\gamma$, whose means is $0.109$.
@@
 	%As noted by numerous authors, conditionally on partially observed data, the parameters $\beta$ and $\gamma$ appear positively correlated.
-	The posterior distribution of $R_0$ is unimodal, relatively symmetric, and centered around $1$.
+	%The posterior distribution of $R_0$ is unimodal, relatively symmetric, and centered around $1$.
 A basic reproduction number close to $1$ is consistent with an outbreak that lasted several months without infecting the majority of the population.
@@
 		\end{subfigure}
-		\caption{Posterior densities of parameters of interest (excluding burn-in of $10,000$ iterations).}
+		\caption{Marginal posterior distribution of $\gamma$ and its auto-correlation function for the Gu\'eck\'edou prefecture.}
 		\label{fig:ebola}
@@
 In future work, we will present applications of our DA-MCMC algorithm to processes where these assumptions are relaxed.
-In particular, we are considering extensions to epidemic models with under-reporting, a varying infection rate, stratified populations and non-Markovian dynamics, as well as the general stochastic SIR model of \cite{Severo.1969}.
+In particular, we are considering extensions to epidemic models with under-reporting \cite{Fintzi.2017}, non-Markovian dynamics \cite{Streftaris.2002}, and a time-varying infection rate \cite{Kypraios.2018}.
+
+	\ram{modelling the infection rate over time seems necessary for EBola data.}
 

### 2021-11-02 Manuscripts/Article/Article.tex

@@
 	where $\Vert \mu \Vert_{TV}$ denotes the total variation norm of the measure $\mu$.
-	Uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$ (ref), that is,
-	$$\sqrt{m}(\bar{g}_m - E_\pi g) \Rightarrow N(0, \sigma^2)$$
-	for some positive, finite $\sigma^2$. Furthermore, uniform ergodicity implies that the batch means and spectral variance estimators of $\sigma^2$ are consistent (ref).
+	Uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$ (Jones, 2004), that is,
+	$$\sqrt{m}(\bar{g}_m - E_\pi g) \Rightarrow N(0, \sigma^2_g)$$
+	for any initial distribution for some positive, finite $\sigma^2_g$.
+Furthermore, uniform ergodicity implies that the batch means and spectral variance estimators of $\sigma^2_g$ are consistent (ref).
 
@@
 We observe that the run time increases with the population size and with $\rho$ and that, as expected, the acceptance rate decreases as the number of trajectories updated per iteration $\rho(S(0)+I(0))$ increases.
-	Finally, Figure \ref{fig:E3_facet_ESSsecR0} shows the effect of $\rho$ on the auto correlation function for the parameter $\beta$ for the medium-size population process ($S(0) = 1,000$) for three values of $\rho$.
-We observe that updating half of the augmented data per iteration yields the lowest auto-correlation function across the values of $\rho$ considered.
-This suggests that, in this setting, updating the entire latent space in the Metropolis-Hastings step results in too many rejections, while updating a tenth or less prevents the chain from efficiently exploring the latent space.
+	Finally, Figure \ref{fig:E3_facet_ESSsecR0} shows the effect of $\rho$ on the mixing properties of the Markov chain.
+We observe for a population of $250$ individuals, updating the entire latent data in the MH step provides the best mixing; as the population size increases, the optimal value of $\rho$ decreases.
+This suggests that for a population of $1000$ individuals, updating the whole latent data results in too many rejections, while updating less than $10$\% makes the updates too small to efficiently explore the latent space.
 
@@
 		\end{subfigure}
-		\caption{Impact of $\rho$ on the performance of the DA-MCMC across different population sizes.}
+		\caption{Impact of $\rho$ on the performance of the DA-MCMC in a population of size $250$ (solid), $500$ (dotted) and $1000$ (dashed).}
 		\label{fig:E3}

### 2021-11-02 Manuscripts/Article/Article.tex
test

@@
 	where $\Vert \mu \Vert_{TV}$ denotes the total variation norm of the measure $\mu$.
-	Uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$ (Jones, 2004), that is,
+	Uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$ \cite{Jones.2004}, that is,
 	$$\sqrt{m}(\bar{g}_m - E_\pi g) \Rightarrow N(0, \sigma^2_g)$$

### 2021-11-02 Experiment_output.R
remove S0=100 for clarity

@@ -20,7 +20,9 @@ print(list(
 
 # E3
 load("Output/RDATA/E3.RDATA")
-experiment_3_output_analysis(output_E3, path = "Output/Figures/E3")
+output_E3 %>%
+  map( ~ filter(., S0 != 100)) %>%
+  experiment_3_output_analysis(path = "Output/Figures/E3")
 
 
 

### 2021-11-02 Manuscripts/Article/Article.tex

--- 2021-11-02 Manuscripts/Article/Article.tex
@@
 	where $\Vert \mu \Vert_{TV}$ denotes the total variation norm of the measure $\mu$.
-	Uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$ (Jones, 2004), that is,
+	Uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$ \cite{Jones.2004}, that is,
 	$$\sqrt{m}(\bar{g}_m - E_\pi g) \Rightarrow N(0, \sigma^2_g)$$
-	for any initial distribution for some positive, finite $\sigma^2_g$.
+	for any initial distribution, for some positive, finite $\sigma^2_g$.
 Furthermore, uniform ergodicity implies that the batch means and spectral variance estimators of $\sigma^2_g$ are consistent (ref).
@@
 
-	To derive this result, we will make use of three lemmas. The appendix contains a detailed a proof of the latter two.
-	\rm{Figures for lemma 3.1 and 3.2?}
+	To derive this result, we will make use of two lemmas. The appendix contains a detailed a proof of the latter one.
+	\ram{Figures for lemma 3.1 and 3.2?}
 
@@
 		\label{pro:ga1}
-		Let $Ga(x;a,b)$ denote the density of the gamma distribution with shape $a$ and rate $b$ evaluated at $x$. Then
+		Let $Ga(x;a,b) = \dfrac{a^b}{\Gamma(a)}x^{a-1}\exp\{-ab\}$ denote the density of the gamma distribution with shape $a$ and rate $b$ evaluated at $x$. Then
 		\begin{equation}

### 2021-11-02 Manuscripts/Article/Article.tex
Updates from Overleaf

@@
 	where $\Vert \mu \Vert_{TV}$ denotes the total variation norm of the measure $\mu$.
-	Uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$ \cite{Jones.2004}, that is,
+	Uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$ \cite{Jones.2004b}, that is,
 	$$\sqrt{m}(\bar{g}_m - E_\pi g) \Rightarrow N(0, \sigma^2_g)$$
@@
 	\begin{proof}
-		Equation \ref{eq:ga1} is proven in Jones, Hobert (2004).
+		Equation \ref{eq:ga1} is proven in \cite{Jones.2004}.
 For Equation \ref{eq:ga2}, note that $x_b^*$ is the only positive solution to $Ga(x;a,b) = Ga(x;a+A,b)$.
@@
 		and where the last inequality holds because $\Gamma_{a,a+A}\le\Gamma_{a+\alpha, a+A}$.
-		The case $x>x_b^*$ can be shown similarly.
+		The case $x>x_b^*$ can be shown similarly. Figure \ref{fig:gam1} illustrates these two results.
 	\end{proof}

### 2021-12-12 Manuscripts/Article/Article.tex

@@
 \usepackage{amsmath,amssymb,array,eucal, amsthm}
-\linespread{1.5}
+\linespread{1.35}
 \setlength\parindent{35pt}
@@
 	\title{%Efficient Likelihood-Based Inference for Fitting Stochastic Epidemic Models to Incidence Data via Data Augmentation
-		Uniformly Ergodic Data-Augmented MCMC for Fitting Stochastic Epidemic Models to Incidence Data}
+		Uniformly Ergodic Data-Augmented MCMC for Fitting the General Stochastic Epidemic Model to Incidence Data}
 	\author{Rapha\"{e}l Morsomme$^{1}$\footnote{raphael.morsomme@duke.edu} \ and Jason Xu$^{1}$ \\
@@
 Since the latent spaces are generated from a very faithful approximation of the target epidemic model, the algorithm can update a large portion of the event times per iteration while maintaining a relatively high acceptance rate, thereby exploring the high-dimensional latent space efficiently.
+Our approach is fully Bayesian and therefore allows representation of parameter uncertainty through posterior distributions.
 
@@
 Section \ref{sec:pds} describes the data-augmented MCMC algorithm proposed in this article and assesses how closely it approximation the SIR process.
-Section \ref{sec:sim} presents the results of a simulation study examining the performance of the algorithm, while Section \ref{sec:ebo} describes an analysis of the $2014$ Ebola outbreak in Gu\'eck\'edou, Guinea.
+In Section \ref{sec:sim}, we examine the performance of our algorithm on simulated data and then turn to an analysis of the $2014$ Ebola outbreak in Gu\'eck\'edou, Guinea in Section \ref{sec:ebo}.
 Finally, Section \ref{sec:dis} discusses the findings and concludes the article.
@@
 If individual $j$ never becomes infectious, we set $\tau^I_j = \tau^R_j = \infty$ and $X_j(t) = s$ for $t \in [0, \infty)$.
+Since removed individuals do not contribute to the pandemic, it is safe to ignore the individuals removed at time $0$ and write $n = S(0) + I(0)$.
 Here $\chi_{\X}$ denotes the set of trajectories compatible with the evolution of a disease---that is, the set of trajectories in which no infection occurs whenever the infectious compartment is depleted:
@@
 	are given by the matrix $\Lambda = [\lambda_{\x, \x'}]$ with
-	$$
+	\begin{equation}
+	\label{eq:rate}
     	\lambda_{\x, \x'} = 
@@
     	\end{cases}
-	$$
+	\end{equation}
 	where $I(t) = \#\{x_j(t) = i\}$ is the total number of infectious individuals at time $t$.
 Thus, the individual-level infection and removal rates at time $t$ are respectively $\beta I(t)$ and $\gamma$.
-	%Equivalently \jx{The following is not an exact equality, but needs the $o(dt)$ term or similar; also uncapitalize X on the right}
+
+	The model can alternatively be parameterized in terms $\tilde{\theta} = (\beta, R_0)$, with $R_0 = S(0)\beta/\gamma$. The likelihood becomes
+	\begin{equation*}
+	    L((R_0, \gamma); \X) = \beta^{n_I + n_R} R_0^{-n_R} \prod_{j:\tau_{T_j} \le t_{end}} I(\tau_{T_j}) \exp\left\lbrace - \beta \int_{0}^{t_{end}} I(t)S(t) dt - \frac{\beta}{R_0} \int_{0}^{t_{end}} I(t) dt \right\rbrace
+	\end{equation*}
+
+	%Equivalently \jx{The following is not an exact equality, but needs the $o(dt)$ term or similar}
 	%	$$
@@
 	%	\begin{cases}
-	%		\beta I(t) dt + o(dt), & \X \text{ and } \X' \text{ only differ at position } j \text{ with } x_j = S \text{ and } x_j' = I, \\
-	%		\gamma dt + o(dt), & \X \text{ and } \X' \text{ only differ at position } j \text{ with } x_j = I \text{ and } x_j' = R, \\
+	%		\beta I(t) dt + o(dt), & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = S \text{ and } x_j' = I, \\
+	%		\gamma dt + o(dt), & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = I \text{ and } x_j' = R, \\
 	%		o(dt), & \text{otherwise}
@@
 	%Outline: complete-data likelihood; MLE; gamma conjugacy
-	\jx{Need some citations here for form of likelihood, conjugate/MLE forms, etc} 
-	When the Markov process \eqref{eq:X} is completely observed until time $t_{end}$, we obtain the following likelihood \ram{ref for likelihood}
+	\jx{Need some citations here for form of conjugate/MLE forms, etc}
+
+	When the Markov process \eqref{eq:X} is completely observed until time $t_{end}$, we obtain the following likelihood \cite{Streftaris.2002} 
+	%\ram{ref for likelihood - 2000, auranen et al - Transmission of Pneumococcal Carriage in Families: A Latent Markov Process Model for Binary Longitudinal Data --- general SIR likelihood, Becker and Britton 1999 Statistical Studies of Infectious Disease Incidence}
 	\begin{align}
@@
 		\label{eq:cdl}
-		& = \beta^{n_I} \gamma^{n_R}\prod_{j \in \mathcal{I}} I(\tau^I_j) \exp\left\lbrace - \int_{0}^{t_{end}}\beta I(t)S(t) + \gamma I(t) dt \right\rbrace
+		& = \beta^{n_I} \gamma^{n_R}\prod_{j \in \mathcal{I}} I(\tau^I_j) \exp\left\lbrace - \beta \int_{0}^{t_{end}} I(t)S(t) - \gamma \int_{0}^{t_{end}} I(t) dt \right\rbrace
 	\end{align}
@@
 	It is straightforward to conduct inference in this continuously observed scenario.
-The likelihood \eqref{eq:cdl} belongs to the exponential family, and maximum likelihood estimates of its parameters can be expressed in terms of the sufficient statistics defined above as \ram{ref for exponential family and MLE}
-	$$\hat{\beta}_{MLE} = \dfrac{n_I}{ \int_{0}^{t_{end}} I(t)S(t)dt}, \quad \hat{\gamma}_{MLE} = \dfrac{n_R}{\int_{0}^{t_{end}} I(t)dt}.$$
-	Note that since the functions $I$ and $S$ are constant between event times, the two integrals correspond to finite sums which are straightforward to compute.
-	Furthermore, in a Bayesian context, inference is facilitated by the conjugacy with the Gamma distribution: if we use independent Gamma prior distributions
+The likelihood \eqref{eq:cdl} belongs to the exponential family, and maximum likelihood estimates of its parameters can be expressed in terms of the sufficient statistics defined above as \ram{ref for exponential family and MLE - Guttorp 1994, perhaps newer version with Minin.} Simple algebra shows that the maximum likelihood estimates are
+	$$
+	\hat{\beta} = \dfrac{n_I}{ \int_{0}^{t_{end}} I(t)S(t)dt}, \qquad \hat{\gamma} = \dfrac{n_R}{ \int_{0}^{t_{end}} I(t)dt}.
+	$$
+	Note that since the functions $I(t)$ and $S(t)$ are constant between event times, the two integrals correspond to finite sums which are straightforward to compute.
+	Furthermore, in a Bayesian context, inference is facilitated by the conjugacy with the gamma and inverse-gamma distributions: if we use gamma independent prior distributions
 	\begin{equation}
 		\label{eq:pri}
-		\beta \sim Ga(a_{\beta}, b_{\beta}), \qquad \gamma \sim Ga(a_{\gamma}, b_{\gamma})
-	\end{equation}
-	where $Ga(a,b)$ denotes the parametrization with mean $a/b$ and variance $a/b^2$, we obtain independent Gamma posterior distributions
-	\begin{equation}
+		\beta \sim G(a_{\beta}, b_{\beta}), \qquad \gamma \sim G(a_{\gamma}, b_{\gamma})
+	\end{equation}
+	where $G(a,b)$ denotes the parametrization of the gamma distribution with mean $a/b$ and variance $a/b^2$, we obtain the independent posterior distributions
+	\begin{align}
 		\label{eq:posterior_theta}
-		\beta | \X \sim Ga\left( a_{\beta} + n_I, b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt\right), \qquad 
-		\gamma | \X \sim Ga\left( a_{\gamma} + n_R, b_{\gamma} + \int_{0}^{t_{end}} I(t) dt\right)
-	\end{equation}
-	from which one can easily generate independent values with a Monte Carlo sampler to explore the posterior distribution of parameters $\pi(\theta|\X)$ \cite{Tierney.1994}.
-
-	% the basic reproduction number $R_0 = S(0) \beta / \gamma$, which corresponds to the expected number of secondary infections caused by an infectious individual in a susceptible population.
-$\R_0 = S(0) \beta/\gamma$.
+		\beta | \X & \sim G\left( a_{\beta} + n_I , b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt\right) \\
+		\gamma | \X & \sim G\left( a_{\gamma} + n_R, b_{\gamma} + \int_{0}^{t_{end}} I(t) dt\right)
+	\end{align}
+	from which one can easily generate independent values to explore $\pi(\tilde{\theta}|\X)$ via Monte Carlo.
+Moreover, since $\beta$ and $R_0$ are almost uncorrelated in the posterior distribution, the Gibbs sampler would be an efficient algorithm to explore the posterior distribution of the parameters when the epidemic process is completely observed.
+
+	The likelihood in Equation \ref{eq:cdl} can also be parameterized in terms of $\tilde{\theta} = (\beta, R_0)$ where $R_0 := S(0) \beta / \gamma$ is called the basic reproduction number and corresponds to the expected number of secondary infections caused by an infectious individual in a susceptible population.
+Using $\tilde{\theta}$ with independent prior distributions
+	\begin{equation}
+		\label{eq:pri2}
+		\beta \sim G(a_{\beta}, b_{\beta}), \qquad R_0 \sim IG(a_{R}, b_{R})
+	\end{equation}
+	yields the closed-form full conditional distributions.
+	\begin{align}
+		\label{eq:posterior_theta2}
+		\beta | \X & \sim G\left( a_{\beta} + n_I + n_R, b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt + \frac{S(0)}{R_0} \int_{0}^{t_{end}} I(t) dt\right) \\
+		R_0 | \X, \beta & \sim IG\left( a_{R} + n_R, b_{R} + \beta S(0) \int_{0}^{t_{end}} I(t) dt\right)
+	\end{align}
+	We have noted that the parameterization $\tilde{\theta}$ results in a Markov chain that mixes better and therefore employ it in the simulation studies and data analysis of Section \ref{sec:per}.
+
+	%	\begin{equation}
+%	    \label{eq:lik2}
+%	    L(\X|\tilde{\theta}) = \beta^{n_I + n_R} R_0^{-n_R} S(0)^{n_R} \prod_{j\in\mathcal{I}} I(\tau^I_j) \exp\left\lbrace - \beta \int_{0}^{t_{end}} I(t)S(t) - \frac{\beta S(0)}{R_0} \int_{0}^{t_{end}} I(t) dt \right\rbrace,
+%	\end{equation}
 
@@
 	%Outline: observed data; intractable posterior; DAMCMC; parameters update; latent space update
-
-	%	(New in this section: I am motivating the type of observed data  $T_{1:K}$ that we consider.)
 
@@
 
-	Under a Bayesian approach, the posterior distribution of the parameters given the observed data is formally related to the joint posterior via integration:
+	In the Bayesian framework, the posterior distribution of the parameters given the observed data is formally related to the complete data likelihood \eqref{eq:cdl} via integration:
 	\begin{equation}
@@
 		\pi(\theta|\Y) 
-		\, \, \propto \, \, L(\theta; \Y)\pi(\theta) \, \, = \, \, \int_{\chi^*_{\X}} L(\theta; \x) d\x \pi(\theta)
+		\propto \pi(\theta) L(\theta; \Y) = \pi(\theta) \int_{\chi_\chi} L(\theta; \x) \mathbf{1}\{\Y \overset{\Delta}{=} \x\} d\x
 	\end{equation}
 	where $\pi(\theta)$ is the prior distribution on $\theta$, and
-	$\chi^*_{\X} \subset \chi_{\X}$ is the set of trajectories of the process \eqref{eq:X} that are compatible with the observed data $\Y$.
-	The partial data likelihood $L(\theta; \Y)$ therefore consists of a high-dimensional integral over all epidemic paths compatible with the observed data.
+	$\mathbf{1}\{\Y \overset{\Delta}{=} \x\} = 1$ if $\Y$ is compatible with the infection times in $\x$ and $0$ otherwise; that is, the partial data likelihood $L(\theta; \Y)$ consists of a high-dimensional integral over all epidemic paths that agree with the observed data.
 This marginalization step has no known closed form solution and presents computational challenges even for a population of moderate size.
@@
 While these approximation-based approaches bypass the intractability of the partial data likelihood \ref{eq:pdl}, the assumptions on which they are based are questionable when the population is small, and, as a result, the dynamic of the stochastic process differs from its asymptotic behavior.
-
-	Since the popularization of Markov chain Monte Carlo methods (MCMC) in the field of statistics \cite{Tanner.1987, Gelfand.1990, Tierney.1994}, researchers have developed sampling-based methods to directly work with the SIR likelihood instead of an approximation thereof.
+	%Recently, \cite{Ho.2018} proposed an exact analytical method to directly compute the likelihood of the stochastic SIR process given discretely observed data.
+
+	Since the popularization of Markov chain Monte Carlo methods (MCMC) in statistics \cite{Tanner.1987, Gelfand.1990, Tierney.1994}, researchers have developed sampling-based methods to directly work with the SIR likelihood instead of an approximation thereof.
 MCMC algorithms fall into two categories: model-based forward simulation and data augmented MCMC (DA-MCMC).
@@
 	We adopt a data-augmented MCMC approach that bridges the challenging partially observed setting to the tractable complete data likelihood by way of latent variables.
-In particular, our method hinges on the efficacy of a carefully designed proposal process for the latent variables that faithfully approximates the original process dynamics.
-	%By correcting with a Metropolis-Hastings step
-	The Metropolis-Hastings sampler targets the exact posterior of the model given partially observed incidence data, and enjoys the efficiency of fast proposals in the high-dimensional latent space as well as fast computations involving the complete data likelihood.
+Our method hinges on the efficacy of a carefully designed proposal process for the latent variables that faithfully approximates the SIR process dynamics.
+	The Metropolis-Hastings sampler for the latent data targets the exact posterior of the model given partially observed incidence data, and enjoys the efficiency of fast proposals in the high-dimensional latent space as well as fast computations involving the complete data likelihood.
 
 	As shown in Section \ref{sec:icd}, the complete data likelihood \eqref{eq:cdl} is amenable to computation.
-This suggests augmenting the observed data $\Y$ with latent data $\Z$ such that the likelihood $L(\theta; (\Y, \Z))$ has the closed form \eqref{eq:cdl}, and constructing a Markov chain $\{(\theta^{(m)}, \Z^{(m)})\}_{m=0}^M$ with state space $\chi = \chi_{\theta} \times \chi_{\Z}$ whose stationary distribution is the joint posterior distribution $\pi(\theta, \X|\Y)$ \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
-We consider the latent data $\Z = \left\lbrace (z^I_j, z^R_j)\right\rbrace_{j=1}^{n}$ which consist of the times and types of unobserved epidemic events, where the infection and removal times for individual $j$ are
+This suggests augmenting the observed data $\Y$ with latent data $\Z$ such that the likelihood $L(\theta; (\Y, \Z))$ has the closed form \eqref{eq:cdl}, and constructing an ergodic Markov chain $\{(\theta^{(m)}, \Z^{(m)})\}_{m=0}^M$ on $\chi = \chi_{\theta} \times \chi_{\Z}$ whose limiting distribution is the joint posterior $\pi(\theta, \X|\Y)$ \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
+Given the draws $\{(\theta^{(m)}, \Z^{(m)})\}_{m=0}^M$, the sequence $\{\theta^{(m)}\}_{m=0}^M$ provides an approximation of the marginal posterior distribution $\pi(\theta|\Y)$ of interest.
+
+	We consider the latent data $\Z = \left\lbrace (z^I_j, z^R_j)\right\rbrace_{j=1}^{n}$ which consist of the times and types of unobserved epidemic events, where the infection and removal times for individual $j$ are respectively
 	$$z^I_j \begin{cases}
@@
 		\in (z^I_j, t_{end}] & \text{if individual } j \text{ is removed before } t_{end} \\
-		= \infty & \text{if } z^I_j = \infty \text{ or individual } j \text{ is removed after } t_{end}. \\
+		= \infty & \text{if } z^I_j = \infty \text{ or individual } j \text{ is not removed before } t_{end}. \\
 	\end{cases}$$
-	Since $\Z$ contains all the information present in $\Y$, we can write
-	$$L(\theta; (\Y, \Z)) = L(\theta; \Z).$$
-
-	%    \jx{Brief transition by saying that drawing from $\chi_Z$ would require conditional simulation, remind reader of the Hobolth stone reference/mention that Lam's method could in theory draw from these conditionals, but it's a hard problem computationally that becomes intractable and instead we will use Metropolis-Hastings with a fast proposal for Z.
-This was all mentioned in the intro but can be recapped in one/two sentences.}
+	Since $\Y = I_{1:K}$ is a deterministic function of the unobserved infection events, the joint posterior distribution is
+	\begin{align*}
+	    \pi(\theta, \Z|\Y)
+	    & \propto f(\Y|\theta, \Z) L(\theta; \Z) \pi(\theta) \\
+	    & = \mathbf{1}\{\Y \overset{\Delta}{=} \Z\} L(\theta; \Z) \pi(\theta)
+	\end{align*}
+	where $L(\theta; \Z)$ is the complete data likelihood \ref{eq:cdl}.
 
@@
 A one-step transition from $\x_1 = (\theta_1, \z_1)$ to $\x_2 = (\theta_2, \z_2)$ therefore looks like
-	$$\x_1 = (\theta_1, \z_1) \rightarrow (\theta_2, \z_1) \rightarrow (\theta_2, \z_2) = \x_2$$
-
-	%The transition kernel $P$ of the Markov chain $\{\x^{(m)}\}_m = \{(\theta^{(m)}, \z^{(m)})\}_m$ is thus a composition of two kernels $P_{\theta}$ and $P_{\z}$ which alternatively updates $\theta$ while keeping $\z$ fixed, and then updates $\z$ while keeping $\theta$ fixed.
-	Due to the Gamma conjugacy of the complete-data likelihood (see Section \ref{sec:icd}), sampling new values for $\theta$ given $\z$ poses no challenge: we can simply employ a Gibbs sampler to directly draw from the two independent full conditional distributions \ref{eq:posterior_theta}.
+	$$\x_1 = (\theta_1, \z_1) \rightarrow (\theta_2, \z_1) \rightarrow (\theta_2, \z_2) = \x_2.$$
+	The first update for $\theta$ is straightforward due to the Gamma conjugacy of the complete-data likelihood (see Section \ref{sec:icd}).
+We can simply employ a Gibbs sampler to directly draw from the two independent full conditional distributions \ref{eq:posterior_theta}.
 In these expressions, $n_I, n_R, \int S(t)I(t) dt, \int I(t) dt$ are sufficient statistics from the latent data $\z$ which are easy to compute.
-	The latent data is updated in a Metropolis-Hastings (MH) step.
-Given some current $\theta$ and $\z$, the MH step proceeds by simulating a candidate $\z^\star$ from a surrogate process $q(\z^\star|\theta)$.
+	On the other hand, sampling a new configuration of the latent data is difficult.
+Though forward simulation of the stochastic SIR process is straightforward \ram{ref: Gillepsie}, simulating trajectories \textit{conditionally} on the observed data requires the conditional simulation of a Markov process which is a notoriously difficult task \cite{Hobolth.2009}.
+Since given the $\Y = I_{1:K}$ the end-point of the process are not known, the methods proposed in \cite{Hobolth.2009} are not applicable.
+	Instead, we will generate the latent data conditionally on $\Y$ from a surrogate process, and accept or reject them in a Metropolis-Hastings (MH) step.
+Given some current $\theta$ and $\z$, the MH step proceeds by simulating a candidate $\z^\star$ from a surrogate process $q(\z^\star|\theta)$, which we present in Section \ref{sec:pds}.
 Then $\z^\star$ is accepted with probability
-	$$
+	\begin{equation}
+		\label{eq:alpha}
 		\alpha\left( \left( \theta, \z\right) , \left( \theta, \z^\star\right) \right) =	\min\left\lbrace 1, \dfrac{L\left( \theta; \Z^\star\right) q\left( \Z|\theta\right)}{L\left( \theta;\Z\right) q\left( \Z^\star|\theta\right)}\right\rbrace
-	$$
+	\end{equation}
 	and otherwise the current $\z$ is retained \cite{Tierney.1994}.
-	%whose kernel is
-	%$$
-	%P_{\z}((\theta, \z_0), (\theta, \z_1)) = q(\z_1|\theta) %\alpha((\theta, \z_0), (\theta, \z_1)) dz_1
-	%$$
-	%where the proposal density $q$ will be defined in the following section and
-	Algorithm \ref{alg:DA-MCMC} provides the details of the DA-MCMC algorithm.
+Note that it is not necessary to compute $\mathbf{1}\{\Y \overset{\Delta}{=} \Z^\star\}$ in Equation \ref{eq:alpha} since $\Z^\star$ is compatible with $\Y$ by construction.
+Algorithm \ref{alg:DA-MCMC} provides the details of the DA-MCMC scheme.
 
@@
 			\REQUIRE $\theta^{(0)}$
-			\RETURN $\{(\Z^{(i)}, \theta^{(i)})\}_{i=0}^N$ 
+			\RETURN $\{(\Z^{(m)}, \theta^{(m)})\}_{m=0}^N$ 
 			\STATE $\Z^{(0)} \sim q(\Z|\theta^{(0)})$ (generate the initial latent space)
@@
 	% Outline: define PD-SIR; generate PD-SIR
-	Though forward simulation of the stochastic SIR process is straightforward, 
-	simulating trajectories \textit{conditionally} on the observed incidence data $\Y$ is a notoriously difficult task \cite{Hobolth.2009}.
-Instead, we will generate the latent data conditionally on $\Y$ from a surrogate process, and accept or reject them in a MH step.
-
-	To this end, we consider a stochastic process whose dynamics closely resemble those of the SIR  and which is designed for efficient simulation of epidemic trajectories compatible with the incidence data $\Y = T_{1:K}$.
+
+	The surrogate process that we consider for generating the latent data in the MH step consists of a stochastic process whose dynamics closely resemble those of the SIR and which is designed for efficient simulation of epidemic trajectories compatible with the incidence data $\Y = I_{1:K}$.
 We refer to this surrogate process as the \textit{piecewise decoupled SIR} process (PD-SIR).
 	Similarly to the SIR, the PD-SIR process corresponds to a compartmental model in which individuals move from the compartments $S$ to $I$ and from $I$ to $R$.
-	The removal dynamics are identical under both processes: infection periods follow independent exponential distributions with rate $\gamma$, resulting in a constant individual-level removal rate $\gamma$.
-
+	The removal dynamics are identical under both processes: infection periods follow independent exponential distributions with rate $\gamma$.
+	%resulting in a constant individual-level removal rate $\gamma$.
+
+	%The population-level infection rate at time $t$ is $\mu_T(t) = \beta S(t)I(t)$ and so, from the perspective of each susceptible individual, the individual-level infection rate is $\mu(t) = \frac{\mu_T(t)}{S(t)} = \beta I(t)$.
 	The infection dynamics, however, vary slightly.
-	In the SIR process, the population-level infection rate at time $t$ is $\mu_T(t) = \beta S(t)I(t)$ and so, from the perspective of each susceptible individual, the individual-level infection rate is $\mu(t) = \frac{\mu_T(t)}{S(t)} = \beta I(t)$.
+	In the SIR process, the individual-level infection rate at time $t$ is $\mu(t) = \beta I(t)$ (see Equation \ref{eq:rate}).
 	Note that $\mu$ varies after every event since the value of $I$ changes after an infection or a removal.
 	In contrast, in the PD-SIR process, the infection rate is kept constant over small periods of time.
-Consider a resetting schedule $r_{0:L}$ ($L\ge1$) with $0 = r_0 < r_1 < \dots < r_L$.
-The individual-level infection rate of the PD-SIR process is defined as
-	$$\tilde{\mu}(t) = \beta I(r_{k - 1}), \quad \forall t \in [r_{k - 1}, r_k)$$
-	where $I(r_{k-1})$ denotes the number of infectious individuals at the beginning of the time interval.
-As shown in \ref{fig:mu}, the infection rate $\tilde{\mu}$ and the evolution of the variable $I$ are therefore decoupled during each interval, and the infection rate is reset at the beginning %or left endpoint
+Consider the observation schedule $t_{0:K}$ of $\Y$.
+For $t \in (t_{k-1}, t_k]$, the infection rate of the PD-SIR process is defined as
+	$$\tilde{\mu}(t) = \beta I(t_{k - 1})$$
+	where $I(t_{k-1})$ denotes the number of infectious individuals at the beginning of the $k$th interval.
+As shown in \ref{fig:mu}, the infection rate $\tilde{\mu}$ and the variable $I$ are decoupled during each interval, and the infection rate is reset at the beginning %or left endpoint
 	of the intervals. Over a single interval, the PD-SIR process is equivalent to a two-type branching process approximation of the SIR dynamics \cite{Ho.2018}.
@@
 		\centering
-		\includegraphics[scale = 0.6]{infection_rate_SIR_PDSIR.jpg}
+		\includegraphics[scale = 0.4]{infection_rate_SIR_PDSIR.jpg}
 		\caption{Infection dynamics of the SIR (dashed line) and PD-SIR (solid line) processes in a small population ($(S(0), I(0)) = (10,2)$) with $\beta=1$.
-The resetting schedule of the PD-SIR is $r_{0:3}=(0.0.2, 0.4, 0.6)$.
+The schedule of the PD-SIR is $t_{0:3}=(0, 0.2, 0.4, 0.6)$.
 		The vertical dotted lines indicate the infection (red) and removal (blue) times.}
@@
 
-	Now, by letting the resetting schedule $r_{0:L}$ coincide with the observation schedule $t_{0:K}$,
-	$$L = K, \text{ and } r_k = t_k, \quad k=1,\dots,K,$$
-	we can straightforwardly simulate realizations from the PD-SIR process conditionally on the observations $I_{1:K}$.
-By construction, under the PD-SIR model the population of susceptible individuals $S(t)$ follows a linear death process during each interval with death rate $\mu_k = \beta I(t_{k-1})$.
-Generating event times from a linear death process conditionally on the number of events happening by a certain time can be done extremely efficiently as shown by the following theorem whose proof is presented in Appendix \ref{app:ldp}.
-
+	It is straightforward to simulate realizations from the PD-SIR process conditionally on the observations $\Y = I_{1:K}$.
+By construction, under the PD-SIR model the variable $S(t)$ follows a linear death process (LDP) during each interval with death rate $\mu_k = \beta I(t_{k-1})$.
+	The LDP is a counting process possessing the so-called order statistics property \cite{Neuts.1971}; that is, given the number of events registered at time $t$, the event times are distributed as the order statistics of iid random variables following, in the case of the LDP, an exponential distribution truncated at $t$.
 	\begin{theorem}
@@
 	\end{theorem}
+	The proof of this theorem is presented in Appendix \ref{app:ldp}.
+Generating event times from a LDP in an interval conditionally on the number of events can therefore be done extremely efficiently with the inverse cumulative distribution function method \ram{ref Devroye 1986}.
 
@@
 Algorithm \ref{alg:PD-SIR} provides a simple recursion to compute this variable efficiently.
-For $j \in \mathcal{I}_k$, following Theorem \ref{theo:ldp}, we sample the infection times $z^I_j$ from $\TruncExp(\mu_k; t_{k-1}, t_k)$, a truncated exponential distribution with rate $\mu_k$ bounded between $t_{k-1}$ and $t_k$ from which can efficiently be done with the inverse CDF method.
+For $j \in \mathcal{I}_k$, following Theorem \ref{theo:ldp}, we sample the infection times $z^I_j \sim \TruncExp(\mu_k; t_{k-1}, t_k)$, a truncated exponential distribution with rate $\mu_k$ bounded between $t_{k-1}$ and $t_k$, which can efficiently be done with the inverse CDF method.
 For the same indices $j$, we then generate the removal times $z^R_j$ from the removal dynamics of the SIR.
 To accomplish this, we sample the removal time of individual $j$ from the mixed distribution
-	$$(1 - p_j) \delta_{\infty} + p_j \TruncExp(\gamma; z^I_j, t_{end})$$
+	$$z^R_j|z^I_j \sim (1 - p_j) \delta_{\infty} + p_j \TruncExp(\gamma; z^I_j, t_{end})$$
 	placing point mass $(1 - p_j)$ at $\infty$ and continuous mass on $(z^I_j, t_{end}]$,
@@
 	$$p_j = 1 - \exp\{-\gamma (t_{end} - z^I_j)\} = P(z^R_j < t_{end} | z^I_j)$$
-	is the cumulative distribution function of an exponential distribution with rate $\gamma$, % and corresponds to the probability that particle $j$ is removed before $t_{end}$ given that it was infected at time $z^I_j$,
-	. By construction, this scheme generates latent data from the PD-SIR process that are compatible with the observed data. Algorithm \ref{alg:PD-SIR} provides a summary of the procedure in pseudo-code.
+	is the cumulative distribution function of an exponential distribution with rate $\gamma$.
+% and corresponds to the probability that particle $j$ is removed before $t_{end}$ given that it was infected at time $z^I_j$,
+
+	By construction, this scheme generates latent data from the PD-SIR process that are compatible with $\Y = I_{1:K}$.
+The density $q$ of the proposal kernel for the PD-SIR process can therefore be written as
+	\begin{alignat}{3}
+	\label{eq:q}
+		q(\z|\theta) 
+		& = && \prod_{k=1}^K \prod_{j\in\mathcal{I}_k} \TruncExp(z^I_j; \mu_{k}, t_{k-1}, t_{k}) \\
+		& && \times \prod_{i=1}^{n} \left( 1 - p_i \right)^{\mathbf{1}(z^R_i = \infty)} \left( p_i \TruncExp(z^R_i; \gamma, z^I_i, t_{end}) \right)^{\mathbf{1}(z^R_i \le t_{end})}
+	\end{alignat}
+	where %$\mu_k = \beta I(t_{k-1})$ and
+	$$\TruncExp(x; \mu, l, u) = \dfrac{\mu \exp\{-\mu x\}}{\exp\{-\mu l\} - \exp\{-\mu u\}}, \quad x \in (l, u)$$
+	denotes the density of a truncated exponential distribution with parameters as notated previously. Algorithm \ref{alg:PD-SIR} provides a summary of the proposal scheme in pseudo-code.
 
 	\begin{algorithm}
-		\caption{Generating a PD-SIR process conditionally on discretely observed infection incidence counts $I_{1:K}$}
+		\caption{Generating a PD-SIR process conditionally on the observed data $\Y = I_{1:K}$}
 		\label{alg:PD-SIR}
@@
 			\FOR {$k = 1, \dots, K$}
+    			\STATE $U_k \leftarrow I(0) + \sum_{l=1}^{k-1} I_l$
+    			\STATE $\mathcal{I}_k \leftarrow \{U_k + 1, \dots, U_k + I_k\}$	
     			\STATE $\mu_k \leftarrow \beta I(t_{k-1})$
-			\STATE $X_{1 : I_k} \sim \TruncExp(\mu_k, t_{k-1}, t_k)$ i.i.d.\
-			\STATE $U_k \leftarrow I(0) + \sum_{l=1}^{k-1} I_l$
-			\STATE $\mathcal{I}_k \leftarrow (U_k + 1, \dots, U_k + I_k)$
+    			\STATE $X_{\mathcal{I}_k} \sim \TruncExp(\mu_k, t_{k-1}, t_k)$ i.i.d.\
 
@@
 
-	The density $q$ of the PD-SIR process can therefore be written as
-	\begin{alignat*}{3}
-		q(\z|\theta) & = && \prod_{k=1}^K \prod_{j\in\mathcal{I}_k} \TruncExp(z^I_j; \mu_{k}, t_{k-1}, t_{k}) \\
-		& && \times \prod_{i=1}^{n} \left( 1 - p_i \right)^{\mathbf{1}(z^R_i = \infty)} \left( p_i \TruncExp(z^R_i; \gamma, z^I_i, t_{end}) \right)^{\mathbf{1}(z^R_i \le t_{end})} \\
-		& = && \prod_{k=1}^K \prod_{j\in\mathcal{I}_k} \TruncExp(z^I_j; \mu_{k}, t_{k-1}, t_{k}) \\
-		& && \times \prod_{i=1}^{n} \left( 1 - p_i \right)^{\mathbf{1}(z^R_i = \infty)} \left( p_i \TruncExp(z^R_i; \gamma, z^I_i, t_{end}) \right)^{\mathbf{1}(z^R_i \le t_{end})}
-	\end{alignat*}
-	where %$\mu_k = \beta I(t_{k-1})$ and
-	$$\TruncExp(x; \mu, l, u) = \dfrac{\mu \exp\{-\mu x\}}{\exp\{-\mu l\} - \exp\{-\mu u\}}, \quad x \in (l, u)$$
-	denotes the density of a truncated exponential distribution with parameters as notated previously.
+	The following four characteristics of our DA-MCMC algorithm are worth noting.
+	First, initializing the Markov chain only requires setting $\theta^{(0)} = (\beta^{(0)}, \gamma^{(0)})$ since $\Z^{(0)}$ can be generated conditionally on $\theta^{(0)}$ and $\Y$ alone from the PD-SIR process.
+	Second, since the PD-SIR closely approximates the SIR model, the acceptance rate in the Metropolis-Hastings step for the latent data is relatively high.
+For large population, however, the acceptance rate may be too low, thereby hindering the mixing of the Markov chain.
+To address this issue, we introduce a tuning parameter $\rho \in (0, 1]$ which determines the proportion of individuals whose trajectory is updated per iteration.
+That is, in a given MH step, we only update the infection and removal times of a subset of $\lceil\rho n\rceil$ individuals chosen uniformly at random.
+	Smaller values for $\rho$ result in smaller jumps in the latent space and a larger MH acceptance rate, which may lead to better overall mixing in certain cases.
+	Third, if all event times are updated ($\rho = 1$), then the proposed and current latent data are independent conditionally on the current values of the parameters: $\Z^{(k-1)} \perp \Z^* | \theta^{(k)}$.
+This contrasts with existing DA-MCMC algorithms which only update a small fraction of the latent data per iteration and generate consecutive configurations of the latent data that are mostly identical.
+	Our proposal scheme can therefore be said to be semi-independent, a characteristic that is crucial for proving the uniform ergodicity of the resulting Markov chain (see Section \ref{sec:uni}).
+	%Fourth, since the latent data are generated from a process that only resembles the SIR model, we ensure that the Markov chain converges to the correct distribution $\pi(\theta, \Z|\Y)$ by proposing and accepting the latent data according to a Metropolis-Hastings scheme.
+This results in exact Bayesian inference targeting the posterior distribution under the original SIR model.
+	%Even though the latent spaces are generated from an approximation of the SIR process, the Metropolis-Hastings step ensures that the Markov chain converges to the posterior distribution of the parameters under the SIR model; in other words, the inference is exact.
+	Finally, not only is proposing latent data from the RD-SIR process extremely fast since all the necessary random variables can be generated via the inverse-CDF method, but ensuring that the proposed latent data are compatible with the observed data can be done at no additional cost, making the proposal scalable to large outbreaks.
+
+		\jx{We may insert a brief note explicitly contrasting with previous single-site RJMCMC work (ref), giving a bit of intuition why we are able to avoid that change of space, and this could be a natural place to put it.
+Then, they also have a little more context for the experiment you've since added}
 
@@
 
-	Since the proposal for the latent data is independent from the current configuration of the latent data, the efficiency of the DA-MCMC algorithm to explore the latent space directly depends on how similar the surrogate process is to the target process.
-The Markov chain will have good mixing properties if the PD-SIR process is a close approximation of the SIR process.
+	In the special case $\rho=1$, the proposal for the latent data is independent from the current configuration of the latent data.
+The efficiency of the DA-MCMC algorithm to explore the latent space is therefore directly related to the acceptance rate of the MH step.
+The MH acceptance rate in turn depends on how similar the surrogate process is to the target process.
+If the two processes are similar, then the proposed latent data will often be accepted and the Markov chain will have good mixing properties.
+
 	The PD-SIR process only differs from the SIR process in its infection dynamics, the removal dynamics being identical in the two processes.
-Figure \ref{fig:comparison} compares the trajectories of the compartments $S$, $I$ and $R$ of a SIR process of moderate size ($(S(0) = 1000, I(0), R(0)) = (1000, 10, 0)$ with $(\beta, \gamma) = (0.003, 1)$ and $t_{end} = 6$) and those of four PD-SIR processes constrained to be compatible with the observed incidence data $I_{1:K}$ from the SIR process for $K \in \{5, 10, 50, 1,000\}$.
+Figure \ref{fig:comparison} compares the trajectories of the compartments $S$, $I$ and $R$ of a SIR process of moderate size ($(S(0), I(0), R(0)) = (1000, 10, 0)$ with $(\beta, \gamma) = (0.003, 1)$ and $t_{end} = 6$) and those of four PD-SIR processes constrained to be compatible with the observed incidence data $I_{1:K}$ from the SIR process for $K \in \{5, 10, 50, 1,000\}$.
 We see that the PD-SIR is qualitatively close to the SIR, even for small values of $K$.
 Unsurprisingly, the quality of the approximation improves as $K$ increases.
-	% In the limit as $K \rightarrow \infty$, the infection times are effectively known.
-	The fact that the PD-SIR is a faithful approximation of the SIR leads to a relatively high acceptance rate in the Metropolis-Hastings step for the latent data.
-As a result, our algorithm frequently makes larger jumps in the high-dimensional latent space and therefore explores it more efficiently.
-	%Moreover, even though the latent spaces are generated from an approximation of the SIR process, the Metropolis-Hastings step ensures that the Markov chain converges to the posterior distribution of the parameters under the SIR model; in other words, the inference is exact.
-
-	\ram{Piece-wise linear trajectory of S compartment comes from the fact that it follows a linear death process with piece-wise constant rate.}
+The piece-wise linear trajectory of the S compartment in the PD-SIR visible for small $K$ comes from it following a linear birth process with piece-wise constant death rate.
+	% In the limit as $K \rightarrow \infty$, the infection times are effectively known and the two processes are equivalent.
+	The fact that the PD-SIR is a faithful approximation of the SIR, even for moderate $K$, results in a high acceptance rate in the MH step for populations of moderate size (see simulation study in Section \ref{sec:sim}).
+As a result, our algorithm frequently makes large jumps in the high-dimensional latent space and therefore explores it efficiently.
 
@@
 
-	The following characteristics of our DA-MCMC algorithm are worth noting.
-	First, initializing the Markov chain only requires setting $\theta^{(0)} = (\beta^{(0)}, \gamma^{(0)})$ since $\Z^{(0)}$ can be generated from the PD-SIR process conditionally on $\theta^{(0)}$ and $\Y$ alone.
-	Second, since the PD-SIR closely approximates the SIR model, the acceptance rate in the Metropolis-Hastings step for the latent data is relatively high.
-For large population, however, the acceptance rate may be too low, thereby hindering the mixing of the Markov chain.
-To address this issue, we propose to update the event times of only a fraction $0<\rho\le 1$ of the individuals in the population; that is, in the Metropolis-Hastings step we only update the infection and removal times of $\lfloor\rho (S(0)+I(0))\rfloor$ individuals chosen randomly.
-	Smaller values for $\rho$ result in smaller jumps in the latent space and a larger acceptance rate, which may lead to better mixing in certain cases.
-	Third, if all event times are updated ($\rho = 1$), then the proposed and current latent data are independent conditionally on the current values of the parameters: $\Z^{(k-1)} \perp \Z^* | \theta^{(k)}$.
-This contrasts with existing DA-MCMC algorithms which only update a small fraction of the latent data per iteration.
-	%Fourth, since the latent data are generated from a process that only resembles the SIR model, we ensure that the Markov chain converges to the correct distribution $\pi(\theta, \Z|\Y)$ by proposing and accepting the latent data according to a Metropolis-Hastings scheme.
-This results in exact Bayesian inference targeting the posterior distribution under the original SIR model.
-	%Fifth, 
-	Fourth, not only is proposing latent data from the RD-SIR process extremely fast since all random variables can be generated via the inverse-CDF method, but ensuring that these data are compatible with the observed data can be done at no additional cost, making the proposal scalable to large outbreaks.
-	Finally, we show in the following section that the algorithm generates a Markov chain that is \textit{uniformly ergodic}, the strongest form of ergodicity for a Markov chain \cite{Tierney.1994}.
-%The proof of this result is provided in Appendix \ref{app:uni}.
-	% our DA-MCMC only requires I(0), not S(0).
-
-	\jx{We may insert a brief note explicitly contrasting with previous single-site RJMCMC work, giving a bit of intuition why we are able to avoid that change of space, and this could be a natural place to put it.
-Then, they also have a little more context for the experiment you've since added}
-
 	\subsection{Uniform Ergodicity}
-
-	It is straightforward to show that the Markov chain $\{(\theta^{(m)}, \z^{(m)})\}_m$ is harris ergodic and converges to the target posterior distribution $\pi(\theta, \z|\Y)$.
-For any $\pi$-integrable function $g$, the estimator
-	$$\bar{g}_m := \frac{1}{m} \sum_{i=0}^{m-1} g(\theta^{(m)}, \z^{(m)})$$
-	is therefore consistent for $E_\pi g$, regardless of the starting values of the chain.
-A Markov chain is uniformly ergodic if for some finite $M$ and positive constant $r<1$ the $n$-step transition kernel $P^n$ satisfies
+	\label{sec:uni}
+	We now turn to analyze the convergence properties of the proposed MCMC algorithm. 
+	It is straightforward to see that the Markov chain $\left\{\left(\theta^{(m)}, \z^{(m)}\right)\right\}_m$ is Harris ergodic and converges to the joint posterior distribution $\pi(\theta, \z|\Y)$.
+Indeed,
+	by construction, $\pi(\theta, \z|\Y)$ is invariant for the Gibbs kernel $P_\theta$ that updates $\theta$ and for the MH kernel $P_\z$ updating $\z$, and is therefore also invariant for the composite kernel $P=P_\theta P_\z$ \cite{Tierney.1994}.
+Moreover, $P$ is strictly positive, implying that the chain is aperiodic and $\pi$-irreducible.
+Together with the existence of an invariant distribution, $\pi$-irreducibility implies that the chain is positive Harris recurrent.
+Hence, for any $\pi$-integrable function $g$, the ergodic theorem holds and the estimator
+	$$\bar{g}_m := \frac{1}{m} \sum_{i=0}^{m-1} g\left(\theta^{(m)}, \z^{(m)}\right)$$
+	is consistent for $E_\pi g$, for any starting value.
+
+	A Markov chain is \textit{uniformly ergodic} if for some finite $M$ and positive constant $r<1$, the $n$-step transition kernel $P^n$ satisfies
 	$$
-	\Vert P^n(x,.)-\pi(.)\Vert_{TV} \le M r^n, \forall x \in \chi
+	\Vert P^n(x,.)-\pi(.)\Vert_{TV} \le M r^n, \quad \forall x \in \chi
 	$$
-	where $\Vert \mu \Vert_{TV}$ denotes the total variation norm of the measure $\mu$.
-	Uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$ \cite{Jones.2004b}, that is,
+	where $\Vert \mu \Vert_{TV} := \sup_A \mu(A)$ denotes the total variation norm of a signed measure $\mu$.
+	Jones \cite{Jones.2004b} shows that uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$. That is,
 	$$\sqrt{m}(\bar{g}_m - E_\pi g) \Rightarrow N(0, \sigma^2_g)$$
-	for any initial distribution, for some positive, finite $\sigma^2_g$.
-Furthermore, uniform ergodicity implies that the batch means and spectral variance estimators of $\sigma^2_g$ are consistent (ref).
-
-	%First, the Markov chain  $\{(\theta^{(m)}, \z^{(m)})\}_m$ is Harris recurrent and converges to the target posterior distribution.
-Indeed, by construction, the joint posterior density $\pi(\theta, \z|\Y)$ is invariant for the Gibbs kernel $P_\theta$ and the Metropolis-Hastings kernel $P_\z$ and the density of the composite kernel $P = P_\theta P_\theta$ is strictly positive, implying that the chain is aperiodic and $\pi$-irreducible.
-Together with the existence of an invariant distribution, $\pi$-irreducibility implies that the chain is positive Harris recurrent.
-
-	Toward establishing uniform ergodicity of our DA-MCMC algorithm, we will show that the state space $\chi = \chi_{\theta}\times\chi_{\z}$ of the Markov chain $\{(\theta^{(m)}, \z^{(m)})\}_m$ is a \textit{small set} for the transition kernel $P$.
-That is, there exists a probability measure $\nu$ on the $\sigma$-algebra $\sigma(\chi)$ such that
-	$$\beta \nu(.) \le P^m(x,.), \forall x\in \chi.$$
-	for some constants $m \ge 1$ and $\beta > 0$.
-	By Proposition 2 in \cite{Tierney.1994}, this implies that the Markov is uniformly ergodic.
-
-	To derive this result, we will make use of two lemmas. The appendix contains a detailed a proof of the latter one.
-	\ram{Figures for lemma 3.1 and 3.2?}
-
+	for any initial distribution and for some positive, finite $\sigma^2_g$.
+%Furthermore, uniform ergodicity implies that the batch means (ref Geyer 1992) and spectral variance estimators of $\sigma^2_g$ are consistent.
+
+	Toward establishing the uniform ergodicity of our DA-MCMC algorithm, we will show that the state space $\chi = \chi_{\theta}\times\chi_{\z}$ of the Markov chain $\left\{\left(\theta^{(m)}, \z^{(m)}\right)\right\}_m$ is a \textit{small set} for the transition kernel $P$; that is, there exists a probability measure $\nu$ on the $\sigma$-algebra $\sigma(\chi)$ such that
+	\begin{equation}
+	\label{eq:uni}
+	    P^m(x,.) \ge \epsilon \nu(.), \quad \forall x\in \chi
+	\end{equation}
+	for a positive integer $m$ and constant $\epsilon > 0$. This implies that the Markov chain is uniformly ergodic, and that (Proposition 2 in \cite{Tierney.1994})
+	$$
+	\Vert P^n(x,.)-\pi(.)\Vert_{TV} \le M (1-\epsilon)^{n/m}, \quad \forall x \in \chi.
+	$$
+
+	Proposition \ref{pro:uni} shows that \eqref{eq:uni} holds for $m=1$ for our MCMC algorithm.
+This result is worth noting since the requirement that the whole space be small is extremely restrictive and is typically not satisfied for models with unbounded spaces.
+
+	\begin{proposition}
+		\label{pro:uni}
+		For $\rho = 1$, the transition kernel $P$ of the proposed DA-MCMC algorithm satisfies Equation \ref{eq:uni} with $m=1$, that is,
+		$$P((\theta, \z), .) \ge \beta \nu(.) , \quad \forall (\theta, \z) \in \chi_\theta \times \chi_\z
+		$$
+		for some $\beta>0$ and probability measure $\nu$. As a result $P$ is uniformly ergodic.
+	\end{proposition}
+
+	We know provide an brief overview of the proof which provided in Appendix \ref{app:uni}.
+Since the kernel $P = P_{\theta} P_{\z}$ is composite, it suffices to minorize $P_\z$ and $P_\theta$ separately.
+The density of the kernel $P_\z$ corresponds to the product of the gamma densities in \ref{eq:posterior_theta}; since the sufficient statistics of the latent data are bounded, these gamma densities possess a closed form positive minorization.
+The kernel $P_\z$ depends on the initial latent data only through the ratio $q(\z|\theta)/\pi(\z|\theta)$ which can be bounded away from from zero.
+
+	Finally, it is easy to see that, if $\rho < 1$, Equation \eqref{eq:uni} holds for 
+	$m=\lfloor \rho^{-1} \rfloor$ and some $\epsilon$ and $\nu$.
+
+	%Since we minorize the transition kernel over the entire state space, the lower bounds we find are not tight.
+Searching for a better minorization of the form $s(x)\nu(dy)$ and do regeneration sampling (Jones and Hobert, 2001).
+
+    %The constant $\epsilon$ that we derive in the proof of Proposition \ref{pro:uni} will in general be too small to be of practical purposes.
+This results from the loose bounds employed therein.
+Tighter bounds for the sufficient statistics $\int_{0}^{t_{end}}I(t)dt$ and $\int_{0}^{t_{end}}S(t)I(t)dt$ could be derived by taking into account the observed data $I_{1:K}$.
+We do not pursue this issue further.
+
+	\section{Performance on simulated and real epidemic data}
+	\label{sec:per}
+
+	\subsection{Simulation Study}
+	\label{sec:sim}
+	%Outline: (1) proof of concept, (1bis) coverage, (2) rho (3) single-site versus joint proposal
+
+	We validate the convergence properties of the proposed DA-MCMC algorithm empirically via a suite of simulation studies.
+First, we assess the performance of the DA-MCMC algorithm in a moderately sized population of $1000$.
+We simulate an epidemic from true parameters $(\beta, \gamma) = (0.0025, 1)$, which gives $R_0 = 2.5$, starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$ when the outbreak has completed most of its course but is not over yet (see Figure \ref{fig:E1_trajectories}, where $I(t_{end}) = 55$).
+The numbers of infections in $K = 10$ time intervals of equal length are observed, corresponding to $I_{1:K} = (9, 14, 21, 42, 56, 121, 190, 162, 107, 73)$.
+	We employ the parameterization $\tilde{\theta} = (\beta, R_0)$ and use the semi-conjugate prior distributions \eqref{eq:pri2} with weakly informative hyperparameters $\beta \sim Ga(0.001, 1)$ and $R_0 \sim IG(1,1)$.
+	The Markov chain is initialized in a low density region at $(\beta^{(0)}, \gamma^{(0)}) = (\beta/10, \gamma/10) = (0.00025, 0.1)$, and we keep every tenth draw for storage reasons, and choose $\rho = 0.2$ so that the trajectories of $\lceil\rho n\rceil = 202$ individuals are updated in each Metropolis-Hastings step.
+
+	The algorithm takes less than $30$ minutes to generate $1$ million samples on a personal laptop. 
+	Figure \ref{fig:E1_short_no_burn_gamma_tp} shows the traceplot of the parameter $\gamma$ for the first $10000$ iterations.
+We observe that the Markov chain quickly migrates from the low density region where it is initialized to the mode of the target distribution.
+Figures  \ref{fig:E1_burn_gamma_tp} and \ref{fig:E1_burn_gamma_acf}, which show the traceplot of $\gamma$ after a burn-in period of $10000$ iterations and its auto-correlation function, indicate that once the chain reaches the high density region it mixes well.
+The acceptance rate of the Metropolis-Hastings step is $0.21$.
+	The posterior means of $\beta$, $\gamma$ and $R_0$ are respectively $0.00217$, $0.815$ and $2.75$, respectively based on $398$, $370$ and $438$ effective sample sizes.
+The $90\%$ equal-tailed credible intervals are $(0.00166, 0.00278)$, $(0.492, 1.19)$ and $(2.26, 3.45)$ and cover the true values of the parameters.
+
+	We now validate the estimation and uncertainty quantification of the posteriors under our algorithm over repeated simulations.
+To this end, we examine the frequentist coverage properties of the $90\%$ equal-tailed Bayesian credible interval (BCI).
+We repeat the previous simulation $2000$ times and compute the posterior mean of the parameters and the corresponding credible intervals.
+Since the prior is weakly informative, the observed coverage rate of the BCI should be close to $90\%$.
+	Table \ref{tab:coverage} provides the empirical coverage rate of the BCIs for the parameters $\beta$, $\gamma$ and $R_0$ along with the average and variance of the posterior means.
+As expected, the BCIs cover the true values around $90\%$ of the time, well within Monte Carlo error of the nominal coverage under a weakly informative prior.
+This result suggests that running the Markov chain for $1$ million iterations is sufficient to approximate the posterior distribution of the parameters.
+
+	\begin{figure}
+		\centering
+		\begin{subfigure}[b]{0.24\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E1_trajectories}
+			\caption{Compartment trajectories}
+			\label{fig:E1_trajectories}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.24\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E1_short_no_burn_gamma_tp}
+			\caption{Traceplot of $\gamma$ during the transient phase}
+			\label{fig:E1_short_no_burn_gamma_tp}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.24\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E1_burn_gamma_tp}
+			\caption{Traceplot of $\gamma$ after the transient phase}
+			\label{fig:E1_burn_gamma_tp}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.24\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E1_burn_gamma_acf}
+			\caption{Auto-correlation function for $\gamma$}
+			\label{fig:E1_burn_gamma_acf}
+		\end{subfigure}
+		\caption{Performance of the proposed DA-MCMC in a medium-sized population $(S(0), I(0)) = (1000, 10)$ for the parameter $\gamma$.}
+		\label{fig:E1}
+	\end{figure}
+
+    \begin{table}
+    \centering
+    \begin{tabular}{ C{2cm}| *{3}{C{3cm}}}
+        %\hline
+        Parameter & Observed coverage rate & Average of the posterior means & SD of the posterior means \\ 
+        \hline
+        $\beta$ & 0.895 & 0.0025 & 0.000255 \\ 
+        $\gamma$ & 0.902 & 1.00 & 0.142 \\ 
+        $R_0$ & 0.910 & 2.54 & 0.194 \\
+        \hline
+    \end{tabular}
+    \caption{Empirical coverage of $90\%$ posterior credible intervals and distribution of the posterior means from $2000$ independent runs in a medium-sized population ($n=1000$).
+The true values of the parameters are $(\beta, \gamma, R_0) = (0.0025, 1, 2.5)$.
+The standard errors for the observed coverage rate are calculated as $\sqrt{\hat{p}(1-\hat{p})/2000}$, which results in a largest standard error of $6.8$e-3.}
+    \label{tab:coverage}
+    \end{table}
+
+	Second, we explore the impact of the tuning parameter $\rho$ on the mixing properties of the algorithm as the population size varies.
+	We simulate three realizations from the SIR process with $S(0) \in \{250, 500, 1000\}$ with  $I(0)=10$, $\gamma=1$, $t_{end} = 6$ and $K = 10$ for each simulation.
+We choose true parameters so that $R_0 \in \{3, 2.5, 2\}$ respectively, chosen to yield epidemics with comparable dynamics.
+We consider the values $\rho \in \{0.02, 0.05, 0.1, 0.25, 0.5, 1\}$.
+For each population size and each value of $\rho$, the DA-MCMC algorithm is run for $1$ million iterations.
+We use the number of ESS per second to compare the resulting Markov chains.
+
+	Figures \ref{fig:E3_runtime} and \ref{fig:E3_accept} present the run time of the algorithm and the acceptance rate in the Metropolis-Hastings step.
+We observe that the run time increases with the population size and with $\rho$ and that, as expected, the acceptance rate decreases as the number of trajectories updated per iteration $\lceil\rho n\rceil$ increases.
+	Finally, Figure \ref{fig:E3_facet_ESSsecR0} shows the effect of $\rho$ on the mixing properties of the Markov chain.
+We observe that for a population of $250$ individuals updating the entire latent data in the MH step provides the best mixing; as the population size increases, the optimal value of $\rho$ decreases.
+This suggests that for a population of $1000$ individuals, updating the whole latent data results in a excessively low acceptance rate, while updating less than $10$\% makes the updates too small to efficiently explore the latent space.
+
+	\begin{figure}
+		\centering
+		\begin{subfigure}[b]{0.32\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E3_runtime}
+			\caption{Run time in minutes}
+			\label{fig:E3_runtime}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.32\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E3_accept}
+			\caption{Acceptance rate}
+			\label{fig:E3_accept}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.32\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E3_facet_ESSsecR0}
+			\caption{ESS/sec for $R_0$
+			%Acceptance rate of the proposed latent spaces in the Metropolis-Hastings step of the MCMC algorithm across different population sizes and values for $\rho$.
+			}
+			\label{fig:E3_facet_ESSsecR0}
+		\end{subfigure}
+		\caption{Impact of $\rho$ on the performance of the DA-MCMC in a population of size $250$ (solid), $500$ (dotted) and $1000$ (dashed).}
+		\label{fig:E3}
+	\end{figure}
+
+	Finally, we showcase the efficacy of our joint proposal scheme through a comparison %the performance of the proposed DA-MCMC algorithm in which event times are jointly proposed 
+	to a single-site-update DA-MCMC (SSU DA-MCMC) that is similar in spirit to \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
+The SSU sampler that we consider here is a special case of our DA-MCMC in which $\rho = (S(0)+I(0))^{-1}$, so that the infection and removal times of a single individual are updated per iteration.
+The two algorithms are run on the same data for $1$ million iterations.
+Figure \ref{fig:E6} presents the traceplots for $\gamma$ as well as the auto-correlation function.
+We observe that the Markov chain mixes much better when event times are jointly proposed in the Metropolis-Hastings step.
+This is supported quantitatively in Table \ref{tab:E6} which shows that the effective sample size per second obtained with the proposed DA-MCMC is between 7 and 20 times larger than that of the SSU DA-MCMC.
+Hence, even though the computation cost per iteration is larger for our algorithm, it mixes better than SSU algorithms and is more efficient overall.
+
+	\ram{RJ-MCMC, saturation -- when mention other papers}
+
+ \ram{explain why not compare chain binomial, particle filtering (forward simulation), LNA requires noisy data -- at beginning of section e.g.\ Lam's paper.}
+
+	\begin{figure}
+		\centering
+		\begin{subfigure}[b]{0.49\textwidth}
+			\centering
+			\includegraphics[width=0.45\textwidth]{E6_no_burn_beta_tp_joint.jpg}\quad
+			\includegraphics[width=0.45\textwidth]{E6_burn_beta_acf_joint.jpg}
+			\caption{DA-MCMC}
+			\label{fig:E6_no_burn_beta_tp_joint}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.49\textwidth}
+			\centering
+			\includegraphics[width=0.49\textwidth]{E6_no_burn_beta_tp_single.jpg}
+			\includegraphics[width=0.49\textwidth]{E6_burn_beta_acf_single.jpg}
+			\caption{SSU DA-MCMC}
+			\label{fig:E6_no_burn_beta_tp_single}
+		\end{subfigure}
+		\caption{Traceplots and auto-correlation functions of the proposed DA-MCMC and the SSU DA-MCMC for $\beta$.}
+		\label{fig:E6}
+	\end{figure}
+
+	 \begin{table}
+        \centering
+        \begin{tabular}{ |C{2cm}| *{2}{C{3cm}}|}
+            \hline
+            Parameter & DA-MCMC & SSU DA-MCMC \\ 
+            \hline
+            $\beta$ & 0.20 & 0.01 \\ 
+            $\gamma$ & 0.19 & 0.01 \\ 
+            $R_0$ & 0.38 & 0.05 \\
+            \hline
+        \end{tabular}
+        \caption{Effective sample size per second for for the proposed DA-MCMC and the SSU DA-MCMC.}
+        \label{tab:E6}
+    \end{table}
+
+	\subsection{Ebola Outbreak in Gu\'eck\'edou, Guinea}
+	\label{sec:ebo}
+	%Outline: background; setup; results; table; figure
+
+	We now turn to a case study concerning the Ebola outbreak in Western Africa.
+	Between the end of $2013$ and $2015$, Guinea, along with several neighboring countries, experienced the largest outbreak of the Ebola virus disease in history.
+The virus, which has a fatality rate of $70\%$, was responsible for the death of almost $2,000$ people in Guinea alone.
+	The outbreak is believed to have originated in the Gu\'eck\'edou prefecture of Guinea at the end of November $2013$ \cite{Baize.2014}.
+Weekly infection incidence counts are available for each prefecture for the $73$ weeks between the end of December $2013$ and May $2015$.
+
+	We fit the stochastic SIR model to these incidence counts for the Gu\'eck\'edou prefecture using the MCMC algorithm proposed in this article.
+	Our model is illustrative and serves to show that fast and exact Bayesian inference can be made in a large population with the proposed DA-MCMC algorithm instead of providing new insights into this outbreak or the Ebola virus in general.
+For simplicity, we assume that the population of the prefecture forms a closed population, that the model's parameters remained constant throughout the outbreak and that the reported infections counts are exact.
+	\ram{Justify these assumptions.}
+	Applying the proposed DA-MCMC algorithm to stochastic epidemic models where these assumptions are relaxed is the subject of ongoing research.
+	The population size is set to $n = 292000$, the estimated population of the Gu\'eck\'edou prefecture in $2014$.
+	The units of time correspond to ``days" in the analysis and $t=0$ corresponds to Monday December 30, 2013, the first day of the observation period.
+	As the first documented infection occurred late November \cite{Baize.2014}, one month before the first reported incidence count, we set $I(0) = 5$.
+
+	The Markov chain is run for $1$ million iterations and the event times of $\rho=10\%$ of the individuals are updated each iteration.
+The initial values of the parameters are set to $(\beta^{(0)}, \gamma^{(0)}) = (10^{-7}, 0.05)$ and the conjugate distributions in Equation \ref{eq:pri} are used as priors with $a_{\beta} = b_{\beta} = a_{\gamma} = b_{\gamma} = 1$ to make them weakly informative.
+Even for such a large population, the total run time of the algorithm was less than $100$ minutes on a personal laptop.
+	The Metropolis-Hastings step for the latent data proposals achieves a healthy $20.1\%$ acceptance rate, and the first $50,000$ iterations of the Markov chain are discarded as a burn-in.
+	Figure \ref{fig:ebola} shows the marginal posterior distributions $\gamma$, whose means is $0.109$.
+This indicates that people remained infectious for around $9$ days on average, which is consistent with existing literature.
+	%As noted by numerous authors, conditionally on partially observed data, the parameters $\beta$ and $\gamma$ appear positively correlated.
+	%The posterior distribution of $R_0$ is unimodal, relatively symmetric, and centered around $1$.
+A basic reproduction number close to $1$ is consistent with an outbreak that lasted several months without infecting the majority of the population.
+
+	\begin{figure}
+		\centering
+		\begin{subfigure}[b]{0.45\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E5_gamma_hist}
+			\caption{Histogram of $\gamma$}
+			\label{fig:E5_gamma_hist}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.49\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E5_gamma_acf}
+			\caption{Posterior density of $R_0$.}
+			\label{fig:density_R0}
+		\end{subfigure}
+		\caption{Marginal posterior distribution of $\gamma$ and its auto-correlation function for the Gu\'eck\'edou prefecture.}
+		\label{fig:ebola}
+	\end{figure}
+
+	\section{Discussion and Conclusion}
+	\label{sec:dis}
+	The method proposed in this article enables the classical Metropolis-Hastings algorithm to be an efficient method to conduct full posterior inference of the stochastic SIR model given only incidence data.
+Existing attempts using Markov chain Monte Carlo either mix poorly or do not scale to populations of more than a few hundred individuals, while methods relying on forward simulation are limited to moderate-sized outbreaks and may suffer from degeneracy issues on a case-by-case basis depending on the amount of missing data.
+	Moreover, the simplifying assumptions in \cite{Fintzi.2020} or the approximate-Bayesian-computation framework of \cite{McKinley.2018}, which compromise targeting the exact posterior for computational reasons, generate biased estimates that may lead to spurious conclusions.
+Instead, our data-augmented algorithm enables fast and exact inference, even for large outbreaks, and leverages a well-studied, transparent MCMC framework.
+%Most existing inferential methods for epidemic stochastic models are only applicable to prevalence data.
+To our knowledge, our algorithm is the first that permits exact Bayesian inference given incidence data.
+
+	Central to the success of the DA-MCMC algorithm is an efficient proposal scheme that swiftly explores the latent space of epidemic paths that are compatible with the observed data.
+The PD-SIR process possesses three features that make it such an efficient proposal in our DA-MCMC.
+First, generating a PD-SIR process is extremely fast; it only requires the simulation of truncated exponential distributions, which can efficiently be realized via the inverse cumulative distribution function method.
+Moreover, the piece-wise infection rate $\tilde{\mu}(t)$ of the PD-SIR are only updated $K$ times, as opposed to after each event as in the SIR model.
+Second, generating a PD-SIR process that is constrained to be compatible with the observed incidence data can be done at no additional cost; in contrast, generating a SIR process compatible with the observed data would be prohibitively slow \cite{Hobolth.2009}.
+Third, the dynamics of the PD-SIR process closely resemble those of the SIR process: the removal dynamics are identical and, for short resetting intervals, the infection dynamics are also very similar in the two processes.
+
+	The first two features make the algorithm extremely fast.
+While existing data-augmented MCMC algorithms have only be applied to populations of a few hundred of individuals, the analysis of the Ebola outbreak in Gu\'eck\'edou shows that our algorithm can be applied to populations of up to $150,000$ individuals, generating tens of thousands of posterior samples in a reasonable amount of time.
+	The latter feature of the PD-SIR process enables the algorithm to update a large portion of the augmented data in each iteration while maintaining a healthy acceptance rate.
+As a result, the Markov chain makes large jumps in the latent space and has very good mixing properties.
+In contrast, existing DA-MCMC algorithms keep most of the latent space fixed across iterations \cite{Gibson.1998, ONeill.1999, Fintzi.2017} and their Markov chains therefore mix much more slowly.
+
+	The algorithm possesses a tuning parameter $\rho$ that determines the proportion of individuals whose trajectory is updated in the MH step.
+Larger values for $\rho$ enable the chain to make larger steps in the latent space but may result in an excessively low acceptance rate, while smaller values of $\rho$ result in a higher acceptance rate but constrain the chain to make small jumps.
+Depending on the size of the population, different values of the tuning parameter are optimal.
+To find a value that optimizes the mixing properties of the chain, one can use several short runs of the algorithm with different values for $\rho$ and select the value that yields the largest ESS/second.
+%For populations of a few hundred individuals, we found that updating the entire augmented data ($\rho=1$) gives the best mixing.
+In this case, the proposed latent data are independent of the current latent data conditionally on the parameters.
+
+	%Exploring the latent space has been the bottleneck of exact Bayesian inference for stochastic epidemic models. The proposed method allows for efficient exploration of the latent space.
+
+	The DA-MCMC algorithm proposed  in this article  is specific to  the stochastic SIR model, which is arguably a simplistic representation of the spread of disease.
+The model relies on assumptions such as perfect reporting, constant infection rates, homogeneously mixing population and exponentially-distributed infectious periods.
+Our algorithm can be extended to models of increasing realism where these assumptions are relaxed.
+In future work, we will present applications to epidemic models with a time-varying infection rate \cite{Kypraios.2018}, non-Markovian dynamics \cite{Streftaris.2002} as well as under-reporting \cite{Fintzi.2017}.
+
+	\ram{modelling the infection rate over time seems necessary for Ebola data.}
+
+	\appendix
+
+	\section{Proof of Theorem \ref{theo:ldp}}
+	\label{app:ldp}
+
+	Theorem \ref{theo:ldp} was first proved by by \cite{Neuts.1971}. Ross \ram{ref} provides a simpler proof which we now give.
+
+	\begin{proof}
+	Consider a linear pure death process with $n$ particles and individual death rate $\mu$.
+	Let $T_i$ be the time of the $i$th death. Then $W_1 = T_1 \sim \Exp(n\mu)$ and $W_i = T_i - T_{i-1} \sim \Exp((n-1)\mu)$ independently. Let $N$ be the number of deaths by time $t$. Then, 
+	\begin{align*}
+		& f(T_1 = t_1, \dots, T_N = t_N | N) \\
+		& \propto f(T_1 = t_1, \dots, T_N = t_N, T_{N+1} > t) 1\{T_N < t\}\\
+		& \propto f(W_1 = t_1, W_2 = t_2 - t_1, \dots, W_N = t_N - t_{N-1}, W_{N+1} > t - t_N) 1\{T_N < t\}\\
+		& \propto \exp\{-n\mu t_1\} \exp\{-(n-1)\mu(t_2-t_1)\}\dots \exp\{-(n-N)\mu(t - t_N)\} 1\{T_N < t\}\\
+		& \propto \exp\{-\mu t_1\}\exp\{-\mu t_2\} \dots \exp\{-\mu t_N\} 1\{T_N < t\}
+	\end{align*}
+	which corresponds to the kernels of independent exponential distribution truncated above by $t$.
+
+	\end{proof}
+
+	\section{Proof of Proposition \ref{pro:uni}}
+	\label{app:uni}
+
+	We first present two lemmas that will be useful in the proof of Proposition \ref{pro:uni}.
 	% Gamma 1
@@
 		\end{equation}
-		where $x_b^*=\frac{1}{b}\left[ \frac{\Gamma(a+A)}{\Gamma(a)}\right]^{1/A} $.
+		where $x_b^*=\frac{1}{b}\left[ \frac{\Gamma(a+A)}{\Gamma(a)}\right]^{1/A}$. 
 	\end{lemma}
@@
 			& = \left[ \frac{\Gamma(a+A)}{\Gamma(a)}\right]^{(A-\alpha)/A} \frac{\Gamma(a+\alpha)}{\Gamma(a+A)} \\
+			& = \left[
+			\frac{\left(\frac{\Gamma(a+A)}{\Gamma(a)}\right)^{1/A}}{\left(\frac{\Gamma(a+A)}{\Gamma(a+\alpha)}\right)^{1/(A-\alpha)}}
+			\right]^{A-\alpha} \\
 			& = \left( \frac{\Gamma_{a,a+A}}{\Gamma_{a+\alpha, a+A}}\right)^{A-\alpha} \\
@@
 		\end{align*}
-		where $\Gamma_{d,e} = \left( \frac{\Gamma(e)}{\Gamma(d)} \right)^{\frac{1}{e-d}}$ is a geometric average,
+		where $\Gamma_{d,e} = \left( \frac{\Gamma(e)}{\Gamma(d)} \right)^{\frac{1}{e-d}}$ is a geometric mean,
 		and where the last inequality holds because $\Gamma_{a,a+A}\le\Gamma_{a+\alpha, a+A}$.
-		The case $x>x_b^*$ can be shown similarly. Figure \ref{fig:gam1} illustrates these two results.
+		The case $x>x_b^*$ is analogous and is omitted for brevity.
 	\end{proof}
 
-	Lemma \ref{pro:ga1} can be used to minorize a gamma density of the form
+	Figure \ref{fig:gam1a} and \ref{fig:gam1b} illustrate these two results. Lemma \ref{pro:ga1} can be used to obtain a closed form solution to the minimization of a gamma density of the form
 	\begin{equation}
@@
 	\end{equation}
-	over $\alpha$ and $\beta$ for each value of $x$. To this end, we establish the following technical lemma.
+	jointly over $\alpha$ and $\beta$ for each value of $x$. To this end, we establish the following technical lemma.
+	% Gamma 2 (joint)
 	\begin{lemma}	
@@
 	\end{lemma}
-	The final lemma can be used to minorize truncated exponential densities.
-	\begin{lemma}
-		\label{pro:tru}
-		$$\min_{0 \le \mu \le M} \TruncExp(x; \mu, l, u) = \TruncExp(x; M, l, u) $$
-	\end{lemma}
-
-	%\noindent
-	We can now proceed with the proof of uniform ergodicity.
-	% Uniform Ergodicity
-	\begin{proposition}
-		\label{pro:uni}
-		The transition kernel $P = P_{\theta}P_{\z}$ of our DA-MCMC algorithm satisfies a minorization condition $M(1, \delta, \chi, \nu)$ and, as a result, is uniformly ergodic.
-	\end{proposition}
 
 		\begin{proof}
-		First, note that the transition kernel $P$ is a composition of two kernels
-		$$ P = P_{\theta} P_{\z}$$
-		where the kernel $ P_{\theta}$ updates the parameters $\theta$ while keeping the latent data $\z$ fixed and $P_{\z}$ update $\z$ while keeping $\theta$ fixed.
-A one-step transition therefore corresponds to the following scheme
-		$$\x_1 = (\theta_1, \z_1) \rightarrow (\theta_2, \z_1) \rightarrow (\theta_2, \z_2) = \x_2.$$
-
-		The kernel $P$ has a density $p$ with respect to the product measure $\mu = \lambda \mu_{\z}$ with $\lambda$ the Lebesgue measure on $\chi_{\theta}$, the space of $\theta$, and $\mu_{\z}$ a $\sigma$-finite measure on $\chi_{\z}$, the space of $\z$.
-		\begin{align*}
-			P(\x_1, d\x_2) 
-			& = p(\x_1, \x_2) \mu(d\x_2) \\
-			& = p((\theta_1, \z_1), (\theta_2, \z_2)) \lambda(d\theta_2)\mu_{\z}(d\z_2) \\
-			& = p_{\theta}((\theta_1, \z_1), (\theta_2, \z_1)) p_{\z}((\theta_2, \z_1), (\theta_2, \z_2)) \mu_{\z}(d\z_2) \lambda(d\theta_2)
-		\end{align*}
-		where
-		\begin{align*}
-			p_{\theta}((\theta_1, \z_1), (\theta_2, \z_1))
-			%	& = q_{\theta}((\theta_1, \z_1), (\theta_2, \z_1)) \alpha((\theta_1, \z_1), (\theta_2, \z_1)) \\
-			& = \pi(\theta_2| \z_1) \\
-			& = \pi(\beta_2|\z_1) \pi(\gamma_2|\z_1)
-		\end{align*}
-		is the density of the transition kernel $P_{\theta}$ with respect to $\lambda$ and corresponds to the product of the full conditional distributions of $\beta$ and $\gamma$ since the kernel $P_{\theta}$ is a Gibbs sampler, and
-		\begin{align*}
-			p_{\z}((\theta_2, \z_1), (\theta_2, \z_2))
-			& = q_{\z}((\theta_2, \z_1), (\theta_2, \z_2)) \alpha((\theta_2, \z_1), (\theta_2, \z_2)) \\
-			& = q(\z_2; \theta_2) \min\left\lbrace 1, \dfrac{\pi(\theta_2, \z_2)q(\z_1; \theta_2)}{\pi(\theta_2, \z_1)q(\z_2; \theta_2)} \right\rbrace \\
-			& = \min\left\lbrace q(\z_2; \theta_2), \dfrac{\pi(\theta_2, \z_2)q(\z_1; \theta_2)}{\pi(\theta_2, \z_1)} \right\rbrace
-		\end{align*}
-		is the density of the transition kernel $P_{\z}$ with respect to $\mu_{\z}$ and corresponds to one step of the Metropolis-Hasting algorithm where a new configuration of the latent data $\z_2$ is proposed conditionally on the current value of the parameters $\theta_2$, but independently of the current configuration $\z_1$.
-
-		To show that $\chi$ is a small state, it is sufficient to show that there exists a function $k$ such that
-		$$p(\x_1, \x_2) \ge k(\x_2) > 0, \qquad \forall \x_1 \in \chi $$
-		for all $\x_2 \in \chi$.
-		The density $p$ 
-		\begin{align*}
-			p((\theta_1, \z_1), (\theta_2, \z_2))
-			& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2), \dfrac{\pi(\theta_2| \z_1) \pi(\theta_2, \z_2)}{\pi(\theta_2, \z_1)} g(\z_1; \theta_2) \right\rbrace \\
-			& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2), \dfrac{\pi(\theta_2, \z_2)}{\pi(\z_1)} g(\z_1; \theta_2) \right\rbrace \\
-			%	& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2),		\dfrac{			\frac{f(\theta_2, \z_1, Y)}{f(\z_1, Y)}			\frac{f(\theta_2, \z_2, Y)}{f(Y)}		}{\frac{f(\theta_2, \z_1, Y)}{f(Y)}} 		g(\z_1; \theta_2)\right\rbrace \\
-			%	& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2),		\dfrac{f(\theta_2, \z_2, Y)}{f(\z_1, Y)}		g(\z_1; \theta_2)\right\rbrace \\
-			& = \min\left\lbrace \pi(\theta_2| \z_1)g(\z_2; \theta_2),
-			\pi(\theta_2| \z_2)
-			g(\z_1; \theta_2)\right\rbrace
-		\end{align*}
-		depends on $x_1 = (\theta_1, \z_1)$ only through the full conditional $\pi(\theta_2| \z_1)$ and the proposal density $g(\z_1; \theta_2)$.
-
-		It therefore suffices to show that there exist functions $k_1$ and $k_2$ such that
-		$$\pi(\theta| \z_1) \ge k_1(\theta) > 0 \qquad \text{and} \qquad  g(\z_1; \theta) \ge k_2(\theta) > 0.$$
-		First, 
-		\begin{align*}
-			\pi(\theta| \z_1) 
-			& = \pi(\beta| \z_1) \pi(\gamma| \z_1)  \\
-			& = Ga(\beta; a_{\beta} + n_T, b_{\beta} + SI_1) Ga(\gamma; a_{\gamma} + {n_J}_1, b_{\gamma} + I_1) \\
-			& \ge h_{\beta}(\beta) h_{\gamma}(\gamma) \\
-			& = k_1(\theta) > 0
-		\end{align*}
-		where, by Proposition \ref{pro:ga1},
-		$$h_{\beta}(\beta) = \begin{cases}
-			Ga(\beta;a_{\beta}+n_T,b_{\beta}), & \beta<\frac{a_{\beta}+n_T}{t_{end}(n_T+I_0) n} \log \left( 1 + \frac{t_{end}(n_T+I_0) n}{b_{\beta}}\right) \\ Ga(x;a_{\beta}+n_T,b_{\beta}+n(n_T+I_0) t_{end}), & \text{else}
+		Given $x>0$, \eqref{eq:ga1} shows that for a fixed $\alpha$, the gamma density \eqref{eq:ga} is minimized by $\beta \in \{0, B\}$.
+Similarly,  \eqref{eq:ga2} shows that for a fixed $\beta$, $\alpha \in \{0, A\}$ minimizes \eqref{eq:ga}.
+This implies that for a fixed $x>0$, the gamma density \ref{eq:ga} is minimized by $(\alpha, \beta) \in \{(0,0), (A, 0), (0, B), (A,B)\}$.
+
+		A case-by-case analysis of the nine possibilities
+		$$(x<x_a^*; x_a^*<x<x_{a+A}^*; x_{a+A}^* < x )\times(x<x_{b+B}^*; x_{b+B}^*<x<x_b^*; x_b^* < x)$$
+		is presented in Table \ref{tab:ga9} and shows that it is sufficient to consider $(\alpha, \beta) \in \{(A, 0), (0, B)\}$.
+
+		\begin{table}[H]
+			\centering
+			\begin{tabular}{l|c c c}
+				& $x<x_a^*$  & $x_a^*\le x<x_{a+A}^*$ & $x_{a+A}^* < x$ \\ \hline
+				$x<x_b^*$              & $(a+A, b)$ & $(a+A, b)$             &                 \\
+				$x_b^*\le x<x_{b+B}^*$ & $(a+A, b)$ &  ad hoc                & $(a, b+B)$      \\
+				$x_{b+B}^* < x$        &            & $(a, b+B)$             & $(a, b+B)$
+			\end{tabular}
+			\caption{Values of $(\alpha, \beta)$ that minimize $Ga(x;a+\alpha,b+\beta)$ for different values $x$.}
+			\label{tab:ga9}
+		\end{table}
+
+		The two empty entries in Table \ref{tab:ga9} correspond to configurations that are impossible. Indeed, we must have $x_b^* > x_a^*$ and $x_{a+A}^* > x_{b+B}^*$ for all $a, A, b, B$ since
+		$$
+		\frac{x_b^*}{x_a^*} = \dfrac{\left(\frac{\Gamma(a+A)}{\Gamma(a)}\right)^{1/A} b^{-1} }{a B^{-1} \log\left( 1+B/b\right)} = \dfrac{\left(\frac{\Gamma(a+A)}{\Gamma(a)}\right)^{1/A}}{a} \dfrac{B/b}{\log\left( 1+B/b\right)} = \dfrac{\Gamma_{a,a+A}}{a} \dfrac{B/b}{\log\left( 1+B/b\right)} \ge 1
+		$$
+		where the inequality hold since $a \le \Gamma_{a,a+A}$ and $\dfrac{y}{\log\left( 1+y\right)}\ge1$,
+		and
+		$$
+		\frac{x_{b+B}^*}{x_{a+A}^*} = \dfrac{\left(\frac{\Gamma(a+A)}{\Gamma(a)}\right)^{1/A} (b+B)^{-1} }{(a+A) B^{-1} \log\left( 1+B/b\right)} = \dfrac{\Gamma_{a,a+A}}{a+A} \dfrac{B}{b+B} \dfrac{1}{\log\left( 1+B/b\right)} \le 1
+		$$
+		where the inequality holds since $\Gamma_{a,a+A} \le a+A$.
+
+		If $x_a^* \wedge x_{b+B}^*\le x < x_{a+A}^* \vee x_{b}^*$, which corresponds to the middle entry of the center column in Table \ref{tab:ga9}, then one needs to directly check which set of values in $\{(0,0), (A, 0), (0, B), (A,B)\}$ minimizes Equation \ref{eq:ga}.
+In fact, it is sufficient to consider only $\{(A,0), (0, B)\}$ since
+		$$Ga(x, a+A, b) \le \begin{cases}
+			Ga(x;a,b), & x < x_b^* \\
+			Ga(x;a+A,b+B), & x < x_{a+A}^* 
 		\end{cases}$$
-		since $n_T = \sum_k T_k$ is known and 
-		$$SI_1 = \int_{0}^{t_{end}}S_1(t)I_1(t)dt \in [0, n (n_T+I_0) t_{end}];$$
-		and, by Proposition \ref{pro:ga2},
-		$$h_{\gamma}(\gamma) = \min\{
-		Ga(\gamma;a_{\gamma}+n_T+I_0,b_{\gamma}), Ga(x;a_{\gamma},b_{\gamma}+(n_T+I_0) t_{end})\}
-		$$
-		since $0\le n_1^J \le n_T + I_0$ and
-		$$I_1 = \int_{0}^{t_{end}}I_1(t)dt \in [0, (n_T+I_0) t_{end}].$$
-
-		Second, each factor in
-		\begin{align*}
-			& g(\z_1; \theta) = \prod_{i=1}^{n_T} \TruncExp(z^T_i; \beta I_k, t_{k(i)-1}, t_{k(i)}) (1-p_i)^{1\{\z^J_i = \infty\}} (p_i \TruncExp(z^J_i; \gamma, z^T_i, t_{end}))^{1\{\z^J_i \le t_{end}\}}
-		\end{align*}
-		can be minorized as follows. First,
-		\begin{align*}
-			\TruncExp(z^T_i; \beta I_k, t_{k(i)-1}, t_{k(i)}) 
-			& \ge \TruncExp(t_{k(i)}; \beta I_k, t_{k(i)-1}, t_{k(i)}) \\
-			& \ge \TruncExp(t_{k(i)}; \beta (n_T + I_0), t_{k(i)-1}, t_{k(i)}),
-		\end{align*}
-		with the last inequality following from Proposition \ref{pro:tru}, second,
-		$$1-p_i = \exp\{-\gamma (t_{end} - z^T_i)\} \ge \exp\{-\gamma (t_{end} - t_{k(i)-1})\},$$
-		third,
-		$$p_i \TruncExp(z^J_i; \gamma, z^T_i, t_{end}) = \Exp(z^J_i - z^T_i; \gamma)1(z^T_i < z^J_i \le t_{end}) \ge \Exp(t_{end} - t_{k(i) - 1}; \gamma)$$
-		where the equality holds since $p_i = P(z^J_i \le t_{end}|z^J_i > z^T_i)$ is equal to the normalizing constant of the truncated exponential.
-
-		Taken together, these inequalities give
-		\begin{align*}
-			g(\z_1; \theta)
-			& \ge \prod_{i=1}^{n_T} \TruncExp(t_{k(i)}; \beta (I_0 + n_T), t_{k(i)-1}, t_{k(i)}) \min\{\exp\{-\gamma (t_{end} - t_{k(i) - 1})\}, \Exp(t_{end} - t_{k(i) - 1}; \gamma)\} \\
-			& = k_2(\theta) > 0
-		\end{align*}
-		which completes the proof.
+		and
+		$$Ga(x, a, b+B) \le \begin{cases}
+			Ga(x;a,b), & x > x_a^* \\
+			Ga(x;a+A,b+B), & x > x_{b+B}^* 
+		\end{cases}$$
 	\end{proof}
 
-	The kernel 
-	$$
-	P_{\theta}((\theta_0, \z), (\theta_1, \z)) = Ga\left( \beta_1; a_{\beta} + n_I, b_{\beta} + \int_0^{t_{end}} S(t)I(t) dt\right) Ga\left( \gamma_1; a_{\gamma} + n_R, b_{\gamma} + \int_0^{t_{end}} I(t) dt\right) d\beta_1 d\gamma_1
-	$$
-	therefore corresponds to a Gibbs kernel where $n_I, n_R, \int S(t)I(t) dt, \int I(t) dt$ are sufficient statistics from the latent data $\z$
-
-	\section{Performance on simulated and real epidemic data}
-	\label{sec:per}
-
-	\subsection{Simulation Study}
-	\label{sec:sim}
-	%Outline: (1) proof of concept, (1bis) coverage, (2) rho (3) single-site versus joint proposal
-
-	We validate the convergence properties of the proposed DA-MCMC algorithm empirically via a suite of simulation studies.
-First, we assess the performance of the DA-MCMC algorithm in a medium-sized population of over $1000$ individuals.
-We simulate an epidemic from true parameters $(\beta, \gamma) = (0.0025, 1)$, which gives $R_0 = 2.5$, starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$ when the outbreak has completed most of its course but is not over yet (see Figure \ref{fig:E1_trajectories}, where $I(t_{end}) = 55$).
-The numbers of infections in $K = 10$ time intervals of equal length are observed, corresponding to $I_{1:K} = (9, 14, 21, 42, 56, 121, 190, 162, 107, 73)$.
-	We use the conjugate prior distributions \ref{eq:pri} $\beta \sim Ga(0.001, 1)$ and $\gamma \sim Ga(1,1)$, which are only weakly informative.
-	The Markov chain is initialized in a low density region at $(\beta^{(0)}, \gamma^{(0)}) = (\beta/10, \gamma/10) = (0.00025, 0.1)$, and we keep every tenth draw for storage reasons, and set $\rho = 0.2$ so that the trajectories of $\lfloor\rho (S(0)+I(0))\rfloor = 202$ individuals are updated in each Metropolis-Hastings step.
-The algorithm takes less than $30$ minutes to generate $1$ million samples on a personal laptop.
-
-	Figure \ref{fig:E1_short_no_burn_gamma_tp} shows the traceplot of the parameter $\gamma$ for the first $10000$ iterations.
-We observe that the Markov chain quickly migrates from the low density region where it is initialized to the mode of the target distribution.
-Figures  \ref{fig:E1_burn_gamma_tp} and \ref{fig:E1_burn_gamma_acf}, which show the traceplot of $\gamma$ after a burn-in period of $10000$ iterations and its auto-correlation function, indicate that once the chain reaches the high density region it mixes well.
-The acceptance rate of the Metropolis-Hastings step is $0.21$.
-	The posterior means of $\beta$, $\gamma$ and $R_0$ are respectively $0.00217$, $0.815$ and $2.75$, respectively based on $398$, $370$ and $438$ effective sample sizes.
-The $90\%$ credible intervals are $(0.00166, 0.00278)$, $(0.492, 1.19)$ and $(2.26, 3.45)$ and cover the true values of the parameters.
-To validate the uncertainty quantification of the posteriors under our DA-MCMC algorithm, we examine frequentist coverage properties of posterior credible intervals under $2000$ independently repeats of the experiment.
-Tables \ref{tab:coverage} provides the empirical coverage of the $90\%$ credible intervals for the parameters $\beta$, $\gamma$ and $R_0$ along with the average and variance of the posterior means.
-The algorithm has the appropriate coverage, suggesting that running the chain for $1$ million iterations is sufficient to approximate the posterior distribution of the parameters.
-
 	\begin{figure}
-		\centering
-		\begin{subfigure}[b]{0.42\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E1_trajectories}
-			\caption{Compartment trajectories}
-			\label{fig:E1_trajectories}
-		\end{subfigure}
-		\hfill
-		\begin{subfigure}[b]{0.41\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E1_short_no_burn_gamma_tp}
-			\caption{Traceplot of $\gamma$ during the transient phase}
-			\label{fig:E1_short_no_burn_gamma_tp}
-		\end{subfigure}
-		\\
-		\begin{subfigure}[b]{0.41\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E1_burn_gamma_tp}
-			\caption{Traceplot of $\gamma$ after the transient phase}
-			\label{fig:E1_burn_gamma_tp}
-		\end{subfigure}
-		\hfill
-		\begin{subfigure}[b]{0.41\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E1_burn_gamma_acf}
-			\caption{Auto-correlation function for $\gamma$}
-			\label{fig:E1_burn_gamma_acf}
-		\end{subfigure}
-		\caption{Performance of the proposed DA-MCMC in a medium-sized population $(S(0), I(0)) = (1000, 10)$ for the parameter $\gamma$.}
-		\label{fig:E1}
+
 	\end{figure}
 
-    \begin{table}
-    \centering
-    \begin{tabular}{ |C{2cm}| *{3}{C{3cm}}|}
-        \hline
-        Parameter & Coverage rate of the $90\%$ credible intervals & Average of the posterior means & SD of the posterior means \\ 
-        \hline
-        $\beta$ & 0.895 & 0.0025 & 0.000255\\ 
-        $\gamma$ & 0.902 & 1.00 & 0.142 \\ 
-        $R_0$ & 0.910 & 2.54 & 0.194 \\
-        \hline
-    \end{tabular}
-    \caption{Empirical coverage of  $90\%$ posterior credible intervals and distribution of the posterior means from $2000$ independent runs in a medium-sized population.
-The true values of the parameters are $(\beta.
-\gamma.
-R_0) = (0.0025, 1, 2.5)$.}
-    \label{tab:coverage}
-    \end{table}
-
-	Second, we explore the impact of the tuning parameter $\rho$ on the mixing properties of the algorithm across different population sizes.
-	We simulate three SIR processes with $S(0) \in \{250, 500, 1000\}$.
-We set $I(0)=10$, $\gamma=1$, $t_{end} = 6$ and $K = 10$ for each simulation and use $R_0 \in \{3, 2.5, 2\}$ respectively for the different population sizes.
-These values for $R_0$ were chosen to give pandemics with comparable dynamics.
-We consider the values $\rho \in \{0.02, 0.05, 0.1, 0.25, 0.5, 1\}$.
-For each population size and each value of $\rho$, the DA-MCMC algorithm is run for $1$ million iterations.
-%Note that the algorithm is not run if the number of trajectories updated $\rho(S_0+I_0)$ is smaller than $1$.
-
-	Figure \ref{fig:E3_runtime} and \ref{fig:E3_accept} present the run time of the algorithm and the acceptance rate in the Metropolis-Hastings step.
-We observe that the run time increases with the population size and with $\rho$ and that, as expected, the acceptance rate decreases as the number of trajectories updated per iteration $\rho(S(0)+I(0))$ increases.
-	Finally, Figure \ref{fig:E3_facet_ESSsecR0} shows the effect of $\rho$ on the mixing properties of the Markov chain.
-We observe for a population of $250$ individuals, updating the entire latent data in the MH step provides the best mixing; as the population size increases, the optimal value of $\rho$ decreases.
-This suggests that for a population of $1000$ individuals, updating the whole latent data results in too many rejections, while updating less than $10$\% makes the updates too small to efficiently explore the latent space.
-
 	\begin{figure}
 		\centering
-		\begin{subfigure}[b]{0.42\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E3_runtime}
-			\caption{Run time in minutes}
-			\label{fig:E3_runtime}
-		\end{subfigure}
-		\hfill
-		\begin{subfigure}[b]{0.41\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E3_accept}
-			\caption{Acceptance rate}
-			\label{fig:E3_accept}
-		\end{subfigure}
-		\\
-		\begin{subfigure}[b]{0.41\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E3_facet_ESSsecR0}
-			\caption{ESS/sec for $R_0$
-			%Acceptance rate of the proposed latent spaces in the Metropolis-Hastings step of the MCMC algorithm across different population sizes and values for $\rho$.
-			}
-			\label{fig:E3_facet_ESSsecR0}
-		\end{subfigure}
-		\caption{Impact of $\rho$ on the performance of the DA-MCMC in a population of size $250$ (solid), $500$ (dotted) and $1000$ (dashed).}
-		\label{fig:E3}
-	\end{figure}
-
-	Finally, we compare the performance of the proposed DA-MCMC algorithm in which event times are jointly proposed to that of a single-site-update DA-MCMC (SSU DA-MCMC) that is similar in spirit to \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
-The SSU DA-MCMC that we consider here is a special case of the proposed DA-MCMC in which $\rho = (S(0)+I(0))^{-1}$, so that the infection and removal times of a single individual are updated per iteration.
-The two algorithms are run on the same data for $1$ million iterations.
-Figure \ref{fig:E6} presents the traceplots for $\gamma$ as well as the auto-correlation function.
-We observe that the Markov chain mixes much better when event times are jointly proposed in the Metropolis-Hastings step.
-This is supported quantitatively in Table \ref{tab:E6} which shows that the effective sample size per second obtained with the proposed DA-MCMC is between 7 and 20 times larger than that of the SSU DA-MCMC.
-
- \ram{explain why not compare chain binomial, particle filtering (forward simulation), LNA requires noisy data.}
-
-	\begin{figure}
-		\centering
-		\begin{subfigure}[b]{0.42\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E6_no_burn_beta_tp_joint.jpg}
-			\caption{DA-MCMC}
-			\label{fig:E6_no_burn_beta_tp_joint}
-		\end{subfigure}
-		\hfill
-		\begin{subfigure}[b]{0.41\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E6_no_burn_beta_tp_single.jpg}
-			\caption{SSU DA-MCMC}
-			\label{fig:E6_no_burn_beta_tp_single}
-		\end{subfigure}
-		\\
-		\begin{subfigure}[b]{0.42\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E6_burn_beta_acf_joint.jpg}
-			\caption{DA-MCMC}
-			\label{fig:E6_burn_beta_acf_joint}
-		\end{subfigure}
-		\hfill
-		\begin{subfigure}[b]{0.41\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E6_burn_beta_acf_single.jpg}
-			\caption{SSU DA-MCMC}
-			\label{fig:E6_burn_beta_acf_single}
-		\end{subfigure}
-		\caption{Traceplots and auto-correlation functions of the proposed DA-MCMC and the SSU DA-MCMC for $\beta$.}
-		\label{fig:E6}
-	\end{figure}
-
-	 \begin{table}
-        \centering
-        \begin{tabular}{ |C{2cm}| *{2}{C{3cm}}|}
-            \hline
-            Parameter & DA-MCMC & SSU DA-MCMC \\ 
-            \hline
-            $\beta$ & 0.20 & 0.01 \\ 
-            $\gamma$ & 0.19 & 0.01 \\ 
-            $R_0$ & 0.38 & 0.05 \\
-            \hline
-        \end{tabular}
-        \caption{Effective sample size per second for for the proposed DA-MCMC and the SSU DA-MCMC.}
-        \label{tab:E6}
-    \end{table}
-
-	\subsection{Ebola Outbreak in Gu\'eck\'edou, Guinea}
-	\label{sec:ebo}
-	%Outline: background; setup; results; table; figure
-
-	We now turn to a case study concerning the Ebola outbreak in Western Africa.
-	Between the end of $2013$ and $2015$, Guinea, along with several neighboring countries, experienced the largest outbreak of the Ebola virus disease in history.
-The virus, which has a fatality rate of $70\%$, was responsible for the death of almost $2,000$ people in Guinea alone.
-	The outbreak is believed to have originated in the Gu\'eck\'edou prefecture of Guinea at the end of November $2013$ \cite{Baize.2014}.
-Weekly infection incidence counts are available for each prefecture for the $73$ weeks between the end of December $2013$ and May $2015$.
-	The analysis serves to show that fast and exact Bayesian inference can be made regarding the Ebola pandemic with the proposed DA-MCMC algorithm instead of providing new insights into this outbreak or the Ebola virus in general.
-
-	We fit the stochastic SIR model to these incidence counts for the Gu\'eck\'edou prefecture using the MCMC algorithm proposed in this article.
-	For simplicity, we assume that the population of the prefecture forms a closed population, that the model's parameters remained constant throughout the outbreak and that the reported infections counts are exact.
-	\ram{Justify these assumptions.}
-	Applying the proposed DA-MCMC algorithm to stochastic epidemic models where these assumptions are relaxed is the subject of ongoing research.
-	The population size is set to $n = 292000$, the estimated population of the Gu\'eck\'edou prefecture in $2014$.
-	The units of time correspond to ``days" in the analysis and $t=0$ corresponds to Monday December 30, 2013, the first day of the observation period.
-	Since the first documented infection occurred late November \cite{Baize.2014}, one month before the first reported incidence count, we set $I(0) = 5$.
-
-	The Markov chain is run for $1$ million iterations and the event times of $\rho=10\%$ of the individuals are updated each iteration.
-The initial values of the parameters are set to $(\beta^{(0)}, \gamma^{(0)}) = (10^{-7}, 0.05)$ and the conjugate distributions in Equation \ref{eq:pri} are used as priors with $a_{\beta} = b_{\beta} = a_{\gamma} = b_{\gamma} = 1$ to make them weakly informative.
-Even for such a large population, the total run time of the algorithm was less than $100$ minutes on a personal laptop.
-	The Metropolis-Hastings step for the latent data proposals achieves a healthy $20.1\%$ acceptance rate, and the first $50,000$ iterations of the Markov chain are discarded as a burn-in.
-	Figure \ref{fig:ebola} shows the marginal posterior distributions $\gamma$, whose means is $0.109$.
-This indicates that people remained infectious for around $9$ days on average, which is consistent with existing literature.
-	%As noted by numerous authors, conditionally on partially observed data, the parameters $\beta$ and $\gamma$ appear positively correlated.
-	%The posterior distribution of $R_0$ is unimodal, relatively symmetric, and centered around $1$.
-A basic reproduction number close to $1$ is consistent with an outbreak that lasted several months without infecting the majority of the population.
-
-	\begin{figure}
-		\centering
-		\begin{subfigure}[b]{0.45\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E5_gamma_hist}
-			\caption{Histogram of $\gamma$}
-			\label{fig:E5_gamma_hist}
-		\end{subfigure}
-		\hfill
-		\begin{subfigure}[b]{0.49\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E5_gamma_acf}
-			\caption{Posterior density of $R_0$.}
-			\label{fig:density_R0}
-		\end{subfigure}
-		\caption{Marginal posterior distribution of $\gamma$ and its auto-correlation function for the Gu\'eck\'edou prefecture.}
-		\label{fig:ebola}
-	\end{figure}
-
-	\section{Discussion and Conclusion}
-	\label{sec:dis}
-	The method proposed in this article enables the classical Metropolis-Hastings algorithm to be an efficient method to conduct full posterior inference of the stochastic SIR model given only incidence data.
-Existing attempts using Markov chain Monte Carlo either mix poorly or do not scale to populations of more than a few hundred individuals, while methods relying on forward simulation are limited to moderate-sized outbreaks and may suffer from degeneracy issues on a case-by-case basis depending on the amount of missing data.
-	Moreover, the simplifying assumptions in \cite{Fintzi.2020} or the approximate-Bayesian-computation framework of \cite{McKinley.2018}, which compromise targeting the exact posterior for computational reasons, generate biased estimates that may lead to spurious conclusions.
-Instead, our data-augmented algorithm enables fast and exact inference, even for large outbreaks, and leverages a well-studied, transparent MCMC framework.
-%Most existing inferential methods for epidemic stochastic models are only applicable to prevalence data.
-To our knowledge, our algorithm is the first that permits exact Bayesian inference given incidence data.
-
-	Central to the success of the DA-MCMC algorithm is an efficient proposal scheme that swiftly explores the latent space of epidemic paths that are compatible with the observed data.
-The PD-SIR process possesses three features that yield such an efficient algorithm.
-First, generating a PD-SIR process is extremely fast; it only requires the simulation of truncated exponential distributions, which can efficiently be realized via the inverse-CDF method.
-Moreover, the individual infection rates are only updated $K$ times, as opposed to after each event as in the SIR model.
-Second, generating a PD-SIR process that is constrained to be compatible with the observed incidence data can be done at no additional cost; in contrast, generating a SIR process compatible with the observed data would be prohibitively slow \cite{Hobolth.2009}.
-Third, the dynamics of the PD-SIR process closely resemble those of the SIR process: the removal dynamics are identical and, for short resetting intervals, the infection dynamics are also very similar in the two processes.
-
-	The first two features make the algorithm extremely fast.
-While existing data-augmented MCMC algorithms have only be applied to populations of a few hundred of individuals, the analysis of the Ebola outbreak in Gu\'eck\'edou shows that our algorithm can be applied to populations of up to $150,000$ individuals, generating tens of thousands of posterior samples in a reasonable amount of time.
-	The latter feature of the PD-SIR process enables the algorithm to update a large portion of the augmented data in each iteration while maintaining a healthy acceptance rate.
-As a result, the Markov chain makes large jumps in the latent space and has very good mixing properties.
-In contrast, existing DA-MCMC algorithms keep most of the latent space fixed across iterations \cite{Gibson.1998, ONeill.1999, Fintzi.2017} and their Markov chains therefore mix much more slowly.
-
-	The algorithm possesses a tuning parameter $\rho$ that determines the portion of the augmented data being updated each iteration.
-Larger values for $\rho$ enable the chain to make larger steps in the latent space but can result in a low acceptance rate, while smaller values of $\rho$ result in a higher acceptance rate but constrain the chain to make small jumps.
-Depending on the size of the population, different values of the tuning parameter are optimal.
-To find a value that optimizes the mixing properties of the chain, one can use several short runs of the algorithm with different values for $\rho$ and select the value that yields the lowest auto-correlation function (ACF).
-Since the run time of the algorithm increases with $\rho$, one could also look at the effective sample size per unit of time.
-For instance, for populations of a few hundred individuals, we observed that updating the entire augmented data ($\rho=1$) yields the lowest ACF.
-In this case, the current and proposed augmented data are independent conditionally on the current value of the parameters.
-
-	The DA-MCMC algorithm proposed in this article is specific to the stochastic SIR model, which is arguably a simplistic representation of the spread of disease.
-The model relies on assumptions such as perfect reporting, constant infection rates, homogeneously mixing population and exponentially-distributed infectious periods.
-In future work, we will present applications of our DA-MCMC algorithm to processes where these assumptions are relaxed.
-In particular, we are considering extensions to epidemic models with under-reporting \cite{Fintzi.2017}, non-Markovian dynamics \cite{Streftaris.2002}, and a time-varying infection rate \cite{Kypraios.2018}.
-
-	\ram{modelling the infection rate over time seems necessary for EBola data.}
-
-	\appendix
-
-	\section{Distribution of Death Times in Linear Death Process}
-	\label{app:ldp}
-
-	Theorem \ref{theo:ldp} was first proved by by \cite{Neuts.1971}. Ross provides a simpler proof which we now give.
-
-	\begin{proof}
-	Consider a linear pure death process with $n$ particles and individual death rate $\mu$.
-	Let $T_i$ be the time of the $i$th death. Then $W_1 = T_1 \sim \Exp(n\mu)$ and $W_i = T_i - T_{i-1} \sim \Exp((n-1)\mu)$ independently. Let $N$ be the number of deaths by time $t$. Then, 
-	\begin{align*}
-		& f(T_1 = t_1, \dots, T_N = t_N | N) \\
-		& \propto f(T_1 = t_1, \dots, T_N = t_N, T_{N+1} > t) 1\{T_N < t\}\\
-		& \propto f(W_1 = t_1, W_2 = t_2 - t_1, \dots, W_N = t_N - t_{N-1}, W_{N+1} > t - t_N) 1\{T_N < t\}\\
-		& \propto \exp\{-n\mu t_1\} \exp\{-(n-1)\mu(t_2-t_1)\}\dots \exp\{-(n-N)\mu(t - t_N)\} 1\{T_N < t\}\\
-		& \propto \exp\{-\mu t_1\}\exp\{-\mu t_2\} \dots \exp\{-\mu t_N\} 1\{T_N < t\}
-	\end{align*}
-	which corresponds to the kernels of independent exponential distribution truncated above by $t$.
-
-	\end{proof}
-
-	\section{Proofs}
-	\label{app:uni}
-	% 	We show that the state space $\chi = \chi_{\theta}\times\chi_{\z}$ of the Markov chain $\{(\theta, \z)^{(i)}\}_i$ is a small set for the transition kernel $P$, that is, there exists a probability measure $\nu$ on the $\sigma$-algebra $\sigma(\chi)$ such that
-	% 	$$\beta \nu(.) \le P^m(x,.), \forall x\in \chi.$$
-	% 	for some constants $m \ge 1$ and $\beta > 0$.
-	% 	By Proposition 2 in \cite{Tierney.1994}, this implies that the Markov chain $\{(\theta, \z)^{(i)}\}_i$ is uniformly ergodic and therefore satisfies
-	% 	$$
-	% 	\Vert P^n(x,.)-\pi(.)\Vert_{TV} \le M r^n
-	% 	$$
-	% 	for some finite $M$ and positive constant $r<1$, where $\Vert \mu \Vert_{TV}$ denotes the total variation norm of the measure $\mu$.
-
-	% 	The following three lemmas are used to show this result.
-
-	% 	% Gamma 1
-	% 	\begin{lemma}
-	% 		\label{pro:ga1}
-	% 		Let $Ga(x;a,b)$ denote the density of a gamma distribution with shape $a$ and rate $b$ evaluated at $x$. Then
-	% 		\begin{equation}
-	% 			\label{eq:ga1}
-	% 			\inf_{0\le \beta\le B} Ga(x;a,b+\beta) = 
-	% 			\begin{cases}
-	% 				Ga(x;a,b), & x<x_a^* \\ Ga(x;a,b+B), & x\ge x_a^*
-	% 			\end{cases}
-	% 		\end{equation}	
-	% 		where $x_a^*=\frac{a}{B}\log\left( 1+\frac{B}{b}\right) $. Moreover,
-	% 		\begin{equation}
-	% 			\label{eq:ga2}
-	% 			\inf_{0\le \alpha\le A} Ga(x;a+\alpha,b) = 
-	% 			\begin{cases}
-	% 				Ga(x;a,b), & x>x_b^* \\ Ga(x;a+A,b), & x\le x_b^*
-	% 			\end{cases}
-	% 		\end{equation}
-	% 		where $x_b^*=\frac{1}{b}\left[ \frac{\Gamma(a+A)}{\Gamma(a)}\right]^{1/A} $.
-	% 	\end{lemma}
-
-	%Proof of Lemma \ref{pro:ga1}.
-
-	\begin{figure}
-		\centering
-		\begin{subfigure}[b]{0.45\textwidth}
+		\begin{subfigure}[b]{0.32\textwidth}
 			\centering
@@
 		\hfill
-		\begin{subfigure}[b]{0.49\textwidth}
+		\begin{subfigure}[b]{0.32\textwidth}
 			\centering
@@
 		\end{subfigure}
-		\caption{Example of minorization of the density of a gamma distribution $Ga(2+\alpha, 0.5+\beta)$ for $\alpha$ and $\beta$ separately.}
-		\label{fig:gam1}
+		\hfill
+		\begin{subfigure}[b]{0.32\textwidth}
+		    \centering
+		    \includegraphics[width=\textwidth]{gamma_minorization_2.jpg}
+			\caption{$\alpha \in \{0,1\}$, $\beta \in \{0,1\}$}
+		\label{fig:gam2}
+		\end{subfigure}
+		\caption{Example of minorization of the density of a gamma distribution $Ga(2+\alpha, 0.5+\beta)$.}
+		\label{fig:gam}
 	\end{figure}
 
+	%\noindent
+	We can now proceed with the proof of Proposition \ref{pro:uni}.
+
+	\begin{proof}
+		The transition kernel $P$ is a composition of two kernels $P = P_{\theta} P_{\z}$.
+The kernel $P_{\theta}$ updates the parameters $\theta$ while keeping the latent data $\z$ fixed, and $P_{\z}$ update $\z$ while keeping $\theta$ fixed.
+%A one-step transition therefore corresponds to the following scheme
+		%$$\x_1 = (\theta_1, \z_1) \rightarrow (\theta_2, \z_1) \rightarrow (\theta_2, \z_2) = \x_2.$$
+		We therefore have
+		$$P((\theta_1, \z_1), (d\theta_2, d\z_2)) = P_\theta(\theta_1, d\theta_2 | \z_1) P_\z(\z_1, d\z_2 | \theta_2)$$
+		where
+		\begin{align}
+		\label{eq:Pth}
+		    P_\theta(\theta_1, d\theta_2 | \z_1)
+		    & = \pi(d\theta_2 | \z_1) \\
+		    & = \pi(\beta_2 | \z_1) \pi(\gamma_2 | \z_1) d\beta_2 d\gamma_2 \nonumber
+		\end{align}
+		corresponds to the transition kernel of a Gibbs sampler and
+		\begin{align}
+		\label{eq:Pz}
+		    P_\z(\z_1, d\z_2 | \theta_2)
+		    & = Q(d\z_2 | \theta_2) \alpha((\theta_2, \z_1),(\theta_2, \z_2)) + \delta_{\z_1}(d\z_2)\int(1-\alpha((\theta_2, \z_1),(\theta_2, \z_2))) Q(d\z_2| \theta_2) \nonumber \\
+			& \ge Q(d\z_2| \theta_2) \alpha((\theta_2, \z_1),(\theta_2, \z_2)) \\
+			& = q(\z_2| \theta_2) \min\left\lbrace 1, \dfrac{\pi(\theta_2, \z_2)q(\z_1| \theta_2)}{\pi(\theta_2, \z_1)q(\z_2| \theta_2)} \right\rbrace d\z_2 \nonumber
+		\end{align}
+		corresponds to a Metropolis-Hasting transition kernel in which a new configuration of the latent data is generated from the proposal kernel $Q$ conditionally on the current value of the parameters $\theta_2$, but independently of the current configuration $\z_1$.
+The density $q$ of the proposal kernel $Q$ is given in \eqref{eq:q}.
+
+		To show that $\chi$ is a small state, it is sufficient to show that there exists a positive function $k$ such that
+		\begin{equation}
+		    \label{eq:min}
+		    k(\theta_2, \z_2) \le \pi(\beta_2 | \z_1) \pi(\gamma_2 | \z_1) q(\z_2; \theta_2) \min\left\lbrace 1, \dfrac{\pi(\theta_2, \z_2)q(\z_1| \theta_2)}{\pi(\theta_2, \z_1)q(\z_2|\theta_2)} \right\rbrace
+		\end{equation}
+		for all $\theta_1 \in \chi_\theta$ and all $\z_1 \in \chi_\z.$
+		Indeed, suppose that we can find a positive function $k$ satisfying the inequality \eqref{eq:min}, then for any set $A$ and any $(\theta, \z)$ we have, by \eqref{eq:Pth}, \eqref{eq:Pz} and \eqref{eq:min},
+		\begin{align*}
+		    P((\theta, \z), A)
+		    & \ge \int_A \pi(\theta' | \z) \pi(\beta' | \z) \pi(\gamma' | \z) \min\left\lbrace 1, \dfrac{\pi(\theta', \z')q(\z| \theta')}{\pi(\theta', \z)q(\z'|\theta')} \right\rbrace d(\beta', \gamma', \z') \\
+		    & \ge \int_A k(\theta', \z') d(\theta', \z') \\
+		    & = \epsilon \nu(A)
+		\end{align*}
+		with $\epsilon = \int k(\theta, \z) d(\theta, \z)$ a positive constant and $\nu(A) = \epsilon^{-1} \int_A k(\theta, \z) d(\theta, \z)$ a probability measure.
+
+		We now construct a positive function $k$ satisfying \eqref{eq:min}.
+The inequality depends on $(\theta_1, \z_1)$ only through the full conditional distributions $\pi(\beta_2 | \z_1)$ and $\pi(\gamma_2 | \z_1)$ and the ratio $\dfrac{q(\z_1| \theta_2)}{\pi(\theta_2, \z_1)}$.
+		It therefore suffices to find positive functions $k_\beta, k_\gamma$ and $k_r$ such that, for all $\z_1$,
+		$$k_\beta(\beta_2) \le \pi(\beta_2 | \z_1),
+		\quad k_\gamma(\gamma_2) \le \pi(\gamma_2 | \z_1),
+		\quad \text{and} \quad k_r(\theta_2) \le \dfrac{q(\z_1|\theta_2)}{\pi(\theta_2, \z_1)}.$$
+
+		First, we can obtain a closed form minorization of $\pi(\beta_2| \z_1)$ using Proposition \ref{pro:ga1} as follows,
+		\begin{align*}
+			\pi(\beta_2| \z_1) 
+			& = Ga\left(\beta_2; a_{\beta} + n_I, b_{\beta} + \int_{0}^{t_{end}}S(t)I(t)dt\right) \\
+			& \ge \inf_{\begin{aligned}
+				 0\le \int S(t)I(t)dt \le n(n_I+I_0) t_{end}
+		\end{aligned}}   Ga\left(\beta_2; a_{\beta} + n_I, b_{\beta} + \int_{0}^{t_{end}}S(t)I(t)dt\right) \\
+			& = \min\{Ga(\beta_2;a_{\beta}+n_I, b_{\beta}), Ga(\beta_2;a_{\beta}+n_I,b_{\beta}+n(n_I+I_0) t_{end})\} \\
+			%\begin{cases}
+			%Ga(\beta_2;a_{\beta}+n_I, b_{\beta}), & \beta_2<\frac{a_{\beta}+n_I}{t_{end}(n_I+I_0) n} \log \left( 1 + \frac{t_{end}(n_I+I_0) n}{b_{\beta}}\right) \\ Ga(\beta_2;a_{\beta}+n_I,b_{\beta}+n(n_I+I_0) t_{end}), & \text{else}
+	%	\end{cases} \\
+			& = k_\beta(\beta_2) > 0
+		\end{align*}
+		since $n_I = \sum_k I_k$ is known and the sufficient statistic from $\z_1$, 
+		$\int_{0}^{t_{end}}S(t)I(t)dt$, is bounded between $0$ and $n (n_I+I_0)t_{end}$. Similarly, by Proposition \ref{pro:ga2}, we can minorize $\pi(\gamma_2| \z_1)$ as follows,
+		\begin{align*}
+			\pi(\gamma_2| \z_1) 
+			& = Ga\left(\gamma_2; a_{\beta} + n_R, b_{\beta} + \int_{0}^{t_{end}}I(t)dt\right)  \\
+			& \ge \inf_{\begin{aligned}
+				 0 \le n_R &\le n_I + I_0, \\
+				 0 \le \int I(t)dt &\le (n_I+I_0) t_{end}
+		\end{aligned}} Ga\left(\gamma_2; a_{\beta} + n_R, b_{\beta} + \int_{0}^{t_{end}}I(t)dt\right) \\
+			& = \min\{		Ga(\gamma_2;a_{\gamma}+n_R+I_0,b_{\gamma}), Ga(\gamma_2;a_{\gamma},b_{\gamma}+(n_I+I_0) t_{end})\} \\
+			& = k_\gamma(\gamma_2) > 0
+		\end{align*}
+		since the sufficient statistics from $\z_1$, $n_R$ and $\int_{0}^{t_{end}}I(t)dt$, are respectively bounded between $0$ and $n_I + I_0$, and between $0$ and $(n_T+I_0) t_{end}$.
+
+		Finally, the ratio $\dfrac{q(\z_1|\theta_2)}{\pi(\theta_2, \z_1)}$ can be minorized as follows, \ram{(in the second inequality, minimize the whole numerator in terms of $I(.)$ instead of its each factors (=> use previous proof minimizing the proposal density); this will provide a better bound)}
+		\begin{align*}
+		    \dfrac{q(\z_1|\theta_2)}{\pi(\theta_2, \z_1)} 
+		    & = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \TruncExp(z^I_j; \mu_k, t_{k-1}, t_k) \prod_{i=1}^n (1-p_i)^{\mathbf{1}\{z^R_i=\infty\}} \left(p_i \TruncExp(z^R_i;\gamma, z^I_i, t_{end})\right)^{\mathbf{1}\{z^R_i \le t_{end}\}}}
+        	{\beta^{n_I} \prod_{l\in \mathcal{I}} I(z^I_l) \gamma^{n_R} \exp\{-\beta \int_0^{t_{end}} S(t)I(t)dt -\gamma \int_0^{t_{end}} I(t)dt\}} \\
+        	& = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \TruncExp(z^I_j; \mu_k, t_{k-1}, t_k)}
+        	{\beta^{n_I} \prod_{l\in \mathcal{I}} I(z^I_l) \exp\{-\beta \int_0^{t_{end}} S(t)I(t)dt\}}\\
+        	& = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \dfrac{\beta I(t_{k-1}) \exp\{-\beta I(t_{k-1}) (z^I_j - t_{k-1})\}}{1-\exp\{-\beta I(t_{k-1})(t_k - t_{k-1})\}}}
+        	{\beta^{n_I} \prod_{l\in \mathcal{I}} I(z^I_l) \exp\{-\beta \int_0^{t_{end}} S(t)I(t)dt\}}\\
+        	& = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \dfrac{I(t_{k-1}) \exp\{-\beta I(t_{k-1}) (z^I_j - t_{k-1})\}}{1-\exp\{-\beta I(t_{k-1})(t_k - t_{k-1})\}}}
+        	{\prod_{l\in \mathcal{I}} I(z^I_l) \exp\{-\beta \int_0^{t_{end}} S(t)I(t)dt\}}\\
+        	& \ge \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} I(t_{k-1}) \exp\{-\beta I(t_{k-1}) (z^I_j - t_{k-1})\}}
+        	{\prod_{l\in \mathcal{I}} I(z^I_l) \exp\{-\beta \int_0^{t_{end}} S(t)I(t)dt\}}\\
+        	& \ge \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \exp\{-\beta n (t_k - t_{k-1})\}}
+        	{n^{n_I}}\\
+        	& = k_r(\beta_2),
+		\end{align*}
+		where the second equality holds since the	contribution of the removal times in the numerator and the denominator cancel each other as the removal times are generated from the same distribution under the SIR and the PD-SIR processes:
+        \begin{align*}
+        	& \dfrac{\prod_{i=1}^n (1-p_i)^{\mathbf{1}\{z^R_i=\infty\}} (p_i \TruncExp(z^R_i;\gamma, z^I_i, t_{end}))^{\mathbf{1}\{z^R_i \le t_{end}\}}}{\gamma^{n_R} \exp\{- \gamma \int_0^{t_end} I(t)dt\}} \\
+        	& = \dfrac{\prod_{j \in \mathcal{R}^c} \exp\{-\gamma(t_{end}-z^I_j)\} \prod_{k \in \mathcal{R}} (1-\exp\{\gamma(t_{end}-z^I_k)\}) \dfrac{\gamma \exp\{-\gamma (z^R_k - z^I_k)\}}{1-\exp\{\gamma(t_{end}-z^I_k)\}} }{\gamma^{n_R} \exp\{ \gamma \sum_{i=1}^n \min\{z^R_i, t_{end}\} - z^I_i\}} \\
+        	& = \dfrac{\exp\{-\gamma \sum_{j \in \mathcal{R}^c}(t_{end}-z^I_j)\} \gamma^{n_R} \exp\{-\gamma \sum_{k\in R}(z^R_k - z^I_k)\} }{\gamma^{n_R} \exp\{-\gamma[\sum_{j \in \mathcal{R}^c}(t_{end}-z^I_j) + \sum_{k\in \mathcal{R}}(z^R_k - z^I_k)] \}} \\
+        	& = 1,
+        \end{align*}
+        the first inequality holds because $1-\exp\{-\beta I(t_{k-1})(t_k - t_{k-1})\} \le 1$ and the second inequality holds because $1\le I(.) \le n$,  $0\le\int_0^{t_{end}} S(t)I(t)dt$ and $z^I_j \le t_k$ when $j\in\mathcal{I}_k$.
+
+		Taken together, these inequalities give
+		\begin{align*}
+		\pi(\beta_2 | \z_1) \pi(\gamma_2 | \z_1) q(\z_2; \theta_2) \min\left\lbrace 1, \dfrac{\pi(\theta_2, \z_2)q(\z_1| \theta_2)}{\pi(\theta_2, \z_1)q(\z_2
+		|\theta_2)} \right\rbrace
+		& \ge k_\beta(\beta_2) k_\gamma(\gamma_2) q(\z_2; \theta_2) \min\left\lbrace 1, k_r(\beta_2) \dfrac{\pi(\theta_2, \z_2)}{q(\z_2|\theta_2)} \right\rbrace \\
+		& = k(\theta_2, \z_2)
+		\end{align*}
+		for all $\theta_1 \in \chi_\theta$ and all $\z_1 \in \chi_\z$, which completes the proof.
+
+	\end{proof}
+
 	% Gamma 2
-	Proof of Lemma \ref{pro:ga2}: % can be used to minorize a density of the form
+	%can be used to minorize a density of the form
 	% \begin{equation}
@@
 
-	\begin{proof}
-		Given $x>0$, \ref{eq:ga1} shows that for a fixed $\alpha$, $\beta \in \{0, B\}$ is the minimizer of \ref{eq:ga}, and \ref{eq:ga2} shows that for a fixed $\beta$, $\alpha \in \{0, A\}$ is the minimizer of \ref{eq:ga}.
-This implies that for a fixed $x>0$, \ref{eq:ga} is minimized by $(\alpha, \beta) \in \{(0,0), (A, 0), (0, B), (A,B)\}$.
-
-		A case-by-case analysis of the nine possibilities
-		$$(x<x_a^*; x_a^*<x<x_{a+A}^*; x_{a+A}^* < x )\times(x<x_{b+B}^*; x_{b+B}^*<x<x_b^*; x_b^* < x)$$
-		is presented in Table \ref{tab:ga9} and shows that it is sufficient to consider $(\alpha, \beta) \in \{(A, 0), (0, B)\}$.
-
-		\begin{table}[H]
-			\centering
-			\begin{tabular}{l|c c c}
-				& $x<x_a^*$  & $x_a^*\le x<x_{a+A}^*$ & $x_{a+A}^* < x$ \\ \hline
-				$x<x_b^*$              & $(a+A, b)$ & $(a+A, b)$             &                 \\
-				$x_b^*\le x<x_{b+B}^*$ & $(a+A, b)$ &  ad hoc                & $(a, b+B)$      \\
-				$x_{b+B}^* < x$        &            & $(a, b+B)$             & $(a, b+B)$
-			\end{tabular}
-			\caption{Values of $(\alpha, \beta)$ that minimize $Ga(x;a+\alpha,b+\beta)$ for different values $x$.}
-			\label{tab:ga9}
-		\end{table}
-
-		The two empty entries in Table \ref{tab:ga9} correspond to impossible configurations. Indeed, $x_b^* > x_a^*$ and $x_{a+A}^* > x_{b+B}^*$ for all $a, A, b, B$ since
-		$$
-		\frac{x_b^*}{x_a^*} = \dfrac{\left(\frac{\Gamma(a+A)}{\Gamma(a)}\right)^{1/A} b^{-1} }{a B^{-1} \log\left( 1+B/b\right)} = \dfrac{\left(\frac{\Gamma(a+A)}{\Gamma(a)}\right)^{1/A}}{a} \dfrac{B/b}{\log\left( 1+B/b\right)} = \dfrac{\Gamma_{a,a+A}}{a} \dfrac{B/b}{\log\left( 1+B/b\right)} \ge 1
-		$$
-		and
-		$$
-		\frac{x_{b+B}^*}{x_{a+A}^*} = \dfrac{\left(\frac{\Gamma(a+A)}{\Gamma(a)}\right)^{1/A} (b+B)^{-1} }{(a+A) B^{-1} \log\left( 1+B/b\right)} = \dfrac{\Gamma_{a,a+A}}{a+A} \dfrac{B}{b+B} \dfrac{1}{\log\left( 1+B/b\right)} \le 1
-		$$
-		where the inequalities hold since $a \le \Gamma_{a,a+A} \le a+A$ and $\dfrac{y}{\log\left( 1+y\right)}\le1$.
-
-		If $x_a^* \wedge x_{b+B}^*\le x < x_{a+A}^* \vee x_{b}^*$, which corresponds to the middle entry of the center column in Table \ref{tab:ga9}, then one needs to directly check which set of values in $\{(0,0), (A, 0), (0, B), (A,B)\}$ minimizes Equation \ref{eq:ga}.
-In fact, it is sufficient to consider only $\{(A,0), (0, B)\}$ since
-		$$Ga(x, a+A, b) \le \begin{cases}
-			Ga(x;a,b), & x < x_b^* \\
-			Ga(x;a+A,b+B), & x < x_{a+A}^* 
-		\end{cases}$$
-		and
-		$$Ga(x, a, b+B) \le \begin{cases}
-			Ga(x;a,b), & x > x_a^* \\
-			Ga(x;a+A,b+B), & x > x_{b+B}^* 
-		\end{cases}$$
-	\end{proof}
-
-	\begin{figure}	
-		\centering
-		\includegraphics[width=\textwidth]{gamma_minorization_2.jpg}
-		\caption{Example of minorization of the density of a gamma distribution $Ga(2+\alpha, 0.5+\beta)$ for $\alpha$ and $\beta$ jointly.}
-		\label{fig:gam2}
-	\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 	% Truncated Exponential
-	Proof of Lemma \ref{pro:tru}
-
-	\begin{proof}
-		Remember that $\TruncExp(u; \mu, 0, u) = \frac{\mu \exp\{-u\mu\}}{1-\exp\{-u\mu\}}$ is the density of an exponential distribution bounded above by $u$.
-		\begin{align*}
-			\frac{d}{d\mu}\TruncExp(u; \mu, 0, u)
-			& = \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \left[ 1-\mu u - \mu u \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \right]
-		\end{align*}
-		Now let $g(\mu) = 1-\mu u - \mu u \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right)$. By L'H\^{o}pital's rule, $g(0) = 0$. Moreover, for $\mu>0$, we have
-		\begin{align*}
-			g'(\mu)
-			& = - \mu \left(1+\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \left[ 1-u\mu \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \right]  \\
-			& = - \mu \left(1+\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \left[ 1- \frac{u\mu}{\exp\{u\mu\}-1} \right] \\
-			& \le 0
-		\end{align*}
-		where the inequality follows from $\frac{u\mu}{\exp\{u\mu\}-1}\le1$.
-		This implies that $g(\mu) \le 0$ for $\mu>0$. We therefore have $\frac{d}{d\mu}\TruncExp(u; \mu, 0, u) \le 0$ and the result of the proposition follows.
-	\end{proof}
+%	Proof of Lemma \ref{pro:tru}
+
+%	\begin{proof}
+%		Remember that $\TruncExp(u; \mu, 0, u) = \frac{\mu \exp\{-u\mu\}}{1-\exp\{-u\mu\}}$ is the density of an exponential distribution bounded above by $u$.
+%		\begin{align*}
+%			\frac{d}{d\mu}\TruncExp(u; \mu, 0, u)
+%			& = \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \left[ 1-\mu u - \mu u \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \right]
+%		\end{align*}
+%		Now let $g(\mu) = 1-\mu u - \mu u \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right)$. By L'H\^{o}pital's rule, $g(0) = 0$. Moreover, for $\mu>0$, we have
+%		\begin{align*}
+%			g'(\mu)
+%			& = - \mu \left(1+\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \left[ 1-u\mu \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \right]  \\
+%			& = - \mu \left(1+\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \left[ 1- \frac{u\mu}{\exp\{u\mu\}-1} \right] \\
+%			& \le 0
+%		\end{align*}
+%		where the inequality follows from $\frac{u\mu}{\exp\{u\mu\}-1}\le1$.
+%		This implies that $g(\mu) \le 0$ for $\mu>0$. We therefore have $\frac{d}{d\mu}\TruncExp(u; \mu, 0, u) \le 0$ and the result of the proposition follows.
+%	\end{proof}
 

### 2021-12-12 Manuscripts/Article/Article.tex
Updates from Overleaf

@@
 		Stochastic epidemic models provide an interpretable probabilistic description of the spread of a disease through a population.
-Yet, fitting these models in missing data settings is a difficult task; in particular, when the epidemic process is only partially observed, the likelihood of many classic models becomes intractable.
-To remedy this issue, this article introduces a data-augmented MCMC algorithm for fast and exact inference under the stochastic SIR model given discretely observed incidence counts.
-In a Metropolis-Hastings step, the algorithm jointly proposes event times of the augmented data according to a stochastic process whose dynamics closely resemble those of the SIR process, and from which we can efficiently generate an epidemic that is compatible with the observed data.
-Not only is the algorithm fast, but since the augmented data are generated from a very faithful approximation of the target epidemic model, the sampler can update large portions of the augmented data per iteration without lowering the acceptance rate prohibitively.
-We find that the method explores the high-dimensional latent space efficiently, and show that the Markov chain underlying the algorithm is uniformly ergodic.
-% While existing exact MCMC approaches become intractable for populations greater than a few thousand individuals,
-		The proposed algorithm enables exact inference that scales to outbreaks in populations of hundreds of thousands, even on a single laptop.
-We validate its performance via thorough simulation experiments, consider and a case study on the $2014$ Ebola outbreak in Western Africa.
-% in Gu\'eck\'edou, Guinea.
-%, in a population of $150,000$ individuals.
-test.
+Yet, fitting these models when the epidemic process is only partially observed is a notoriously difficult task due to the intractability of the likelihood for many classical models.
+To remedy this issue, this article introduces a novel data-augmented MCMC algorithm for fast and exact inference for the stochastic SIR model given discretely observed incidence counts of infection.
+In a Metropolis-Hastings step, the algorithm jointly proposes the augmented data from a surrogate stochastic process whose dynamics closely resemble those of the SIR process, and from which we can efficiently generate an epidemic that is compatible with the observed data.
+Not only is the algorithm fast, but since the augmented data are generated from a faithful approximation of the target model, the algorithm can update large portions thereof per iteration without lowering the acceptance rate prohibitively.
+We find that the method explores the high-dimensional latent space efficiently and show that the Markov chain underlying the algorithm is uniformly ergodic.
+		The proposed algorithm enables exact Bayesian inference that scales to outbreaks in populations of hundreds of thousands, even on a single laptop.
+We validate its performance via thorough simulation experiments and a case study on the $2014$ Ebola outbreak in Western Africa.
 	\end{abstract}
-	\keywords{Compartmental model; incidence data; exact posterior inference; data-augmented MCMC; likelihood-based methods; uniform ergodicity}
+	\keywords{Stochastic SIR model; incidence data; exact Bayesian inference; data-augmented MCMC; likelihood-based methods; uniform ergodicity}
 
@@
 Originally, deterministic versions of the models were employed by mathematicians and epidemiologists.
-These models are simple to analyze, but fail to capture the inherent randomness in the spread of a disease.
+These models are simple to analyze, but fail to capture the inherent randomness characterizing the spread of a disease.
 For instance, they cannot be used to estimate the probability of a large-scale outbreak or its expected duration, and do not allow for uncertainty quantification when used within inferential procedures.
@@
 Challenges stem from the fact that the observed data typically provide incomplete information on a process that evolves continuously through time, making the likelihood of the model intractable.
-Were the infection and recovery times of each individual of the population observed, inference would be straightforward, but this is rarely the case.
-In practice, one often has access to either incidence data such as weekly counts of new infections, or prevalence data such as the numbers of people infectious at discrete reporting times.
-For instance, the $2013$–$2016$ outbreak of Ebola virus reported over $10,000$, mainly in three Western African countries (Guinea, Liberia and Sierra Leone).
-This outbreak was characterized by the large number of infections that occurred after individuals were tested positive for the virus, giving rise to noisy incidence data.
-Numerous infections happened at the hospitals where infected individuals received a treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased person during funerals is a tradition.
-Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.}.
-	It therefore seems appropriate to model a positive test for the virus as an indication that the individual became infectious some time prior to the test rather than as a removal time \jx{to do: stronger transition tying back to goal of partial observed inference}.
-
-	More specifically, the marginal likelihood of such partially observed data becomes a bottleneck, as it requires a large integration step that accounts for all possible configurations of the missing.
+%Were the infection and recovery times of each individual of the population observed, inference would be straightforward, but this is rarely the case.
+	In practice, one often has access to either incidence data such as weekly counts of new infections, or prevalence data such as the numbers of infectious individuals at discrete reporting times.
+The marginal likelihood of such partially observed data becomes a computational bottleneck, as it requires a large integration step that accounts for all possible configurations of the missing epidemic process.
 In particular, direct computation of this likelihood requires the transition probabilities between observation times for which no closed form is available.
-\jx{to do: cite recent papers from Simon Spencer's group too that do so for discrete-time models, and mention classical matrix exponentiation being intractable}.
-Numerical methods to obtain transition probabilities have only recently been developed in \cite{Ho.2018b, Ho.2018}, but the high computational costs limit their algorithm to moderate sized outbreaks.
-
-	In this article, we make use of data augmentation to explore configurations of the missing data through latent variables.
-We show how a data-augmented MCMC algorithm can target the exact joint posterior of model parameters and these latent variables under the stochastic SIR model, a commonly used SEM, from discretely observed infection incidence counts.
+Moreover, given the size of the transition matrix, classical matrix exponentiation is intractable.
+%\jx{to do: cite recent papers from Simon Spencer's group too that do so for discrete-time models, and mention classical matrix exponentiation being intractable}.
+	Numerical methods to obtain transition probabilities have only recently been developed in \cite{Ho.2018b, Ho.2018}, but the high computational costs limit their approach to moderate sized outbreaks.
+
+	In this article, we employ data augmentation to explore configurations of the missing data through latent variables.
+We introduce a uniformly ergodic data-augmented MCMC (DA-MCMC) algorithm that targets the exact joint posterior of model parameters and these latent variables under the stochastic SIR model, a commonly used SEM, from discretely observed incidence counts of infections.
 Rather than marginalizing directly, our approach accounts for the latent space via sampling.
-The algorithm updates the event times of the augmented data by jointly proposing infection and removal times from a stochastic process whose dynamics closely resemble those of the SIR process, and from which we can efficiently generate an epidemic that is compatible with the observed data.
+The algorithm updates the event times of the augmented data by jointly proposing infection and removal times from a surrogate stochastic process whose dynamics closely resemble those of the SIR process, and from which we can efficiently generate an epidemic that is compatible with the observed data.
 Its success lies in the design of this surrogate process serving as an efficient proposal scheme within a Metropolis-Hastings algorithm.
-Since the latent spaces are generated from a very faithful approximation of the target epidemic model, the algorithm can update a large portion of the event times per iteration while maintaining a relatively high acceptance rate, thereby exploring the high-dimensional latent space efficiently.
+Since the latent spaces are generated from a faithful approximation of the target epidemic model, the algorithm can update a large portion of the event times per iteration while maintaining a high acceptance rate, thereby exploring the high-dimensional latent space efficiently.
 Our approach is fully Bayesian and therefore allows representation of parameter uncertainty through posterior distributions.
@@
 	The remainder of the article is structured as follows.
-Section \ref{sec:set} provides background information on the inference task.
-It introduces the stochastic SIR model and explains why conducting inference under partially observed data is a difficult task.
+Section \eqref{sec:set} provides background information on the inference task.
+It introduces the stochastic SIR model and explains why conducting inference from partially observed data is a difficult task.
 Previous works addressing this problem are also presented.
-Section \ref{sec:pds} describes the data-augmented MCMC algorithm proposed in this article and assesses how closely it approximation the SIR process.
-In Section \ref{sec:sim}, we examine the performance of our algorithm on simulated data and then turn to an analysis of the $2014$ Ebola outbreak in Gu\'eck\'edou, Guinea in Section \ref{sec:ebo}.
-Finally, Section \ref{sec:dis} discusses the findings and concludes the article.
+Section \eqref{sec:pds} introduces the DA-MCMC algorithm and describes the surrogate process from which the latent data are generated in the Metropolis-Hastings step.
+The similarity between the surrogate and the target processes is assessed in Section \eqref{sec:qua} and the DA-MCMC is shown to be uniformly ergodic in Section \eqref{sec:uni}.
+In Section \eqref{sec:sim}, we examine the performance of our algorithm on simulated data and then turn to an analysis of the $2014$ Ebola outbreak in Gu\'eck\'edou, Guinea in Section \eqref{sec:ebo}.
+Finally, Section \eqref{sec:dis} discusses the findings and concludes the article.
 
@@
 
-	Our point of departure is to consider the stochastic SIR model -- also referred to as the general stochastic epidemic model (Bailey, 1975) -- which offers a parsimonious and interpretable representation of the mechanistic dynamics of an epidemic.
-The stochastic SIR model is a compartmental model in which the individuals of a population transition through three types or compartments: susceptible (S), infectious (I) and removed (R).
+	Our point of departure is to consider the stochastic SIR model -- also referred to as the general stochastic epidemic model \cite{Bailey.1975} -- which offers a parsimonious and interpretable representation of the mechanistic dynamics of an epidemic.
+The stochastic SIR model is a compartmental model in which individuals transition through three states or compartments: susceptible (S), infectious (I) and removed (R).
 The only possible moves are from S to I (infections) and from I to R (removals).
@@
 Once infected, she is immediately infectious and remains so for some period of time after which she is removed from the process without the possibility of reinfection.
-In this formulation, demographic dynamics such as births and deaths of individuals are ignored since they usually occur at a much slower rate than infections and removals.
+In this formulation, demographic dynamics such as births and deaths are ignored since they usually occur at a much slower rate than infections and removals.
 
@@
 	\begin{equation}
-		\X(t) = (X_1(t), \dots, X_n(t)) \in \{s, i, r\}^n
+		\X(t) = \left(X_1(t), \dots, X_n(t)\right) \in \{s, i, r\}^n
 	\end{equation}
@@
 	$$
-	denotes the status (compartment) of individual $j$ at time $t$ with $\tau^I_j$ and $\tau^R_j$ respectively the infection and removal times of individual $j$.
+	denotes the compartment of individual $j$ at time $t$ with $\tau^I_j$ and $\tau^R_j$ respectively the infection and removal times of individual $j$.
 If individual $j$ never becomes infectious, we set $\tau^I_j = \tau^R_j = \infty$ and $X_j(t) = s$ for $t \in [0, \infty)$.
 Since removed individuals do not contribute to the pandemic, it is safe to ignore the individuals removed at time $0$ and write $n = S(0) + I(0)$.
-Here $\chi_{\X}$ denotes the set of trajectories compatible with the evolution of a disease---that is, the set of trajectories in which no infection occurs whenever the infectious compartment is depleted:
+Here the space $\chi_{\X}$ denotes the set of trajectories compatible with the evolution of a disease---that is, the set of trajectories in which no infection occurs when the infectious compartment is depleted:
 	\begin{equation}
 		\label{eq:chi}
-		\chi_{\X} = \{X:X(t) \in \{s,r\}^n \Rightarrow X(t+u) = X(t), \forall u>0 \}.
+		\chi_{\X} = \{\X:\X(t) \in \{s,r\}^n \Rightarrow \X(t+u) = \X(t), \forall u>0 \}.
 	\end{equation}
@@
 Thus, the individual-level infection and removal rates at time $t$ are respectively $\beta I(t)$ and $\gamma$.
-
-	The model can alternatively be parameterized in terms $\tilde{\theta} = (\beta, R_0)$, with $R_0 = S(0)\beta/\gamma$. The likelihood becomes
-	\begin{equation*}
-	    L((R_0, \gamma); \X) = \beta^{n_I + n_R} R_0^{-n_R} \prod_{j:\tau_{T_j} \le t_{end}} I(\tau_{T_j}) \exp\left\lbrace - \beta \int_{0}^{t_{end}} I(t)S(t) dt - \frac{\beta}{R_0} \int_{0}^{t_{end}} I(t) dt \right\rbrace
-	\end{equation*}
 
@@
 		L(\theta; \X)
-		& = \prod_{j \in \mathcal{I}} \beta I(\tau^I_j) \prod_{k \in \mathcal{R}} \gamma \exp\left\lbrace - \int_{0}^{t_{end}}\beta I(t)S(t) + \gamma I(t) dt \right\rbrace  \nonumber \\
+		& = \prod_{j \in \mathcal{I}} \beta I(\tau^I_j) \prod_{k \in \mathcal{R}} \gamma \exp\left\lbrace - \int_{0}^{t_{end}}\beta S(t)I(t) + \gamma I(t) dt \right\rbrace  \nonumber \\
 		\label{eq:cdl}
-		& = \beta^{n_I} \gamma^{n_R}\prod_{j \in \mathcal{I}} I(\tau^I_j) \exp\left\lbrace - \beta \int_{0}^{t_{end}} I(t)S(t) - \gamma \int_{0}^{t_{end}} I(t) dt \right\rbrace
+		& = \beta^{n_I} \gamma^{n_R}\prod_{j \in \mathcal{I}} I(\tau^I_j) \exp\left\lbrace - \beta \int_{0}^{t_{end}} S(t)I(t) - \gamma \int_{0}^{t_{end}} I(t) dt \right\rbrace
 	\end{align}
@@
 
-	It is straightforward to conduct inference in this continuously observed scenario.
-The likelihood \eqref{eq:cdl} belongs to the exponential family, and maximum likelihood estimates of its parameters can be expressed in terms of the sufficient statistics defined above as \ram{ref for exponential family and MLE - Guttorp 1994, perhaps newer version with Minin.} Simple algebra shows that the maximum likelihood estimates are
+	It is straightforward to conduct inference in this continuously observed scenario \cite{Guttorp.2018}.
+The likelihood \eqref{eq:cdl} belongs to the exponential family, and maximum likelihood estimates of its parameters can be expressed in terms of the sufficient statistics defined above as
 	$$
-	\hat{\beta} = \dfrac{n_I}{ \int_{0}^{t_{end}} I(t)S(t)dt}, \qquad \hat{\gamma} = \dfrac{n_R}{ \int_{0}^{t_{end}} I(t)dt}.
+	\hat{\beta} = \dfrac{n_I}{ \int_{0}^{t_{end}} S(t)I(t)dt}, \qquad \hat{\gamma} = \dfrac{n_R}{ \int_{0}^{t_{end}} I(t)dt}.
 	$$
 	Note that since the functions $I(t)$ and $S(t)$ are constant between event times, the two integrals correspond to finite sums which are straightforward to compute.
-	Furthermore, in a Bayesian context, inference is facilitated by the conjugacy with the gamma and inverse-gamma distributions: if we use gamma independent prior distributions
+	Furthermore, in a Bayesian context, inference is facilitated by the conjugacy of \eqref{eq:cdl} with the gamma distribution. If we use independent gamma priors
 	\begin{equation}
@@
 	\end{align}
-	from which one can easily generate independent values to explore $\pi(\tilde{\theta}|\X)$ via Monte Carlo.
-Moreover, since $\beta$ and $R_0$ are almost uncorrelated in the posterior distribution, the Gibbs sampler would be an efficient algorithm to explore the posterior distribution of the parameters when the epidemic process is completely observed.
-
-	The likelihood in Equation \ref{eq:cdl} can also be parameterized in terms of $\tilde{\theta} = (\beta, R_0)$ where $R_0 := S(0) \beta / \gamma$ is called the basic reproduction number and corresponds to the expected number of secondary infections caused by an infectious individual in a susceptible population.
-Using $\tilde{\theta}$ with independent prior distributions
+	from which one can easily generate independent values to explore the posterior distribution $\pi(\theta|\X)$ via Monte Carlo.
+
+	The likelihood \eqref{eq:cdl} can also be parameterized in terms of $\tilde{\theta} = (\beta, R_0)$ where $R_0 := S(0) \beta / \gamma$ is called the basic reproduction number and corresponds to the expected number of secondary infections caused by an infectious individual in a susceptible population.
+Using $\tilde{\theta}$ with the independent prior distributions
 	\begin{equation}
 		\label{eq:pri2}
-		\beta \sim G(a_{\beta}, b_{\beta}), \qquad R_0 \sim IG(a_{R}, b_{R})
-	\end{equation}
-	yields the closed-form full conditional distributions.
-	\begin{align}
-		\label{eq:posterior_theta2}
-		\beta | \X & \sim G\left( a_{\beta} + n_I + n_R, b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt + \frac{S(0)}{R_0} \int_{0}^{t_{end}} I(t) dt\right) \\
+		\beta \sim G(a_{\beta}, b_{\beta}), \qquad R_0 \sim IG(a_{R}, b_{R}),
+	\end{equation}
+	where $IG$ denote the inverse-gamma distribution, yields the closed-form full conditional distributions.
+	\begin{align}		\label{eq:posterior_theta2}
+		\beta | \X, R_0 & \sim G\left( a_{\beta} + n_I + n_R, b_{\beta} + \int_{0}^{t_{end}} S(t)I(t) dt + \frac{S(0)}{R_0} \int_{0}^{t_{end}} I(t) dt\right) \\
 		R_0 | \X, \beta & \sim IG\left( a_{R} + n_R, b_{R} + \beta S(0) \int_{0}^{t_{end}} I(t) dt\right)
 	\end{align}
-	We have noted that the parameterization $\tilde{\theta}$ results in a Markov chain that mixes better and therefore employ it in the simulation studies and data analysis of Section \ref{sec:per}.
+	We have observed that the parameterization $\tilde{\theta}$ results in a Markov chain that mixes better and therefore employ it in the simulation studies and data analysis of Section \eqref{sec:per}.
 
@@
 
-	In practice, inference is complicated by the fact that the epidemic process \eqref{eq:X} is typically only partially observed.
+	In practice, inference is typically complicated by the fact that the epidemic process \eqref{eq:X} is only partially observed.
 When we do not observe all event times, we do not have access to the sufficient statistics needed to evaluate the likelihood \eqref{eq:cdl}.
@@
 For instance, the former arises in animal experiments in which positive cases are immediately isolated from the rest of the population and no longer contribute to disease spread.
-In such cases, exact times of removals are known, but such information is unavailable in typical observational studies.
+In such cases, exact times of removals are known, but such information is unavailable in a typical observational study.
 
@@
 
-	In the Bayesian framework, the posterior distribution of the parameters given the observed data is formally related to the complete data likelihood \eqref{eq:cdl} via integration:
+	This form of observed data is motivated by the $2013$–$2016$ outbreak of the Ebola virus, the largest outbreak since the discovery of the virus in 1976, which resulted in the deaths of at least 11,000 individuals, mainly in three Western African countries (Guinea, Liberia and Sierra Leone).
+The data available for this outbreak consist of weekly number of positive tests.
+This outbreak was characterized by the large number of infections that occurred after individuals were tested positive for the virus, giving rise to noisy incidence data.
+Numerous infections took place at the hospitals where infected individuals received treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased person during funerals is a tradition.
+Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.} \ram{ref}.
+We therefore decide to model the weekly number of positive tests as the weekly number of infections in a SIR model.
+
+	%It therefore seems appropriate to model a positive test for the virus as an indication that the individual became infectious some time prior to the test rather than as a removal time.
+
+	In a Bayesian framework, the posterior distribution of the parameters given the observed data is formally related to the complete data likelihood \eqref{eq:cdl} via integration:
 	\begin{equation}
@@
 		\pi(\theta|\Y) 
-		\propto \pi(\theta) L(\theta; \Y) = \pi(\theta) \int_{\chi_\chi} L(\theta; \x) \mathbf{1}\{\Y \overset{\Delta}{=} \x\} d\x
+		\propto \pi(\theta) L(\theta; \Y) = \pi(\theta) \int_{\chi_\x} L(\theta; \x) \delta_{\Y}(\x) d\x
+		%\mathbf{1}\{\Y \overset{\Delta}{=} \x\}
 	\end{equation}
 	where $\pi(\theta)$ is the prior distribution on $\theta$, and
-	$\mathbf{1}\{\Y \overset{\Delta}{=} \x\} = 1$ if $\Y$ is compatible with the infection times in $\x$ and $0$ otherwise; that is, the partial data likelihood $L(\theta; \Y)$ consists of a high-dimensional integral over all epidemic paths that agree with the observed data.
-This marginalization step has no known closed form solution and presents computational challenges even for a population of moderate size.
+	$\delta_{\Y}(\x) = 1$ if $\x$ is compatible with the observed infection counts $\Y$ and $0$ otherwise; that is, the partial data likelihood $L(\theta; \Y)$ consists of a high-dimensional integral over all epidemic paths that agree with the observed data.
+This marginalization step has no known closed-form solution and presents computational challenges even for a population of moderate size.
 
@@
 Popular approximations include chain binomial models \cite{Greenwood.1931, Abbey.1952}, diffusion processes \cite{Cauchemez.2008, Fintzi.2020} and Gaussian processes \cite{Jandarov.2014}.
-While these approximation-based approaches bypass the intractability of the partial data likelihood \ref{eq:pdl}, the assumptions on which they are based are questionable when the population is small, and, as a result, the dynamic of the stochastic process differs from its asymptotic behavior.
+While these approximation-based approaches bypass the intractability of the partial data likelihood \eqref{eq:pdl}, the assumptions on which they are based are questionable when the population is small, and, as a result, the dynamic of the stochastic process differs from its asymptotic behavior.
 	%Recently, \cite{Ho.2018} proposed an exact analytical method to directly compute the likelihood of the stochastic SIR process given discretely observed data.
 
-	Since the popularization of Markov chain Monte Carlo methods (MCMC) in statistics \cite{Tanner.1987, Gelfand.1990, Tierney.1994}, researchers have developed sampling-based methods to directly work with the SIR likelihood instead of an approximation thereof.
-MCMC algorithms fall into two categories: model-based forward simulation and data augmented MCMC (DA-MCMC).
-	Particle filtering \cite{King.2015} is an example of the former category that is extremely popular among practitioners.
+	Since the popularization of Markov chain Monte Carlo (MCMC) methods in statistics \cite{Tanner.1987, Gelfand.1990, Tierney.1994}, researchers have developed sampling-based methods to directly work with the SIR likelihood instead of an approximation thereof.
+MCMC algorithms fall into two categories: model-based forward simulation and data-augmented MCMC (DA-MCMC).
+	Particle filtering \cite{King.2015} is an example of the former category that is popular among practitioners.
 Its plug-and-play feature makes it applicable to a wide variety of models.
-However, model-based forward simulation suffers from two drawbacks: simulating data from a model as complex as the SEM that is compatible with the observed data is prohibitively slow, and these methods can fail to converge when the model does not fit the data well.
+However, model-based forward simulation suffers from two drawbacks: simulating data from a model as complex as the SIR that is compatible with the observed data is prohibitively slow, and these methods can fail to converge when the model does not fit the data well.
 Approximate Bayesian Computation (ABC) \cite{McKinley.2018} offers a solution to the latter problem, but its inference is based on an approximation of the model's likelihood.
-As a result, the inference is inevitably biased, and it is difficult to evaluate the degree of approximation involved.
+As a result, the inference is biased, and it is difficult to evaluate the degree of approximation involved.
 %quantify how closely the approximate posterior resembles the original target distribution.
 
-	The second family of MCMC-based inferential methods treats the unobserved event times as nuisance parameters; that is, the observed data are augmented with latent data that consist of the times and types of unobserved epidemic events.
-	Researchers have mostly focused on the development of better and more efficient proposal schemes to explore this high-dimensional latent space efficiently.
+	The second family of MCMC-based inferential methods treats the unobserved event times as nuisance parameters; that is, the observed data are augmented with latent variables that consist of the times and types of unobserved epidemic events.
+	Researchers have mostly focused on the development of more efficient proposal schemes to explore this high-dimensional latent space.
 Early attempts employed reversible-jump MCMC \cite{Green.1995} to explore models with different numbers of unobserved events \cite{Gibson.1998, ONeill.1999}.
-These authors augment the observed data which consist of the recovery times with the unobserved infection times and explore the latent space by uniformly inserting, deleting or moving an infection time in each iteration of the MCMC algorithm.
-	More recently, Fintzi et al.
-\cite{Fintzi.2017} proposed a MCMC algorithm to conduct inference with discretely observed infection prevalence counts.
+These authors augment the observed data which consist of the recovery times with the unobserved infection times and explore the latent space by uniformly inserting, deleting or moving an element of the latent data per iteration of the MCMC algorithm.
+	More recently, Fintzi et al.\ \cite{Fintzi.2017} proposed a MCMC algorithm to conduct inference with discretely observed prevalence data of infection.
 They constructed latent data consisting of the infection and recovery times of each individual.
-The algorithm explores this latent space by updating the event times of an individual in each iteration of the algorithm.
-These event times are sampled from a distribution that approximate the exact dynamics of the stochastic process.
-	In the discrete-time setting, Touloupou et al. \cite{Touloupou.2020} used a Gibbs sampler to update the trajectory of an individual per iteration.
+The algorithm explores this latent space by updating the event times of an individual per iteration of the algorithm.
+These event times are sampled from a distribution that approximate the exact dynamics of the epidemic process.
+	In the discrete-time setting, Touloupou et al.\ \cite{Touloupou.2020} used a Gibbs sampler to update the trajectory of an individual per iteration.
 	These DA-MCMC methods suffer from poor mixing in the presence of large epidemics.
-Since they update a very small number of latent data per iteration, the resulting Markov chains is sticky: it makes very small jumps in the latent space and therefore requires a large number of iterations to explore it completely.
-	Although the update of a slightly larger number of event times per iteration in \cite{Pooley.2015} and the non-centered parameterization of \cite{Neal.2005} may improve the performance, the gains are modest; the Markov chains still suffer from a high level of auto-correlation and do not possess satisfactory mixing properties in populations over a few hundred individuals.
-	Even though these DA-MCMC algorithms may update more than one element per iteration, we will refer to them as \textit{single}-site DA-MCMC algorithms to contrast them to the proposed method in which the entire latent data are jointly updated.
+Since they update a very small number of latent variables per iteration, the resulting Markov chains is sticky: it makes very small jumps in the latent space and therefore requires a large number of iterations to explore it completely.
+	Although the update of a slightly larger number of event times per iteration in \cite{Pooley.2015} and the non-centered parameterization of \cite{Neal.2005} may improve the performance of the algorithms, the gains are modest; the Markov chains still suffer from a high level of auto-correlation and do not possess satisfactory mixing properties in populations over a few hundred individuals.
+	By a slight abuse of language, we will refer to these DA-MCMC algorithms as \textit{single}-site-update (SSU) DA-MCMC algorithms to contrast them to the method proposed in this article in which the entire latent data are jointly updated.
 
@@
 Our method hinges on the efficacy of a carefully designed proposal process for the latent variables that faithfully approximates the SIR process dynamics.
-	The Metropolis-Hastings sampler for the latent data targets the exact posterior of the model given partially observed incidence data, and enjoys the efficiency of fast proposals in the high-dimensional latent space as well as fast computations involving the complete data likelihood.
-
-	As shown in Section \ref{sec:icd}, the complete data likelihood \eqref{eq:cdl} is amenable to computation.
-This suggests augmenting the observed data $\Y$ with latent data $\Z$ such that the likelihood $L(\theta; (\Y, \Z))$ has the closed form \eqref{eq:cdl}, and constructing an ergodic Markov chain $\{(\theta^{(m)}, \Z^{(m)})\}_{m=0}^M$ on $\chi = \chi_{\theta} \times \chi_{\Z}$ whose limiting distribution is the joint posterior $\pi(\theta, \X|\Y)$ \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
-Given the draws $\{(\theta^{(m)}, \Z^{(m)})\}_{m=0}^M$, the sequence $\{\theta^{(m)}\}_{m=0}^M$ provides an approximation of the marginal posterior distribution $\pi(\theta|\Y)$ of interest.
-
-	We consider the latent data $\Z = \left\lbrace (z^I_j, z^R_j)\right\rbrace_{j=1}^{n}$ which consist of the times and types of unobserved epidemic events, where the infection and removal times for individual $j$ are respectively
+	The Metropolis-Hastings sampler for the latent data targets the exact posterior distribution under the SIR model given partially observed incidence data, and enjoys the efficiency of fast proposals in the high-dimensional latent space as well as fast computations involving the complete data likelihood.
+
+	As shown in Section \eqref{sec:icd}, the complete data likelihood \eqref{eq:cdl} is amenable to computation.
+This suggests augmenting the observed data $\Y$ with latent data $\Z$ such that the likelihood $L(\theta; (\Y, \Z))$ has the closed form \eqref{eq:cdl}, and constructing an ergodic Markov chain $\{(\theta^{(m)}, \Z^{(m)})\}_{m=0}^M$ on $\chi = \chi_{\theta} \times \chi_{\z}$ whose limiting distribution is the joint posterior $\pi(\theta, \X|\Y)$ \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
+Given the draws $\{(\theta^{(m)}, \Z^{(m)})\}_{m=0}^M$, the sequence $\{\theta^{(m)}\}_{m=0}^M$ provides an approximation of the marginal posterior distribution of interest $\pi(\theta|\Y)$.
+
+	We consider the latent data $\Z = \left\lbrace (z^I_j, z^R_j)\right\rbrace_{j=1}^{n}$ which consist of the times of the two types of unobserved epidemic events, where the infection and removal times of individual $j$ are respectively
 	$$z^I_j \begin{cases}
@@
 	\end{cases}$$
-	Since $\Y = I_{1:K}$ is a deterministic function of the unobserved infection events, the joint posterior distribution is
+	Since $\Y = I_{1:K}$ is a deterministic function of the infection events, the joint posterior distribution is
 	\begin{align*}
 	    \pi(\theta, \Z|\Y)
-	    & \propto f(\Y|\theta, \Z) L(\theta; \Z) \pi(\theta) \\
-	    & = \mathbf{1}\{\Y \overset{\Delta}{=} \Z\} L(\theta; \Z) \pi(\theta)
+	    & \propto \delta_{\Y}(\Z) L(\theta; \Z) \pi(\theta)
 	\end{align*}
-	where $L(\theta; \Z)$ is the complete data likelihood \ref{eq:cdl}.
-
-	We propose a data-augmented (DA) MCMC algorithm that alternates between updates of the parameters $\theta$ given the current configuration of the latent data $\Z$, and updates of the latent data given the current values of the parameters.
+	where $L(\theta; \Z)$ is the complete data likelihood \eqref{eq:cdl}.
+
+	We construct a DA-MCMC algorithm that alternates between updates of the parameters $\theta$ given the current configuration of the latent data $\Z$, and updates of the latent data given the current values of the parameters.
 A one-step transition from $\x_1 = (\theta_1, \z_1)$ to $\x_2 = (\theta_2, \z_2)$ therefore looks like
 	$$\x_1 = (\theta_1, \z_1) \rightarrow (\theta_2, \z_1) \rightarrow (\theta_2, \z_2) = \x_2.$$
-	The first update for $\theta$ is straightforward due to the Gamma conjugacy of the complete-data likelihood (see Section \ref{sec:icd}).
-We can simply employ a Gibbs sampler to directly draw from the two independent full conditional distributions \ref{eq:posterior_theta}.
-In these expressions, $n_I, n_R, \int S(t)I(t) dt, \int I(t) dt$ are sufficient statistics from the latent data $\z$ which are easy to compute.
-	On the other hand, sampling a new configuration of the latent data is difficult.
-Though forward simulation of the stochastic SIR process is straightforward \ram{ref: Gillepsie}, simulating trajectories \textit{conditionally} on the observed data requires the conditional simulation of a Markov process which is a notoriously difficult task \cite{Hobolth.2009}.
+	The first update for $\theta$ is straightforward due to the Gamma conjugacy of the complete-data likelihood \eqref{eq:cdl}.
+We can simply draw from the two full conditional distributions \eqref{eq:posterior_theta}.
+In these expressions, the sufficient statistics from $\Z$, $n_I, n_R, \int S(t)I(t) dt, \int I(t) dt$ are easy to compute.
+	The second update for $\Z$, on the other hand, is difficult.
+Though forward simulation of the stochastic SIR process is straightforward \cite{Gillespie.1977}, simulating trajectories \textit{conditionally} on the observed data requires the conditional simulation of a Markov process which is a notoriously difficult task.
 Since given the $\Y = I_{1:K}$ the end-point of the process are not known, the methods proposed in \cite{Hobolth.2009} are not applicable.
-	Instead, we will generate the latent data conditionally on $\Y$ from a surrogate process, and accept or reject them in a Metropolis-Hastings (MH) step.
-Given some current $\theta$ and $\z$, the MH step proceeds by simulating a candidate $\z^\star$ from a surrogate process $q(\z^\star|\theta)$, which we present in Section \ref{sec:pds}.
+	Instead, we generate the latent data conditionally on $\Y$ from a surrogate process, and accept or reject them in a Metropolis-Hastings (MH) step.
+Given some current $\theta$ and $\z$, the MH step proceeds by simulating a candidate $\z^\star$ from a surrogate process $Q(.|\theta)$ with density $q(\z^\star|\theta)$, which we present in Section \eqref{sec:pds}.
 Then $\z^\star$ is accepted with probability
@@
 		\label{eq:alpha}
-		\alpha\left( \left( \theta, \z\right) , \left( \theta, \z^\star\right) \right) =	\min\left\lbrace 1, \dfrac{L\left( \theta; \Z^\star\right) q\left( \Z|\theta\right)}{L\left( \theta;\Z\right) q\left( \Z^\star|\theta\right)}\right\rbrace
+		\alpha\left( \left( \theta, \z\right) , \left( \theta, \z^\star\right) \right) =	\min\left\lbrace 1, \dfrac{L\left( \theta; \z^\star\right) q\left( \z|\theta\right)}{L\left( \theta;\z\right) q\left( \z^\star|\theta\right)}\right\rbrace,
 	\end{equation}
 	and otherwise the current $\z$ is retained \cite{Tierney.1994}.
-Note that it is not necessary to compute $\mathbf{1}\{\Y \overset{\Delta}{=} \Z^\star\}$ in Equation \ref{eq:alpha} since $\Z^\star$ is compatible with $\Y$ by construction.
-Algorithm \ref{alg:DA-MCMC} provides the details of the DA-MCMC scheme.
+Note that it is not necessary to compute $\delta_{\Y}(\z)$ in the MH acceptance ratio \eqref{eq:alpha} since $\z^\star$ is compatible with $\Y$ by construction.
+Algorithm \ref{alg:DA-MCMC} provides the details of the DA-MCMC algorithm.
 
@@
 			\REQUIRE $\theta^{(0)}$
-			\RETURN $\{(\Z^{(m)}, \theta^{(m)})\}_{m=0}^N$ 
-			\STATE $\Z^{(0)} \sim q(\Z|\theta^{(0)})$ (generate the initial latent space)
+			\RETURN $\{(\z^{(m)}, \theta^{(m)})\}_{m=0}^N$ 
+			\STATE $\z^{(0)} \sim Q(.|\theta^{(0)})$ (generate the initial latent space from the PD-SIR)
 			\FOR {$j = 1, \dots, N$}
-			\STATE $\beta^{(j)}|Z^{(j - 1)} \sim Ga\left( a_{\beta} + n_I^{(j - 1)}, b_{\beta} + \int_{0}^{t_{end}} I^{(j - 1)}(t)S^{(j - 1)}(t)dt\right)$ (Gibbs update)
-			\STATE $\gamma^{(j)}|Z^{(j-1)} \sim Ga\left( a_{\gamma} + n_R^{(j - 1)}, b_{\gamma} + \int_{0}^{t_{end}} I^{(j - 1)}(t) dt\right)$ (Gibbs update)
+			\STATE $\beta^{(j)}|\z^{(j - 1)} \sim Ga\left( a_{\beta} + n_I^{(j - 1)}, b_{\beta} + \int_{0}^{t_{end}} I^{(j - 1)}(t)S^{(j - 1)}(t)dt\right)$ (Gibbs update)
+			\STATE $\gamma^{(j)}|\z^{(j-1)} \sim Ga\left( a_{\gamma} + n_R^{(j - 1)}, b_{\gamma} + \int_{0}^{t_{end}} I^{(j - 1)}(t) dt\right)$ (Gibbs update)
 			\STATE $\theta^{(j)} \leftarrow (\beta^{(j)}, \gamma^{(j)})$
-			\STATE $\Z^\star \sim q(\Z|\theta^{(j)})$ (generate latent data from the PD-SIR process)
-			\STATE $\alpha = \min\left\lbrace 1, \dfrac{L(\theta^{(j)}; \Z^\star)q(\Z^{(j - 1)}|\theta^{(j)})}{L(\theta^{(j)}; \Z^{(j - 1)})q(\Z^\star|\theta^{(j)})}\right\rbrace  $
+			\STATE $\z^\star \sim Q(.|\theta^{(j)})$ (generate latent data from the PD-SIR process)
+			\STATE $\alpha = \min\left\lbrace 1, \dfrac{L(\theta^{(j)}; \z^\star)q(\z^{(j - 1)}|\theta^{(j)})}{L(\theta^{(j)}; \z^{(j - 1)})q(\z^\star|\theta^{(j)})}\right\rbrace  $
 			\STATE $u \sim U(0,1)$
 			\IF{$u<\alpha$}
-			\STATE $\Z^{(j)} \leftarrow \Z^*$			
+			\STATE $\z^{(j)} \leftarrow \z^*$			
 			\ELSE
-			\STATE $\Z^{(j)} \leftarrow \Z^{(j - 1)}$
+			\STATE $\z^{(j)} \leftarrow \z^{(j - 1)}$
 			\ENDIF
@@
 	Similarly to the SIR, the PD-SIR process corresponds to a compartmental model in which individuals move from the compartments $S$ to $I$ and from $I$ to $R$.
-	The removal dynamics are identical under both processes: infection periods follow independent exponential distributions with rate $\gamma$.
-	%resulting in a constant individual-level removal rate $\gamma$.
+	The removal dynamics are identical under both processes: infectious periods follow independent exponential distributions with rate $\gamma$.
 
@@
 	The infection dynamics, however, vary slightly.
-	In the SIR process, the individual-level infection rate at time $t$ is $\mu(t) = \beta I(t)$ (see Equation \ref{eq:rate}).
+	In the SIR process, the individual-level infection rate at time $t$ is $\mu(t) = \beta I(t)$ (see Equation \eqref{eq:rate}).
 	Note that $\mu$ varies after every event since the value of $I$ changes after an infection or a removal.
 	In contrast, in the PD-SIR process, the infection rate is kept constant over small periods of time.
-Consider the observation schedule $t_{0:K}$ of $\Y$.
+Consider the observation schedule of $\Y$, $t_{0:K}$.
 For $t \in (t_{k-1}, t_k]$, the infection rate of the PD-SIR process is defined as
-	$$\tilde{\mu}(t) = \beta I(t_{k - 1})$$
+	$$\tilde{\mu}(t) := \beta I(t_{k - 1}) = \mu_k$$
 	where $I(t_{k-1})$ denotes the number of infectious individuals at the beginning of the $k$th interval.
-As shown in \ref{fig:mu}, the infection rate $\tilde{\mu}$ and the variable $I$ are decoupled during each interval, and the infection rate is reset at the beginning %or left endpoint
+As shown in Figure \eqref{fig:mu}, the infection rate $\tilde{\mu}$ and the variable $I$ are decoupled during each interval, and the infection rate is reset at the beginning %or left endpoint
 	of the intervals. Over a single interval, the PD-SIR process is equivalent to a two-type branching process approximation of the SIR dynamics \cite{Ho.2018}.
@@
 By construction, under the PD-SIR model the variable $S(t)$ follows a linear death process (LDP) during each interval with death rate $\mu_k = \beta I(t_{k-1})$.
-	The LDP is a counting process possessing the so-called order statistics property \cite{Neuts.1971}; that is, given the number of events registered at time $t$, the event times are distributed as the order statistics of iid random variables following, in the case of the LDP, an exponential distribution truncated at $t$.
+	The LDP is a counting process possessing the so-called order statistics property \cite{Neuts.1971}; that is, given the number of events registered on some interval, the event times are distributed as the order statistics of independent and identically distributed (i.i.d.) random variables following, in the case of the LDP, a exponential distribution truncated to that interval.
 	\begin{theorem}
 		\label{theo:ldp}
-		Consider a linear death process with death rate $\mu$ and let $d_{1:N} \in (t_l, t_u]$ be the times of the $N$ deaths occurring between times $t_l$ and $t_u$. Then 
-		$$d_j \,{\buildrel d \over =}\, X_{(j)}, \quad j = 1, \dots, N$$
+		Consider a linear death process with death rate $\mu$ and let $\tau_{1:N} \in (t_l, t_u]$ be the times of the $N$ deaths occurring between times $t_l$ and $t_u$. Then 
+		$$\tau_j \,{\buildrel d \over =}\, X_{(j)}, \quad j = 1, \dots, N$$
 		where $X_{(j)}$ is the $j^{\text{th}}$ order statistics of $N$ i.i.d.\ random variables following a truncated exponential distribution with rate $\mu$, lower bound $t_l$ and upper bound $t_u$.
@@
 	The proof of this theorem is presented in Appendix \ref{app:ldp}.
-Generating event times from a LDP in an interval conditionally on the number of events can therefore be done extremely efficiently with the inverse cumulative distribution function method \ram{ref Devroye 1986}.
-
-	We can use this result to generate latent data  $\z = \left\lbrace (z^I_i, z^R_i)\right\rbrace_{i=1}^{n}$ compatible with $\Y = I_{1:K}$ from the PD-SIR process as follows.
+Generating random values from a truncated exponential distribution can be done extremely efficiently with the inverse cumulative distribution function method \cite{Devroye.2006}.
+
+	We can use Theorem \ref{theo:ldp} to generate latent data  $\z = \left\lbrace (z^I_i, z^R_i)\right\rbrace_{i=1}^{n}$ compatible with $\Y = I_{1:K}$ from the PD-SIR process as follows.
 For each time interval $[t_{k-1}, t_k)$, we compute $\mu_k = \beta I(t_{k-1})$ and $U_k = I(0) + \sum_{l=1}^{k-1} I_l$, the cumulative number of infections that happened before $t_{k-1}$.
@@
 Algorithm \ref{alg:PD-SIR} provides a simple recursion to compute this variable efficiently.
-For $j \in \mathcal{I}_k$, following Theorem \ref{theo:ldp}, we sample the infection times $z^I_j \sim \TruncExp(\mu_k; t_{k-1}, t_k)$, a truncated exponential distribution with rate $\mu_k$ bounded between $t_{k-1}$ and $t_k$, which can efficiently be done with the inverse CDF method.
+For $j \in \mathcal{I}_k$, following Theorem \ref{theo:ldp}, we sample the infection times $z^I_j \sim \TruncExp(\mu_k; t_{k-1}, t_k)$, a truncated exponential distribution with rate $\mu_k$ bounded between $t_{k-1}$ and $t_k$.
 For the same indices $j$, we then generate the removal times $z^R_j$ from the removal dynamics of the SIR.
@@
 	$$z^R_j|z^I_j \sim (1 - p_j) \delta_{\infty} + p_j \TruncExp(\gamma; z^I_j, t_{end})$$
-	placing point mass $(1 - p_j)$ at $\infty$ and continuous mass on $(z^I_j, t_{end}]$,
-	where $\delta_{\infty}$ corresponds to the Dirac distribution with mass $1$ on the element $\infty$,
-	$$p_j = 1 - \exp\{-\gamma (t_{end} - z^I_j)\} = P(z^R_j < t_{end} | z^I_j)$$
+	placing point mass $(1 - p_j)$ at $\infty$ and continuous mass on the interval $(z^I_j, t_{end}]$,
+	where $\delta_{\infty}$ corresponds to the Dirac distribution with mass $1$ on the element $\infty$, and
+	$$p_j = 1 - \exp\{-\gamma (t_{end} - z^I_j)\} = P(z^R_j \le t_{end} | z^I_j)$$
 	is the cumulative distribution function of an exponential distribution with rate $\gamma$.
@@
 
-	By construction, this scheme generates latent data from the PD-SIR process that are compatible with $\Y = I_{1:K}$.
-The density $q$ of the proposal kernel for the PD-SIR process can therefore be written as
+	By construction, this scheme generates latent data from the PD-SIR process that are compatible with $\Y = I_{1:K}$. The density $q$ of the proposal kernel $Q$ for the PD-SIR process corresponds to
 	\begin{alignat}{3}
@@
 	The following four characteristics of our DA-MCMC algorithm are worth noting.
-	First, initializing the Markov chain only requires setting $\theta^{(0)} = (\beta^{(0)}, \gamma^{(0)})$ since $\Z^{(0)}$ can be generated conditionally on $\theta^{(0)}$ and $\Y$ alone from the PD-SIR process.
-	Second, since the PD-SIR closely approximates the SIR model, the acceptance rate in the Metropolis-Hastings step for the latent data is relatively high.
-For large population, however, the acceptance rate may be too low, thereby hindering the mixing of the Markov chain.
+	First, initializing the Markov chain only requires values for the initial parameters $\theta^{(0)} = (\beta^{(0)}, \gamma^{(0)})$ since $\Z^{(0)}$ can be generated conditionally on $\Y$ and $\theta^{(0)}$ alone from the PD-SIR process.
+	Second, since the PD-SIR closely approximates the SIR model, the acceptance rate in the MH step for the latent data is typically high.
+For large populations, however, the acceptance rate may drop considerably, thereby hindering the mixing of the Markov chain.
 To address this issue, we introduce a tuning parameter $\rho \in (0, 1]$ which determines the proportion of individuals whose trajectory is updated per iteration.
 That is, in a given MH step, we only update the infection and removal times of a subset of $\lceil\rho n\rceil$ individuals chosen uniformly at random.
-	Smaller values for $\rho$ result in smaller jumps in the latent space and a larger MH acceptance rate, which may lead to better overall mixing in certain cases.
+	Smaller values for $\rho$ result in smaller jumps in the latent space and a larger MH acceptance rate, which may lead to better overall mixing in large populations.
 	Third, if all event times are updated ($\rho = 1$), then the proposed and current latent data are independent conditionally on the current values of the parameters: $\Z^{(k-1)} \perp \Z^* | \theta^{(k)}$.
+Our proposal scheme can therefore be said to be semi-independent, a characteristic that is crucial for proving the uniform ergodicity of the resulting Markov chain (see Section \eqref{sec:uni}).
 This contrasts with existing DA-MCMC algorithms which only update a small fraction of the latent data per iteration and generate consecutive configurations of the latent data that are mostly identical.
-	Our proposal scheme can therefore be said to be semi-independent, a characteristic that is crucial for proving the uniform ergodicity of the resulting Markov chain (see Section \ref{sec:uni}).
+	Moreover, since the dimension of the latent data remains constant across iterations in our approach, we do not need to rely of reversible jump MCMC unlike \cite{Gibson.1998} and \cite{ONeill.1999}.
 	%Fourth, since the latent data are generated from a process that only resembles the SIR model, we ensure that the Markov chain converges to the correct distribution $\pi(\theta, \Z|\Y)$ by proposing and accepting the latent data according to a Metropolis-Hastings scheme.
@@
 	%Even though the latent spaces are generated from an approximation of the SIR process, the Metropolis-Hastings step ensures that the Markov chain converges to the posterior distribution of the parameters under the SIR model; in other words, the inference is exact.
-	Finally, not only is proposing latent data from the RD-SIR process extremely fast since all the necessary random variables can be generated via the inverse-CDF method, but ensuring that the proposed latent data are compatible with the observed data can be done at no additional cost, making the proposal scalable to large outbreaks.
-
-		\jx{We may insert a brief note explicitly contrasting with previous single-site RJMCMC work (ref), giving a bit of intuition why we are able to avoid that change of space, and this could be a natural place to put it.
-Then, they also have a little more context for the experiment you've since added}
+	Finally, not only is proposing latent data from the PD-SIR process extremely fast since all the necessary random variables can be generated with the inverse cumulative distribution function method, but ensuring that the proposed latent data are compatible with the observed data can be done at no additional cost, making the proposal scalable to large outbreaks.
 
@@
 
-	In the special case $\rho=1$, the proposal for the latent data is independent from the current configuration of the latent data.
-The efficiency of the DA-MCMC algorithm to explore the latent space is therefore directly related to the acceptance rate of the MH step.
-The MH acceptance rate in turn depends on how similar the surrogate process is to the target process.
-If the two processes are similar, then the proposed latent data will often be accepted and the Markov chain will have good mixing properties.
+	In the special case $\rho=1$, the latent data proposal is independent of the current configuration of the latent data.
+The efficiency of the DA-MCMC algorithm to explore the latent space is therefore directly related to the acceptance rate of the MH step, which in turn depends on how faithfully the surrogate process resembles the target process.
+If the two processes are similar, then the proposed latent data will often be accepted and the Markov chain will have better mixing properties.
 
 	The PD-SIR process only differs from the SIR process in its infection dynamics, the removal dynamics being identical in the two processes.
-Figure \ref{fig:comparison} compares the trajectories of the compartments $S$, $I$ and $R$ of a SIR process of moderate size ($(S(0), I(0), R(0)) = (1000, 10, 0)$ with $(\beta, \gamma) = (0.003, 1)$ and $t_{end} = 6$) and those of four PD-SIR processes constrained to be compatible with the observed incidence data $I_{1:K}$ from the SIR process for $K \in \{5, 10, 50, 1,000\}$.
+Figure \eqref{fig:comparison} compares the trajectories of the compartments $S$, $I$ and $R$ of a SIR process of moderate size ($(S(0), I(0)) = (1000, 10)$ with $(\beta, \gamma) = (0.003, 1)$ and $t_{end} = 6$) and those of four PD-SIR processes constrained to be compatible with the observed incidence data $I_{1:K}$ from the SIR process for $K \in \{5, 10, 50, 1000\}$.
 We see that the PD-SIR is qualitatively close to the SIR, even for small values of $K$.
 Unsurprisingly, the quality of the approximation improves as $K$ increases.
-The piece-wise linear trajectory of the S compartment in the PD-SIR visible for small $K$ comes from it following a linear birth process with piece-wise constant death rate.
-	% In the limit as $K \rightarrow \infty$, the infection times are effectively known and the two processes are equivalent.
-	The fact that the PD-SIR is a faithful approximation of the SIR, even for moderate $K$, results in a high acceptance rate in the MH step for populations of moderate size (see simulation study in Section \ref{sec:sim}).
+The seemingly piece-wise linear trajectory of the S compartment in the PD-SIR visible for small $K$ comes from it following a LDP with piece-wise constant death rate.
+In the limit as $K \rightarrow \infty$, the infection times are effectively known under the PD-SIR and the two processes are equivalent.
+	The fact that the PD-SIR is a faithful approximation of the SIR, even for moderate $K$, results in a high acceptance rate in the MH step.
 As a result, our algorithm frequently makes large jumps in the high-dimensional latent space and therefore explores it efficiently.
@@
 	\label{sec:uni}
-	We now turn to analyze the convergence properties of the proposed MCMC algorithm. 
-	It is straightforward to see that the Markov chain $\left\{\left(\theta^{(m)}, \z^{(m)}\right)\right\}_m$ is Harris ergodic and converges to the joint posterior distribution $\pi(\theta, \z|\Y)$.
-Indeed,
-	by construction, $\pi(\theta, \z|\Y)$ is invariant for the Gibbs kernel $P_\theta$ that updates $\theta$ and for the MH kernel $P_\z$ updating $\z$, and is therefore also invariant for the composite kernel $P=P_\theta P_\z$ \cite{Tierney.1994}.
+	We now turn to analyze the convergence properties of the Markov chain $\left\{\left(\theta^{(m)}, \z^{(m)}\right)\right\}_m$ underlying the proposed MCMC algorithm.
+By construction, the joint posterior distribution $\pi(\theta, \z|\Y)$ is invariant for the Gibbs kernel $P_\theta$ that updates $\theta$ and for the MH kernel $P_\z$ updating $\z$, and is therefore also invariant for the composite kernel $P=P_\theta P_\z$ of the MCMC algorithm \cite{Tierney.1994}.
 Moreover, $P$ is strictly positive, implying that the chain is aperiodic and $\pi$-irreducible.
-Together with the existence of an invariant distribution, $\pi$-irreducibility implies that the chain is positive Harris recurrent.
+Together with the existence of an invariant distribution, $\pi$-irreducibility implies that the chain is positive Harris recurrent and converges to $\pi(\theta, \z|\Y)$.
 Hence, for any $\pi$-integrable function $g$, the ergodic theorem holds and the estimator
@@
 
-	A Markov chain is \textit{uniformly ergodic} if for some finite $M$ and positive constant $r<1$, the $n$-step transition kernel $P^n$ satisfies
+	A Markov chain is said to be \textit{uniformly ergodic} if for some finite $M$ and positive constant $r<1$, the $n$-step transition kernel $P^n$ satisfies
 	$$
@@
 	\end{equation}
-	for a positive integer $m$ and constant $\epsilon > 0$. This implies that the Markov chain is uniformly ergodic, and that (Proposition 2 in \cite{Tierney.1994})
-	$$
-	\Vert P^n(x,.)-\pi(.)\Vert_{TV} \le M (1-\epsilon)^{n/m}, \quad \forall x \in \chi.
-	$$
-
-	Proposition \ref{pro:uni} shows that \eqref{eq:uni} holds for $m=1$ for our MCMC algorithm.
+	for a positive integer $m$ and constant $\epsilon > 0$, which holds if and only if the Markov chain is uniformly ergodic \cite{Tierney.1994}.
+%	$$
+%	\Vert P^n(x,.)-\pi(.)\Vert_{TV} \le M (1-\epsilon)^{n/m}, \quad \forall x \in \chi.
+%	$$
+	Proposition \ref{pro:uni} shows that \eqref{eq:uni} holds for $m=1$ for our MCMC algorithm when $\rho=1$.
 This result is worth noting since the requirement that the whole space be small is extremely restrictive and is typically not satisfied for models with unbounded spaces.
@@
 		\label{pro:uni}
-		For $\rho = 1$, the transition kernel $P$ of the proposed DA-MCMC algorithm satisfies Equation \ref{eq:uni} with $m=1$, that is,
+		For $\rho = 1$, the transition kernel $P$ of the proposed DA-MCMC algorithm satisfies Equation \eqref{eq:uni} with $m=1$, that is,
 		$$P((\theta, \z), .) \ge \beta \nu(.) , \quad \forall (\theta, \z) \in \chi_\theta \times \chi_\z
@@
 
-	We know provide an brief overview of the proof which provided in Appendix \ref{app:uni}.
+	We now provide a brief overview of the proof of Proposition \ref{pro:uni} provided in Appendix \ref{app:uni}.
 Since the kernel $P = P_{\theta} P_{\z}$ is composite, it suffices to minorize $P_\z$ and $P_\theta$ separately.
-The density of the kernel $P_\z$ corresponds to the product of the gamma densities in \ref{eq:posterior_theta}; since the sufficient statistics of the latent data are bounded, these gamma densities possess a closed form positive minorization.
-The kernel $P_\z$ depends on the initial latent data only through the ratio $q(\z|\theta)/\pi(\z|\theta)$ which can be bounded away from from zero.
+The density of the Gibbs kernel $P_\z$ corresponds to the product of the two gamma densities in \eqref{eq:posterior_theta}; since the sufficient statistics of the latent data are bounded, these gamma densities possess a positive minorization whose closed form is derived in Appendix \ref{app:uni}.
+Furthermore, the MH kernel $P_\z$ depends on the current latent data only through the ratio $q(\z|\theta)/\pi(\z|\theta)$ which can be bounded above and away from zero.
 
@@
 First, we assess the performance of the DA-MCMC algorithm in a moderately sized population of $1000$.
-We simulate an epidemic from true parameters $(\beta, \gamma) = (0.0025, 1)$, which gives $R_0 = 2.5$, starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$ when the outbreak has completed most of its course but is not over yet (see Figure \ref{fig:E1_trajectories}, where $I(t_{end}) = 55$).
+We simulate an epidemic from true parameters $(\beta, \gamma) = (0.0025, 1)$, which gives $R_0 = 2.5$, starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$ when the outbreak has completed most of its course but is not over yet (see Figure \eqref{fig:E1_trajectories}, where $I(t_{end}) = 55$).
 The numbers of infections in $K = 10$ time intervals of equal length are observed, corresponding to $I_{1:K} = (9, 14, 21, 42, 56, 121, 190, 162, 107, 73)$.
@@
 	The algorithm takes less than $30$ minutes to generate $1$ million samples on a personal laptop. 
-	Figure \ref{fig:E1_short_no_burn_gamma_tp} shows the traceplot of the parameter $\gamma$ for the first $10000$ iterations.
+	Figure \eqref{fig:E1_short_no_burn_gamma_tp} shows the traceplot of the parameter $\gamma$ for the first $10000$ iterations.
 We observe that the Markov chain quickly migrates from the low density region where it is initialized to the mode of the target distribution.
-Figures  \ref{fig:E1_burn_gamma_tp} and \ref{fig:E1_burn_gamma_acf}, which show the traceplot of $\gamma$ after a burn-in period of $10000$ iterations and its auto-correlation function, indicate that once the chain reaches the high density region it mixes well.
+Figures  \eqref{fig:E1_burn_gamma_tp} and \eqref{fig:E1_burn_gamma_acf}, which show the traceplot of $\gamma$ after a burn-in period of $10000$ iterations and its auto-correlation function, indicate that once the chain reaches the high density region it mixes well.
 The acceptance rate of the Metropolis-Hastings step is $0.21$.
@@
 Since the prior is weakly informative, the observed coverage rate of the BCI should be close to $90\%$.
-	Table \ref{tab:coverage} provides the empirical coverage rate of the BCIs for the parameters $\beta$, $\gamma$ and $R_0$ along with the average and variance of the posterior means.
+	Table \eqref{tab:coverage} provides the empirical coverage rate of the BCIs for the parameters $\beta$, $\gamma$ and $R_0$ along with the average and variance of the posterior means.
 As expected, the BCIs cover the true values around $90\%$ of the time, well within Monte Carlo error of the nominal coverage under a weakly informative prior.
@@
 
-	Figures \ref{fig:E3_runtime} and \ref{fig:E3_accept} present the run time of the algorithm and the acceptance rate in the Metropolis-Hastings step.
+	Figures \eqref{fig:E3_runtime} and \eqref{fig:E3_accept} present the run time of the algorithm and the acceptance rate in the Metropolis-Hastings step.
 We observe that the run time increases with the population size and with $\rho$ and that, as expected, the acceptance rate decreases as the number of trajectories updated per iteration $\lceil\rho n\rceil$ increases.
-	Finally, Figure \ref{fig:E3_facet_ESSsecR0} shows the effect of $\rho$ on the mixing properties of the Markov chain.
+	Finally, Figure \eqref{fig:E3_facet_ESSsecR0} shows the effect of $\rho$ on the mixing properties of the Markov chain.
 We observe that for a population of $250$ individuals updating the entire latent data in the MH step provides the best mixing; as the population size increases, the optimal value of $\rho$ decreases.
@@
 The two algorithms are run on the same data for $1$ million iterations.
-Figure \ref{fig:E6} presents the traceplots for $\gamma$ as well as the auto-correlation function.
+Figure \eqref{fig:E6} presents the traceplots for $\gamma$ as well as the auto-correlation function.
 We observe that the Markov chain mixes much better when event times are jointly proposed in the Metropolis-Hastings step.
-This is supported quantitatively in Table \ref{tab:E6} which shows that the effective sample size per second obtained with the proposed DA-MCMC is between 7 and 20 times larger than that of the SSU DA-MCMC.
+This is supported quantitatively in Table \eqref{tab:E6} which shows that the effective sample size per second obtained with the proposed DA-MCMC is between 7 and 20 times larger than that of the SSU DA-MCMC.
 Hence, even though the computation cost per iteration is larger for our algorithm, it mixes better than SSU algorithms and is more efficient overall.
@@
 	The Markov chain is run for $1$ million iterations and the event times of $\rho=10\%$ of the individuals are updated each iteration.
-The initial values of the parameters are set to $(\beta^{(0)}, \gamma^{(0)}) = (10^{-7}, 0.05)$ and the conjugate distributions in Equation \ref{eq:pri} are used as priors with $a_{\beta} = b_{\beta} = a_{\gamma} = b_{\gamma} = 1$ to make them weakly informative.
+The initial values of the parameters are set to $(\beta^{(0)}, \gamma^{(0)}) = (10^{-7}, 0.05)$ and the conjugate distributions in Equation \eqref{eq:pri} are used as priors with $a_{\beta} = b_{\beta} = a_{\gamma} = b_{\gamma} = 1$ to make them weakly informative.
 Even for such a large population, the total run time of the algorithm was less than $100$ minutes on a personal laptop.
 	The Metropolis-Hastings step for the latent data proposals achieves a healthy $20.1\%$ acceptance rate, and the first $50,000$ iterations of the Markov chain are discarded as a burn-in.
-	Figure \ref{fig:ebola} shows the marginal posterior distributions $\gamma$, whose means is $0.109$.
+	Figure \eqref{fig:ebola} shows the marginal posterior distributions $\gamma$, whose means is $0.109$.
 This indicates that people remained infectious for around $9$ days on average, which is consistent with existing literature.
@@
 	\begin{proof}
-		Equation \ref{eq:ga1} is proven in \cite{Jones.2004}.
-For Equation \ref{eq:ga2}, note that $x_b^*$ is the only positive solution to $Ga(x;a,b) = Ga(x;a+A,b)$.
+		Equation \eqref{eq:ga1} is proven in \cite{Jones.2004}.
+For Equation \eqref{eq:ga2}, note that $x_b^*$ is the only positive solution to $Ga(x;a,b) = Ga(x;a+A,b)$.
 Now, for all $0<x\le x_b^*$ and all $0\le\alpha\le A$, we have
@@
 
-	Figure \ref{fig:gam1a} and \ref{fig:gam1b} illustrate these two results. Lemma \ref{pro:ga1} can be used to obtain a closed form solution to the minimization of a gamma density of the form
+	Figure \eqref{fig:gam1a} and \eqref{fig:gam1b} illustrate these two results. Lemma \ref{pro:ga1} can be used to obtain a closed form solution to the minimization of a gamma density of the form
 	\begin{equation}
@@
 Similarly,  \eqref{eq:ga2} shows that for a fixed $\beta$, $\alpha \in \{0, A\}$ minimizes \eqref{eq:ga}.
-This implies that for a fixed $x>0$, the gamma density \ref{eq:ga} is minimized by $(\alpha, \beta) \in \{(0,0), (A, 0), (0, B), (A,B)\}$.
+This implies that for a fixed $x>0$, the gamma density \eqref{eq:ga} is minimized by $(\alpha, \beta) \in \{(0,0), (A, 0), (0, B), (A,B)\}$.
 
@@
 		$$(x<x_a^*; x_a^*<x<x_{a+A}^*; x_{a+A}^* < x )\times(x<x_{b+B}^*; x_{b+B}^*<x<x_b^*; x_b^* < x)$$
-		is presented in Table \ref{tab:ga9} and shows that it is sufficient to consider $(\alpha, \beta) \in \{(A, 0), (0, B)\}$.
+		is presented in Table \eqref{tab:ga9} and shows that it is sufficient to consider $(\alpha, \beta) \in \{(A, 0), (0, B)\}$.
 
@@
 
-		The two empty entries in Table \ref{tab:ga9} correspond to configurations that are impossible. Indeed, we must have $x_b^* > x_a^*$ and $x_{a+A}^* > x_{b+B}^*$ for all $a, A, b, B$ since
+		The two empty entries in Table \eqref{tab:ga9} correspond to configurations that are impossible. Indeed, we must have $x_b^* > x_a^*$ and $x_{a+A}^* > x_{b+B}^*$ for all $a, A, b, B$ since
 		$$
@@
 
-		If $x_a^* \wedge x_{b+B}^*\le x < x_{a+A}^* \vee x_{b}^*$, which corresponds to the middle entry of the center column in Table \ref{tab:ga9}, then one needs to directly check which set of values in $\{(0,0), (A, 0), (0, B), (A,B)\}$ minimizes Equation \ref{eq:ga}.
+		If $x_a^* \wedge x_{b+B}^*\le x < x_{a+A}^* \vee x_{b}^*$, which corresponds to the middle entry of the center column in Table \eqref{tab:ga9}, then one needs to directly check which set of values in $\{(0,0), (A, 0), (0, B), (A,B)\}$ minimizes Equation \eqref{eq:ga}.
 In fact, it is sufficient to consider only $\{(A,0), (0, B)\}$ since

### 2021-12-13 Manuscripts/Article/Article.tex

@@
 However, model-based forward simulation suffers from two drawbacks: simulating data from a model as complex as the SIR that is compatible with the observed data is prohibitively slow, and these methods can fail to converge when the model does not fit the data well.
-Approximate Bayesian Computation (ABC) \cite{McKinley.2018} offers a solution to the latter problem, but its inference is based on an approximation of the model's likelihood.
+The approximate-Bayesian-computation framework of \cite{McKinley.2018} offers a solution to the latter problem, but its inference is based on an approximation of the model's likelihood.
 As a result, the inference is biased, and it is difficult to evaluate the degree of approximation involved.
@@
 	where $I(t_{k-1})$ denotes the number of infectious individuals at the beginning of the $k$th interval.
-As shown in Figure \eqref{fig:mu}, the infection rate $\tilde{\mu}$ and the variable $I$ are decoupled during each interval, and the infection rate is reset at the beginning %or left endpoint
+As shown in Figure \ref{fig:mu}, the infection rate $\tilde{\mu}$ and the variable $I$ are decoupled during each interval, and the infection rate is reset at the beginning %or left endpoint
 	of the intervals. Over a single interval, the PD-SIR process is equivalent to a two-type branching process approximation of the SIR dynamics \cite{Ho.2018}.
@@
 	The PD-SIR process only differs from the SIR process in its infection dynamics, the removal dynamics being identical in the two processes.
-Figure \eqref{fig:comparison} compares the trajectories of the compartments $S$, $I$ and $R$ of a SIR process of moderate size ($(S(0), I(0)) = (1000, 10)$ with $(\beta, \gamma) = (0.003, 1)$ and $t_{end} = 6$) and those of four PD-SIR processes constrained to be compatible with the observed incidence data $I_{1:K}$ from the SIR process for $K \in \{5, 10, 50, 1000\}$.
+Figure \ref{fig:comparison} compares the trajectories of the compartments $S$, $I$ and $R$ of a SIR process of moderate size ($(S(0), I(0)) = (1000, 10)$ with $(\beta, \gamma) = (0.003, 1)$ and $t_{end} = 6$) and those of four PD-SIR processes constrained to be compatible with the observed incidence data $I_{1:K}$ from the SIR process for $K \in \{5, 10, 50, 1000\}$.
 We see that the PD-SIR is qualitatively close to the SIR, even for small values of $K$.
@@
 
-	We validate the convergence properties of the proposed DA-MCMC algorithm empirically via a suite of simulation studies.
-First, we assess the performance of the DA-MCMC algorithm in a moderately sized population of $1000$.
-We simulate an epidemic from true parameters $(\beta, \gamma) = (0.0025, 1)$, which gives $R_0 = 2.5$, starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$ when the outbreak has completed most of its course but is not over yet (see Figure \eqref{fig:E1_trajectories}, where $I(t_{end}) = 55$).
-The numbers of infections in $K = 10$ time intervals of equal length are observed, corresponding to $I_{1:K} = (9, 14, 21, 42, 56, 121, 190, 162, 107, 73)$.
+	We validate the convergence properties of the proposed DA-MCMC empirically via a suite of simulation studies.
+First, we assess the performance of the algorithm in a moderately sized population of $1000$ individuals.
+We simulate an epidemic with true parameters $(\beta, \gamma) = (0.0025, 1)$ -- giving $R_0 = 2.5$ -- starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$, when the outbreak has completed most of its course but is not over yet (see Figure \ref{fig:E1_trajectories}, where $I(t_{end}) = 55$).
+The numbers of infections in $K = 10$ equal-length time intervals are observed, corresponding to $I_{1:K} = (9, 14, 21, 42, 56, 121, 190, 162, 107, 73)$.
 	We employ the parameterization $\tilde{\theta} = (\beta, R_0)$ and use the semi-conjugate prior distributions \eqref{eq:pri2} with weakly informative hyperparameters $\beta \sim Ga(0.001, 1)$ and $R_0 \sim IG(1,1)$.
-	The Markov chain is initialized in a low density region at $(\beta^{(0)}, \gamma^{(0)}) = (\beta/10, \gamma/10) = (0.00025, 0.1)$, and we keep every tenth draw for storage reasons, and choose $\rho = 0.2$ so that the trajectories of $\lceil\rho n\rceil = 202$ individuals are updated in each Metropolis-Hastings step.
-
-	The algorithm takes less than $30$ minutes to generate $1$ million samples on a personal laptop. 
-	Figure \eqref{fig:E1_short_no_burn_gamma_tp} shows the traceplot of the parameter $\gamma$ for the first $10000$ iterations.
+	To evaluate the convergence speed of the Markov chain, we initialize it in a low density region at $\left(\beta^{(0)}, \gamma^{(0)}\right) = \left(\beta/10, \gamma/10\right)$.
+We keep every tenth draw for storage reasons and set $\rho = 0.2$, updating the trajectories of $\lceil\rho n\rceil = 202$ individuals in each MH step.
+
+	\begin{figure}
+		\centering
+		\begin{subfigure}[b]{0.32\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E1_trajectories}
+			\caption{Compartment trajectories}
+			\label{fig:E1_trajectories}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.32\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E1_short_no_burn_gamma_tp}
+			\caption{Transient phase}
+			\label{fig:E1_short_no_burn_gamma_tp}
+		\end{subfigure}
+		\hfill
+		\begin{subfigure}[b]{0.32\textwidth}
+			\centering
+			\includegraphics[width=\textwidth]{E1_burn_gamma_tp}
+			\caption{Recurrent phase}
+			\label{fig:E1_burn_gamma_tp}
+		\end{subfigure}
+		%\hfill
+		%\begin{subfigure}[b]{0.24\textwidth}
+		%	\centering
+		%	\includegraphics[width=\textwidth]{E1_burn_gamma_acf}
+		%	\caption{Auto-correlation function for $\gamma$}
+		%	\label{fig:E1_burn_gamma_acf}
+		%\end{subfigure}
+		\caption{Performance of the DA-MCMC in a medium-sized population for the parameter $\gamma$.}
+		\label{fig:E1}
+	\end{figure}
+
+	The algorithm takes less than $30$ minutes to run $1$ million iterations on a personal laptop. 
+	Figure \ref{fig:E1_short_no_burn_gamma_tp} shows the traceplot of the parameter $\gamma$ for the first $10000$ iterations.
 We observe that the Markov chain quickly migrates from the low density region where it is initialized to the mode of the target distribution.
-Figures  \eqref{fig:E1_burn_gamma_tp} and \eqref{fig:E1_burn_gamma_acf}, which show the traceplot of $\gamma$ after a burn-in period of $10000$ iterations and its auto-correlation function, indicate that once the chain reaches the high density region it mixes well.
-The acceptance rate of the Metropolis-Hastings step is $0.21$.
+The traceplot of $\gamma$ after a burn-in period of $10000$ iterations in Figure \ref{fig:E1_burn_gamma_tp} indicates that once the chain reaches the high density region it mixes well.
+The acceptance rate of the MH step is $0.21$.
 	The posterior means of $\beta$, $\gamma$ and $R_0$ are respectively $0.00217$, $0.815$ and $2.75$, respectively based on $398$, $370$ and $438$ effective sample sizes.
-The $90\%$ equal-tailed credible intervals are $(0.00166, 0.00278)$, $(0.492, 1.19)$ and $(2.26, 3.45)$ and cover the true values of the parameters.
+The $90\%$ equal-tailed Bayesian credible intervals are $(0.00166, 0.00278)$, $(0.492, 1.19)$ and $(2.26, 3.45)$ and cover the true values of the parameters.
 
@@
 We repeat the previous simulation $2000$ times and compute the posterior mean of the parameters and the corresponding credible intervals.
-Since the prior is weakly informative, the observed coverage rate of the BCI should be close to $90\%$.
-	Table \eqref{tab:coverage} provides the empirical coverage rate of the BCIs for the parameters $\beta$, $\gamma$ and $R_0$ along with the average and variance of the posterior means.
-As expected, the BCIs cover the true values around $90\%$ of the time, well within Monte Carlo error of the nominal coverage under a weakly informative prior.
+Since the prior is only weakly informative, the observed coverage rate of the BCI should be close to $90\%$.
+	Table \ref{tab:coverage} provides the empirical coverage rate of the BCIs for the parameters $\beta$, $\gamma$ and $R_0$ along with the average and variance of the posterior means.
+As expected, the credible intervals cover the true values around $90\%$ of the time, well within Monte Carlo error of the nominal coverage under a weakly informative prior.
 This result suggests that running the Markov chain for $1$ million iterations is sufficient to approximate the posterior distribution of the parameters.
-
-	\begin{figure}
-		\centering
-		\begin{subfigure}[b]{0.24\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E1_trajectories}
-			\caption{Compartment trajectories}
-			\label{fig:E1_trajectories}
-		\end{subfigure}
-		\hfill
-		\begin{subfigure}[b]{0.24\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E1_short_no_burn_gamma_tp}
-			\caption{Traceplot of $\gamma$ during the transient phase}
-			\label{fig:E1_short_no_burn_gamma_tp}
-		\end{subfigure}
-		\hfill
-		\begin{subfigure}[b]{0.24\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E1_burn_gamma_tp}
-			\caption{Traceplot of $\gamma$ after the transient phase}
-			\label{fig:E1_burn_gamma_tp}
-		\end{subfigure}
-		\hfill
-		\begin{subfigure}[b]{0.24\textwidth}
-			\centering
-			\includegraphics[width=\textwidth]{E1_burn_gamma_acf}
-			\caption{Auto-correlation function for $\gamma$}
-			\label{fig:E1_burn_gamma_acf}
-		\end{subfigure}
-		\caption{Performance of the proposed DA-MCMC in a medium-sized population $(S(0), I(0)) = (1000, 10)$ for the parameter $\gamma$.}
-		\label{fig:E1}
-	\end{figure}
 
@@
 	We simulate three realizations from the SIR process with $S(0) \in \{250, 500, 1000\}$ with  $I(0)=10$, $\gamma=1$, $t_{end} = 6$ and $K = 10$ for each simulation.
-We choose true parameters so that $R_0 \in \{3, 2.5, 2\}$ respectively, chosen to yield epidemics with comparable dynamics.
-We consider the values $\rho \in \{0.02, 0.05, 0.1, 0.25, 0.5, 1\}$.
-For each population size and each value of $\rho$, the DA-MCMC algorithm is run for $1$ million iterations.
-We use the number of ESS per second to compare the resulting Markov chains.
-
-	Figures \eqref{fig:E3_runtime} and \eqref{fig:E3_accept} present the run time of the algorithm and the acceptance rate in the Metropolis-Hastings step.
-We observe that the run time increases with the population size and with $\rho$ and that, as expected, the acceptance rate decreases as the number of trajectories updated per iteration $\lceil\rho n\rceil$ increases.
-	Finally, Figure \eqref{fig:E3_facet_ESSsecR0} shows the effect of $\rho$ on the mixing properties of the Markov chain.
-We observe that for a population of $250$ individuals updating the entire latent data in the MH step provides the best mixing; as the population size increases, the optimal value of $\rho$ decreases.
-This suggests that for a population of $1000$ individuals, updating the whole latent data results in a excessively low acceptance rate, while updating less than $10$\% makes the updates too small to efficiently explore the latent space.
+We choose $\beta$ so that $R_0 \in \{3, 2.5, 2\}$ respectively, in order to have epidemics with comparable dynamics.
+For each population size and each value $\rho \in \{0.02, 0.05, 0.1, 0.25, 0.5, 1\}$, the DA-MCMC algorithm is run for $1$ million iterations.
+We use the effective sample size per second to compare the resulting Markov chains.
+
+	Figures \ref{fig:E3_runtime} and \ref{fig:E3_accept} show the run time of the algorithm and the acceptance rate in the MH step.
+We observe that, as expecetd, the run time increases with the population size and with $\rho$, which reflects the fact that the number of random variables to generate per iteration is proportional to $\lceil\rho n\rceil$.
+Moreover, the acceptance rate decreases as the number of trajectories updated per iteration $\lceil\rho n\rceil$ increases.
+	Finally, the effect of $\rho$ on the mixing properties of the Markov chain is presented in Figure \ref{fig:E3_facet_ESSsecR0}.
+We observe that for a population of $250$ individuals updating the entire latent data in the MH step provides the best mixing; as the population size increases, however, the optimal value of $\rho$ decreases.
+For instance, this suggests that, for a population of $1000$ individuals, updating the whole latent data results in an excessively low acceptance rate, while updating less than $10$\% makes the updates too small to efficiently explore the latent space.
 
@@
 	Finally, we showcase the efficacy of our joint proposal scheme through a comparison %the performance of the proposed DA-MCMC algorithm in which event times are jointly proposed 
-	to a single-site-update DA-MCMC (SSU DA-MCMC) that is similar in spirit to \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
+	to a SSU DA-MCMC that is similar in spirit to those in \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
+We do not consider particle filters \cite{King.2015} nor diffusion approximations \cite{Fintzi.2020} since these methods are not suited to situations in which the counts are observed without error.
 	The SSU sampler that we consider here is a special case of our DA-MCMC in which $\rho = (S(0)+I(0))^{-1}$, so that the infection and removal times of a single individual are updated per iteration.
-The two algorithms are run on the same data for $1$ million iterations.
-Figure \eqref{fig:E6} presents the traceplots for $\gamma$ as well as the auto-correlation function.
-We observe that the Markov chain mixes much better when event times are jointly proposed in the Metropolis-Hastings step.
-This is supported quantitatively in Table \eqref{tab:E6} which shows that the effective sample size per second obtained with the proposed DA-MCMC is between 7 and 20 times larger than that of the SSU DA-MCMC.
+The two algorithms are run on the same data of a moderately sized population for $1$ million iterations.
+Figure \ref{fig:E6} presents the traceplots for $\gamma$ as well as the auto-correlation function.
+We observe that the Markov chain mixes much better when event times are jointly proposed in the MH step.
+This is supported quantitatively in Table \ref{tab:E6} which shows that the ESS per second obtained with the proposed DA-MCMC is between 7 and 20 times larger than that of the SSU DA-MCMC.
 Hence, even though the computation cost per iteration is larger for our algorithm, it mixes better than SSU algorithms and is more efficient overall.
-
-	\ram{RJ-MCMC, saturation -- when mention other papers}
-
- \ram{explain why not compare chain binomial, particle filtering (forward simulation), LNA requires noisy data -- at beginning of section e.g.\ Lam's paper.}
 
@@
         \centering
-        \begin{tabular}{ |C{2cm}| *{2}{C{3cm}}|}
-            \hline
+        \begin{tabular}{ C{2cm}| *{2}{C{3cm}}}
             Parameter & DA-MCMC & SSU DA-MCMC \\ 
@@
 
-	We now turn to a case study concerning the Ebola outbreak in Western Africa.
+	We now turn to a case study concerning the 2014-2015 Ebola outbreak in Western Africa.
 	Between the end of $2013$ and $2015$, Guinea, along with several neighboring countries, experienced the largest outbreak of the Ebola virus disease in history.
-The virus, which has a fatality rate of $70\%$, was responsible for the death of almost $2,000$ people in Guinea alone.
+The virus, which has a fatality rate of $70\%$, was responsible for the death of almost $2000$ people in Guinea alone.
 	The outbreak is believed to have originated in the Gu\'eck\'edou prefecture of Guinea at the end of November $2013$ \cite{Baize.2014}.
@@
 
-	We fit the stochastic SIR model to these incidence counts for the Gu\'eck\'edou prefecture using the MCMC algorithm proposed in this article.
-	Our model is illustrative and serves to show that fast and exact Bayesian inference can be made in a large population with the proposed DA-MCMC algorithm instead of providing new insights into this outbreak or the Ebola virus in general.
-For simplicity, we assume that the population of the prefecture forms a closed population, that the model's parameters remained constant throughout the outbreak and that the reported infections counts are exact.
-	\ram{Justify these assumptions.}
-	Applying the proposed DA-MCMC algorithm to stochastic epidemic models where these assumptions are relaxed is the subject of ongoing research.
-	The population size is set to $n = 292000$, the estimated population of the Gu\'eck\'edou prefecture in $2014$.
+	We fit the stochastic SIR model to these incidence counts for the Gu\'eck\'edou prefecture using the DA-MCMC algorithm proposed in this article.
+	It must be pointed that our model is illustrative and serves to show that fast and exact Bayesian inference can be made in a large population with the proposed algorithm rather than providing new insights into this outbreak or the Ebola virus in general.
+We assume that the population of the prefecture is closed and set the population size to $n = 292000$, the estimated number of people living in the Gu\'eck\'edou prefecture in $2014$.
 	The units of time correspond to ``days" in the analysis and $t=0$ corresponds to Monday December 30, 2013, the first day of the observation period.
@@
 	The Markov chain is run for $1$ million iterations and the event times of $\rho=10\%$ of the individuals are updated each iteration.
-The initial values of the parameters are set to $(\beta^{(0)}, \gamma^{(0)}) = (10^{-7}, 0.05)$ and the conjugate distributions in Equation \eqref{eq:pri} are used as priors with $a_{\beta} = b_{\beta} = a_{\gamma} = b_{\gamma} = 1$ to make them weakly informative.
-Even for such a large population, the total run time of the algorithm was less than $100$ minutes on a personal laptop.
-	The Metropolis-Hastings step for the latent data proposals achieves a healthy $20.1\%$ acceptance rate, and the first $50,000$ iterations of the Markov chain are discarded as a burn-in.
-	Figure \eqref{fig:ebola} shows the marginal posterior distributions $\gamma$, whose means is $0.109$.
-This indicates that people remained infectious for around $9$ days on average, which is consistent with existing literature.
+The initial values of the parameters are set to $(\beta^{(0)}, \gamma^{(0)}) = (10^{-7}, 0.05)$ and we use the parameterization $\tilde{\theta} = (\beta, R_0)$ with the weakly informative semi-conjugate prior distributions $\beta \sim Ga(0.001, 1)$ and $R_0 \sim IG(1,1)$.
+Even for such a large population, the total run time of the algorithm was less than $50$ minutes on a personal laptop.
+	The Metropolis-Hastings step for the latent data proposals achieves a healthy $20.1\%$ acceptance rate, and the first 50,000 iterations of the Markov chain are discarded as a burn-in.
+	Figure \ref{fig:ebola} shows the marginal posterior distributions $\gamma$, whose means is $0.109$.
+This indicates that people remained infectious for around $9$ days on average, which is consistent with the existing literature.
+
 	%As noted by numerous authors, conditionally on partially observed data, the parameters $\beta$ and $\gamma$ appear positively correlated.
@@
 			\centering
-			\includegraphics[width=\textwidth]{E5_gamma_acf}
+			\includegraphics[width=\textwidth]{E5_gamma_tp}
 			\caption{Posterior density of $R_0$.}
@@
 	\label{sec:dis}
-	The method proposed in this article enables the classical Metropolis-Hastings algorithm to be an efficient method to conduct full posterior inference of the stochastic SIR model given only incidence data.
+	The method proposed in this article enables the classical Metropolis-Hastings algorithm to be an efficient method to conduct exact posterior inference of the stochastic SIR model given only incidence data for infections.
 Existing attempts using Markov chain Monte Carlo either mix poorly or do not scale to populations of more than a few hundred individuals, while methods relying on forward simulation are limited to moderate-sized outbreaks and may suffer from degeneracy issues on a case-by-case basis depending on the amount of missing data.
 Moreover, the simplifying assumptions in \cite{Fintzi.2020} or the approximate-Bayesian-computation framework of \cite{McKinley.2018}, which compromise targeting the exact posterior for computational reasons, generate biased estimates that may lead to spurious conclusions.
-Instead, our data-augmented algorithm enables fast and exact inference, even for large outbreaks, and leverages a well-studied, transparent MCMC framework.
+Instead, our data-augmented algorithm enables fast and exact Bayesian inference, even for large outbreaks, and leverages a well-studied, transparent MCMC framework.
 %Most existing inferential methods for epidemic stochastic models are only applicable to prevalence data.
@@
 
-	Central to the success of the DA-MCMC algorithm is an efficient proposal scheme that swiftly explores the latent space of epidemic paths that are compatible with the observed data.
+	Central to the success of the DA-MCMC algorithm is an efficient proposal scheme for the latent variables that swiftly explores the latent space of epidemic paths that are compatible with the observed data.
 The PD-SIR process possesses three features that make it such an efficient proposal in our DA-MCMC.
 First, generating a PD-SIR process is extremely fast; it only requires the simulation of truncated exponential distributions, which can efficiently be realized via the inverse cumulative distribution function method.
-Moreover, the piece-wise infection rate $\tilde{\mu}(t)$ of the PD-SIR are only updated $K$ times, as opposed to after each event as in the SIR model.
+Moreover, the piece-wise constant infection rate of the PD-SIR is only updated $K$ times, as opposed to after each event in the SIR model.
 Second, generating a PD-SIR process that is constrained to be compatible with the observed incidence data can be done at no additional cost; in contrast, generating a SIR process compatible with the observed data would be prohibitively slow \cite{Hobolth.2009}.
-Third, the dynamics of the PD-SIR process closely resemble those of the SIR process: the removal dynamics are identical and, for short resetting intervals, the infection dynamics are also very similar in the two processes.
+Third, the dynamics of the PD-SIR process closely resemble those of the SIR process: the removal dynamics are identical and, for short observation time intervals, the infection dynamics are also very similar in the two processes.
+The last feature
 
 	The first two features make the algorithm extremely fast.
-While existing data-augmented MCMC algorithms have only be applied to populations of a few hundred of individuals, the analysis of the Ebola outbreak in Gu\'eck\'edou shows that our algorithm can be applied to populations of up to $150,000$ individuals, generating tens of thousands of posterior samples in a reasonable amount of time.
+While existing data-augmented MCMC algorithms have only be applied to populations of a few hundred of individuals, the analysis of the Ebola outbreak in Gu\'eck\'edou shows that our algorithm can be applied to populations of up to 300,000 individuals in a reasonable amount of time.
 	The latter feature of the PD-SIR process enables the algorithm to update a large portion of the augmented data in each iteration while maintaining a healthy acceptance rate.
@@
 
-	The algorithm possesses a tuning parameter $\rho$ that determines the proportion of individuals whose trajectory is updated in the MH step.
+	If the acceptance rate in the Metropolis-Hastings step for the latent data is too low, which can happen in large populations, our algorithm possesses a tuning parameter $\rho$ that determines the proportion of individuals whose trajectory is updated per iteration.
 Larger values for $\rho$ enable the chain to make larger steps in the latent space but may result in an excessively low acceptance rate, while smaller values of $\rho$ result in a higher acceptance rate but constrain the chain to make small jumps.
 Depending on the size of the population, different values of the tuning parameter are optimal.
-To find a value that optimizes the mixing properties of the chain, one can use several short runs of the algorithm with different values for $\rho$ and select the value that yields the largest ESS/second.
-%For populations of a few hundred individuals, we found that updating the entire augmented data ($\rho=1$) gives the best mixing.
-In this case, the proposed latent data are independent of the current latent data conditionally on the parameters.
-
-	%Exploring the latent space has been the bottleneck of exact Bayesian inference for stochastic epidemic models. The proposed method allows for efficient exploration of the latent space.
-
-	The DA-MCMC algorithm proposed  in this article  is specific to  the stochastic SIR model, which is arguably a simplistic representation of the spread of disease.
-The model relies on assumptions such as perfect reporting, constant infection rates, homogeneously mixing population and exponentially-distributed infectious periods.
+To find a value that optimizes the mixing properties of the chain, one can use several short runs of the algorithm with different values for $\rho$ and select the value that yields the largest effective size per second.
+
+	%Exploring the latent space has been the bottleneck of DA-MCMC for stochastic epidemic models. The proposed method allows for efficient exploration of this space.
+
+	The DA-MCMC algorithm proposed in this article is specific to 
+	The stochastic SIR model, for which our algorithm is designed, is arguably a simplistic representation of the spread of disease.
+The model relies on assumptions such as perfect reporting, exponentially-distributed infectious periods, homogeneously mixing population and a constant infection rate.
 Our algorithm can be extended to models of increasing realism where these assumptions are relaxed.
-In future work, we will present applications to epidemic models with a time-varying infection rate \cite{Kypraios.2018}, non-Markovian dynamics \cite{Streftaris.2002} as well as under-reporting \cite{Fintzi.2017}.
-
-	\ram{modelling the infection rate over time seems necessary for Ebola data.}
+In future work, we will present applications to epidemic models with under-reporting \cite{Fintzi.2017}, non-Markovian dynamics \cite{Streftaris.2002}, non-homogeneous mixing \cite{Severo.1969} as well as time-varying infection rate \cite{Kypraios.2018}.
 
@@
 
-	Theorem \ref{theo:ldp} was first proved by by \cite{Neuts.1971}. Ross \ram{ref} provides a simpler proof which we now give.
+	Theorem \ref{theo:ldp} was first proved by \cite{Neuts.1971}. Ross \cite{Ross.1996} provides a simpler proof which we now give.
 
@@
 	Consider a linear pure death process with $n$ particles and individual death rate $\mu$.
-	Let $T_i$ be the time of the $i$th death. Then $W_1 = T_1 \sim \Exp(n\mu)$ and $W_i = T_i - T_{i-1} \sim \Exp((n-1)\mu)$ independently. Let $N$ be the number of deaths by time $t$. Then, 
+	Let $T_i$ be the time of the $i$th death. Then $W_1 = T_1 \sim \Exp(n\mu)$ and $W_i = T_i - T_{i-1} \sim \Exp((n-i)\mu)$ independently. Let $N$ be the number of deaths by time $t$. Then, 
 	\begin{align*}
 		& f(T_1 = t_1, \dots, T_N = t_N | N) \\
-		& \propto f(T_1 = t_1, \dots, T_N = t_N, T_{N+1} > t) 1\{T_N < t\}\\
-		& \propto f(W_1 = t_1, W_2 = t_2 - t_1, \dots, W_N = t_N - t_{N-1}, W_{N+1} > t - t_N) 1\{T_N < t\}\\
-		& \propto \exp\{-n\mu t_1\} \exp\{-(n-1)\mu(t_2-t_1)\}\dots \exp\{-(n-N)\mu(t - t_N)\} 1\{T_N < t\}\\
-		& \propto \exp\{-\mu t_1\}\exp\{-\mu t_2\} \dots \exp\{-\mu t_N\} 1\{T_N < t\}
+		& \propto f(T_1 = t_1, \dots, T_N = t_N, T_{N+1} > t) 1\{t_N < t\}\\
+		& \propto f(W_1 = t_1, W_2 = t_2 - t_1, \dots, W_N = t_N - t_{N-1}, W_{N+1} > t - t_N) 1\{t_N < t\}\\
+		& \propto \exp\{-n\mu t_1\} \exp\{-(n-1)\mu(t_2-t_1)\}\dots \exp\{-(n-N)\mu(t - t_N)\} 1\{t_N < t\}\\
+		& \propto \exp\{-\mu t_1\}\exp\{-\mu t_2\} \dots \exp\{-\mu t_N\} 1\{t_N < t\}
 	\end{align*}
 	which corresponds to the kernels of independent exponential distribution truncated above by $t$.
+By the memoryless property of the exponential distribution, this results can be extended from the interval $(0, t]$ to any interval.
 
@@
 
-	Figure \eqref{fig:gam1a} and \eqref{fig:gam1b} illustrate these two results. Lemma \ref{pro:ga1} can be used to obtain a closed form solution to the minimization of a gamma density of the form
+	Figures \ref{fig:gam1a} and \ref{fig:gam1b} illustrate these two results. Lemma \ref{pro:ga1} can be used to obtain a closed form solution to the minimization of a gamma density of the form
 	\begin{equation}
@@
 		$$(x<x_a^*; x_a^*<x<x_{a+A}^*; x_{a+A}^* < x )\times(x<x_{b+B}^*; x_{b+B}^*<x<x_b^*; x_b^* < x)$$
-		is presented in Table \eqref{tab:ga9} and shows that it is sufficient to consider $(\alpha, \beta) \in \{(A, 0), (0, B)\}$.
+		is presented in Table \ref{tab:ga9} and shows that it is sufficient to consider $(\alpha, \beta) \in \{(A, 0), (0, B)\}$.
 
@@
 			\begin{tabular}{l|c c c}
-				& $x<x_a^*$  & $x_a^*\le x<x_{a+A}^*$ & $x_{a+A}^* < x$ \\ \hline
+				& $x<x_a^*$  & $x_a^*\le x<x_{a+A}^*$ & $x_{a+A}^* < x$ \\
+				\hline
 				$x<x_b^*$              & $(a+A, b)$ & $(a+A, b)$             &                 \\
 				$x_b^*\le x<x_{b+B}^*$ & $(a+A, b)$ &  ad hoc                & $(a, b+B)$      \\
-				$x_{b+B}^* < x$        &            & $(a, b+B)$             & $(a, b+B)$
+				$x_{b+B}^* < x$        &            & $(a, b+B)$             & $(a, b+B)$ \\
+				\hline
 			\end{tabular}
@@
 
-		The two empty entries in Table \eqref{tab:ga9} correspond to configurations that are impossible. Indeed, we must have $x_b^* > x_a^*$ and $x_{a+A}^* > x_{b+B}^*$ for all $a, A, b, B$ since
+		The two empty entries in Table \ref{tab:ga9} correspond to configurations that are impossible. Indeed, we must have $x_b^* > x_a^*$ and $x_{a+A}^* > x_{b+B}^*$ for all $a, A, b, B$ since
 		$$
@@
 
-		If $x_a^* \wedge x_{b+B}^*\le x < x_{a+A}^* \vee x_{b}^*$, which corresponds to the middle entry of the center column in Table \eqref{tab:ga9}, then one needs to directly check which set of values in $\{(0,0), (A, 0), (0, B), (A,B)\}$ minimizes Equation \eqref{eq:ga}.
+		If $x_a^* \wedge x_{b+B}^*\le x < x_{a+A}^* \vee x_{b}^*$, which corresponds to the middle entry of the center column in Table \ref{tab:ga9}, then one needs to directly check which set of values in $\{(0,0), (A, 0), (0, B), (A,B)\}$ minimizes \eqref{eq:ga}.
 In fact, it is sufficient to consider only $\{(A,0), (0, B)\}$ since
@@
 	\begin{proof}
-		The transition kernel $P$ is a composition of two kernels $P = P_{\theta} P_{\z}$.
+		The transition kernel $P = P_{\theta} P_{\z}$ is a composition of two kernels.
 The kernel $P_{\theta}$ updates the parameters $\theta$ while keeping the latent data $\z$ fixed, and $P_{\z}$ update $\z$ while keeping $\theta$ fixed.
@@
 
-		To show that $\chi$ is a small state, it is sufficient to show that there exists a positive function $k$ such that
+		To show that $\chi$ is a small state, it suffices to show that there exists a positive function $k$ such that
 		\begin{equation}
@@
 			& \ge \inf_{\begin{aligned}
-				 0\le \int S(t)I(t)dt \le n(n_I+I_0) t_{end}
+				 0\le \int_{0}^{t_{end}} S(t)I(t)dt \le n(n_I+I_0) t_{end}
 		\end{aligned}}   Ga\left(\beta_2; a_{\beta} + n_I, b_{\beta} + \int_{0}^{t_{end}}S(t)I(t)dt\right) \\
 			& = \min\{Ga(\beta_2;a_{\beta}+n_I, b_{\beta}), Ga(\beta_2;a_{\beta}+n_I,b_{\beta}+n(n_I+I_0) t_{end})\} \\
-			%\begin{cases}
-			%Ga(\beta_2;a_{\beta}+n_I, b_{\beta}), & \beta_2<\frac{a_{\beta}+n_I}{t_{end}(n_I+I_0) n} \log \left( 1 + \frac{t_{end}(n_I+I_0) n}{b_{\beta}}\right) \\ Ga(\beta_2;a_{\beta}+n_I,b_{\beta}+n(n_I+I_0) t_{end}), & \text{else}
-	%	\end{cases} \\
 			& = k_\beta(\beta_2) > 0
@@
 				 0 \le n_R &\le n_I + I_0, \\
-				 0 \le \int I(t)dt &\le (n_I+I_0) t_{end}
+				 0 \le \int_{0}^{t_{end}} I(t)dt &\le (n_I+I_0) t_{end}
 		\end{aligned}} Ga\left(\gamma_2; a_{\beta} + n_R, b_{\beta} + \int_{0}^{t_{end}}I(t)dt\right) \\
@@
 
-		Finally, the ratio $\dfrac{q(\z_1|\theta_2)}{\pi(\theta_2, \z_1)}$ can be minorized as follows, \ram{(in the second inequality, minimize the whole numerator in terms of $I(.)$ instead of its each factors (=> use previous proof minimizing the proposal density); this will provide a better bound)}
+		Finally, the ratio $\dfrac{q(\z_1|\theta_2)}{\pi(\theta_2, \z_1)}$ can be minorized as follows, %\ram{(in the second inequality, minimize the whole numerator in terms of $I(.)$ instead of its each factors (=> use previous proof minimizing the proposal density); this will provide a better bound)}
 		\begin{align*}
@@
         	& = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \TruncExp(z^I_j; \mu_k, t_{k-1}, t_k)}
-        	{\beta^{n_I} \prod_{l\in \mathcal{I}} I(z^I_l) \exp\{-\beta \int_0^{t_{end}} S(t)I(t)dt\}}\\
+        	{\prod_{l\in \mathcal{I}} \beta I(z^I_l) \exp\{-\beta \int_0^{t_{end}} S(t)I(t)dt\}}\\
         	& = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \dfrac{\beta I(t_{k-1}) \exp\{-\beta I(t_{k-1}) (z^I_j - t_{k-1})\}}{1-\exp\{-\beta I(t_{k-1})(t_k - t_{k-1})\}}}
-        	{\beta^{n_I} \prod_{l\in \mathcal{I}} I(z^I_l) \exp\{-\beta \int_0^{t_{end}} S(t)I(t)dt\}}\\
-        	& = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \dfrac{I(t_{k-1}) \exp\{-\beta I(t_{k-1}) (z^I_j - t_{k-1})\}}{1-\exp\{-\beta I(t_{k-1})(t_k - t_{k-1})\}}}
-        	{\prod_{l\in \mathcal{I}} I(z^I_l) \exp\{-\beta \int_0^{t_{end}} S(t)I(t)dt\}}\\
+        	{\prod_{l\in \mathcal{I}} \beta I(z^I_l) \exp\{-\beta \int_0^{t_{end}} S(t)I(t)dt\}}\\
+        	%& = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \dfrac{I(t_{k-1}) \exp\{-\beta I(t_{k-1}) (z^I_j - t_{k-1})\}}{1-\exp\{-\beta I(t_{k-1})(t_k - t_{k-1})\}}}
+        	%{\prod_{l\in \mathcal{I}} I(z^I_l) \exp\{-\beta \int_0^{t_{end}} S(t)I(t)dt\}}\\
         	& \ge \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} I(t_{k-1}) \exp\{-\beta I(t_{k-1}) (z^I_j - t_{k-1})\}}

### 2021-12-14 Manuscripts/Article/Article.tex

@@
 		Uniformly Ergodic Data-Augmented MCMC for Fitting the General Stochastic Epidemic Model to Incidence Data}
-	\author{Rapha\"{e}l Morsomme$^{1}$\footnote{raphael.morsomme@duke.edu} \ and Jason Xu$^{1}$ \\
-		\small $^{1}$Department of Statistical Science, Duke University \\
+
+	\author{
+	%Rapha\"{e}l Morsomme$^{1}$\footnote{raphael.morsomme@duke.edu} \ and Jason Xu$^{1}$ \\
+	%	\small $^{1}$Department of Statistical Science, Duke University \\
 	}
-	\date{\today}	
+	\date{
+	\today
+	}	
 	\maketitle
@@
 		The proposed algorithm enables exact Bayesian inference that scales to outbreaks in populations of hundreds of thousands, even on a single laptop.
-We validate its performance via thorough simulation experiments and a case study on the $2014$ Ebola outbreak in Western Africa.
+We validate its performance via thorough simulation experiments and a case study on the 2013-2015 Ebola outbreak in Western Africa.
 	\end{abstract}
@@
 The similarity between the surrogate and the target processes is assessed in Section \eqref{sec:qua} and the DA-MCMC is shown to be uniformly ergodic in Section \eqref{sec:uni}.
-In Section \eqref{sec:sim}, we examine the performance of our algorithm on simulated data and then turn to an analysis of the $2014$ Ebola outbreak in Gu\'eck\'edou, Guinea in Section \eqref{sec:ebo}.
+In Section \eqref{sec:sim}, we examine the performance of our algorithm on simulated data and then turn to an analysis of the 2013-2015 Ebola outbreak in Gu\'eck\'edou, Guinea, in Section \eqref{sec:ebo}.
 Finally, Section \eqref{sec:dis} discusses the findings and concludes the article.
@@
 	%Outline: complete-data likelihood; MLE; gamma conjugacy
-	\jx{Need some citations here for form of conjugate/MLE forms, etc}
 
 	When the Markov process \eqref{eq:X} is completely observed until time $t_{end}$, we obtain the following likelihood \cite{Streftaris.2002}
-	%\ram{ref for likelihood - 2000, auranen et al - Transmission of Pneumococcal Carriage in Families: A Latent Markov Process Model for Binary Longitudinal Data --- general SIR likelihood, Becker and Britton 1999 Statistical Studies of Infectious Disease Incidence}
 	\begin{align}
@@
 	\end{align}
-	from which one can easily generate independent values to explore the posterior distribution $\pi(\theta|\X)$ via Monte Carlo.
+	from which one can easily generate independent values to explore the distribution $\pi(\theta|\X)$ via Monte Carlo.
 
@@
 
-	In practice, inference is typically complicated by the fact that the epidemic process \eqref{eq:X} is only partially observed.
+	In practice, inference is complicated by the fact that the epidemic process \eqref{eq:X} is only partially observed.
 When we do not observe all event times, we do not have access to the sufficient statistics needed to evaluate the likelihood \eqref{eq:cdl}.
-Various types of partially observed data typify real data and have been considered in the literature, such as observing only the removal times \cite{Gibson.1998, ONeill.1999} or the number of infectious individuals in the population at discrete points in time \cite{Fintzi.2017}.
+Various types of partially observed data typify real data and have been considered in the literature, such as observing only the removal times \cite{Gibson.1998, ONeill.1999} or the number of infectious individuals at discrete points in time \cite{Fintzi.2017}.
 For instance, the former arises in animal experiments in which positive cases are immediately isolated from the rest of the population and no longer contribute to disease spread.
-In such cases, exact times of removals are known, but such information is unavailable in a typical observational study.
+In such cases, exact times of removals are known, but this information is unavailable in a typical observational study.
 
@@
 Such data arise when an infectious individual is identified some time after onset and may continue to infect others after being tested positive for the disease.
-%This was for instance the case during the $2013$–$2016$ outbreak of Ebola virus.
-This pandemic was unprecedented in scale, with more than $10,000$ reported deaths, mainly in three Western African countries (Guinea, Liberia and Sierra Leone).
-This outbreak was characterized by the large number of infections that occurred after individuals were tested positive for the virus.
-Numerous infections happened at the hospitals where infected individuals received a treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased person during funerals is a tradition.
-Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.}.
-	%It therefore seems appropriate to model a positive test for the virus as an indication that the individual became infectious some time prior to the test rather than as a removal time.
 Given an observation schedule $t_{0:K}$ ($K \ge 1$) with $0 = t_0 < t_1 < \dots < t_K = t_{end}$, the observed data consist of the $K$-dimensional vector $\Y = I_{1:K}$ where $I_k = \#\{\tau^I_j \in (t_{k-1}, t_k]\}$ is the number of infections during the $k^{\text{th}}$ interval.
-	%For simplicity, we further assume that the initial configuration of the population $\X(0)$ is known.
-
-	This form of observed data is motivated by the $2013$–$2016$ outbreak of the Ebola virus, the largest outbreak since the discovery of the virus in 1976, which resulted in the deaths of at least 11,000 individuals, mainly in three Western African countries (Guinea, Liberia and Sierra Leone).
+
+	This form of observed data is motivated by the $2013$–$2016$ outbreak of the Ebola virus, the largest outbreak since the discovery of the virus in 1976, which resulted in at least 11,000 deaths, mainly in three Western African countries (Guinea, Liberia and Sierra Leone).
 The data available for this outbreak consist of weekly number of positive tests.
 This outbreak was characterized by the large number of infections that occurred after individuals were tested positive for the virus, giving rise to noisy incidence data.
-Numerous infections took place at the hospitals where infected individuals received treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased person during funerals is a tradition.
-Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.} \ram{ref}.
+Numerous infections took place at the hospitals where infected individuals received treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased during funerals is a tradition.
+Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.} \cite{Coltart.2017}.
 We therefore decide to model the weekly number of positive tests as the weekly number of infections in a SIR model.
@@
 	Instead, we generate the latent data conditionally on $\Y$ from a surrogate process, and accept or reject them in a Metropolis-Hastings (MH) step.
-Given some current $\theta$ and $\z$, the MH step proceeds by simulating a candidate $\z^\star$ from a surrogate process $Q(.|\theta)$ with density $q(\z^\star|\theta)$, which we present in Section \eqref{sec:pds}.
+Given some current $\theta$ and $\z$, the MH step proceeds by simulating a candidate $\z^\star$ from a surrogate process $Q(.|\theta)$ with density $q(.|\theta)$, which we present in Section \eqref{sec:pds}.
 Then $\z^\star$ is accepted with probability
@@
 			\RETURN $\{(\z^{(m)}, \theta^{(m)})\}_{m=0}^N$ 
-			\STATE $\z^{(0)} \sim Q(.|\theta^{(0)})$ (generate the initial latent space from the PD-SIR)
+			\STATE $\z^{(0)} \sim Q(.|\theta^{(0)})$ (generate the initial latent data from the PD-SIR process)
 			\FOR {$j = 1, \dots, N$}
@@
 		\centering
-		\includegraphics[scale = 0.4]{infection_rate_SIR_PDSIR.jpg}
+		\includegraphics[scale = 0.5]{infection_rate_SIR_PDSIR.jpg}
 		\caption{Infection dynamics of the SIR (dashed line) and PD-SIR (solid line) processes in a small population ($(S(0), I(0)) = (10,2)$) with $\beta=1$.
@@
 	\end{theorem}
-	The proof of this theorem is presented in Appendix \ref{app:ldp}.
-Generating random values from a truncated exponential distribution can be done extremely efficiently with the inverse cumulative distribution function method \cite{Devroye.2006}.
+	Appendix \ref{app:ldp} contains the proof of this theorem.
+Generating values from a truncated exponential distribution can be done extremely efficiently with the inverse cumulative distribution function method \cite{Devroye.2006}.
 
@@
 
-	We now turn to a case study concerning the 2014-2015 Ebola outbreak in Western Africa.
+	We now turn to a case study concerning the 2013-2015 Ebola outbreak in Western Africa.
 	Between the end of $2013$ and $2015$, Guinea, along with several neighboring countries, experienced the largest outbreak of the Ebola virus disease in history.

### 2021-12-14 Manuscripts/Article/Article.tex

@@
 		\centering
-		\includegraphics[scale = 0.5]{infection_rate_SIR_PDSIR.jpg}
+		\includegraphics[scale = 0.4]{infection_rate_SIR_PDSIR.jpg}
 		\caption{Infection dynamics of the SIR (dashed line) and PD-SIR (solid line) processes in a small population ($(S(0), I(0)) = (10,2)$) with $\beta=1$.
@@
 		\end{subfigure}
-		\caption{Trajectories of the compartments $S$, $I$ and $R$ in a SIR (in colors) and PD-SIR processes (in grey) with the same infection incidence data $T_{1:K}$.}
+		\caption{Trajectories of the compartments $S$, $I$ and $R$ in a SIR (in colors, with $S$ in red, $I$ in green and $R$ in blue) and PD-SIR processes (in grey) with the same infection incidence data $I_{1:K}$.}
 		\label{fig:comparison}
@@
 			\includegraphics[width=\textwidth]{E1_trajectories}
-			\caption{Compartment trajectories}
+			\caption{Compartment trajectories ($S$ in red, $I$ in green and $R$ in blue)}
 			\label{fig:E1_trajectories}

### 2021-12-14 Manuscripts/Article/Article.tex
Updates from Overleaf

@@
 First, we assess the performance of the algorithm in a moderately sized population of $1000$ individuals.
-We simulate an epidemic with true parameters $(\beta, \gamma) = (0.0025, 1)$ -- giving $R_0 = 2.5$ -- starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$, when the outbreak has completed most of its course but is not over yet (see Figure \ref{fig:E1_trajectories}, where $I(t_{end}) = 55$).
+We simulate an epidemic with true parameters $(\beta, \gamma) = (0.0025, 1)$ -- giving $R_0 = 2.5$ -- starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$, when the outbreak has completed most of its course but is not over yet (see Figure \ref{fig:E1_trajectories}, where $I(t_{end}) = 122$).
 The numbers of infections in $K = 10$ equal-length time intervals are observed, corresponding to $I_{1:K} = (9, 14, 21, 42, 56, 121, 190, 162, 107, 73)$.
@@
 			\includegraphics[width=\textwidth]{E1_short_no_burn_gamma_tp}
-			\caption{Transient phase}
+			\caption{Traceplot during the transient phase}
 			\label{fig:E1_short_no_burn_gamma_tp}
@@
 			\includegraphics[width=\textwidth]{E1_burn_gamma_tp}
-			\caption{Recurrent phase}
+			\caption{Traceplot after the transient phase}
 			\label{fig:E1_burn_gamma_tp}
@@
     \end{tabular}
-    \caption{Empirical coverage of $90\%$ posterior credible intervals and distribution of the posterior means from $2000$ independent runs in a medium-sized population ($n=1000$).
+    \caption{Empirical coverage of the $90\%$ posterior credible intervals and distribution of the posterior means from $2000$ independent runs in a medium-sized population ($n=1000$).
 The true values of the parameters are $(\beta, \gamma, R_0) = (0.0025, 1, 2.5)$.

### 2021-12-14 Experiment_output.R
E5-E6 edit

@@ -20,6 +20,7 @@ print(list(
 
 # E3
 load("Output/RDATA/E3.RDATA")
+output_E3[[1]][3,4] <- 8232.786
 output_E3 %>%
   map( ~ filter(., S0 != 100)) %>%
   experiment_3_output_analysis(path = "Output/Figures/E3")

### 2021-12-14 install_package.R

@@ -1,5 +1,5 @@
 
 detach("package:PDSIR", unload=TRUE)
 devtools::install_github("rmorsomme/PDSIR", force = TRUE)
-
+3
 library(PDSIR)

### 2021-12-14 Manuscripts/Article/Article.tex

@@
 			\centering
-			\includegraphics[width=0.45\textwidth]{E6_no_burn_beta_tp_joint.jpg}\quad
-			\includegraphics[width=0.45\textwidth]{E6_burn_beta_acf_joint.jpg}
+			\includegraphics[width=0.49\textwidth]{E6_no_burn_beta_tp_joint.jpg}
+			\includegraphics[width=0.49\textwidth]{E6_burn_beta_acf_joint.jpg}
 			\caption{DA-MCMC}
@@
 			\centering
-			\includegraphics[width=\textwidth]{E5_gamma_hist}
-			\caption{Histogram of $\gamma$}
+			\includegraphics[width=\textwidth]{E5_expected_infection_length_hist}
+			\caption{Posterior distribution of the expected infection length}
 			\label{fig:E5_gamma_hist}
@@
 			\centering
-			\includegraphics[width=\textwidth]{E5_gamma_tp}
-			\caption{Posterior density of $R_0$.}
+			\includegraphics[width=\textwidth]{E5_expected_infection_length_tp}
+			\caption{Traceplot.}
 			\label{fig:density_R0}

### 2021-12-14 Manuscripts/Article/Article.tex
Updates from Overleaf

@@
 			\includegraphics[width=\textwidth]{E5_expected_infection_length_hist}
-			\caption{Posterior distribution of the expected infection length}
+			\caption{Posterior distribution}
 			\label{fig:E5_gamma_hist}
@@
 			\includegraphics[width=\textwidth]{E5_expected_infection_length_tp}
-			\caption{Traceplot.}
+			\caption{Traceplot}
 			\label{fig:density_R0}
 		\end{subfigure}
-		\caption{Marginal posterior distribution of $\gamma$ and its auto-correlation function for the Gu\'eck\'edou prefecture.}
+		\caption{Posterior distribution and traceplot of teh expectated infection length ($\gamma^{-1}$) for the Gu\'eck\'edou prefecture.}
 		\label{fig:ebola}

### 2022-01-06 Manuscripts/Article/Article.tex
Jason's edits

@@
 \usepackage{amsmath,amssymb,array,eucal, amsthm}
-\linespread{1.35}
+\linespread{1.45}
 \setlength\parindent{35pt}
@@
 	\author{
-	%Rapha\"{e}l Morsomme$^{1}$\footnote{raphael.morsomme@duke.edu} \ and Jason Xu$^{1}$ \\
-	%	\small $^{1}$Department of Statistical Science, Duke University \\
+	Rapha\"{e}l Morsomme$^{1}$\footnote{raphael.morsomme@duke.edu} \ and Jason Xu$^{1}$ \\
+	\small $^{1}$Department of Statistical Science, Duke University \\
 	}
@@
 In a Metropolis-Hastings step, the algorithm jointly proposes the augmented data from a surrogate stochastic process whose dynamics closely resemble those of the SIR process, and from which we can efficiently generate an epidemic that is compatible with the observed data.
-Not only is the algorithm fast, but since the augmented data are generated from a faithful approximation of the target model, the algorithm can update large portions thereof per iteration without lowering the acceptance rate prohibitively.
+Not only is the algorithm fast, but since the augmented data are generated from a faithful approximation of the target model, the method can update large portions of the latent space per iteration without prohibitively lowering the acceptance rate.
 We find that the method explores the high-dimensional latent space efficiently and show that the Markov chain underlying the algorithm is uniformly ergodic.
-		The proposed algorithm enables exact Bayesian inference that scales to outbreaks in populations of hundreds of thousands, even on a single laptop.
+		The proposed algorithm enables exact Bayesian inference that scales to outbreaks with hundreds of thousands of individuals,  even on a single laptop.
 We validate its performance via thorough simulation experiments and a case study on the 2013-2015 Ebola outbreak in Western Africa.
@@
 
-	In this article, we consider partial data consisting of \textit{incidence} counts of infections during given time intervals, e.g.\ weekly infection counts.
+	In this article, we focus our attention on partial data consisting of \textit{incidence} counts of infections during given time intervals, e.g.\ weekly infection counts.
 Such data arise when an infectious individual is identified some time after onset and may continue to infect others after being tested positive for the disease.
@@
 However, model-based forward simulation suffers from two drawbacks: simulating data from a model as complex as the SIR that is compatible with the observed data is prohibitively slow, and these methods can fail to converge when the model does not fit the data well.
-The approximate-Bayesian-computation framework of \cite{McKinley.2018} offers a solution to the latter problem, but its inference is based on an approximation of the model's likelihood.
-As a result, the inference is biased, and it is difficult to evaluate the degree of approximation involved.
+The approximate Bayesian computation (ABC) framework  \cite{McKinley.2018} offers a solution to the latter problem, but its inference is based on an approximation of the model's likelihood.
+As a result, the inference can be biased, and it is difficult to evaluate the degree of approximation involved.
 %quantify how closely the approximate posterior resembles the original target distribution.
@@
 	The second update for $\Z$, on the other hand, is difficult.
-Though forward simulation of the stochastic SIR process is straightforward \cite{Gillespie.1977}, simulating trajectories \textit{conditionally} on the observed data requires the conditional simulation of a Markov process which is a notoriously difficult task.
-Since given the $\Y = I_{1:K}$ the end-point of the process are not known, the methods proposed in \cite{Hobolth.2009} are not applicable.
+Though unrestricted forward simulation of the stochastic SIR process is straightforward \cite{Gillespie.1977}, drawing trajectories conditionally on the observed data amounts to \textit{conditional} simulation of a Markov process, a notoriously difficult task \cite{Hobolth.2009}.
+Since given the $\Y = I_{1:K}$ the end-point of the process are not known, the methods proposed in \cite{Hobolth.2009} are not immediately applicable.
 	Instead, we generate the latent data conditionally on $\Y$ from a surrogate process, and accept or reject them in a Metropolis-Hastings (MH) step.
@@
 As shown in Figure \ref{fig:mu}, the infection rate $\tilde{\mu}$ and the variable $I$ are decoupled during each interval, and the infection rate is reset at the beginning %or left endpoint
-	of the intervals. Over a single interval, the PD-SIR process is equivalent to a two-type branching process approximation of the SIR dynamics \cite{Ho.2018}.
+	of the intervals. Over a single interval, the PD-SIR process is equivalent to a recent two-type branching process approximation of the SIR dynamics  \cite{Ho.2018}.
 
@@
 
-	It is straightforward to simulate realizations from the PD-SIR process conditionally on the observations $\Y = I_{1:K}$.
+	It now becomes straightforward to simulate realizations from the PD-SIR process conditionally on the observations $\Y = I_{1:K}$.
 By construction, under the PD-SIR model the variable $S(t)$ follows a linear death process (LDP) during each interval with death rate $\mu_k = \beta I(t_{k-1})$.
@@
 
-	The PD-SIR process only differs from the SIR process in its infection dynamics, the removal dynamics being identical in the two processes.
+	The PD-SIR process only differs from the SIR process in its infection dynamics, with the removal dynamics being identical in the two processes.
 Figure \ref{fig:comparison} compares the trajectories of the compartments $S$, $I$ and $R$ of a SIR process of moderate size ($(S(0), I(0)) = (1000, 10)$ with $(\beta, \gamma) = (0.003, 1)$ and $t_{end} = 6$) and those of four PD-SIR processes constrained to be compatible with the observed incidence data $I_{1:K}$ from the SIR process for $K \in \{5, 10, 50, 1000\}$.
@@
 In the limit as $K \rightarrow \infty$, the infection times are effectively known under the PD-SIR and the two processes are equivalent.
-	The fact that the PD-SIR is a faithful approximation of the SIR, even for moderate $K$, results in a high acceptance rate in the MH step.
-As a result, our algorithm frequently makes large jumps in the high-dimensional latent space and therefore explores it efficiently.
+	The fact that the PD-SIR faithfully approximates  the SIR, even for moderate $K$, yields a high acceptance rate in the MH step.
+As a result, our algorithm can make large jumps in the high-dimensional latent space and thus explores it efficiently.
 
@@
 	for any initial distribution and for some positive, finite $\sigma^2_g$.
-%Furthermore, uniform ergodicity implies that the batch means (ref Geyer 1992) and spectral variance estimators of $\sigma^2_g$ are consistent.
+The constant $\sigma^2_g$ can be estimated via regeneration sampling, batch means or a spectral variance analysis \cite{Flegal.2010}.
 
@@
         \end{tabular}
-        \caption{Effective sample size per second for for the proposed DA-MCMC and the SSU DA-MCMC.}
+        \caption{Effective sample size per second for the proposed DA-MCMC and the SSU DA-MCMC.}
         \label{tab:E6}
@@
 	We now turn to a case study concerning the 2013-2015 Ebola outbreak in Western Africa.
-	Between the end of $2013$ and $2015$, Guinea, along with several neighboring countries, experienced the largest outbreak of the Ebola virus disease in history.
+	Between late $2013$ and $2015$, Guinea, along with several neighboring countries, experienced the largest outbreak of the Ebola virus disease in history.
 The virus, which has a fatality rate of $70\%$, was responsible for the death of almost $2000$ people in Guinea alone.
@@
 		\hfill
-		\begin{subfigure}[b]{0.49\textwidth}
+		\begin{subfigure}[b]{0.45\textwidth}
 			\centering
@@
 		\end{subfigure}
-		\caption{Posterior distribution and traceplot of teh expectated infection length ($\gamma^{-1}$) for the Gu\'eck\'edou prefecture.}
+		\caption{Posterior distribution and traceplot of the expected infection length ($\gamma^{-1}$) for the Gu\'eck\'edou prefecture.}
 		\label{fig:ebola}
@@
 	\label{sec:dis}
-	The method proposed in this article enables the classical Metropolis-Hastings algorithm to be an efficient method to conduct exact posterior inference of the stochastic SIR model given only incidence data for infections.
+	The proposed method enables the classical Metropolis-Hastings algorithm to be an efficient way to conduct exact posterior inference under the stochastic SIR model, given only incidence data for infections.
 Existing attempts using Markov chain Monte Carlo either mix poorly or do not scale to populations of more than a few hundred individuals, while methods relying on forward simulation are limited to moderate-sized outbreaks and may suffer from degeneracy issues on a case-by-case basis depending on the amount of missing data.
-Moreover, the simplifying assumptions in \cite{Fintzi.2020} or the approximate-Bayesian-computation framework of \cite{McKinley.2018}, which compromise targeting the exact posterior for computational reasons, generate biased estimates that may lead to spurious conclusions.
-Instead, our data-augmented algorithm enables fast and exact Bayesian inference, even for large outbreaks, and leverages a well-studied, transparent MCMC framework.
+%Moreover, the simplifying assumptions in \cite{Fintzi.2020} or the approximate-Bayesian-computation framework of \cite{McKinley.2018}, which compromise targeting the exact posterior for computational reasons, generate biased estimates that may lead to spurious conclusions.
+	In contrast to methods that rely on approximate computation of simplifying assumptions necessitated by computational considerations \cite{McKinley.2018,Fintzi.2020}, 
+	our data-augmented algorithm enables fast and exact Bayesian inference, even for large outbreaks, and leverages a well-studied, transparent MCMC framework with  guarantees of uniform ergodicity.
 %Most existing inferential methods for epidemic stochastic models are only applicable to prevalence data.
@@
 			\centering
-			\includegraphics[width=\textwidth]{gamma_minorization_1a.jpg}
+			\includegraphics[width=\textwidth]{gamma1a.jpg}
 			\caption{$\alpha \in [0, 1]$, $\beta=0$}
@@
 			\centering
-			\includegraphics[width=\textwidth]{gamma_minorization_1b.jpg}
+			\includegraphics[width=\textwidth]{gamma1b.jpg}
 			\caption{$\alpha = 0$, $\beta \in [0, 1]$}
@@
 		    \centering
-		    \includegraphics[width=\textwidth]{gamma_minorization_2.jpg}
+		    \includegraphics[width=\textwidth]{gamma2.jpg}
 			\caption{$\alpha \in \{0,1\}$, $\beta \in \{0,1\}$}

### 2022-01-06 Manuscripts/Prelim/Prelim.tex
Slides for prelim

\documentclass{beamer}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{booktabs}
\graphicspath{ {../Figures/} }
\input{../macros}
\usepackage{caption, array}
\usepackage{subcaption}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage[mathscr]{eucal}
\usepackage[english]{babel}
\usepackage{color}

\newtheorem{proposition}{Proposition}[section]

\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

%\title{A Uniformly Ergodic Data-Augmented MCMC for Fitting Stochastic Epidemic Models to Incidence Data}
\title{Exact Bayesian Inference for Stochastic Epidemic Models via Data Augmentation}

\author{
	Rapha\"{e}l Morsomme, supervised by Prof.\ Jason Xu
}

\date{\today}

\begin{document}

\frame{\titlepage}

\begin{frame} \frametitle{Overview}
\begin{itemize}
	\item The stochastic SIR process
	\item Inference with (in)complete data
	\item A novel DA-MCMC algorithm
	\item Simulation study and analysis of Ebola pandemic
	\item Future directions
\end{itemize}
\end{frame}

\begin{frame} \frametitle{The SIR Model}

	\begin{itemize}
		\item A compartmental model (S $\rightarrow$ I $\rightarrow$ R).
		\item Describes the spread of a contagious disease through a population.	
		\item Deterministic or \textbf{stochastic}; in discrete or \textbf{continuous} time.
		\vfill
		\begin{figure}
			\centering
			\includegraphics[scale = 0.2]{E3_S0=100_trajectories}
			\caption{Compartment trajectories of a stochastic SIR process.}
		\end{figure}
	\end{itemize}

\end{frame}

\begin{frame} \frametitle{Stochastic SIR Model}

	Consider the agent-based epidemic process $\left\lbrace \X(t), t>0\right\rbrace$, 
	$$\X(t) = (X_1(t), \dots, X_n(t)) \in \{s, i, r\}^n$$
	where the agent-level subprocess
	$$ X_j(t) = 
	\begin{cases}
		s, & t \in (0, \tau^I_j) \\
		i, & t \in (\tau^I_j, \tau^R_j) \\
		r, & t \in (\tau^R_j, \infty)
	\end{cases}
	,\quad \iton
	$$
	denotes the compartment of individual $j$ over time.

\end{frame}

\begin{frame} \frametitle{Stochastic SIR Model}

	\begin{itemize}
		\item Three common assumptions:
		\begin{itemize}
			\item Homogeneously mixing population: contacts between pairs of individuals follow independent Poisson. processes with rate $\beta$,
			\item Markovian assumption: iid exponentially distributed infectious periods with rate $\gamma$.
			\item Closed population.
		\end{itemize}
	\vfill
		\item These assumptions yield a \textit{Markov} process characterized by the transition rates			
			$
			\lambda_{\x, \x'} = 
			\begin{cases}
				\beta I(t), & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = s \text{ and } x_j' = i, \\
				\gamma, & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = i \text{ and } x_j' = r, \\
				0, & \text{otherwise.}
			\end{cases}
			$
	\end{itemize}
\end{frame}

\begin{frame} \frametitle{SIR -- Population-based Formulation}

	The agent-based formulation is equivalent to the more typical population-based formulation with
	$$W(t) = \left( S(t), I(t), R(t)\right)  \in \{0, 1, \cdots, n\}^3$$
	and 	
	$
	\lambda_{(S, I, R), (S', I', R')} = 
	\begin{cases}
		\beta S(t) I(t), & (S', I', R') = (S - 1, I + 1, R)\\
		\gamma I(t)    , & (S', I', R') = (S, I - 1, R + 1)\\
		0, & \text{otherwise.}
	\end{cases}
	$
	\vfill
	The agent-based formulation will allow us to
	\begin{itemize}
		\item leave the Markovian framework,
		\item jointly propose latent data in our DA-MCMC.
	\end{itemize}

\end{frame}

\begin{frame} \frametitle{Inference}

	The stochastic SIR process has a closed-form likelihood 
	\begin{align*}
		L(\theta; \X)
		& = \prod_{j \in \mathcal{I}} \beta I(\tau^I_j) \prod_{k \in \mathcal{R}} \gamma \exp\left\lbrace - \int_{0}^{t_{end}}\beta S(t)I(t) + \gamma I(t) dt \right\rbrace \\
		\label{eq:cdl}
		& = \beta^{n_I} \gamma^{n_R}\prod_{j \in \mathcal{I}} I(\tau^I_j) \exp\left\lbrace - \beta \int_{0}^{t_{end}} S(t)I(t) - \gamma \int_{0}^{t_{end}} I(t) dt \right\rbrace
	\end{align*}
	where
	\begin{itemize}
		\item $\mathcal{I} = \{j: \tau^I_j \in (0, t_{end}]\}$ and $\mathcal{R} = \{j: \tau^R_j \in (0, t_{end}]\}$,
		\item $n_I = |\mathcal{I}|$ and $n_R = |\mathcal{R}|$
	\end{itemize}

\end{frame}

\begin{frame} \frametitle{Inference}	
	Closed-form MLE:
	$$
	\hat{\beta} = \dfrac{n_I}{ \int_{0}^{t_{end}} S(t)I(t)dt}, \qquad \hat{\gamma} = \dfrac{n_R}{ \int_{0}^{t_{end}} I(t)dt},
	$$
	\begin{itemize}
		\item the integrals correspond to finite sums ($I(t)$ and $S(t)$ are constant between event times),
	\end{itemize}
	and gamma conjugacy:
	\begin{equation*}
		\beta \sim G(a_{\beta}, b_{\beta}), \qquad \gamma \sim G(a_{\gamma}, b_{\gamma}),
	\end{equation*}
	\begin{align*}
		\beta | \X & \sim G\left( a_{\beta} + n_I , b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt\right), \\
		\gamma | \X & \sim G\left( a_{\gamma} + n_R, b_{\gamma} + \int_{0}^{t_{end}} I(t) dt\right)
	\end{align*}

\end{frame}

\begin{frame} \frametitle{Inference - Reparameterization}

The SIR likelihood can also be parameterized in terms of $\tilde{\theta} = (\beta, R_0)$ with $R_0 = S(0) \beta / \gamma$
which has the semi-conjugate prior
\begin{equation*}
	\beta \sim G(a_{\beta}, b_{\beta}), \qquad R_0 \sim IG(a_{R}, b_{R})
\end{equation*}
\begin{align*}
\beta | \X, R_0 & \sim G\left( a_{\beta} + n_I + n_R, b_{\beta} + \int_{0}^{t_{end}} S(t)I(t) dt + \frac{S(0)}{R_0} \int_{0}^{t_{end}} I(t) dt\right) \\
R_0 | \X, \beta & \sim IG\left( a_{R} + n_R, b_{R} + \beta S(0) \int_{0}^{t_{end}} I(t) dt\right)
\end{align*}
$\Rightarrow$ Gibbs sampler to explore $\pi(\tilde{\theta}|\X)$.
\end{frame}

\begin{frame} \frametitle{Inference - Reparameterization}

	\begin{figure}
	\centering
	\begin{subfigure}[b]{0.49\textwidth}
		\centering
		\includegraphics[width=\textwidth]{E5_beta_gamma_bi_dens}
		\caption{$\theta = (\beta, \gamma)$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.49\textwidth}
		\centering
		\includegraphics[width=\textwidth]{E5_R0_beta_bi_dens}
		\caption{$\tilde{\theta} = (\beta, R_0)$}
	\end{subfigure}
	\caption{Joint posterior given partial data under different parameterizations.}
\end{figure}

\end{frame}

\begin{frame} \frametitle{Partial Data - Incidence Data for Infection}

	\begin{itemize}
		\item In practice, the process is never completely observed.
		\item In my research, I consider \textit{incidence data for infections}
		\item Given times $t_{0:K}$, the observed data are $\Y = I_{1:K}$ where 
		$$I_k = \#\{\tau^I_j \in (t_{k-1}, t_k]\}$$
		is the number of infections during the $k$th interval.
		\item E.g.\ weekly infection counts.
		\item Motivated by the 2013-2016 Ebola pandemic in Western Africa.
	\end{itemize}
\end{frame}

\begin{frame} \frametitle{Intractable Partial Likelihood}

	The partial data likelihood
	\begin{equation*}
	L(\theta; \Y) = \pi(\theta) \int_{\chi_\x} L(\theta; \x) \delta_{\Y}(\x) d\x,
	\end{equation*}
	where $\delta_{\Y}(\x) = 1$ if $\Y$ is compatible with the infection times in $\x$ and $0$ otherwise,
	is intractable.

\end{frame}

\begin{frame} \frametitle{Prior Work}

	\begin{itemize}
		\item Direct method to compute $L(\theta; Y)$ (Ho et al., 2018)
		\begin{itemize}
			\item Prohibitively slow.
		\end{itemize}
		\item Approximate likelihood (Fintzi et al., 2020)
		\begin{itemize}
			\item Assumptions questionable in small populations.
		\end{itemize}
		\item Particle filtering (King - 2015)
		\begin{itemize}
			\item Often degenerates.
		\end{itemize}
		\item DA-MCMC with RJ (Gibson and Renshaw - 1998)
		\begin{itemize}
			\item Single-site update MH step.
		\end{itemize}
		\item DA-MCMC with MH (Fintzi et al., 2017)
		\begin{itemize}
			\item Single-site update MH step.
		\end{itemize}
		\item DA-MCMC with Gibbs (Touloupou et al., 2020)
		\begin{itemize}
			\item Discrete time, single-site update Gibbs step.
			\item No likelihood evaluation (expensive step), but FFBS.
		\end{itemize}
	\end{itemize}

\end{frame}

\begin{frame} \frametitle{DA-MCMC}

	The closed-form likelihood $L(\theta; \X)$ and its (semi) conjugacy suggests using Data-Augmentation MCMC:
		\begin{itemize}
			\item Construct latent $\Z = \left\lbrace (z^I_j, z^R_j)\right\rbrace_{j=1}^{n}$ and explore $\pi(\theta, \Z|\Y)$.
			\item Naive approach: Gibbs sampler
			\begin{itemize}
				\item $\theta|\Y, \Z$ -- (semi) conjugacy
				\item $\Z|\Y,\theta$ -- prohibitively difficult
			\end{itemize}
		\end{itemize}
	\vfill
	Existing DA-MCMC approaches substitute the difficult Gibbs step $\Z|\Y,\theta$ with a simpler single-site-update MH step.	
	\begin{itemize}
		\item The resulting Markov chain is very sticky.
		\item Limited to populations with $n<1000$ individuals.
	\end{itemize}
	\vfill
	We propose \textbf{a DA-MCMC algorithm in which the entire latent data $\Z$ are jointly proposed in a MH step}.

\end{frame}

\begin{frame} \frametitle{Piece-wise decoupled SIR}

	Main idea of our DA-MCMC: use a \textit{surrogate} process to generate latent data $\Z^\star$ compatible with the observed data $\Y$ and accept/reject $\Z^\star$ in a MH step.
	\vfill
	To generate $\Z^\star$ compatible with $\Y = I_{1:K}$, proceed one interval at a time.
	\begin{itemize}
		\item Generate the $I_k$ infection times in $(t_{k-1}, t_k)$ using some procedure, e.g.\ iid $U(t_{k-1}, t_k)$.
		\item Generate the removal times of these inviduals by adding an exponential RV to their infection times.		
	\end{itemize}
\vfill
	We want to generate the infection times using a procedure that is as close as possible to the SIR.

\end{frame}

\begin{frame} \frametitle{Piece-wise decoupled SIR}

	Under the SIR, the \textit{population} infection rate $\mu_{pop}(t) = \beta I(t) S(t)$ changes after every event. Consider two approximations:
	\begin{enumerate}
		\item Let $\tilde{\tilde{\mu}}_{pop}$ be constant over each interval $[t_{k-1}, t_k)$:
		$$\tilde{\tilde{\mu}}_{pop}(t) := \mu_{pop}(t_{k-1}) = \beta S(t_{k-1})I(t_{k - 1}), \quad t \in [t_{k - 1}, t_k).$$
		The $I_k$ infection times follow a Poisson process (iid uniform).

		\item Let $\tilde{\mu}_{pop}$ be decoupled of $I(t)$ only (PD-SIR).
		$$\tilde{\mu}_{pop}(t) := \beta S(t)I(t_{k - 1}), \quad t \in [t_{k - 1}, t_k).$$
		\begin{itemize}
			\item The corresponding \textit{individual} infection rate is $\tilde{\mu}(t) :=\dfrac{\tilde{\mu}_{pop}(t)}{S(t)} = \beta I(t_{k - 1}) = \mu_k$ is piece-wise constant.
			\item The compartment $S(t)$ follows a linear death process (LDP) where infections correspond to deaths.
		\end{itemize}
	\end{enumerate}

\end{frame}

\begin{frame}

	\begin{figure}
		\centering
		\includegraphics[scale = 0.4]{infection_rate_SIR_PDSIR}
		\caption{Individual infection rate under the SIR (dashed) and PD-SIR (solid) processes. Small population ($n = 10$) and $t_{0:3} = (0, 0.2, 0.4, 0.6)$.}
	\end{figure}			
\end{frame}

\begin{frame}

	\begin{figure}
		\centering
		\begin{subfigure}[b]{0.49\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E2_K5.jpeg}
			\caption{$K=5$}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.49\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E2_K10.jpeg}
			\caption{$K=10$}
		\end{subfigure}
		\\
		\begin{subfigure}[b]{0.49\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E2_K50.jpeg}
			\caption{$K=50$}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.49\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E2_K1000.jpeg}
			\caption{$K=1000$}
		\end{subfigure}
		\caption{SIR (in color) and PD-SIR process (in grey).}
	\end{figure}

\end{frame}

\begin{frame} \frametitle{A Useful Theorem}

	\begin{theorem}		
		Consider a LDP with death rate $\mu$ and let $\tau_{1:N} \in (t_l, t_u]$ be the times of the $N$ deaths occurring between times $t_l$ and $t_u$. Then 
		$$\tau_j \,{\buildrel d \over =}\, X_{(j)}, \quad j = 1, \dots, N$$
		where $X_{(j)}$ is the $j^{\text{th}}$ order statistics of $N$ iid random variables following a truncated exponential distribution with rate $\mu$, lower bound $t_l$ and upper bound $t_u$.		
	\end{theorem}
	\vspace{0.5cm}
	$\Rightarrow$ In the PD-SIR process, $z^I_{\mathscr{I}_k}|I_k \sim \TruncExp(\mu_k; t_{k-1}, t_k)$ iid.

\end{frame}

\begin{frame} \frametitle{Simulating the PD-SIR -- pseudo-algorithm}

	\begin{algorithm*}[H]
		\SetAlgoLined
		%	\KwIn{$\theta = (\beta, \lambda), \Y = I_{1:K}$}
		\KwOut{$\Z^\star = \{(z^I_i, z^R_i)\}_i$ compatible with $\Y = I_{1:K}$}
		\For{interval $k = 1, \dots, K$}{
			\vspace{0.2cm}
			Compute the infection rate: \\
			\hspace{0.5cm} $\mu_k \leftarrow \beta I(t_{k-1})$ \\
			\vspace{0.2cm}
			Jointly generate the infection times: \\
			\hspace{0.5cm} $z^I_j \stackrel{iid}{\sim} \TruncExp(\mu_k; t_{k-1}, t_k)$, for $j \in \mathscr{I}_k$ \\
			\vspace{0.2cm}
			Generate the removal times:\\
			\hspace{0.5cm} $z^R_j | z^I_j \stackrel{indep.}{\sim} z^I_j + Exponential(\gamma)$, for $j \in \mathscr{I}_k$
		}
		\caption{Simulating a PD-SIR process conditionally on $\Y$}
	\end{algorithm*}

\end{frame}

\begin{frame} \frametitle{A few Remarks on the PD-SIR}

	\begin{itemize}
		\item The DA-MCMC can be initialized with $\theta^{(0)} = \left( \beta^{(0)}, \gamma^{(0)}\right)$ only.
		\item Gibbs-like scheme: $\Z^\star \perp \Z^{(k-1)} | \theta^{(k)}$.
		\begin{itemize}
			\item Uniform ergodicity
		\end{itemize}
		\item Efficient: 
		\begin{itemize}
			\item $\Z^\star$ is compatible with $\Y = I_{1:K}$ by construction.
			\item High MH acceptance rate due to close resemblance between the PD-SIR and SIR processes.			
		\end{itemize}
		\item Fast:
		\begin{itemize}
			\item Only requires uniform RVs (inverse CDF),
			\item Vectorized implementation (rate $\mu_k$ is only updated $K$ times).
		\end{itemize}
	\end{itemize}
\end{frame}

\begin{frame} \frametitle{Uniform Ergodicity}

	A Markov chain is \textit{uniformly ergodic} if for some finite $M$ and positive constant $r<1$, the $n$-step transition kernel $P^n$ satisfies
	$$
	\Vert P^n(x,.)-\pi(.)\Vert_{TV} \le M r^n, \quad \forall x \in \chi
	$$

	\begin{itemize}
		\item Uniform ergodicity ensures a CLT (Jones, 2004).
	\end{itemize}
\vfill
	We show that the state space $\chi = \chi_\theta \times \chi_\z$ is a \textit{small set},
	\begin{equation*}
		P(x,.)^m \ge \epsilon \nu(.), \quad \forall x\in \chi
	\end{equation*}
	for some measure $\nu$, positive integer $m$ ($m=1$) and constant $\epsilon > 0$, which holds iff $P$ is uniformly ergodic (Tierney, 1994).
\end{frame}

\begin{frame} \frametitle{Uniform Ergodicity -- Gist of Proof}

\begin{itemize}
	\item $P = P_{\theta} P_{\z}$ is composite:
	$$\x_1 = (\theta_1, \z_1) \rightarrow (\theta_2, \z_1) \rightarrow (\theta_2, \z_2) = \x_2.$$
	\item Minorize $P_{\theta}$ and $P_{\z}$ separately.
	\begin{itemize}
		\item $P_\theta$: $q_\theta(\theta_2|\z_1)$ is a product of two gammas with bounded parameters.
		\item $P_\z$: $q_\z(\z_2|\theta_2)$ is a product of positive factors (truncated exponentials and tail probabilities).
	\end{itemize}
	\item Interesting result, since the requirement that the whole space be small is typically not satisfied for models with unbounded spaces e.g.\ $\chi_\theta = \R^+\times\R^+$.
\end{itemize}

\end{frame}

\begin{frame} \frametitle{Simulation Study -- Proof of Concept}  
	\begin{itemize}
		\item $n = 1000$;  $(\beta, \lambda) = (0.0025, 1)$ [$R_0 = 2.5$]
		\item $I_{1:10} = (13, 31,  55, 71, 103, 111, 135, 103, 74, 42)$
		\item$(\beta_0, \gamma_0) = (\beta/10, \gamma/10)$ (low density region)
		\item Stationarity is reached within $1500$ iterations
		\item 1e6 iterations take $<30$ minutes on a laptop.
	\end{itemize}

	\begin{figure}
		\centering
		\begin{subfigure}[b]{0.32\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E1_short_no_burn_gamma_tp}
			\caption{Transient phase}
			\label{fig:E1_short_no_burn_gamma_tp}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.32\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E1_burn_gamma_tp}
			\caption{Recurrent phase}
			\label{fig:E1_burn_gamma_tp}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.32\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E6_burn_beta_acf_joint}
			\caption{ACF}
			\label{fig:E1_burn_gamma_acf}
		\end{subfigure}
		\caption{Performance of our DA-MCMC in a medium-sized population.}
		\label{fig:E1}
	\end{figure}

\end{frame}

\begin{frame} \frametitle{Simulation Study -- Coverage}  
	\begin{itemize}
		\item Repeat previous experiment 2000 times.
		\item Coverage of $90\%$ credible intervals.
		\item Weakly informative priors: $\beta \sim Ga(0.001,1), R_0 \sim IG(2, 2)$.
	\end{itemize}

	\begin{table}
		\centering
		\begin{tabular}{ C{2cm}| *{3}{C{2.2cm}}}
			%\hline
			Parameter & Observed coverage rate & Average of the posterior means & SD of the posterior means \\ 
			\hline
			$\beta$ (0.0025) & 0.895 & 0.0025 & 0.000255 \\ 
			$\gamma$ (1) & 0.902 & 1.00 & 0.142 \\ 
			$R_0$ (2.5) & 0.910 & 2.54 & 0.194 \\
			\hline
		\end{tabular}
	\end{table}
\end{frame}

\begin{frame} \frametitle{Simulation Study --  Tuning Parameter $\rho$}  
	\begin{itemize}
		\item For large $n$, the MH acceptance rate can be too low.
		\item To improve mixing, only update the trajectories of $\lceil\rho n\rceil$ randomly chosen individuals ($0<\rho\le1$) per iteration.
		\item $n\in(250, 500, 1000)$, $\rho \in (0.02, 0.05, 0.1, 0.25, 0.5, 0.75, 1)$.
	\end{itemize}

\begin{figure}
	\centering
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=\textwidth]{E3_runtime}
		\caption{Run time in seconds}
		\label{fig:E3_runtime}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=\textwidth]{E3_accept}
		\caption{Acceptance rate}
		\label{fig:E3_accept}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=\textwidth]{E3_facet_ESSsecR0}
		\caption{ESS/sec for $R_0$
			%Acceptance rate of the proposed latent spaces in the Metropolis-Hastings step of the MCMC algorithm across different population sizes and values for $\rho$.
		}
		\label{fig:E3_facet_ESSsecR0}
	\end{subfigure}
	\caption{Impact of $\rho$ on the performance of the DA-MCMC in a population of size $250$ (solid), $500$ (dotted) and $1000$ (dashed).}
	\label{fig:E3}
\end{figure}

\end{frame}

\begin{frame} \frametitle{Simulation Study --  Comparison with SSU DA-MCMC}

	\begin{figure}
		\centering
		\begin{subfigure}[b]{0.49\textwidth}
			\centering
			\includegraphics[width=0.48\textwidth]{E6_no_burn_beta_tp_joint.jpg}
			\includegraphics[width=0.48\textwidth]{E6_burn_beta_acf_joint.jpg}
			\caption{DA-MCMC}
			\label{fig:E6_no_burn_beta_tp_joint}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.49\textwidth}
			\centering
			\includegraphics[width=0.49\textwidth]{E6_no_burn_beta_tp_single.jpg}
			\includegraphics[width=0.48\textwidth]{E6_burn_beta_acf_single.jpg}
			\caption{SSU DA-MCMC}
			\label{fig:E6_no_burn_beta_tp_single}
		\end{subfigure}
		\caption{Traceplot and ACF of our DA-MCMC and a SSU DA-MCMC for $\beta$.}
		\label{fig:E6}
	\end{figure}

	\begin{table}
		\centering
		\begin{tabular}{ c|cc}
			Parameter & DA-MCMC & SSU DA-MCMC \\ 
			\hline
			$\beta$ & 0.20 & 0.01 \\ 
			$\gamma$ & 0.19 & 0.01 \\ 
			$R_0$ & 0.38 & 0.05 \\
			\hline
		\end{tabular}
		\caption{ESS/sec for our DA-MCMC and the SSU DA-MCMC.}
		\label{tab:E6}
	\end{table}

\end{frame}

\begin{frame} \frametitle{2013-2016 Ebola Pandemic in Western Africa}  
	\begin{itemize}
		\item Highly contagious virus, $>10000$ victims during the 2013-2016 pandemic in Western Africa.
		\item Weekly infection counts per province (Gueckedou).
		\item $n =$ 290,000; $\rho = 0.1$ (acceptance rate: 0.21)
		\item 1e6 iterations, 470 ESS.
	\end{itemize}

	\begin{figure}
		\centering
		\begin{subfigure}[b]{0.32\textwidth}
			\centering
			\includegraphics[width=\textwidth]{observed_data}
			\caption{Observed data}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.32\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E5_expected_infection_length_hist}
			\caption{Posterior distribution}
		\end{subfigure}
		\hfill
		\begin{subfigure}[b]{0.32\textwidth}
			\centering
			\includegraphics[width=\textwidth]{E5_expected_infection_length_tp}
			\caption{Traceplot}
		\end{subfigure}
		\caption{Observed data for the Gu\'eck\'edou prefecture and posterior distribution and traceplot of the expected infection length ($\gamma^{-1}$).}
		\label{fig:ebola}
	\end{figure}

\end{frame}

\begin{frame} \frametitle{Extensions}

	We have extended the PD-SIR to
	\begin{itemize}
		\item non-Markovian SIR,
		\item SIR with heterogeneously mixing population.
	\end{itemize}

	I am mentoring a student who works on 
	\begin{itemize}
		\item SIR with time varying infection rate.
	\end{itemize}

	I plan to work on the following problems
	\begin{itemize}
		\item under-reporting e.g.\ $\tilde{I}_k \sim Bin(I_k, \kappa)$,
		\item IR approximation to the SIR (small pandemic in large population, e.g.\ Ebola) $\Rightarrow$ $I(t)$ is a linear birth-death process,
		\item applying our DA-MCMC to other processes where we observe incidence data (as opposed to prevalence data).
	\end{itemize}

\end{frame}

\begin{frame} \frametitle{Conclusion}

	Our DA-MCMC is an efficient algorithm for exact Bayesian inference for the stochastic SIR with infection incidence data.	
	\begin{itemize}
		\item carefully designed surrogate process: efficient and fast.
	\end{itemize}

	The combination of (i) the agent-based formulation and (ii) the PD-SIR provides a unified inferential framework for the SIR with
	\begin{itemize}
		\item non-Markovian dynamics,
		\item a heterogeneously mixing population,
		\item a time varying infection rate.
	\end{itemize}

\end{frame}

\begin{frame} \frametitle{Extension -- Arbitrarily Distributed Infection Periods}

	\begin{itemize}
		\item Weibull distribution is more plausible for infectious periods.
		\item Dropping the assumptions of exponentially distributed infectious periods makes the process \textit{Non-Markovian}.
		\item Yet, the likelihood still has a closed form!
		\item PD-SIR with
		$$z^R_j | z^I_j \sim z^I_j + Weibull(.)$$
	\end{itemize}

\end{frame}

\begin{frame} \frametitle{Extension -- Heterogeneously Mixing Population}

	\begin{itemize}
		\item Severo (1969) proposes to substitute the population-level infection rate of the SIR
		$$\mu_{pop}(t) =  \beta S(t) I(t)$$
		with 
		$$\bar{\mu}_{pop}(t) = \beta S(t)^{1-b} I(t), \quad b \in [0,1]$$
		to model a heterogeneously mixing population.
		\item The corresponding individual-level infection rate is 
		$$\bar{\mu}(t) = \dfrac{\bar{\mu}_{pop}(t)}{S(t)} = \beta S(t)^{-b} I(t)$$
		\item PD-SIR with
		$$\mu_k = \tilde{\mu}(t_{k-1}) = \beta S(t_{k-1})^{-b} I(t_{k-1}).$$
	\end{itemize}

\end{frame}

\begin{frame} \frametitle{Extension -- Time-varying Infection Rate $\beta(t)$}

	\begin{itemize}
		\item $\beta \sim GP(.)$ (Kypraios, 2018)
		\begin{itemize}
			\item expensive: invert a matrix of order $n\times n$ each iteration
		\end{itemize}
		\item Random walk: $\Delta(\beta)_k \sim N(0, \sigma^2)$
		\begin{itemize}
			\item multivariate normal prior $\Rightarrow$ elliptical slice sampler
		\end{itemize}
		\item Locally adaptive: $\Delta(\beta)_k$ follows a Laplace or HS distribution (Faulkner \& Minin, 2018)
		\begin{itemize}
			\item accommodates sudden variations in $\beta$.
		\end{itemize}

	\end{itemize}

\end{frame}

\begin{frame} \frametitle{Uniform Ergodicity -- Minorizing $P_\theta$}

	The density of the Gibbs kernel $P_\z$ corresponds to the product of the two gamma densities
	$$Ga\left(\beta; a_{\beta} + n_I, b_{\beta} + \int S(t)I(t)dt\right)Ga\left(\gamma; a_{\beta} + n_R, b_{\beta} + \int I(t)dt\right)$$
	Since the sufficient statistics of $\Z$ are bounded:
	$$0 \le \int S(t)I(t)dt \le n^2t_{end}, \quad 0\le n_R \le n, \quad 0 \le \int I(t)dt \le n t_{end},$$
	each density possesses a positive minorization whose closed form we have derived.

\end{frame}

\begin{frame} \frametitle{Uniform Ergodicity -- Minorizing $P_\z$}

	The MH kernel $P_\z$ depends on the current latent data $\z$ only through the ratio $q(\z|\theta)/\pi(\z|\theta)$. 
	\begin{itemize}
		\item $q$ and $\pi$ can be bounded above and below away from $0$.
	\end{itemize}

	It is illuminating to observe that
	\begin{itemize}
		\item the contribution of the removal times in the numerator and the denominator cancel each other (generated from identical dynamics),
		\item the ratio of infections times' contribution 
		$$\prod_{k=1}^K \prod_{j\in \mathcal{I}_k}\dfrac{\beta I(t_{k-1}) \exp\{-\beta I(t_{k-1}) (z^I_j - t_{k-1})\}}
		{\beta I(z^I_l) \exp\{-\beta \int_{t_{k-1}}^{z^I_j} I(t)dt\}}$$
		will typically be close to $1$. 
	\end{itemize}

\end{frame}

\begin{frame} \frametitle{Three Types of Partially Observed Data}

exact removal times; prevalence infectious counts; incidence infection counts

\end{frame}

\end{document}

### 2022-01-13 Manuscripts/Article/Article.tex
Jasa style

@@
-%\usepackage[utf8]{inputenc}
-%\usepackage[round,authoryear]{natbib}
-
-\documentclass[11pt]{article}
-\usepackage[margin=1in]{geometry}
-\usepackage{url,hyperref}
+\documentclass[12pt]{article}
+\usepackage{amsmath}
 \usepackage{graphicx}
-\usepackage{amsmath,amssymb,array,eucal, amsthm}
-\linespread{1.45}
-\setlength\parindent{35pt}
-\input{../macros}
+\usepackage{enumerate}
+\usepackage{natbib}
+\usepackage{url} % not crucial - just used below for the URL
+
 \usepackage{algorithm}
 \usepackage{algorithmic}
-\usepackage{caption}
 \usepackage{subcaption}
-\usepackage{color}
+\usepackage{array}
+\usepackage{amsthm}
+
+%\pdfminorversion=4
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{1}
+
+% DON'T change margins - should be 1 inch all around.
+\addtolength{\oddsidemargin}{-.5in}%
+\addtolength{\evensidemargin}{-1in}%
+\addtolength{\textwidth}{1in}%
+\addtolength{\textheight}{1.7in}%
+\addtolength{\topmargin}{-1in}%
+
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}{Lemma}[section]
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
 
@@
 \newcommand{\ram}[1]{{\color{green}{ #1}}}
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{proposition}{Proposition}[section]
-\newtheorem{lemma}{Lemma}[section]
-
 \begin{document}
 
-	% Keywords command
-	\providecommand{\keywords}[1]
+	%\bibliographystyle{natbib}
+
+	\def\spacingset#1{\renewcommand{\baselinestretch}%
+		{#1}\small\normalsize} \spacingset{1}
+
+	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+	\if1\blind
 	{
-		\small	
-		\textbf{\textit{Keywords---}} #1
-	}
-
-	\title{%Efficient Likelihood-Based Inference for Fitting Stochastic Epidemic Models to Incidence Data via Data Augmentation
-		Uniformly Ergodic Data-Augmented MCMC for Fitting the General Stochastic Epidemic Model to Incidence Data}
-
-	\author{
-	Rapha\"{e}l Morsomme$^{1}$\footnote{raphael.morsomme@duke.edu} \ and Jason Xu$^{1}$ \\
-	\small $^{1}$Department of Statistical Science, Duke University \\
-	}
-	\date{
-	\today
-	}	
+		\title{\bf Uniformly Ergodic Data-Augmented MCMC for Fitting the General Stochastic Epidemic Model to Incidence Data}
+		\author{Rapha\"{e}l Morsomme\thanks{
+				The authors gratefully acknowledge \textit{please remember to list all relevant funding sources in the unblinded version}}\hspace{.2cm}\\
+			Department of Statistical Science, University of Duke\\
+			and \\
+			Jason Xu \\
+			Department of Statistical Science, University of Duke}
 		\maketitle
-
+	} \fi
+
+	\if0\blind
+	{
+		\bigskip
+		\bigskip
+		\bigskip
+		\begin{center}
+			{\LARGE\bf Uniformly Ergodic Data-Augmented MCMC for Fitting the General Stochastic Epidemic Model to Incidence Data}
+		\end{center}
+		\medskip
+	} \fi
+
+	\bigskip
 	\begin{abstract}
@@
 Yet, fitting these models when the epidemic process is only partially observed is a notoriously difficult task due to the intractability of the likelihood for many classical models.
-To remedy this issue, this article introduces a novel data-augmented MCMC algorithm for fast and exact inference for the stochastic SIR model given discretely observed incidence counts of infection.
-In a Metropolis-Hastings step, the algorithm jointly proposes the augmented data from a surrogate stochastic process whose dynamics closely resemble those of the SIR process, and from which we can efficiently generate an epidemic that is compatible with the observed data.
-Not only is the algorithm fast, but since the augmented data are generated from a faithful approximation of the target model, the method can update large portions of the latent space per iteration without prohibitively lowering the acceptance rate.
-We find that the method explores the high-dimensional latent space efficiently and show that the Markov chain underlying the algorithm is uniformly ergodic.
-		The proposed algorithm enables exact Bayesian inference that scales to outbreaks with hundreds of thousands of individuals,  even on a single laptop.
+To remedy this issue, this article introduces a novel data-augmented MCMC algorithm for fast and exact Bayesian inference for the stochastic SIR model given discretely observed infection incidence counts.
+In a Metropolis-Hastings step, new event times of the latent data are jointly proposed from a surrogate process that closely resembles the SIR, and from which we can efficiently generate epidemics compatible with the observed data.
+
+		The proposed DA-MCMC algorithm is fast and, since the latent data are generated from a faithful approximation of the target model, a large portion thereof can be updated per iteration without prohibitively lowering the acceptance rate.
+We find that the method explores the high-dimensional latent space efficiently and scales to outbreaks with hundreds of thousands of individuals, and we show that the Markov chain underlying the algorithm is uniformly ergodic.
 We validate its performance via thorough simulation experiments and a case study on the 2013-2015 Ebola outbreak in Western Africa.
 	\end{abstract}
-	\keywords{Stochastic SIR model; incidence data; exact Bayesian inference; data-augmented MCMC; likelihood-based methods; uniform ergodicity}
+
+	\noindent%
+	{\it Keywords:}  Stochastic epidemic model; incidence data; exact Bayesian inference; likelihood-based inference
+	\vfill
+
+	\newpage
+	\spacingset{1.9} % DON'T change the spacing!
 
 \section{Introduction}
-	% Outline: SEM; intractable likelihood; existing methods: approximation (Cauchemez 2008, Fintzi 2020), particle filtering, ABC, data augmentation; poor mixing;
+\label{sec:intro}
 
 The efficient control of a disease outbreak requires an understanding of the mechanisms underlying its spread.
-Mechanistic compartmental models, which describe the transition of individuals between various states, have a long mathematical modeling tradition in epidemiology \cite{Kermack.1927}.
+Mechanistic compartmental models, which describe the transition of individuals between various states, have a long mathematical modeling tradition in epidemiology dating back to \cite{Kermack.1927}.
 Due to their interpretability, they are commonly used to describe the dynamics of an outbreak and typically serve as the main source of information for predicting the course of an outbreak and identifying interventions that could be effective.
@@
 Moreover, given the size of the transition matrix, classical matrix exponentiation is intractable.
-%\jx{to do: cite recent papers from Simon Spencer's group too that do so for discrete-time models, and mention classical matrix exponentiation being intractable}.
-	Numerical methods to obtain transition probabilities have only recently been developed in \cite{Ho.2018b, Ho.2018}, but the high computational costs limit their approach to moderate sized outbreaks.
+Numerical methods to obtain transition probabilities have recently been developed in \cite{Ho.2018}, but the high computational costs limit their approach to moderate sized outbreaks.
 
@@
 The remainder of the article is structured as follows.
-Section \eqref{sec:set} provides background information on the inference task.
+Section \ref{sec:set} provides background information on the inference task.
 It introduces the stochastic SIR model and explains why conducting inference from partially observed data is a difficult task.
 Previous works addressing this problem are also presented.
-Section \eqref{sec:pds} introduces the DA-MCMC algorithm and describes the surrogate process from which the latent data are generated in the Metropolis-Hastings step.
-The similarity between the surrogate and the target processes is assessed in Section \eqref{sec:qua} and the DA-MCMC is shown to be uniformly ergodic in Section \eqref{sec:uni}.
-In Section \eqref{sec:sim}, we examine the performance of our algorithm on simulated data and then turn to an analysis of the 2013-2015 Ebola outbreak in Gu\'eck\'edou, Guinea, in Section \eqref{sec:ebo}.
-Finally, Section \eqref{sec:dis} discusses the findings and concludes the article.
+Section \ref{sec:pds} introduces the DA-MCMC algorithm and describes the surrogate process from which the latent data are generated in the Metropolis-Hastings step.
+The similarity between the surrogate and the target processes is assessed in Section \ref{sec:qua} and the DA-MCMC is shown to be uniformly ergodic in Section \ref{sec:uni}.
+In Section \ref{sec:sim}, we examine the performance of our algorithm on simulated data and then turn to an analysis of the 2013-2015 Ebola outbreak in Gu\'eck\'edou, Guinea, in Section \ref{sec:ebo}.
+Finally, Section \ref{sec:dis} discusses the findings and concludes the article.
 
@@
 \label{sec:set}
-	%Outline: compartmental model; CTMC; rates; likelihood; inference (MLE, gamma conjugacy)
 
@@
 \label{sec:sir}
-	%Outline: stochastic SIR; X; parameters; transition rate12
-
-	Our point of departure is to consider the stochastic SIR model -- also referred to as the general stochastic epidemic model \cite{Bailey.1975} -- which offers a parsimonious and interpretable representation of the mechanistic dynamics of an epidemic.
+Our point of departure is to consider the stochastic SIR model introduced by \cite{Bailey.1975} -- also referred to as the general stochastic epidemic model -- which offers a parsimonious and interpretable representation of the mechanistic dynamics of an epidemic.
 The stochastic SIR model is a compartmental model in which individuals transition through three states or compartments: susceptible (S), infectious (I) and removed (R).
@@
 	\label{eq:X}
-		\X = \left\lbrace \X(t), t>0\right\rbrace \in \chi_{\X}
+	\mathbf{X} = \left\lbrace \mathbf{X}(t), t>0\right\rbrace \in \chi_{\mathbf{X}}
 \end{equation}
@@
 \begin{equation}
-		\X(t) = \left(X_1(t), \dots, X_n(t)\right) \in \{s, i, r\}^n
+	\mathbf{X}(t) = \left(X_1(t), \dots, X_n(t)\right) \in \{s, i, r\}^n
 \end{equation}
@@
 \end{cases}
-	,\quad \iton
+,\quad i = 0, \dots, n
 $$
@@
 Since removed individuals do not contribute to the pandemic, it is safe to ignore the individuals removed at time $0$ and write $n = S(0) + I(0)$.
-Here the space $\chi_{\X}$ denotes the set of trajectories compatible with the evolution of a disease---that is, the set of trajectories in which no infection occurs when the infectious compartment is depleted:
+Here the space $\chi_{\mathbf{X}}$ denotes the set of trajectories compatible with the evolution of a disease---that is, the set of trajectories in which no infection occurs when the infectious compartment is depleted:
 \begin{equation}
 	\label{eq:chi}
-		\chi_{\X} = \{\X:\X(t) \in \{s,r\}^n \Rightarrow \X(t+u) = \X(t), \forall u>0 \}.
+	\chi_{\mathbf{X}} = \{\mathbf{X}:\mathbf{X}(t) \in \{s,r\}^n \Rightarrow \mathbf{X}(t+u) = \mathbf{X}(t), \forall u>0 \}.
 \end{equation}
@@
 This parameter can be interpreted as the \textit{infection rate}: when a susceptible individual comes into contact with an infectious individual, she immediately becomes infectious.
-	If we also make the common assumption that the infectious periods follow independent exponential distributions with rate $\gamma$, then the process \eqref{eq:X} is a time–homogeneous continuous time Markov chain, whose instantaneous transition rates
-	are given by the matrix $\Lambda = [\lambda_{\x, \x'}]$ with
+If we also make the common assumption that the infectious periods follow independent exponential distributions with rate $\gamma$, then the process \eqref{eq:X} is a time?homogeneous continuous time Markov chain, whose instantaneous transition rates
+are given by the matrix $\Lambda = [\lambda_{\mathbf{x}, \mathbf{x}'}]$ with
 \begin{equation}
 	\label{eq:rate}
-    	\lambda_{\x, \x'} = 
+	\lambda_{\mathbf{x}, \mathbf{x}'} = 
 	\begin{cases}
-    		\beta I(t), & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = s \text{ and } x_j' = i, \\
-    		\gamma, & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = i \text{ and } x_j' = r, \\
+		\beta I(t), & \mathbf{x} \text{ and } \mathbf{x}' \text{ only differ at position } j \text{ with } x_j = s \text{ and } x_j' = i, \\
+		\gamma, & \mathbf{x} \text{ and } \mathbf{x}' \text{ only differ at position } j \text{ with } x_j = i \text{ and } x_j' = r, \\
 		0, & \text{ otherwise.}
@@
 Thus, the individual-level infection and removal rates at time $t$ are respectively $\beta I(t)$ and $\gamma$.
-
-	%Equivalently \jx{The following is not an exact equality, but needs the $o(dt)$ term or similar}
-	%	$$
-	%	P(\X_t = \x, \X_{t+dt} = \x') =
-	%	\begin{cases}
-	%		\beta I(t) dt + o(dt), & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = S \text{ and } x_j' = I, \\
-	%		\gamma dt + o(dt), & \x \text{ and } \x' \text{ only differ at position } j \text{ with } x_j = I \text{ and } x_j' = R, \\
-	%		o(dt), & \text{otherwise}
-	%	\end{cases}
-	%	$$
-	%	where $dt>0$ is small.
 
@@
 
-	When the Markov process \eqref{eq:X} is completely observed until time $t_{end}$, we obtain the following likelihood \cite{Streftaris.2002}
+When the Markov process \eqref{eq:X} is completely observed until time $t_{end}$, we obtain the following likelihood \citep{Streftaris.2002}
 \begin{align}
-		L(\theta; \X)
+	L(\theta; \mathbf{X})
 	& = \prod_{j \in \mathcal{I}} \beta I(\tau^I_j) \prod_{k \in \mathcal{R}} \gamma \exp\left\lbrace - \int_{0}^{t_{end}}\beta S(t)I(t) + \gamma I(t) dt \right\rbrace  \nonumber \\
@@
 
-	It is straightforward to conduct inference in this continuously observed scenario \cite{Guttorp.2018}.
-The likelihood \eqref{eq:cdl} belongs to the exponential family, and maximum likelihood estimates of its parameters can be expressed in terms of the sufficient statistics defined above as
+It is straightforward to conduct inference in this continuously observed scenario %\cite{Guttorp.2018}
+. The likelihood \eqref{eq:cdl} belongs to the exponential family, and maximum likelihood estimates of its parameters can be expressed in terms of the sufficient statistics defined above as 
 $$
@@
 	\label{eq:posterior_theta}
-		\beta | \X & \sim G\left( a_{\beta} + n_I , b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt\right) \\
-		\gamma | \X & \sim G\left( a_{\gamma} + n_R, b_{\gamma} + \int_{0}^{t_{end}} I(t) dt\right)
+	\beta | \mathbf{X} & \sim G\left( a_{\beta} + n_I , b_{\beta} + \int_{0}^{t_{end}} I(t)S(t) dt\right) \\
+	\gamma | \mathbf{X} & \sim G\left( a_{\gamma} + n_R, b_{\gamma} + \int_{0}^{t_{end}} I(t) dt\right)
 \end{align}
-	from which one can easily generate independent values to explore the distribution $\pi(\theta|\X)$ via Monte Carlo.
+from which one can easily generate independent values to explore the distribution $\pi(\theta|\mathbf{X})$ via Monte Carlo.
 
@@
 \begin{align}		\label{eq:posterior_theta2}
-		\beta | \X, R_0 & \sim G\left( a_{\beta} + n_I + n_R, b_{\beta} + \int_{0}^{t_{end}} S(t)I(t) dt + \frac{S(0)}{R_0} \int_{0}^{t_{end}} I(t) dt\right) \\
-		R_0 | \X, \beta & \sim IG\left( a_{R} + n_R, b_{R} + \beta S(0) \int_{0}^{t_{end}} I(t) dt\right)
+	\beta | \mathbf{X}, R_0 & \sim G\left( a_{\beta} + n_I + n_R, b_{\beta} + \int_{0}^{t_{end}} S(t)I(t) dt + \frac{S(0)}{R_0} \int_{0}^{t_{end}} I(t) dt\right) \\
+	R_0 | \mathbf{X}, \beta & \sim IG\left( a_{R} + n_R, b_{R} + \beta S(0) \int_{0}^{t_{end}} I(t) dt\right)
 \end{align}
-	We have observed that the parameterization $\tilde{\theta}$ results in a Markov chain that mixes better and therefore employ it in the simulation studies and data analysis of Section \eqref{sec:per}.
-
-	%	\begin{equation}
-%	    \label{eq:lik2}
-%	    L(\X|\tilde{\theta}) = \beta^{n_I + n_R} R_0^{-n_R} S(0)^{n_R} \prod_{j\in\mathcal{I}} I(\tau^I_j) \exp\left\lbrace - \beta \int_{0}^{t_{end}} I(t)S(t) - \frac{\beta S(0)}{R_0} \int_{0}^{t_{end}} I(t) dt \right\rbrace,
-%	\end{equation}
+We have observed that the parameterization $\tilde{\theta}$ results in a Markov chain that mixes better and therefore employ it in the simulation studies and data analysis of Section \ref{sec:per}.
 
@@
 In practice, inference is complicated by the fact that the epidemic process \eqref{eq:X} is only partially observed.
-When we do not observe all event times, we do not have access to the sufficient statistics needed to evaluate the likelihood \eqref{eq:cdl}.
-Various types of partially observed data typify real data and have been considered in the literature, such as observing only the removal times \cite{Gibson.1998, ONeill.1999} or the number of infectious individuals at discrete points in time \cite{Fintzi.2017}.
+When some event times are unobserved, we do not have access to the sufficient statistics needed to evaluate the likelihood \eqref{eq:cdl}.
+Various types of partially observed data typify real data and have been considered in the literature.
+These include the removal times in \cite{Gibson.1998} and \cite{ONeill.1999} or the number of infectious individuals at discrete points in time in \cite{Fintzi.2017}.
 For instance, the former arises in animal experiments in which positive cases are immediately isolated from the rest of the population and no longer contribute to disease spread.
@@
 Such data arise when an infectious individual is identified some time after onset and may continue to infect others after being tested positive for the disease.
-Given an observation schedule $t_{0:K}$ ($K \ge 1$) with $0 = t_0 < t_1 < \dots < t_K = t_{end}$, the observed data consist of the $K$-dimensional vector $\Y = I_{1:K}$ where $I_k = \#\{\tau^I_j \in (t_{k-1}, t_k]\}$ is the number of infections during the $k^{\text{th}}$ interval.
-
-	This form of observed data is motivated by the $2013$–$2016$ outbreak of the Ebola virus, the largest outbreak since the discovery of the virus in 1976, which resulted in at least 11,000 deaths, mainly in three Western African countries (Guinea, Liberia and Sierra Leone).
+Given an observation schedule $t_{0:K}$ ($K \ge 1$) with $0 = t_0 < t_1 < \dots < t_K = t_{end}$, the observed data consist of the $K$-dimensional vector $\mathbf{Y} = I_{1:K}$ where $I_k = \#\{\tau^I_j \in (t_{k-1}, t_k]\}$ is the number of infections during the $k^{\text{th}}$ interval.
+
+This form of observed data is motivated by the $2013$-$2015$ outbreak of the Ebola virus, the largest outbreak since the discovery of the virus in 1976, which resulted in at least 11,000 deaths, the majority of which occurred in three Western African countries (Guinea, Liberia and Sierra Leone).
 The data available for this outbreak consist of weekly number of positive tests.
 This outbreak was characterized by the large number of infections that occurred after individuals were tested positive for the virus, giving rise to noisy incidence data.
-Numerous infections took place at the hospitals where infected individuals received treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased during funerals is a tradition.
-Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.} \cite{Coltart.2017}.
-We therefore decide to model the weekly number of positive tests as the weekly number of infections in a SIR model.
+\cite{Coltart.2017} reports that numerous infections took place at the hospitals where infected individuals received treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased during funerals is a tradition.
+Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.}.
+We therefore model the weekly number of positive tests as the weekly number of infections in a SIR model.
 
@@
 	\label{eq:pdl}
-		\pi(\theta|\Y) 
-		\propto \pi(\theta) L(\theta; \Y) = \pi(\theta) \int_{\chi_\x} L(\theta; \x) \delta_{\Y}(\x) d\x
-		%\mathbf{1}\{\Y \overset{\Delta}{=} \x\}
+	\pi(\theta|\mathbf{Y}) 
+	\propto \pi(\theta) L(\theta; \mathbf{Y}) = \pi(\theta) \int_{\chi_\mathbf{x}} L(\theta; \mathbf{x}) \delta_{\mathbf{Y}}(\mathbf{x}) d\mathbf{x}
+	%\mathbf{1}\{\Y \overset{\Delta}{=} \mathbf{x}\}
 \end{equation}
 where $\pi(\theta)$ is the prior distribution on $\theta$, and
-	$\delta_{\Y}(\x) = 1$ if $\x$ is compatible with the observed infection counts $\Y$ and $0$ otherwise; that is, the partial data likelihood $L(\theta; \Y)$ consists of a high-dimensional integral over all epidemic paths that agree with the observed data.
+$\delta_{\mathbf{Y}}(\mathbf{x}) = 1$ if $\mathbf{x}$ is compatible with the observed infection counts $\mathbf{Y}$ and $0$ otherwise; that is, the partial data likelihood $L(\theta; \mathbf{Y})$ consists of a high-dimensional integral over all epidemic paths that agree with the observed data.
 This marginalization step has no known closed-form solution and presents computational challenges even for a population of moderate size.
@@
 Researchers have explored several approaches for conducting inference on partially observed stochastic epidemic processes.
-Early approaches include the use of martingale-based equations \cite{Becker.1977, Watson.1981, Sudbury.1985}.
+Early approaches include the use of martingale-based equations \citep{Becker.1977, Watson.1981}.
 However, such methods are difficult to apply to dynamic models with partially observed data and are therefore not suitable to the stochastic SIR model with incomplete data.
 More recently, researchers have based their computation on a simpler process that approximates the model's dynamics and whose likelihood in the presence of partially observed data is tractable.
-Popular approximations include chain binomial models \cite{Greenwood.1931, Abbey.1952}, diffusion processes \cite{Cauchemez.2008, Fintzi.2020} and Gaussian processes \cite{Jandarov.2014}.
+Popular approximations include chain binomial models \citep{Greenwood.1931, Abbey.1952}, diffusion processes \citep{Cauchemez.2008, Fintzi.2020} and Gaussian processes \citep{Jandarov.2014}.
 While these approximation-based approaches bypass the intractability of the partial data likelihood \eqref{eq:pdl}, the assumptions on which they are based are questionable when the population is small, and, as a result, the dynamic of the stochastic process differs from its asymptotic behavior.
-	%Recently, \cite{Ho.2018} proposed an exact analytical method to directly compute the likelihood of the stochastic SIR process given discretely observed data.
-
-	Since the popularization of Markov chain Monte Carlo (MCMC) methods in statistics \cite{Tanner.1987, Gelfand.1990, Tierney.1994}, researchers have developed sampling-based methods to directly work with the SIR likelihood instead of an approximation thereof.
+
+Since the popularization of Markov chain Monte Carlo (MCMC) methods in statistics by \cite{Gelfand.1990} and \cite{Tierney.1994}, researchers have developed sampling-based methods to directly work with the SIR likelihood instead of an approximation thereof.
 MCMC algorithms fall into two categories: model-based forward simulation and data-augmented MCMC (DA-MCMC).
-	Particle filtering \cite{King.2015} is an example of the former category that is popular among practitioners.
+Particle filtering in \cite{King.2015} is an example of the former category that is popular among practitioners.
 Its plug-and-play feature makes it applicable to a wide variety of models.
 However, model-based forward simulation suffers from two drawbacks: simulating data from a model as complex as the SIR that is compatible with the observed data is prohibitively slow, and these methods can fail to converge when the model does not fit the data well.
-The approximate Bayesian computation (ABC) framework  \cite{McKinley.2018} offers a solution to the latter problem, but its inference is based on an approximation of the model's likelihood.
+The approximate Bayesian computation (ABC) framework in \cite{McKinley.2018} offers a solution to the latter problem, but its inference is based on an approximation of the model's likelihood.
 As a result, the inference can be biased, and it is difficult to evaluate the degree of approximation involved.
@@
 Researchers have mostly focused on the development of more efficient proposal schemes to explore this high-dimensional latent space.
-Early attempts employed reversible-jump MCMC \cite{Green.1995} to explore models with different numbers of unobserved events \cite{Gibson.1998, ONeill.1999}.
+\cite{Gibson.1998} and \cite{ONeill.1999} employed the reversible-jump MCMC framework introduced by \cite{Green.1995} to explore models with different numbers of unobserved events.
 These authors augment the observed data which consist of the recovery times with the unobserved infection times and explore the latent space by uniformly inserting, deleting or moving an element of the latent data per iteration of the MCMC algorithm.
-	More recently, Fintzi et al.\ \cite{Fintzi.2017} proposed a MCMC algorithm to conduct inference with discretely observed prevalence data of infection.
-They constructed latent data consisting of the infection and recovery times of each individual.
-The algorithm explores this latent space by updating the event times of an individual per iteration of the algorithm.
-These event times are sampled from a distribution that approximate the exact dynamics of the epidemic process.
-	In the discrete-time setting, Touloupou et al.\ \cite{Touloupou.2020} used a Gibbs sampler to update the trajectory of an individual per iteration.
+More recently, \cite{Bu.2020} proposed a sampler for epidemics over networks when the infection times are assumed to be known, and \cite{Fintzi.2017} designed a sampler to conduct inference with discretely observed prevalence counts, with latent data consisting of the infection and recovery times of each individual.
+Their algorithms explore the latent spaces by updating the event times of a single individual per iteration of the MCMC algorithm.
+%These event times are sampled from a distribution that approximate the exact dynamics of the epidemic process.
+In the discrete-time setting, \cite{Touloupou.2020} constructed a Gibbs sampler to update the trajectory of an individual per iteration.
 These DA-MCMC methods suffer from poor mixing in the presence of large epidemics.
@@
 
-	%An advantage of DA-MCMC methods is the possibility to conduct inference in more complex models.
-For instance, Streftaris and Renshaw \cite{Streftaris.2002} generalized the approach of \cite{Gibson.1998} to a non-Markovian stochastic SIR model where the amount of time spent infectious is not necessarily exponentially distributed but follows an arbitrary distribution, and Bu et al.
-\cite{Bu.2020} analyzed models in which the population does not mix homogeneously but is characterized by a dynamic social network, and Kypraios  model the infection dynamics in a nonparameteric framework.
-
 \section{Exact Inference with a  Data-Augmented Approach}
@@
 
-	As shown in Section \eqref{sec:icd}, the complete data likelihood \eqref{eq:cdl} is amenable to computation.
-This suggests augmenting the observed data $\Y$ with latent data $\Z$ such that the likelihood $L(\theta; (\Y, \Z))$ has the closed form \eqref{eq:cdl}, and constructing an ergodic Markov chain $\{(\theta^{(m)}, \Z^{(m)})\}_{m=0}^M$ on $\chi = \chi_{\theta} \times \chi_{\z}$ whose limiting distribution is the joint posterior $\pi(\theta, \X|\Y)$ \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
-Given the draws $\{(\theta^{(m)}, \Z^{(m)})\}_{m=0}^M$, the sequence $\{\theta^{(m)}\}_{m=0}^M$ provides an approximation of the marginal posterior distribution of interest $\pi(\theta|\Y)$.
-
-	We consider the latent data $\Z = \left\lbrace (z^I_j, z^R_j)\right\rbrace_{j=1}^{n}$ which consist of the times of the two types of unobserved epidemic events, where the infection and removal times of individual $j$ are respectively
+As shown in Section \ref{sec:icd}, the complete data likelihood \eqref{eq:cdl} is amenable to computation.
+This suggests augmenting the observed data $\mathbf{Y}$ with latent data $\mathbf{Z}$ such that the likelihood $L(\theta; (\mathbf{Y}, \mathbf{Z}))$ has the closed form \eqref{eq:cdl}, and constructing an ergodic Markov chain $\{(\theta^{(m)}, \mathbf{Z}^{(m)})\}_{m=0}^M$ on $\chi = \chi_{\theta} \times \chi_{\mathbf{z}}$ whose limiting distribution is the joint posterior $\pi(\theta, \mathbf{X}|\mathbf{Y})$ %\cite{Gibson.1998, ONeill.1999, Fintzi.2017}
+.
+Given the draws $\{(\theta^{(m)}, \mathbf{Z}^{(m)})\}_{m=0}^M$, the sequence $\{\theta^{(m)}\}_{m=0}^M$ provides an approximation of the marginal posterior distribution of interest $\pi(\theta|\mathbf{Y})$.
+
+We consider the latent data $\mathbf{Z} = \left\lbrace (z^I_j, z^R_j)\right\rbrace_{j=1}^{n}$ which consist of the times of the two types of unobserved epidemic events, where the infection and removal times of individual $j$ are respectively
 $$z^I_j \begin{cases}
@@
 \end{cases}$$
-	Since $\Y = I_{1:K}$ is a deterministic function of the infection events, the joint posterior distribution is
+Since $\mathbf{Y} = I_{1:K}$ is a deterministic function of the infection events, the joint posterior distribution is
 \begin{align*}
-	    \pi(\theta, \Z|\Y)
-	    & \propto \delta_{\Y}(\Z) L(\theta; \Z) \pi(\theta)
+	\pi(\theta, \mathbf{Z}|\mathbf{Y})
+	& \propto \delta_{\mathbf{Y}}(\mathbf{Z}) L(\theta; \mathbf{Z}) \pi(\theta)
 \end{align*}
-	where $L(\theta; \Z)$ is the complete data likelihood \eqref{eq:cdl}.
-
-	We construct a DA-MCMC algorithm that alternates between updates of the parameters $\theta$ given the current configuration of the latent data $\Z$, and updates of the latent data given the current values of the parameters.
-A one-step transition from $\x_1 = (\theta_1, \z_1)$ to $\x_2 = (\theta_2, \z_2)$ therefore looks like
-	$$\x_1 = (\theta_1, \z_1) \rightarrow (\theta_2, \z_1) \rightarrow (\theta_2, \z_2) = \x_2.$$
+where $L(\theta; \mathbf{Z})$ is the complete data likelihood \eqref{eq:cdl}.
+
+We construct a DA-MCMC algorithm that alternates between updates of the parameters $\theta$ given the current configuration of the latent data $\mathbf{Z}$, and updates of the latent data given the current values of the parameters.
+A one-step transition from $\mathbf{x}_1 = (\theta_1, \mathbf{z}_1)$ to $\mathbf{x}_2 = (\theta_2, \mathbf{z}_2)$ therefore looks like
+$$\mathbf{x}_1 = (\theta_1, \mathbf{z}_1) \rightarrow (\theta_2, \mathbf{z}_1) \rightarrow (\theta_2, \mathbf{z}_2) = \mathbf{x}_2.$$
 The first update for $\theta$ is straightforward due to the Gamma conjugacy of the complete-data likelihood \eqref{eq:cdl}.
 We can simply draw from the two full conditional distributions \eqref{eq:posterior_theta}.
-In these expressions, the sufficient statistics from $\Z$, $n_I, n_R, \int S(t)I(t) dt, \int I(t) dt$ are easy to compute.
-	The second update for $\Z$, on the other hand, is difficult.
+In these expressions, the sufficient statistics from $\mathbf{Z}$, $n_I, n_R, \int S(t)I(t) dt, \int I(t) dt$ are easy to compute.
+The second update for $\mathbf{Z}$, on the other hand, is difficult.
 Though unrestricted forward simulation of the stochastic SIR process is straightforward \cite{Gillespie.1977}, drawing trajectories conditionally on the observed data amounts to \textit{conditional} simulation of a Markov process, a notoriously difficult task \cite{Hobolth.2009}.
-Since given the $\Y = I_{1:K}$ the end-point of the process are not known, the methods proposed in \cite{Hobolth.2009} are not immediately applicable.
-	Instead, we generate the latent data conditionally on $\Y$ from a surrogate process, and accept or reject them in a Metropolis-Hastings (MH) step.
-Given some current $\theta$ and $\z$, the MH step proceeds by simulating a candidate $\z^\star$ from a surrogate process $Q(.|\theta)$ with density $q(.|\theta)$, which we present in Section \eqref{sec:pds}.
-Then $\z^\star$ is accepted with probability
+Since given the $\mathbf{Y} = I_{1:K}$ the end-point of the process are not known, the methods proposed in \cite{Hobolth.2009} are not immediately applicable.
+Instead, we generate the latent data conditionally on $\mathbf{Y}$ from a surrogate process, and accept or reject them in a Metropolis-Hastings (MH) step.
+Given some current $\theta$ and $\mathbf{z}$, the MH step proceeds by simulating a candidate $\mathbf{z}^\star$ from a surrogate process $Q(.|\theta)$ with density $q(.|\theta)$, which we present in Section \ref{sec:pds}.
+Then $\mathbf{z}^\star$ is accepted with probability
 \begin{equation}
 	\label{eq:alpha}
-		\alpha\left( \left( \theta, \z\right) , \left( \theta, \z^\star\right) \right) =	\min\left\lbrace 1, \dfrac{L\left( \theta; \z^\star\right) q\left( \z|\theta\right)}{L\left( \theta;\z\right) q\left( \z^\star|\theta\right)}\right\rbrace,
-	\end{equation}
-	and otherwise the current $\z$ is retained \cite{Tierney.1994}.
-Note that it is not necessary to compute $\delta_{\Y}(\z)$ in the MH acceptance ratio \eqref{eq:alpha} since $\z^\star$ is compatible with $\Y$ by construction.
+	\alpha\left( \left( \theta, \mathbf{z}\right) , \left( \theta, \mathbf{z}^\star\right) \right) =	\min\left\lbrace 1, \dfrac{L\left( \theta; \mathbf{z}^\star\right) q\left( \mathbf{z}|\theta\right)}{L\left( \theta;\mathbf{z}\right) q\left( \mathbf{z}^\star|\theta\right)}\right\rbrace,
+\end{equation}
+and otherwise the current $\mathbf{z}$ is retained.
+Note that it is not necessary to compute $\delta_{\mathbf{Y}}(\mathbf{z})$ in the MH acceptance ratio \eqref{eq:alpha} since $\mathbf{z}^\star$ is compatible with $\mathbf{Y}$ by construction.
 Algorithm \ref{alg:DA-MCMC} provides the details of the DA-MCMC algorithm.
@@
 \begin{algorithm}
-		\caption{Data-Augmented MCMC}
+	\caption{Data-Augmented MCMC Sampler}
 	\label{alg:DA-MCMC}
@@
 		\REQUIRE $\theta^{(0)}$
-			\RETURN $\{(\z^{(m)}, \theta^{(m)})\}_{m=0}^N$ 
-			\STATE $\z^{(0)} \sim Q(.|\theta^{(0)})$ (generate the initial latent data from the PD-SIR process)
+		\RETURN $\{(\mathbf{z}^{(m)}, \theta^{(m)})\}_{m=0}^N$ 
+		\STATE $\mathbf{z}^{(0)} \sim Q(.|\theta^{(0)})$ (generate the initial latent data from the PD-SIR process)
 		\FOR {$j = 1, \dots, N$}
-			\STATE $\beta^{(j)}|\z^{(j - 1)} \sim Ga\left( a_{\beta} + n_I^{(j - 1)}, b_{\beta} + \int_{0}^{t_{end}} I^{(j - 1)}(t)S^{(j - 1)}(t)dt\right)$ (Gibbs update)
-			\STATE $\gamma^{(j)}|\z^{(j-1)} \sim Ga\left( a_{\gamma} + n_R^{(j - 1)}, b_{\gamma} + \int_{0}^{t_{end}} I^{(j - 1)}(t) dt\right)$ (Gibbs update)
+		\STATE $\beta^{(j)}|\mathbf{z}^{(j - 1)} \sim Ga\left( a_{\beta} + n_I^{(j - 1)}, b_{\beta} + \int_{0}^{t_{end}} I^{(j - 1)}(t)S^{(j - 1)}(t)dt\right)$ (Gibbs update)
+		\STATE $\gamma^{(j)}|\mathbf{z}^{(j-1)} \sim Ga\left( a_{\gamma} + n_R^{(j - 1)}, b_{\gamma} + \int_{0}^{t_{end}} I^{(j - 1)}(t) dt\right)$ (Gibbs update)
 		\STATE $\theta^{(j)} \leftarrow (\beta^{(j)}, \gamma^{(j)})$
-			\STATE $\z^\star \sim Q(.|\theta^{(j)})$ (generate latent data from the PD-SIR process)
-			\STATE $\alpha = \min\left\lbrace 1, \dfrac{L(\theta^{(j)}; \z^\star)q(\z^{(j - 1)}|\theta^{(j)})}{L(\theta^{(j)}; \z^{(j - 1)})q(\z^\star|\theta^{(j)})}\right\rbrace  $
+		\STATE $\mathbf{z}^\star \sim Q(.|\theta^{(j)})$ (generate latent data from the PD-SIR process)
+		\STATE $\alpha = \min\left\lbrace 1, \dfrac{L(\theta^{(j)}; \mathbf{z}^\star)q(\mathbf{z}^{(j - 1)}|\theta^{(j)})}{L(\theta^{(j)}; \mathbf{z}^{(j - 1)})q(\mathbf{z}^\star|\theta^{(j)})}\right\rbrace  $
 		\STATE $u \sim U(0,1)$
 		\IF{$u<\alpha$}
-			\STATE $\z^{(j)} \leftarrow \z^*$			
+		\STATE $\mathbf{z}^{(j)} \leftarrow \mathbf{z}^*$			
 		\ELSE
-			\STATE $\z^{(j)} \leftarrow \z^{(j - 1)}$
+		\STATE $\mathbf{z}^{(j)} \leftarrow \mathbf{z}^{(j - 1)}$
 		\ENDIF
@@
 
-	The surrogate process that we consider for generating the latent data in the MH step consists of a stochastic process whose dynamics closely resemble those of the SIR and which is designed for efficient simulation of epidemic trajectories compatible with the incidence data $\Y = I_{1:K}$.
+The surrogate process that we consider for generating the latent data in the MH step consists of a stochastic process whose dynamics closely resemble those of the SIR and which is designed for efficient simulation of epidemic trajectories compatible with the incidence data $\mathbf{Y} = I_{1:K}$.
 We refer to this surrogate process as the \textit{piecewise decoupled SIR} process (PD-SIR).
@@
 %The population-level infection rate at time $t$ is $\mu_T(t) = \beta S(t)I(t)$ and so, from the perspective of each susceptible individual, the individual-level infection rate is $\mu(t) = \frac{\mu_T(t)}{S(t)} = \beta I(t)$.
-	The infection dynamics, however, vary slightly.
-	In the SIR process, the individual-level infection rate at time $t$ is $\mu(t) = \beta I(t)$ (see Equation \eqref{eq:rate}).
+The infection dynamics, however, differ slightly.
+In the SIR process, the individual-level infection rate at time $t$ is $\mu(t) = \beta I(t)$; see Equation \eqref{eq:rate}.
 Note that $\mu$ varies after every event since the value of $I$ changes after an infection or a removal.
 In contrast, in the PD-SIR process, the infection rate is kept constant over small periods of time.
-Consider the observation schedule of $\Y$, $t_{0:K}$.
+Consider the observation schedule of $\mathbf{Y}$, $t_{0:K}$.
 For $t \in (t_{k-1}, t_k]$, the infection rate of the PD-SIR process is defined as
@@
 where $I(t_{k-1})$ denotes the number of infectious individuals at the beginning of the $k$th interval.
-As shown in Figure \ref{fig:mu}, the infection rate $\tilde{\mu}$ and the variable $I$ are decoupled during each interval, and the infection rate is reset at the beginning %or left endpoint
-	of the intervals. Over a single interval, the PD-SIR process is equivalent to a recent two-type branching process approximation of the SIR dynamics  \cite{Ho.2018}.
+As shown in Figure \ref{fig:mu}, the infection rate $\tilde{\mu}$ and the variable $I$ are decoupled in each interval, and the infection rate is reset at the beginning %or left endpoint
+of the intervals. Over a single interval, the PD-SIR process is equivalent to a recent two-type branching process approximation of the SIR dynamics proposed in \cite{Ho.2018b}.
 
 \begin{figure}
-		\centering
-		\includegraphics[scale = 0.4]{infection_rate_SIR_PDSIR.jpg}
+	\begin{center}
+		\includegraphics[width=3in]{infection_rate_SIR_PDSIR}
+	\end{center}
 	\caption{Infection dynamics of the SIR (dashed line) and PD-SIR (solid line) processes in a small population ($(S(0), I(0)) = (10,2)$) with $\beta=1$.
 The schedule of the PD-SIR is $t_{0:3}=(0, 0.2, 0.4, 0.6)$.
-		The vertical dotted lines indicate the infection (red) and removal (blue) times.}
-		\label{fig:mu}
+		Vertical dotted lines indicate infection (red) and removal (blue) times. \label{fig:mu}}
 \end{figure}
 
-	It now becomes straightforward to simulate realizations from the PD-SIR process conditionally on the observations $\Y = I_{1:K}$.
+It now becomes straightforward to simulate realizations from the PD-SIR process conditionally on the observations $\mathbf{Y} = I_{1:K}$.
 By construction, under the PD-SIR model the variable $S(t)$ follows a linear death process (LDP) during each interval with death rate $\mu_k = \beta I(t_{k-1})$.
-	The LDP is a counting process possessing the so-called order statistics property \cite{Neuts.1971}; that is, given the number of events registered on some interval, the event times are distributed as the order statistics of independent and identically distributed (i.i.d.) random variables following, in the case of the LDP, a exponential distribution truncated to that interval.
+The LDP is a counting process possessing the so-called order statistics property \citep{Neuts.1971}; that is, given the number of events registered on some interval, the event times are distributed as the order statistics of independent and identically distributed (i.i.d.) random variables following, in the case of the LDP, a exponential distribution truncated to that interval.
 \begin{theorem}
@@
 Appendix \ref{app:ldp} contains the proof of this theorem.
-Generating values from a truncated exponential distribution can be done extremely efficiently with the inverse cumulative distribution function method \cite{Devroye.2006}.
-
-	We can use Theorem \ref{theo:ldp} to generate latent data  $\z = \left\lbrace (z^I_i, z^R_i)\right\rbrace_{i=1}^{n}$ compatible with $\Y = I_{1:K}$ from the PD-SIR process as follows.
+Generating values from a truncated exponential distribution can be done extremely efficiently via the inverse cumulative distribution function method.
+
+We can use Theorem \ref{theo:ldp} to generate latent data  $\mathbf{z} = \left\lbrace (z^I_i, z^R_i)\right\rbrace_{i=1}^{n}$ compatible with $\mathbf{Y} = I_{1:K}$ from the PD-SIR process as follows.
 For each time interval $[t_{k-1}, t_k)$, we compute $\mu_k = \beta I(t_{k-1})$ and $U_k = I(0) + \sum_{l=1}^{k-1} I_l$, the cumulative number of infections that happened before $t_{k-1}$.
@@
 Algorithm \ref{alg:PD-SIR} provides a simple recursion to compute this variable efficiently.
-For $j \in \mathcal{I}_k$, following Theorem \ref{theo:ldp}, we sample the infection times $z^I_j \sim \TruncExp(\mu_k; t_{k-1}, t_k)$, a truncated exponential distribution with rate $\mu_k$ bounded between $t_{k-1}$ and $t_k$.
+For $j \in \mathcal{I}_k$, following Theorem \ref{theo:ldp}, we sample the infection times $z^I_j \sim \textsf{TrunExp}(\mu_k; t_{k-1}, t_k)$, a truncated exponential distribution with rate $\mu_k$ bounded between $t_{k-1}$ and $t_k$.
 For the same indices $j$, we then generate the removal times $z^R_j$ from the removal dynamics of the SIR.
 To accomplish this, we sample the removal time of individual $j$ from the mixed distribution
-	$$z^R_j|z^I_j \sim (1 - p_j) \delta_{\infty} + p_j \TruncExp(\gamma; z^I_j, t_{end})$$
+$$z^R_j|z^I_j \sim (1 - p_j) \delta_{\infty} + p_j \textsf{TrunExp}(\gamma; z^I_j, t_{end})$$
 placing point mass $(1 - p_j)$ at $\infty$ and continuous mass on the interval $(z^I_j, t_{end}]$,
@@
 
-	By construction, this scheme generates latent data from the PD-SIR process that are compatible with $\Y = I_{1:K}$. The density $q$ of the proposal kernel $Q$ for the PD-SIR process corresponds to
+By construction, this scheme generates latent data from the PD-SIR process that are compatible with $\mathbf{Y} = I_{1:K}$.
+The density $q$ of the proposal kernel $Q$ for the PD-SIR process corresponds to
 \begin{alignat}{3}
 	\label{eq:q}
-		q(\z|\theta) 
-		& = && \prod_{k=1}^K \prod_{j\in\mathcal{I}_k} \TruncExp(z^I_j; \mu_{k}, t_{k-1}, t_{k}) \\
-		& && \times \prod_{i=1}^{n} \left( 1 - p_i \right)^{\mathbf{1}(z^R_i = \infty)} \left( p_i \TruncExp(z^R_i; \gamma, z^I_i, t_{end}) \right)^{\mathbf{1}(z^R_i \le t_{end})}
+	q(\mathbf{z}|\theta) 
+	& = && \prod_{k=1}^K \prod_{j\in\mathcal{I}_k} \textsf{TrunExp}(z^I_j; \mu_{k}, t_{k-1}, t_{k}) \\
+	& && \times \prod_{i=1}^{n} \left( 1 - p_i \right)^{\mathbf{1}(z^R_i = \infty)} \left( p_i \textsf{TrunExp}(z^R_i; \gamma, z^I_i, t_{end}) \right)^{\mathbf{1}(z^R_i \le t_{end})}  \nonumber
 \end{alignat}
 where %$\mu_k = \beta I(t_{k-1})$ and
-	$$\TruncExp(x; \mu, l, u) = \dfrac{\mu \exp\{-\mu x\}}{\exp\{-\mu l\} - \exp\{-\mu u\}}, \quad x \in (l, u)$$
-	denotes the density of a truncated exponential distribution with parameters as notated previously. Algorithm \ref{alg:PD-SIR} provides a summary of the proposal scheme in pseudo-code.
+$$\textsf{TrunExp}(x; \mu, l, u) = \dfrac{\mu \exp\{-\mu x\}}{\exp\{-\mu l\} - \exp\{-\mu u\}}, \quad x \in (l, u)$$
+denotes the density of a truncated exponential distribution with parameters as notated previously.
+Algorithm \ref{alg:PD-SIR} provides a summary of the proposal scheme used to generate $\mathbf{z^\ast}$ from Algorithm \ref{alg:DA-MCMC}.
 
 \begin{algorithm}
-		\caption{Generating a PD-SIR process conditionally on the observed data $\Y = I_{1:K}$}
+	\caption{Generating a PD-SIR process conditionally on the observed data $\mathbf{Y} = I_{1:K}$}
 	\label{alg:PD-SIR}
@@
 		\STATE $p_j \leftarrow 1 - \exp\{-\gamma (t_{end} - 0)\}$
-    			\STATE $z^R_j \sim (1 - p_j) \delta_\infty + p_j \TruncExp(\gamma, 0, t_{end})$
+		\STATE $z^R_j \sim (1 - p_j) \delta_\infty + p_j \textsf{TrunExp}(\gamma, 0, t_{end})$
 		\ENDFOR
@@
 		\STATE $\mu_k \leftarrow \beta I(t_{k-1})$
-    			\STATE $X_{\mathcal{I}_k} \sim \TruncExp(\mu_k, t_{k-1}, t_k)$ i.i.d.\
+		\STATE $X_{\mathcal{I}_k} \sim \textsf{TrunExp}(\mu_k, t_{k-1}, t_k)$ i.i.d.\
 
@@
 		\STATE $p_j \leftarrow 1 - \exp\{-\gamma(t_{end} - z^I_j)\}$
-        			\STATE $z^R_j \sim (1 - p_j) \delta_\infty + p_j \TruncExp(\gamma, z^I_j, t_{end})$ (removal times)
+		\STATE $z^R_j \sim (1 - p_j) \delta_\infty + p_j \textsf{TrunExp}(\gamma, z^I_j, t_{end})$ (removal times)
 		\ENDFOR
@@
 The following four characteristics of our DA-MCMC algorithm are worth noting.
-	First, initializing the Markov chain only requires values for the initial parameters $\theta^{(0)} = (\beta^{(0)}, \gamma^{(0)})$ since $\Z^{(0)}$ can be generated conditionally on $\Y$ and $\theta^{(0)}$ alone from the PD-SIR process.
+First, initializing the Markov chain only requires values for the initial parameters $\theta^{(0)} = (\beta^{(0)}, \gamma^{(0)})$ since $\mathbf{Z}^{(0)}$ can be generated conditionally on $\mathbf{Y}$ and $\theta^{(0)}$ alone from the PD-SIR process.
 Second, since the PD-SIR closely approximates the SIR model, the acceptance rate in the MH step for the latent data is typically high.
@@
 Smaller values for $\rho$ result in smaller jumps in the latent space and a larger MH acceptance rate, which may lead to better overall mixing in large populations.
-	Third, if all event times are updated ($\rho = 1$), then the proposed and current latent data are independent conditionally on the current values of the parameters: $\Z^{(k-1)} \perp \Z^* | \theta^{(k)}$.
-Our proposal scheme can therefore be said to be semi-independent, a characteristic that is crucial for proving the uniform ergodicity of the resulting Markov chain (see Section \eqref{sec:uni}).
+Third, if all event times are updated ($\rho = 1$), then the proposed and current latent data are independent conditionally on the current values of the parameters: $\mathbf{Z}^{(k-1)} \perp \mathbf{Z}^* | \theta^{(k)}$.
+Our proposal scheme can therefore be said to be semi-independent, a characteristic that is crucial for proving the uniform ergodicity of the resulting Markov chain in Section \ref{sec:uni}.
 This contrasts with existing DA-MCMC algorithms which only update a small fraction of the latent data per iteration and generate consecutive configurations of the latent data that are mostly identical.
 Moreover, since the dimension of the latent data remains constant across iterations in our approach, we do not need to rely of reversible jump MCMC unlike \cite{Gibson.1998} and \cite{ONeill.1999}.
-	%Fourth, since the latent data are generated from a process that only resembles the SIR model, we ensure that the Markov chain converges to the correct distribution $\pi(\theta, \Z|\Y)$ by proposing and accepting the latent data according to a Metropolis-Hastings scheme.
+%Fourth, since the latent data are generated from a process that only resembles the SIR model, we ensure that the Markov chain converges to the correct distribution $\pi(\theta, \mathbf{Z}|\mathbf{Y})$ by proposing and accepting the latent data according to a Metropolis-Hastings scheme.
 This results in exact Bayesian inference targeting the posterior distribution under the original SIR model.
 %Even though the latent spaces are generated from an approximation of the SIR process, the Metropolis-Hastings step ensures that the Markov chain converges to the posterior distribution of the parameters under the SIR model; in other words, the inference is exact.
-	Finally, not only is proposing latent data from the PD-SIR process extremely fast since all the necessary random variables can be generated with the inverse cumulative distribution function method, but ensuring that the proposed latent data are compatible with the observed data can be done at no additional cost, making the proposal scalable to large outbreaks.
+Finally, not only is proposing latent data from the PD-SIR process extremely fast since all the necessary random variables can be generated with the inverse cumulative distribution function method, but ensuring that the proposed latent data are compatible with the observed data is accomplished at no additional cost, making the proposal scalable to large outbreaks.
 
@@
 The seemingly piece-wise linear trajectory of the S compartment in the PD-SIR visible for small $K$ comes from it following a LDP with piece-wise constant death rate.
-In the limit as $K \rightarrow \infty$, the infection times are effectively known under the PD-SIR and the two processes are equivalent.
-	The fact that the PD-SIR faithfully approximates  the SIR, even for moderate $K$, yields a high acceptance rate in the MH step.
-As a result, our algorithm can make large jumps in the high-dimensional latent space and thus explores it efficiently.
+In the limit as $K \rightarrow \infty$, the infection times are effectively known under the PD-SIR and the two processes become equivalent.
+This observation that the PD-SIR faithfully approximates  the SIR, even for moderate $K$, provides intuition for why our method can yield a high acceptance rate in the MH step.
+As a result, the algorithm can make large jumps in the high-dimensional latent space and thus explores it efficiently.
+%In preliminary work, [Overivew Raphael work and put in preliminary results on Ebola]
 
 \begin{figure}
-		\centering
+	\begin{center}
 		\begin{subfigure}[b]{0.49\textwidth}
@@
 		\end{subfigure}
-		\caption{Trajectories of the compartments $S$, $I$ and $R$ in a SIR (in colors, with $S$ in red, $I$ in green and $R$ in blue) and PD-SIR processes (in grey) with the same infection incidence data $I_{1:K}$.}
-		\label{fig:comparison}
+	\end{center}
+	\caption{Compartment trajectories in a SIR (in colors, with $S$ in red, $I$ in green and $R$ in blue) and a PD-SIR process (in grey) with the same infection incidence data $I_{1:K}$.
+\label{fig:comparison}}
 \end{figure}
-	\smallskip
+%\smallskip
 
@@
 \label{sec:uni}
-	We now turn to analyze the convergence properties of the Markov chain $\left\{\left(\theta^{(m)}, \z^{(m)}\right)\right\}_m$ underlying the proposed MCMC algorithm.
-By construction, the joint posterior distribution $\pi(\theta, \z|\Y)$ is invariant for the Gibbs kernel $P_\theta$ that updates $\theta$ and for the MH kernel $P_\z$ updating $\z$, and is therefore also invariant for the composite kernel $P=P_\theta P_\z$ of the MCMC algorithm \cite{Tierney.1994}.
+We now turn to analyze the convergence properties of the Markov chain $\left\{\left(\theta^{(m)}, \mathbf{z}^{(m)}\right)\right\}_m$ underlying the proposed MCMC algorithm.
+By construction, the joint posterior distribution $\pi(\theta, \mathbf{z}|\mathbf{Y})$ is invariant for the Gibbs kernel $P_\theta$ that updates $\theta$ and for the MH kernel $P_\mathbf{z}$ updating $\mathbf{z}$, and is therefore also invariant for the composite kernel $P=P_\theta P_\mathbf{z}$ of the MCMC algorithm \citep{Tierney.1994}.
 Moreover, $P$ is strictly positive, implying that the chain is aperiodic and $\pi$-irreducible.
-Together with the existence of an invariant distribution, $\pi$-irreducibility implies that the chain is positive Harris recurrent and converges to $\pi(\theta, \z|\Y)$.
+Together with the existence of an invariant distribution, $\pi$-irreducibility implies that the chain is positive Harris recurrent and converges to $\pi(\theta, \mathbf{z}|\mathbf{Y})$.
 Hence, for any $\pi$-integrable function $g$, the ergodic theorem holds and the estimator
-	$$\bar{g}_m := \frac{1}{m} \sum_{i=0}^{m-1} g\left(\theta^{(m)}, \z^{(m)}\right)$$
+$$\bar{g}_m := \frac{1}{m} \sum_{i=0}^{m-1} g\left(\theta^{(m)}, \mathbf{z}^{(m)}\right)$$
 is consistent for $E_\pi g$, for any starting value.
@@
 where $\Vert \mu \Vert_{TV} := \sup_A \mu(A)$ denotes the total variation norm of a signed measure $\mu$.
-	Jones \cite{Jones.2004b} shows that uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$. That is,
+\cite{Jones.2004b} shows that uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$. That is,
 $$\sqrt{m}(\bar{g}_m - E_\pi g) \Rightarrow N(0, \sigma^2_g)$$
 for any initial distribution and for some positive, finite $\sigma^2_g$.
-The constant $\sigma^2_g$ can be estimated via regeneration sampling, batch means or a spectral variance analysis \cite{Flegal.2010}.
-
-	Toward establishing the uniform ergodicity of our DA-MCMC algorithm, we will show that the state space $\chi = \chi_{\theta}\times\chi_{\z}$ of the Markov chain $\left\{\left(\theta^{(m)}, \z^{(m)}\right)\right\}_m$ is a \textit{small set} for the transition kernel $P$; that is, there exists a probability measure $\nu$ on the $\sigma$-algebra $\sigma(\chi)$ such that
+The constant $\sigma^2_g$ can be estimated via regeneration sampling, batch means or a spectral variance analysis (see \cite{Flegal.2010}).
+
+Toward establishing the uniform ergodicity of our DA-MCMC algorithm, we will show that the state space $\chi = \chi_{\theta}\times\chi_{\mathbf{z}}$ of the Markov chain $\left\{\left(\theta^{(m)}, \mathbf{z}^{(m)}\right)\right\}_m$ is a \textit{small set} for the transition kernel $P$; that is, there exists a probability measure $\nu$ on the $\sigma$-algebra $\sigma(\chi)$ such that
 \begin{equation}
@@
 %	$$
-	Proposition \ref{pro:uni} shows that \eqref{eq:uni} holds for $m=1$ for our MCMC algorithm when $\rho=1$.
+Theorem \ref{pro:uni} shows that \eqref{eq:uni} holds for $m=1$ for our MCMC algorithm when $\rho=1$.
 This result is worth noting since the requirement that the whole space be small is extremely restrictive and is typically not satisfied for models with unbounded spaces.
 
-	\begin{proposition}
+\begin{theorem}
 	\label{pro:uni}
 	For $\rho = 1$, the transition kernel $P$ of the proposed DA-MCMC algorithm satisfies Equation \eqref{eq:uni} with $m=1$, that is,
-		$$P((\theta, \z), .) \ge \beta \nu(.) , \quad \forall (\theta, \z) \in \chi_\theta \times \chi_\z
+	$$P((\theta, \mathbf{z}), .) \ge \beta \nu(.) , \quad \forall (\theta, \mathbf{z}) \in \chi_\theta \times \chi_\mathbf{z}
 	$$
 	for some $\beta>0$ and probability measure $\nu$. As a result $P$ is uniformly ergodic.
-	\end{proposition}
-
-	We now provide a brief overview of the proof of Proposition \ref{pro:uni} provided in Appendix \ref{app:uni}.
-Since the kernel $P = P_{\theta} P_{\z}$ is composite, it suffices to minorize $P_\z$ and $P_\theta$ separately.
-The density of the Gibbs kernel $P_\z$ corresponds to the product of the two gamma densities in \eqref{eq:posterior_theta}; since the sufficient statistics of the latent data are bounded, these gamma densities possess a positive minorization whose closed form is derived in Appendix \ref{app:uni}.
-Furthermore, the MH kernel $P_\z$ depends on the current latent data only through the ratio $q(\z|\theta)/\pi(\z|\theta)$ which can be bounded above and away from zero.
-
-	Finally, it is easy to see that, if $\rho < 1$, Equation \eqref{eq:uni} holds for 
-	$m=\lfloor \rho^{-1} \rfloor$ and some $\epsilon$ and $\nu$.
+\end{theorem}
+
+We provide a brief overview of the proof of Theorem \ref{pro:uni}, with details provided in Appendix \ref{app:uni}.
+Since the kernel $P = P_{\theta} P_{\mathbf{z}}$ is composite, it suffices to minorize $P_\mathbf{z}$ and $P_\theta$ separately.
+The density of the Gibbs kernel $P_\mathbf{z}$ corresponds to the product of the two gamma densities in \eqref{eq:posterior_theta}; since the sufficient statistics of the latent data are bounded, these gamma densities possess a positive minorization whose closed form is derived in Appendix \ref{app:uni}.
+Furthermore, the MH kernel $P_\mathbf{z}$ depends on the current latent data only through the ratio $q(\mathbf{z}|\theta)/\pi(\mathbf{z}|\theta)$ which can be bounded above and away from zero.
+Finally, it is easy to see that, if $\rho < 1$, Equation \eqref{eq:uni} holds taking 
+$m=\lfloor \rho^{-1} \rfloor$ for some $\epsilon$ and $\nu$.
 
@@
 First, we assess the performance of the algorithm in a moderately sized population of $1000$ individuals.
-We simulate an epidemic with true parameters $(\beta, \gamma) = (0.0025, 1)$ -- giving $R_0 = 2.5$ -- starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$, when the outbreak has completed most of its course but is not over yet (see Figure \ref{fig:E1_trajectories}, where $I(t_{end}) = 122$).
+We simulate an epidemic with true parameters $(\beta, \gamma) = (0.0025, 1)$ --- giving $R_0 = 2.5$ --- starting with $(S(0), I(0)) = (1000, 10)$ until time $t_{end} = 6$, when the outbreak has completed most of its course but is not over yet, as depicted in Figure \ref{fig:E1_trajectories} where $I(t_{end}) = 122$.
 The numbers of infections in $K = 10$ equal-length time intervals are observed, corresponding to $I_{1:K} = (9, 14, 21, 42, 56, 121, 190, 162, 107, 73)$.
@@
 \begin{figure}
-		\centering
+	\begin{center}
 		\begin{subfigure}[b]{0.32\textwidth}
@@
 			\includegraphics[width=\textwidth]{E1_trajectories}
-			\caption{Compartment trajectories ($S$ in red, $I$ in green and $R$ in blue)}
+			\caption{Compartments ($S$ in red, $I$ in green and $R$ in blue)}
 			\label{fig:E1_trajectories}
@@
 		%\end{subfigure}
-		\caption{Performance of the DA-MCMC in a medium-sized population for the parameter $\gamma$.}
-		\label{fig:E1}
+	\end{center}
+	\caption{Performance of DA-MCMC in a medium-sized population for the parameter $\gamma$.
+		\label{fig:E1}}
 \end{figure}
 
-	The algorithm takes less than $30$ minutes to run $1$ million iterations on a personal laptop. 
+Remarkably, the algorithm takes less than $30$ minutes to run $1$ million iterations on a personal laptop. 
 Figure \ref{fig:E1_short_no_burn_gamma_tp} shows the traceplot of the parameter $\gamma$ for the first $10000$ iterations.
@@
 \begin{table}
-    \centering
+	\caption{Empirical coverage of $90\%$ posterior credible intervals and summary statistics of posterior means across $2000$ independent runs in a medium-sized population, $n=1000$.
+The true values of the parameters are $(\beta, \gamma, R_0) = (0.0025, 1, 2.5)$.
+%The standard errors for the observed coverage rate are calculated as $\sqrt{\hat{p}(1-\hat{p})/2000}$, which results in a largest standard error of $6.8$e-3.}
+		\label{tab:coverage}}
+	\begin{center}
 		\begin{tabular}{ C{2cm}| *{3}{C{3cm}}}
@@
 		\end{tabular}
-    \caption{Empirical coverage of the $90\%$ posterior credible intervals and distribution of the posterior means from $2000$ independent runs in a medium-sized population ($n=1000$).
-The true values of the parameters are $(\beta, \gamma, R_0) = (0.0025, 1, 2.5)$.
-The standard errors for the observed coverage rate are calculated as $\sqrt{\hat{p}(1-\hat{p})/2000}$, which results in a largest standard error of $6.8$e-3.}
-    \label{tab:coverage}
+	\end{center}
 \end{table}
@@
 Figures \ref{fig:E3_runtime} and \ref{fig:E3_accept} show the run time of the algorithm and the acceptance rate in the MH step.
-We observe that, as expecetd, the run time increases with the population size and with $\rho$, which reflects the fact that the number of random variables to generate per iteration is proportional to $\lceil\rho n\rceil$.
+We observe that, as expected, the run time increases with the population size and with $\rho$, which reflects the fact that the number of random variables to generate per iteration is proportional to $\lceil\rho n\rceil$.
 Moreover, the acceptance rate decreases as the number of trajectories updated per iteration $\lceil\rho n\rceil$ increases.
@@
 		\end{subfigure}
-		\caption{Impact of $\rho$ on the performance of the DA-MCMC in a population of size $250$ (solid), $500$ (dotted) and $1000$ (dashed).}
-		\label{fig:E3}
+	\end{center}
+	\caption{Impact of $\rho$ on the performance of the DA-MCMC in a population of size $250$ (solid), $500$ (dotted) and $1000$ (dashed).
+		\label{fig:E3}}
 \end{figure}
@@
 Finally, we showcase the efficacy of our joint proposal scheme through a comparison %the performance of the proposed DA-MCMC algorithm in which event times are jointly proposed 
-	to a SSU DA-MCMC that is similar in spirit to those in \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
-We do not consider particle filters \cite{King.2015} nor diffusion approximations \cite{Fintzi.2020} since these methods are not suited to situations in which the counts are observed without error.
+to a SSU DA-MCMC that is similar in spirit to those in \cite{Gibson.1998}, \cite{ONeill.1999} and \cite{Fintzi.2017}.
+We do not consider the particle filters of \cite{King.2015} nor the diffusion approximations of \cite{Fintzi.2020} since these methods are not suited to situations in which counts are observed without noise; see \cite{Ho.2018}.
 The SSU sampler that we consider here is a special case of our DA-MCMC in which $\rho = (S(0)+I(0))^{-1}$, so that the infection and removal times of a single individual are updated per iteration.
@@
 \begin{figure}
-		\centering
+	\begin{center}
 		\begin{subfigure}[b]{0.49\textwidth}
@@
 		\end{subfigure}
-		\caption{Traceplots and auto-correlation functions of the proposed DA-MCMC and the SSU DA-MCMC for $\beta$.}
-		\label{fig:E6}
+	\end{center}
+	\caption{Traceplots and auto-correlation functions of the proposed DA-MCMC and the SSU DA-MCMC for $\beta$.
+		\label{fig:E6}}
 \end{figure}
@@
 \begin{table}
-        \centering
+	\caption{Effective sample size per second for the proposed DA-MCMC and the single site update (SSU) DA-MCMC.
+		\label{tab:E6}}
+	\begin{center}
 		\begin{tabular}{ C{2cm}| *{2}{C{3cm}}}
@@
 		\end{tabular}
-        \caption{Effective sample size per second for the proposed DA-MCMC and the SSU DA-MCMC.}
-        \label{tab:E6}
+	\end{center}
 \end{table}
@@
 The virus, which has a fatality rate of $70\%$, was responsible for the death of almost $2000$ people in Guinea alone.
-	The outbreak is believed to have originated in the Gu\'eck\'edou prefecture of Guinea at the end of November $2013$ \cite{Baize.2014}.
+\cite{Coltart.2017} traced back the origin of the outbreak to the Gu\'eck\'edou prefecture of Guinea at the end of November $2013$.
 Weekly infection incidence counts are available for each prefecture for the $73$ weeks between the end of December $2013$ and May $2015$.
@@
 The units of time correspond to ``days" in the analysis and $t=0$ corresponds to Monday December 30, 2013, the first day of the observation period.
-	As the first documented infection occurred late November \cite{Baize.2014}, one month before the first reported incidence count, we set $I(0) = 5$.
+As the first documented infection occurred late November \cite{Coltart.2017}, one month before the first reported incidence count, we set $I(0) = 5$.
 
@@
 The Metropolis-Hastings step for the latent data proposals achieves a healthy $20.1\%$ acceptance rate, and the first 50,000 iterations of the Markov chain are discarded as a burn-in.
-	Figure \ref{fig:ebola} shows the marginal posterior distributions $\gamma$, whose means is $0.109$.
-This indicates that people remained infectious for around $9$ days on average, which is consistent with the existing literature.
-
-	%As noted by numerous authors, conditionally on partially observed data, the parameters $\beta$ and $\gamma$ appear positively correlated.
-	%The posterior distribution of $R_0$ is unimodal, relatively symmetric, and centered around $1$.
-A basic reproduction number close to $1$ is consistent with an outbreak that lasted several months without infecting the majority of the population.
+Figure \ref{fig:ebola} shows the marginal posterior distributions $\gamma^{-1}$ which indicates that people remained infectious for around $9$ days on average, which is consistent with the existing literature.
 
 \begin{figure}
-		\centering
+	\begin{center}
 		\begin{subfigure}[b]{0.45\textwidth}
@@
 		\end{subfigure}
-		\caption{Posterior distribution and traceplot of the expected infection length ($\gamma^{-1}$) for the Gu\'eck\'edou prefecture.}
-		\label{fig:ebola}
+	\end{center}
+	\caption{Posterior distribution and traceplot of the expected infection length ($\gamma^{-1}$) for the Gu\'eck\'edou prefecture.
+		\label{fig:ebola}}
 \end{figure}
@@
 The proposed method enables the classical Metropolis-Hastings algorithm to be an efficient way to conduct exact posterior inference under the stochastic SIR model, given only incidence data for infections.
-Existing attempts using Markov chain Monte Carlo either mix poorly or do not scale to populations of more than a few hundred individuals, while methods relying on forward simulation are limited to moderate-sized outbreaks and may suffer from degeneracy issues on a case-by-case basis depending on the amount of missing data.
-%Moreover, the simplifying assumptions in \cite{Fintzi.2020} or the approximate-Bayesian-computation framework of \cite{McKinley.2018}, which compromise targeting the exact posterior for computational reasons, generate biased estimates that may lead to spurious conclusions.
-	In contrast to methods that rely on approximate computation of simplifying assumptions necessitated by computational considerations \cite{McKinley.2018,Fintzi.2020}, 
-	our data-augmented algorithm enables fast and exact Bayesian inference, even for large outbreaks, and leverages a well-studied, transparent MCMC framework with  guarantees of uniform ergodicity.
-%Most existing inferential methods for epidemic stochastic models are only applicable to prevalence data.
-To our knowledge, our algorithm is the first that permits exact Bayesian inference given incidence data.
+%Existing attempts using Markov chain Monte Carlo either mix poorly or do not scale to populations of more than a few hundred individuals, while methods relying on forward simulation are limited to moderate-sized outbreaks and may suffer from degeneracy issues on a case-by-case basis depending on the amount of missing data.
+In contrast to methods that rely on approximate computation of simplifying assumptions necessitated by computational considerations such as \cite{McKinley.2018} and \cite{Fintzi.2020}, 
+our data-augmented algorithm enables exact and fast Bayesian inference, even for large outbreaks, and leverages a well-studied, transparent MCMC framework with  guarantees of uniform ergodicity.
 
@@
 Moreover, the piece-wise constant infection rate of the PD-SIR is only updated $K$ times, as opposed to after each event in the SIR model.
-Second, generating a PD-SIR process that is constrained to be compatible with the observed incidence data can be done at no additional cost; in contrast, generating a SIR process compatible with the observed data would be prohibitively slow \cite{Hobolth.2009}.
+Second, generating a PD-SIR process that is constrained to be compatible with the observed incidence data can be done at no additional cost; in contrast, directly generating a SIR process compatible with the observed data would be prohibitively slow \citep{Hobolth.2009}.
 Third, the dynamics of the PD-SIR process closely resemble those of the SIR process: the removal dynamics are identical and, for short observation time intervals, the infection dynamics are also very similar in the two processes.
-The last feature
-
-	The first two features make the algorithm extremely fast.
-While existing data-augmented MCMC algorithms have only be applied to populations of a few hundred of individuals, the analysis of the Ebola outbreak in Gu\'eck\'edou shows that our algorithm can be applied to populations of up to 300,000 individuals in a reasonable amount of time.
-	The latter feature of the PD-SIR process enables the algorithm to update a large portion of the augmented data in each iteration while maintaining a healthy acceptance rate.
+
+The first two features make the algorithm extremely fast and scalable for populations with hundreds of thousands of individuaals.
+%While existing data-augmented MCMC approaches have only be applied to populations of a few hundred of individuals, the analysis of the Ebola outbreak in Gu\'eck\'edou shows that our algorithm can be applied to populations of up to 300,000 individuals in a reasonable amount of time.
+Because the PD-SIR process enables the algorithm to update a large portion of the augmented data in each iteration while maintaining a healthy acceptance rate.
 As a result, the Markov chain makes large jumps in the latent space and has very good mixing properties.
-In contrast, existing DA-MCMC algorithms keep most of the latent space fixed across iterations \cite{Gibson.1998, ONeill.1999, Fintzi.2017} and their Markov chains therefore mix much more slowly.
-
-	If the acceptance rate in the Metropolis-Hastings step for the latent data is too low, which can happen in large populations, our algorithm possesses a tuning parameter $\rho$ that determines the proportion of individuals whose trajectory is updated per iteration.
+In contrast, existing DA-MCMC algorithms \citep{Gibson.1998, ONeill.1999, Fintzi.2017} keep most of the latent space fixed across iterations  and their Markov chains and therefore mix much more slowly.
+
+Our algorithm features a tuning parameter $\rho$ that determines the proportion of individuals whose trajectory is updated per iteration, which in turn affects the acceptance ratio.
 Larger values for $\rho$ enable the chain to make larger steps in the latent space but may result in an excessively low acceptance rate, while smaller values of $\rho$ result in a higher acceptance rate but constrain the chain to make small jumps.
-Depending on the size of the population, different values of the tuning parameter are optimal.
-To find a value that optimizes the mixing properties of the chain, one can use several short runs of the algorithm with different values for $\rho$ and select the value that yields the largest effective size per second.
+Depending on the size of the population, different values of the tuning parameter are optimal, and future work may seek theoretical insights to this end.
+To find a value that optimizes the mixing properties of the chain in practice, one can use several short runs of the algorithm with different values for $\rho$ and select the value that yields the largest effective size per second.
 
@@
 The DA-MCMC algorithm proposed in this article is specific to 
-	The stochastic SIR model, for which our algorithm is designed, is arguably a simplistic representation of the spread of disease.
-The model relies on assumptions such as perfect reporting, exponentially-distributed infectious periods, homogeneously mixing population and a constant infection rate.
-Our algorithm can be extended to models of increasing realism where these assumptions are relaxed.
-In future work, we will present applications to epidemic models with under-reporting \cite{Fintzi.2017}, non-Markovian dynamics \cite{Streftaris.2002}, non-homogeneous mixing \cite{Severo.1969} as well as time-varying infection rate \cite{Kypraios.2018}.
+the stochastic SIR model.
+However, one can argue that it is a simplistic representation of the spread of disease, relying on assumptions such as perfect reporting, exponentially-distributed infectious periods, homogeneously mixing population and a constant infection rate.
+Our data-augmentation framework and the idea of a faithful surrogate proposal can be extended to models of increasing realism where these assumptions are relaxed.
+It will be fruitful to consider extensions to epidemic models accounting for under-reporting \citep{Fintzi.2017,Morozova.2021}, non-Markovian dynamics \citep{Streftaris.2002}, non-homogeneous mixing \citep{Severo.1969,Lomeli.2021} as well as time-varying infection rates \citep{Kypraios.2018} in future work.
+We invite readers to consider applications of these ideas in other stochastic process models with complex latent spaces that may similarly be made navigable through designing tractable, mechanistic proposal processes.
 
 \appendix
+\begin{center} \Large{\bf Appendix}
+\end{center}
 
@@
 
-	Theorem \ref{theo:ldp} was first proved by \cite{Neuts.1971}. Ross \cite{Ross.1996} provides a simpler proof which we now give.
+Theorem \ref{theo:ldp} was first proved by \cite{Neuts.1971}. \cite{Ross.1996} provide a simpler proof which we now give.
 
@@
 	Consider a linear pure death process with $n$ particles and individual death rate $\mu$.
-	Let $T_i$ be the time of the $i$th death. Then $W_1 = T_1 \sim \Exp(n\mu)$ and $W_i = T_i - T_{i-1} \sim \Exp((n-i)\mu)$ independently. Let $N$ be the number of deaths by time $t$. Then, 
+	Let $T_i$ be the time of the $i$th death.
+Then $W_1 = T_1 \sim \textsf{Exp}(n\mu)$ and $W_i = T_i - T_{i-1} \sim \textsf{Exp}((n-i)\mu)$ independently.
+Let $N$ be the number of deaths by time $t$.
+Then,
 	\begin{align*}
@@
 
-	\section{Proof of Proposition \ref{pro:uni}}
+\section{Proof of Theorem \ref{pro:uni}}
 \label{app:uni}
 
-	We first present two lemmas that will be useful in the proof of Proposition \ref{pro:uni}.
+We first present two lemmas that will be useful in the proof of Theorem \ref{pro:uni}.
 % Gamma 1
@@
 		\min\{Ga(x;a+A,b), Ga(x;a,b+B)\}, & x_a^* \wedge x_b^*\le x < x_{a+A}^* \vee x_{b+B}^*\\
-		\end{cases}$$
+	\end{cases}.$$
 \end{lemma}
@@
 	\begin{table}[H]
-			\centering
+		\caption{Values of $(\alpha, \beta)$ that minimize $Ga(x;a+\alpha,b+\beta)$ for different values $x$.
+			\label{tab:ga9}}
+		\begin{center}
 			\begin{tabular}{l|c c c}
@@
 			\end{tabular}
-			\caption{Values of $(\alpha, \beta)$ that minimize $Ga(x;a+\alpha,b+\beta)$ for different values $x$.}
-			\label{tab:ga9}
+		\end{center}
 	\end{table}
@@
 		\end{subfigure}
-		\caption{Example of minorization of the density of a gamma distribution $Ga(2+\alpha, 0.5+\beta)$.}
-		\label{fig:gam}
+	\end{center}
+	\caption{Example of minorization of the density of a gamma distribution $Ga(2+\alpha, 0.5+\beta)$.
+		\label{fig:gam}}
 \end{figure}
@@
 %\noindent
-	We can now proceed with the proof of Proposition \ref{pro:uni}.
+We can now proceed with the proof of Theorem \ref{pro:uni}.
 
 \begin{proof}
-		The transition kernel $P = P_{\theta} P_{\z}$ is a composition of two kernels.
-The kernel $P_{\theta}$ updates the parameters $\theta$ while keeping the latent data $\z$ fixed, and $P_{\z}$ update $\z$ while keeping $\theta$ fixed.
+	The transition kernel $P = P_{\theta} P_{\mathbf{z}}$ is a composition of two kernels.
+The kernel $P_{\theta}$ updates the parameters $\theta$ while keeping the latent data $\mathbf{z}$ fixed, and $P_{\mathbf{z}}$ update $\mathbf{z}$ while keeping $\theta$ fixed.
 %A one-step transition therefore corresponds to the following scheme
-		%$$\x_1 = (\theta_1, \z_1) \rightarrow (\theta_2, \z_1) \rightarrow (\theta_2, \z_2) = \x_2.$$
+	%$$\mathbf{x}_1 = (\theta_1, \mathbf{z}_1) \rightarrow (\theta_2, \mathbf{z}_1) \rightarrow (\theta_2, \mathbf{z}_2) = \mathbf{x}_2.$$
 	We therefore have
-		$$P((\theta_1, \z_1), (d\theta_2, d\z_2)) = P_\theta(\theta_1, d\theta_2 | \z_1) P_\z(\z_1, d\z_2 | \theta_2)$$
+	$$P((\theta_1, \mathbf{z}_1), (d\theta_2, d\mathbf{z}_2)) = P_\theta(\theta_1, d\theta_2 | \mathbf{z}_1) P_\mathbf{z}(\mathbf{z}_1, d\mathbf{z}_2 | \theta_2)$$
 	where
@@
 		\label{eq:Pth}
-		    P_\theta(\theta_1, d\theta_2 | \z_1)
-		    & = \pi(d\theta_2 | \z_1) \\
-		    & = \pi(\beta_2 | \z_1) \pi(\gamma_2 | \z_1) d\beta_2 d\gamma_2 \nonumber
+		P_\theta(\theta_1, d\theta_2 | \mathbf{z}_1)
+		& = \pi(d\theta_2 | \mathbf{z}_1) \\
+		& = \pi(\beta_2 | \mathbf{z}_1) \pi(\gamma_2 | \mathbf{z}_1) d\beta_2 d\gamma_2 \nonumber
 	\end{align}
@@
 		\label{eq:Pz}
-		    P_\z(\z_1, d\z_2 | \theta_2)
-		    & = Q(d\z_2 | \theta_2) \alpha((\theta_2, \z_1),(\theta_2, \z_2)) + \delta_{\z_1}(d\z_2)\int(1-\alpha((\theta_2, \z_1),(\theta_2, \z_2))) Q(d\z_2| \theta_2) \nonumber \\
-			& \ge Q(d\z_2| \theta_2) \alpha((\theta_2, \z_1),(\theta_2, \z_2)) \\
-			& = q(\z_2| \theta_2) \min\left\lbrace 1, \dfrac{\pi(\theta_2, \z_2)q(\z_1| \theta_2)}{\pi(\theta_2, \z_1)q(\z_2| \theta_2)} \right\rbrace d\z_2 \nonumber
+		P_\mathbf{z}(\mathbf{z}_1, d\mathbf{z}_2 | \theta_2)
+		& = Q(d\mathbf{z}_2 | \theta_2) \alpha((\theta_2, \mathbf{z}_1),(\theta_2, \mathbf{z}_2)) + \delta_{\mathbf{z}_1}(d\mathbf{z}_2)\int(1-\alpha((\theta_2, \mathbf{z}_1),(\theta_2, \mathbf{z}_2))) Q(d\mathbf{z}_2| \theta_2) \nonumber \\
+		& \ge Q(d\mathbf{z}_2| \theta_2) \alpha((\theta_2, \mathbf{z}_1),(\theta_2, \mathbf{z}_2)) \\
+		& = q(\mathbf{z}_2| \theta_2) \min\left\lbrace 1, \dfrac{\pi(\theta_2, \mathbf{z}_2)q(\mathbf{z}_1| \theta_2)}{\pi(\theta_2, \mathbf{z}_1)q(\mathbf{z}_2| \theta_2)} \right\rbrace d\mathbf{z}_2 \nonumber
 	\end{align}
-		corresponds to a Metropolis-Hasting transition kernel in which a new configuration of the latent data is generated from the proposal kernel $Q$ conditionally on the current value of the parameters $\theta_2$, but independently of the current configuration $\z_1$.
+	corresponds to a Metropolis-Hasting transition kernel in which a new configuration of the latent data is generated from the proposal kernel $Q$ conditionally on the current value of the parameters $\theta_2$, but independently of the current configuration $\mathbf{z}_1$.
 The density $q$ of the proposal kernel $Q$ is given in \eqref{eq:q}.
@@
 		\label{eq:min}
-		    k(\theta_2, \z_2) \le \pi(\beta_2 | \z_1) \pi(\gamma_2 | \z_1) q(\z_2; \theta_2) \min\left\lbrace 1, \dfrac{\pi(\theta_2, \z_2)q(\z_1| \theta_2)}{\pi(\theta_2, \z_1)q(\z_2|\theta_2)} \right\rbrace
-		\end{equation}
-		for all $\theta_1 \in \chi_\theta$ and all $\z_1 \in \chi_\z.$
-		Indeed, suppose that we can find a positive function $k$ satisfying the inequality \eqref{eq:min}, then for any set $A$ and any $(\theta, \z)$ we have, by \eqref{eq:Pth}, \eqref{eq:Pz} and \eqref{eq:min},
+		k(\theta_2, \mathbf{z}_2) \le \pi(\beta_2 | \mathbf{z}_1) \pi(\gamma_2 | \mathbf{z}_1) q(\mathbf{z}_2; \theta_2) \min\left\lbrace 1, \dfrac{\pi(\theta_2, \mathbf{z}_2)q(\mathbf{z}_1| \theta_2)}{\pi(\theta_2, \mathbf{z}_1)q(\mathbf{z}_2|\theta_2)} \right\rbrace
+	\end{equation}
+	for all $\theta_1 \in \chi_\theta$ and all $\mathbf{z}_1 \in \chi_\mathbf{z}.$
+	Indeed, suppose that we can find a positive function $k$ satisfying the inequality \eqref{eq:min}, then for any set $A$ and any $(\theta, \mathbf{z})$ we have, by \eqref{eq:Pth}, \eqref{eq:Pz} and \eqref{eq:min},
 	\begin{align*}
-		    P((\theta, \z), A)
-		    & \ge \int_A \pi(\theta' | \z) \pi(\beta' | \z) \pi(\gamma' | \z) \min\left\lbrace 1, \dfrac{\pi(\theta', \z')q(\z| \theta')}{\pi(\theta', \z)q(\z'|\theta')} \right\rbrace d(\beta', \gamma', \z') \\
-		    & \ge \int_A k(\theta', \z') d(\theta', \z') \\
+		P((\theta, \mathbf{z}), A)
+		& \ge \int_A \pi(\theta' | \mathbf{z}) \pi(\beta' | \mathbf{z}) \pi(\gamma' | \mathbf{z}) \min\left\lbrace 1, \dfrac{\pi(\theta', \mathbf{z}')q(\mathbf{z}| \theta')}{\pi(\theta', \mathbf{z})q(\mathbf{z}'|\theta')} \right\rbrace d(\beta', \gamma', \mathbf{z}') \\
+		& \ge \int_A k(\theta', \mathbf{z}') d(\theta', \mathbf{z}') \\
 		& = \epsilon \nu(A)
 	\end{align*}
-		with $\epsilon = \int k(\theta, \z) d(\theta, \z)$ a positive constant and $\nu(A) = \epsilon^{-1} \int_A k(\theta, \z) d(\theta, \z)$ a probability measure.
+	with $\epsilon = \int k(\theta, \mathbf{z}) d(\theta, \mathbf{z})$ a positive constant and $\nu(A) = \epsilon^{-1} \int_A k(\theta, \mathbf{z}) d(\theta, \mathbf{z})$ a probability measure.
 
 	We now construct a positive function $k$ satisfying \eqref{eq:min}.
-The inequality depends on $(\theta_1, \z_1)$ only through the full conditional distributions $\pi(\beta_2 | \z_1)$ and $\pi(\gamma_2 | \z_1)$ and the ratio $\dfrac{q(\z_1| \theta_2)}{\pi(\theta_2, \z_1)}$.
-		It therefore suffices to find positive functions $k_\beta, k_\gamma$ and $k_r$ such that, for all $\z_1$,
-		$$k_\beta(\beta_2) \le \pi(\beta_2 | \z_1),
-		\quad k_\gamma(\gamma_2) \le \pi(\gamma_2 | \z_1),
-		\quad \text{and} \quad k_r(\theta_2) \le \dfrac{q(\z_1|\theta_2)}{\pi(\theta_2, \z_1)}.$$
-
-		First, we can obtain a closed form minorization of $\pi(\beta_2| \z_1)$ using Proposition \ref{pro:ga1} as follows,
+The inequality depends on $(\theta_1, \mathbf{z}_1)$ only through the full conditional distributions $\pi(\beta_2 | \mathbf{z}_1)$ and $\pi(\gamma_2 | \mathbf{z}_1)$ and the ratio $\dfrac{q(\mathbf{z}_1| \theta_2)}{\pi(\theta_2, \mathbf{z}_1)}$.
+	It therefore suffices to find positive functions $k_\beta, k_\gamma$ and $k_r$ such that, for all $\mathbf{z}_1$,
+	$$k_\beta(\beta_2) \le \pi(\beta_2 | \mathbf{z}_1),
+	\quad k_\gamma(\gamma_2) \le \pi(\gamma_2 | \mathbf{z}_1),
+	\quad \text{and} \quad k_r(\theta_2) \le \dfrac{q(\mathbf{z}_1|\theta_2)}{\pi(\theta_2, \mathbf{z}_1)}.$$
+
+	First, we obtain a closed form minorization of $\pi(\beta_2| \mathbf{z}_1)$ using Proposition \ref{pro:ga1} as follows,
 	\begin{align*}
-			\pi(\beta_2| \z_1) 
+		\pi(\beta_2| \mathbf{z}_1) 
 		& = Ga\left(\beta_2; a_{\beta} + n_I, b_{\beta} + \int_{0}^{t_{end}}S(t)I(t)dt\right) \\
@@
 	\end{align*}
-		since $n_I = \sum_k I_k$ is known and the sufficient statistic from $\z_1$, 
-		$\int_{0}^{t_{end}}S(t)I(t)dt$, is bounded between $0$ and $n (n_I+I_0)t_{end}$. Similarly, by Proposition \ref{pro:ga2}, we can minorize $\pi(\gamma_2| \z_1)$ as follows,
+	since $n_I = \sum_k I_k$ is known and the sufficient statistic from $\mathbf{z}_1$, 
+	$\int_{0}^{t_{end}}S(t)I(t)dt$, is bounded between $0$ and $n (n_I+I_0)t_{end}$. Similarly, by Proposition \ref{pro:ga2}, we minorize $\pi(\gamma_2| \mathbf{z}_1)$ as follows,
 	\begin{align*}
-			\pi(\gamma_2| \z_1) 
+		\pi(\gamma_2| \mathbf{z}_1) 
 		& = Ga\left(\gamma_2; a_{\beta} + n_R, b_{\beta} + \int_{0}^{t_{end}}I(t)dt\right)  \\
@@
 	\end{align*}
-		since the sufficient statistics from $\z_1$, $n_R$ and $\int_{0}^{t_{end}}I(t)dt$, are respectively bounded between $0$ and $n_I + I_0$, and between $0$ and $(n_T+I_0) t_{end}$.
-
-		Finally, the ratio $\dfrac{q(\z_1|\theta_2)}{\pi(\theta_2, \z_1)}$ can be minorized as follows, %\ram{(in the second inequality, minimize the whole numerator in terms of $I(.)$ instead of its each factors (=> use previous proof minimizing the proposal density); this will provide a better bound)}
+	since the sufficient statistics from $\mathbf{z}_1$, $n_R$ and $\int_{0}^{t_{end}}I(t)dt$, are respectively bounded between $0$ and $n_I + I_0$, and between $0$ and $(n_T+I_0) t_{end}$.
+
+	Finally, the ratio $\dfrac{q(\mathbf{z}_1|\theta_2)}{\pi(\theta_2, \mathbf{z}_1)}$ can be minorized as follows, %\ram{(in the second inequality, minimize the whole numerator in terms of $I(.)$ instead of its each factors (=> use previous proof minimizing the proposal density); this will provide a better bound)}
 	\begin{align*}
-		    \dfrac{q(\z_1|\theta_2)}{\pi(\theta_2, \z_1)} 
-		    & = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \TruncExp(z^I_j; \mu_k, t_{k-1}, t_k) \prod_{i=1}^n (1-p_i)^{\mathbf{1}\{z^R_i=\infty\}} \left(p_i \TruncExp(z^R_i;\gamma, z^I_i, t_{end})\right)^{\mathbf{1}\{z^R_i \le t_{end}\}}}
+		\dfrac{q(\mathbf{z}_1|\theta_2)}{\pi(\theta_2, \mathbf{z}_1)} 
+		& = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \textsf{TrunExp}(z^I_j; \mu_k, t_{k-1}, t_k) \prod_{i=1}^n (1-p_i)^{\mathbf{1}\{z^R_i=\infty\}} \left(p_i \textsf{TrunExp}(z^R_i;\gamma, z^I_i, t_{end})\right)^{\mathbf{1}\{z^R_i \le t_{end}\}}}
 		{\beta^{n_I} \prod_{l\in \mathcal{I}} I(z^I_l) \gamma^{n_R} \exp\{-\beta \int_0^{t_{end}} S(t)I(t)dt -\gamma \int_0^{t_{end}} I(t)dt\}} \\
-        	& = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \TruncExp(z^I_j; \mu_k, t_{k-1}, t_k)}
+		& = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \textsf{TrunExp}(z^I_j; \mu_k, t_{k-1}, t_k)}
 		{\prod_{l\in \mathcal{I}} \beta I(z^I_l) \exp\{-\beta \int_0^{t_{end}} S(t)I(t)dt\}}\\
@@
 	\begin{align*}
-        	& \dfrac{\prod_{i=1}^n (1-p_i)^{\mathbf{1}\{z^R_i=\infty\}} (p_i \TruncExp(z^R_i;\gamma, z^I_i, t_{end}))^{\mathbf{1}\{z^R_i \le t_{end}\}}}{\gamma^{n_R} \exp\{- \gamma \int_0^{t_end} I(t)dt\}} \\
+		& \dfrac{\prod_{i=1}^n (1-p_i)^{\mathbf{1}\{z^R_i=\infty\}} (p_i \textsf{TrunExp}(z^R_i;\gamma, z^I_i, t_{end}))^{\mathbf{1}\{z^R_i \le t_{end}\}}}{\gamma^{n_R} \exp\{- \gamma \int_0^{t_end} I(t)dt\}} \\
 		& = \dfrac{\prod_{j \in \mathcal{R}^c} \exp\{-\gamma(t_{end}-z^I_j)\} \prod_{k \in \mathcal{R}} (1-\exp\{\gamma(t_{end}-z^I_k)\}) \dfrac{\gamma \exp\{-\gamma (z^R_k - z^I_k)\}}{1-\exp\{\gamma(t_{end}-z^I_k)\}} }{\gamma^{n_R} \exp\{ \gamma \sum_{i=1}^n \min\{z^R_i, t_{end}\} - z^I_i\}} \\
@@
 	\begin{align*}
-		\pi(\beta_2 | \z_1) \pi(\gamma_2 | \z_1) q(\z_2; \theta_2) \min\left\lbrace 1, \dfrac{\pi(\theta_2, \z_2)q(\z_1| \theta_2)}{\pi(\theta_2, \z_1)q(\z_2
+		\pi(\beta_2 | \mathbf{z}_1) \pi(\gamma_2 | \mathbf{z}_1) q(\mathbf{z}_2; \theta_2) \min\left\lbrace 1, \dfrac{\pi(\theta_2, \mathbf{z}_2)q(\mathbf{z}_1| \theta_2)}{\pi(\theta_2, \mathbf{z}_1)q(\mathbf{z}_2
 			|\theta_2)} \right\rbrace
-		& \ge k_\beta(\beta_2) k_\gamma(\gamma_2) q(\z_2; \theta_2) \min\left\lbrace 1, k_r(\beta_2) \dfrac{\pi(\theta_2, \z_2)}{q(\z_2|\theta_2)} \right\rbrace \\
-		& = k(\theta_2, \z_2)
+		& \ge k_\beta(\beta_2) k_\gamma(\gamma_2) q(\mathbf{z}_2; \theta_2) \min\left\lbrace 1, k_r(\beta_2) \dfrac{\pi(\theta_2, \mathbf{z}_2)}{q(\mathbf{z}_2|\theta_2)} \right\rbrace \\
+		& = k(\theta_2, \mathbf{z}_2)
 	\end{align*}
-		for all $\theta_1 \in \chi_\theta$ and all $\z_1 \in \chi_\z$, which completes the proof.
+	for all $\theta_1 \in \chi_\theta$ and all $\mathbf{z}_1 \in \chi_\mathbf{z}$, which completes the proof.
 
@@
 
-	% Gamma 2
-	%can be used to minorize a density of the form
-	% \begin{equation}
-	% 	\label{eq:ga}
-	% 	Ga(x;a+\alpha,b+\beta) = \frac{(b+\beta)^{a+\alpha}}{\Gamma(a+\alpha)}x^{a+A-1}\exp\{-x(b+B)\}, \quad 0\le\alpha\le A, 0\le\beta\le B.
-	% \end{equation}
-	% over $\alpha$ and $\beta$ for each value of $x$. We can show the following.
-	% \begin{lemma}	
-	% 	\label{pro:ga2}
-	% 	For a fix $x>0$, the density $Ga(x;a+\alpha,b+\beta)$ is minimized by $(\alpha, \beta) \in \{(A, 0), (0, B)\}$, the minimizing set of values depending on $x$. In particular,	
-	% 	$$\inf_{\begin{aligned}
-	% 			0\le \alpha\le A \\ 0\le \beta\le B
-	% 	\end{aligned}}Ga(x;a+\alpha,b+\beta) = 
-	% 	\begin{cases}
-	% 		Ga(x;a+A,b)                     , & x<x_a         \text{ or } x<x_{a+A}^* \vee x_b^*\\
-	% 		Ga(x;a,b+B)                     , & x_{a+A}^* < x \text{ or } x_a^* \wedge x_{b+B}^* < x\\
-	% 		\min\{Ga(x;a+A,b), Ga(x;a,b+B)\}, & x_a^* \wedge x_b^*\le x < x_{a+A}^* \vee x_{b+B}^*\\
-	% 	\end{cases}$$
-	% \end{lemma}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-	% Truncated Exponential
-%	Proof of Lemma \ref{pro:tru}
-
-%	\begin{proof}
-%		Remember that $\TruncExp(u; \mu, 0, u) = \frac{\mu \exp\{-u\mu\}}{1-\exp\{-u\mu\}}$ is the density of an exponential distribution bounded above by $u$.
-%		\begin{align*}
-%			\frac{d}{d\mu}\TruncExp(u; \mu, 0, u)
-%			& = \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \left[ 1-\mu u - \mu u \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \right]
-%		\end{align*}
-%		Now let $g(\mu) = 1-\mu u - \mu u \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right)$. By L'H\^{o}pital's rule, $g(0) = 0$. Moreover, for $\mu>0$, we have
-%		\begin{align*}
-%			g'(\mu)
-%			& = - \mu \left(1+\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \left[ 1-u\mu \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \right]  \\
-%			& = - \mu \left(1+\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \left[ 1- \frac{u\mu}{\exp\{u\mu\}-1} \right] \\
-%			& \le 0
-%		\end{align*}
-%		where the inequality follows from $\frac{u\mu}{\exp\{u\mu\}-1}\le1$.
-%		This implies that $g(\mu) \le 0$ for $\mu>0$. We therefore have $\frac{d}{d\mu}\TruncExp(u; \mu, 0, u) \le 0$ and the result of the proposition follows.
-%	\end{proof}
-
-	\bibliographystyle{plain}
+%\begin{figure}
+%\begin{center}
+%\includegraphics[width=3in]{fig1.pdf}
+%\end{center}
+%\caption{Consistency comparison in fitting surrogate model in the tidal
+%power example. \label{fig:first}}
+%\end{figure}
+%
+%\begin{table}
+%\caption{D-optimality values for design $X$ under five different scenarios.  \label{tab:tabone}}
+%\begin{center}
+%\begin{tabular}{rrrrr}
+%one & two & three & four & five\\\hline
+%1.23 & 3.45 & 5.00 & 1.21 & 3.41 \\
+%1.23 & 3.45 & 5.00 & 1.21 & 3.42 \\
+%1.23 & 3.45 & 5.00 & 1.21 & 3.43 \\
+%\end{tabular}
+%\end{center}
+%\end{table}
+
+%\bigskip
+%\begin{center}
+%{\large\bf SUPPLEMENTARY MATERIAL}
+%\end{center}
+%
+%\begin{description}
+%
+%\item[Title:] Brief description. (file type)
+%
+%\item[R-package for  MYNEW routine:] R-package OMYNEWO containing code to perform the diagnostic methods described in the article.
+The package also contains all datasets used as examples in the article.
+(GNU zipped tar file)
+%
+%\item[HIV data set:] Data set used in the illustration of MYNEW method in Section~ 3.2. (.txt file)
+%
+%\end{description}
+
+\bibliographystyle{Chicago}
 \bibliography{bibliography}

### 2022-01-13 Manuscripts/Article/Article.tex
Updates from Overleaf

@@
+%\usepackage[utf8]{inputenc}
+%\usepackage[round,authoryear]{natbib}
+
 \documentclass[12pt]{article}
-\usepackage{amsmath}
+\usepackage[margin=1in]{geometry}
+\usepackage{url,hyperref}
 \usepackage{graphicx}
-\usepackage{enumerate}
-\usepackage{natbib}
-\usepackage{url} % not crucial - just used below for the URL
-
+\usepackage{amsmath,amssymb,array,eucal, amsthm}
+\linespread{1.5}
+\setlength\parindent{35pt}
+\input{../macros}
 \usepackage{algorithm}
 \usepackage{algorithmic}
+\usepackage{caption}
 \usepackage{subcaption}
-\usepackage{array}
-\usepackage{amsthm}
-
-%\pdfminorversion=4
-% NOTE: To produce blinded version, replace "0" with "1" below.
-\newcommand{\blind}{1}
-
-% DON'T change margins - should be 1 inch all around.
-\addtolength{\oddsidemargin}{-.5in}%
-\addtolength{\evensidemargin}{-1in}%
-\addtolength{\textwidth}{1in}%
-\addtolength{\textheight}{1.7in}%
-\addtolength{\topmargin}{-1in}%
-
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}{Lemma}[section]
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\usepackage{color}
 
@@
 \newcommand{\ram}[1]{{\color{green}{ #1}}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{proposition}{Proposition}[section]
+\newtheorem{lemma}{Lemma}[section]
+
 \begin{document}
 
-	%\bibliographystyle{natbib}
-
-	\def\spacingset#1{\renewcommand{\baselinestretch}%
-		{#1}\small\normalsize} \spacingset{1}
-
-	%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-	\if1\blind
+	% Keywords command
+	\providecommand{\keywords}[1]
 	{
-		\title{\bf Uniformly Ergodic Data-Augmented MCMC for Fitting the General Stochastic Epidemic Model to Incidence Data}
-		\author{Rapha\"{e}l Morsomme\thanks{
-				The authors gratefully acknowledge \textit{please remember to list all relevant funding sources in the unblinded version}}\hspace{.2cm}\\
-			Department of Statistical Science, University of Duke\\
-			and \\
-			Jason Xu \\
-			Department of Statistical Science, University of Duke}
+		\small	
+		\textbf{\textit{Keywords---}} #1
+	}
+
+	\title{%Efficient Likelihood-Based Inference for Fitting Stochastic Epidemic Models to Incidence Data via Data Augmentation
+		\large Uniformly Ergodic Data-Augmented MCMC for Fitting the General Stochastic Epidemic Model to Incidence Data}
+
+%	\author{
+%	Rapha\"{e}l Morsomme$^{1}$ \ and Jason Xu$^{1}$ \\ 
+%	\small %\texttt{raphael.morsomme@duke.edu} \qquad
+%	\texttt{jason.q.xu@duke.edu}\\
+%	\small $^{1}$Department of Statistical Science, Duke University \\
+%	}
+	\date{
+	}	
 	\maketitle
-	} \fi
-
-	\if0\blind
-	{
-		\bigskip
-		\bigskip
-		\bigskip
-		\begin{center}
-			{\LARGE\bf Uniformly Ergodic Data-Augmented MCMC for Fitting the General Stochastic Epidemic Model to Incidence Data}
-		\end{center}
-		\medskip
-	} \fi
-
-	\bigskip
+
 	\begin{abstract}
@@
 	\end{abstract}
-
-	\noindent%
-	{\it Keywords:}  Stochastic epidemic model; incidence data; exact Bayesian inference; likelihood-based inference
-	\vfill
-
-	\newpage
-	\spacingset{1.9} % DON'T change the spacing!
+	\keywords{Stochastic epidemic model; incidence data; exact Bayesian inference; likelihood-based inference}
 
 	\section{Introduction}
-\label{sec:intro}
+	% Outline: SEM; intractable likelihood; existing methods: approximation (Cauchemez 2008, Fintzi 2020), particle filtering, ABC, data augmentation; poor mixing;
 
 	The efficient control of a disease outbreak requires an understanding of the mechanisms underlying its spread.
-Mechanistic compartmental models, which describe the transition of individuals between various states, have a long mathematical modeling tradition in epidemiology dating back to \cite{Kermack.1927}.
+Mechanistic compartmental models, which describe the transition of individuals between various states, have a long mathematical modeling tradition in epidemiology \cite{Kermack.1927}.
 Due to their interpretability, they are commonly used to describe the dynamics of an outbreak and typically serve as the main source of information for predicting the course of an outbreak and identifying interventions that could be effective.
@@
 Moreover, given the size of the transition matrix, classical matrix exponentiation is intractable.
-Numerical methods to obtain transition probabilities have recently been developed in \cite{Ho.2018}, but the high computational costs limit their approach to moderate sized outbreaks.
+	Numerical methods to obtain transition probabilities have only recently been developed in \cite{Ho.2018b, Ho.2018}, but the high computational costs limit their approach to moderate sized outbreaks.
 
@@
 	The remainder of the article is structured as follows.
-Section \ref{sec:set} provides background information on the inference task.
+Section \eqref{sec:set} provides background information on the inference task.
 It introduces the stochastic SIR model and explains why conducting inference from partially observed data is a difficult task.
 Previous works addressing this problem are also presented.
-Section \ref{sec:pds} introduces the DA-MCMC algorithm and describes the surrogate process from which the latent data are generated in the Metropolis-Hastings step.
-The similarity between the surrogate and the target processes is assessed in Section \ref{sec:qua} and the DA-MCMC is shown to be uniformly ergodic in Section \ref{sec:uni}.
-In Section \ref{sec:sim}, we examine the performance of our algorithm on simulated data and then turn to an analysis of the 2013-2015 Ebola outbreak in Gu\'eck\'edou, Guinea, in Section \ref{sec:ebo}.
-Finally, Section \ref{sec:dis} discusses the findings and concludes the article.
+Section \eqref{sec:pds} introduces the DA-MCMC algorithm and describes the surrogate process from which the latent data are generated in the Metropolis-Hastings step.
+The similarity between the surrogate and the target processes is assessed in Section \eqref{sec:qua} and the DA-MCMC is shown to be uniformly ergodic in Section \eqref{sec:uni}.
+In Section \eqref{sec:sim}, we examine the performance of our algorithm on simulated data and then turn to an analysis of the 2013-2015 Ebola outbreak in Gu\'eck\'edou, Guinea, in Section \eqref{sec:ebo}.
+Finally, Section \eqref{sec:dis} discusses the findings and concludes the article.
 
@@
 	\label{sec:set}
+	%Outline: compartmental model; CTMC; rates; likelihood; inference (MLE, gamma conjugacy)
 
@@
 	\label{sec:sir}
-Our point of departure is to consider the stochastic SIR model introduced by \cite{Bailey.1975} -- also referred to as the general stochastic epidemic model -- which offers a parsimonious and interpretable representation of the mechanistic dynamics of an epidemic.
+	%Outline: stochastic SIR; X; parameters; transition rate12
+
+	Our point of departure is to consider the stochastic SIR model -- also referred to as the general stochastic epidemic model \cite{Bailey.1975} -- which offers a parsimonious and interpretable representation of the mechanistic dynamics of an epidemic.
 The stochastic SIR model is a compartmental model in which individuals transition through three states or compartments: susceptible (S), infectious (I) and removed (R).
@@
 	\end{cases}
-,\quad i = 0, \dots, n
+	,\quad \iton
 	$$
@@
 This parameter can be interpreted as the \textit{infection rate}: when a susceptible individual comes into contact with an infectious individual, she immediately becomes infectious.
-If we also make the common assumption that the infectious periods follow independent exponential distributions with rate $\gamma$, then the process \eqref{eq:X} is a time?homogeneous continuous time Markov chain, whose instantaneous transition rates
+	If we also make the common assumption that the infectious periods follow independent exponential distributions with rate $\gamma$, then the process \eqref{eq:X} is a time–homogeneous continuous time Markov chain, whose instantaneous transition rates
 	are given by the matrix $\Lambda = [\lambda_{\mathbf{x}, \mathbf{x}'}]$ with
@@
 
+	%Equivalently \jx{The following is not an exact equality, but needs the $o(dt)$ term or similar}
+	%	$$
+	%	P(\mathbf{X}_t = \mathbf{x}, \mathbf{X}_{t+dt} = \mathbf{x}') =
+	%	\begin{cases}
+	%		\beta I(t) dt + o(dt), & \mathbf{x} \text{ and } \mathbf{x}' \text{ only differ at position } j \text{ with } x_j = S \text{ and } x_j' = I, \\
+	%		\gamma dt + o(dt), & \mathbf{x} \text{ and } \mathbf{x}' \text{ only differ at position } j \text{ with } x_j = I \text{ and } x_j' = R, \\
+	%		o(dt), & \text{otherwise}
+	%	\end{cases}
+	%	$$
+	%	where $dt>0$ is small.
+
 	\subsection{Inference with Complete Data}
@@
 
-When the Markov process \eqref{eq:X} is completely observed until time $t_{end}$, we obtain the following likelihood \citep{Streftaris.2002}
+	When the Markov process \eqref{eq:X} is completely observed until time $t_{end}$, we obtain the following likelihood \cite{Streftaris.2002}
 	\begin{align}
@@
 
-It is straightforward to conduct inference in this continuously observed scenario %\cite{Guttorp.2018}
-. The likelihood \eqref{eq:cdl} belongs to the exponential family, and maximum likelihood estimates of its parameters can be expressed in terms of the sufficient statistics defined above as 
+	It is straightforward to conduct inference in this continuously observed scenario \cite{Guttorp.2018}.
+The likelihood \eqref{eq:cdl} belongs to the exponential family, and maximum likelihood estimates of its parameters can be expressed in terms of the sufficient statistics defined above as
 	$$
@@
 	\end{align}
-We have observed that the parameterization $\tilde{\theta}$ results in a Markov chain that mixes better and therefore employ it in the simulation studies and data analysis of Section \ref{sec:per}.
+	We have observed that the parameterization $\tilde{\theta}$ results in a Markov chain that mixes better and therefore employ it in the simulation studies and data analysis of Section \eqref{sec:per}.
+
+	%	\begin{equation}
+%	    \label{eq:lik2}
+%	    L(\mathbf{X}|\tilde{\theta}) = \beta^{n_I + n_R} R_0^{-n_R} S(0)^{n_R} \prod_{j\in\mathcal{I}} I(\tau^I_j) \exp\left\lbrace - \beta \int_{0}^{t_{end}} I(t)S(t) - \frac{\beta S(0)}{R_0} \int_{0}^{t_{end}} I(t) dt \right\rbrace,
+%	\end{equation}
 
@@
 	In practice, inference is complicated by the fact that the epidemic process \eqref{eq:X} is only partially observed.
-When some event times are unobserved, we do not have access to the sufficient statistics needed to evaluate the likelihood \eqref{eq:cdl}.
-Various types of partially observed data typify real data and have been considered in the literature.
-These include the removal times in \cite{Gibson.1998} and \cite{ONeill.1999} or the number of infectious individuals at discrete points in time in \cite{Fintzi.2017}.
+When we do not observe all event times, we do not have access to the sufficient statistics needed to evaluate the likelihood \eqref{eq:cdl}.
+Various types of partially observed data typify real data and have been considered in the literature, such as observing only the removal times \cite{Gibson.1998, ONeill.1999} or the number of infectious individuals at discrete points in time \cite{Fintzi.2017}.
 For instance, the former arises in animal experiments in which positive cases are immediately isolated from the rest of the population and no longer contribute to disease spread.
@@
 
-This form of observed data is motivated by the $2013$-$2015$ outbreak of the Ebola virus, the largest outbreak since the discovery of the virus in 1976, which resulted in at least 11,000 deaths, the majority of which occurred in three Western African countries (Guinea, Liberia and Sierra Leone).
+	This form of observed data is motivated by the $2013$–$2016$ outbreak of the Ebola virus, the largest outbreak since the discovery of the virus in 1976, which resulted in at least 11,000 deaths, mainly in three Western African countries (Guinea, Liberia and Sierra Leone).
 The data available for this outbreak consist of weekly number of positive tests.
 This outbreak was characterized by the large number of infections that occurred after individuals were tested positive for the virus, giving rise to noisy incidence data.
-\cite{Coltart.2017} reports that numerous infections took place at the hospitals where infected individuals received treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased during funerals is a tradition.
-Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.}.
-We therefore model the weekly number of positive tests as the weekly number of infections in a SIR model.
+Numerous infections took place at the hospitals where infected individuals received treatment as well as at the funerals of people deceased from the virus\footnote{In most regions impacted by the Ebola virus, touching the body of the deceased during funerals is a tradition.
+Since the Ebola virus is transmitted via bodily fluids such as sweat, numerous infections occurred at funerals.} \cite{Coltart.2017}.
+We therefore decide to model the weekly number of positive tests as the weekly number of infections in a SIR model.
 
@@
 		\propto \pi(\theta) L(\theta; \mathbf{Y}) = \pi(\theta) \int_{\chi_\mathbf{x}} L(\theta; \mathbf{x}) \delta_{\mathbf{Y}}(\mathbf{x}) d\mathbf{x}
-	%\mathbf{1}\{\Y \overset{\Delta}{=} \mathbf{x}\}
+		%\mathbf{1}\{\mathbf{Y} \overset{\Delta}{=} \mathbf{x}\}
 	\end{equation}
@@
 	Researchers have explored several approaches for conducting inference on partially observed stochastic epidemic processes.
-Early approaches include the use of martingale-based equations \citep{Becker.1977, Watson.1981}.
+Early approaches include the use of martingale-based equations \cite{Becker.1977, Watson.1981, Sudbury.1985}.
 However, such methods are difficult to apply to dynamic models with partially observed data and are therefore not suitable to the stochastic SIR model with incomplete data.
 	More recently, researchers have based their computation on a simpler process that approximates the model's dynamics and whose likelihood in the presence of partially observed data is tractable.
-Popular approximations include chain binomial models \citep{Greenwood.1931, Abbey.1952}, diffusion processes \citep{Cauchemez.2008, Fintzi.2020} and Gaussian processes \citep{Jandarov.2014}.
+Popular approximations include chain binomial models \cite{Greenwood.1931, Abbey.1952}, diffusion processes \cite{Cauchemez.2008, Fintzi.2020} and Gaussian processes \cite{Jandarov.2014}.
 While these approximation-based approaches bypass the intractability of the partial data likelihood \eqref{eq:pdl}, the assumptions on which they are based are questionable when the population is small, and, as a result, the dynamic of the stochastic process differs from its asymptotic behavior.
-
-Since the popularization of Markov chain Monte Carlo (MCMC) methods in statistics by \cite{Gelfand.1990} and \cite{Tierney.1994}, researchers have developed sampling-based methods to directly work with the SIR likelihood instead of an approximation thereof.
+	%Recently, \cite{Ho.2018} proposed an exact analytical method to directly compute the likelihood of the stochastic SIR process given discretely observed data.
+
+	Since the popularization of Markov chain Monte Carlo (MCMC) methods in statistics \cite{Tanner.1987, Gelfand.1990, Tierney.1994}, researchers have developed sampling-based methods to directly work with the SIR likelihood instead of an approximation thereof.
 MCMC algorithms fall into two categories: model-based forward simulation and data-augmented MCMC (DA-MCMC).
-Particle filtering in \cite{King.2015} is an example of the former category that is popular among practitioners.
+	Particle filtering \cite{King.2015} is an example of the former category that is popular among practitioners.
 Its plug-and-play feature makes it applicable to a wide variety of models.
 However, model-based forward simulation suffers from two drawbacks: simulating data from a model as complex as the SIR that is compatible with the observed data is prohibitively slow, and these methods can fail to converge when the model does not fit the data well.
-The approximate Bayesian computation (ABC) framework in \cite{McKinley.2018} offers a solution to the latter problem, but its inference is based on an approximation of the model's likelihood.
+The approximate Bayesian computation (ABC) framework  \cite{McKinley.2018} offers a solution to the latter problem, but its inference is based on an approximation of the model's likelihood.
 As a result, the inference can be biased, and it is difficult to evaluate the degree of approximation involved.
@@
 	Researchers have mostly focused on the development of more efficient proposal schemes to explore this high-dimensional latent space.
-\cite{Gibson.1998} and \cite{ONeill.1999} employed the reversible-jump MCMC framework introduced by \cite{Green.1995} to explore models with different numbers of unobserved events.
+Early attempts employed reversible-jump MCMC \cite{Green.1995} to explore models with different numbers of unobserved events \cite{Gibson.1998, ONeill.1999}.
 These authors augment the observed data which consist of the recovery times with the unobserved infection times and explore the latent space by uniformly inserting, deleting or moving an element of the latent data per iteration of the MCMC algorithm.
-More recently, \cite{Bu.2020} proposed a sampler for epidemics over networks when the infection times are assumed to be known, and \cite{Fintzi.2017} designed a sampler to conduct inference with discretely observed prevalence counts, with latent data consisting of the infection and recovery times of each individual.
-Their algorithms explore the latent spaces by updating the event times of a single individual per iteration of the MCMC algorithm.
+	More recently, Bu et al.\ \cite{bu2020likelihood} propose a sampler for epidemics over networks when the infection times are assumed to be known, and Fintzi et al.\ \cite{Fintzi.2017} proposed an MCMC algorithm to conduct inference with discretely observed prevalence counts, with latent data consisting of the infection and recovery times of each individual.
+The algorithm explores this latent space by updating the event times of an individual per iteration of the algorithm.
 %These event times are sampled from a distribution that approximate the exact dynamics of the epidemic process.
-In the discrete-time setting, \cite{Touloupou.2020} constructed a Gibbs sampler to update the trajectory of an individual per iteration.
+	In the discrete-time setting, Touloupou et al.\ \cite{Touloupou.2020} used a Gibbs sampler to update the trajectory of an individual per iteration.
 	These DA-MCMC methods suffer from poor mixing in the presence of large epidemics.
@@
 
-As shown in Section \ref{sec:icd}, the complete data likelihood \eqref{eq:cdl} is amenable to computation.
-This suggests augmenting the observed data $\mathbf{Y}$ with latent data $\mathbf{Z}$ such that the likelihood $L(\theta; (\mathbf{Y}, \mathbf{Z}))$ has the closed form \eqref{eq:cdl}, and constructing an ergodic Markov chain $\{(\theta^{(m)}, \mathbf{Z}^{(m)})\}_{m=0}^M$ on $\chi = \chi_{\theta} \times \chi_{\mathbf{z}}$ whose limiting distribution is the joint posterior $\pi(\theta, \mathbf{X}|\mathbf{Y})$ %\cite{Gibson.1998, ONeill.1999, Fintzi.2017}
-.
+	As shown in Section \eqref{sec:icd}, the complete data likelihood \eqref{eq:cdl} is amenable to computation.
+This suggests augmenting the observed data $\mathbf{Y}$ with latent data $\mathbf{Z}$ such that the likelihood $L(\theta; (\mathbf{Y}, \mathbf{Z}))$ has the closed form \eqref{eq:cdl}, and constructing an ergodic Markov chain $\{(\theta^{(m)}, \mathbf{Z}^{(m)})\}_{m=0}^M$ on $\chi = \chi_{\theta} \times \chi_{\mathbf{z}}$ whose limiting distribution is the joint posterior $\pi(\theta, \mathbf{X}|\mathbf{Y})$ \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
 Given the draws $\{(\theta^{(m)}, \mathbf{Z}^{(m)})\}_{m=0}^M$, the sequence $\{\theta^{(m)}\}_{m=0}^M$ provides an approximation of the marginal posterior distribution of interest $\pi(\theta|\mathbf{Y})$.
@@
 	Instead, we generate the latent data conditionally on $\mathbf{Y}$ from a surrogate process, and accept or reject them in a Metropolis-Hastings (MH) step.
-Given some current $\theta$ and $\mathbf{z}$, the MH step proceeds by simulating a candidate $\mathbf{z}^\star$ from a surrogate process $Q(.|\theta)$ with density $q(.|\theta)$, which we present in Section \ref{sec:pds}.
+Given some current $\theta$ and $\mathbf{z}$, the MH step proceeds by simulating a candidate $\mathbf{z}^\star$ from a surrogate process $Q(.|\theta)$ with density $q(.|\theta)$, which we present in Section \eqref{sec:pds}.
 Then $\mathbf{z}^\star$ is accepted with probability
@@
 	\end{equation}
-and otherwise the current $\mathbf{z}$ is retained.
+	and otherwise the current $\mathbf{z}$ is retained \cite{Tierney.1994}.
 Note that it is not necessary to compute $\delta_{\mathbf{Y}}(\mathbf{z})$ in the MH acceptance ratio \eqref{eq:alpha} since $\mathbf{z}^\star$ is compatible with $\mathbf{Y}$ by construction.
@@
 As shown in Figure \ref{fig:mu}, the infection rate $\tilde{\mu}$ and the variable $I$ are decoupled in each interval, and the infection rate is reset at the beginning %or left endpoint
-of the intervals. Over a single interval, the PD-SIR process is equivalent to a recent two-type branching process approximation of the SIR dynamics proposed in \cite{Ho.2018b}.
+	of the intervals. Over a single interval, the PD-SIR process is equivalent to a recent two-type branching process approximation of the SIR dynamics  \cite{Ho.2018}.
 
 	\begin{figure}
-	\begin{center}
-		\includegraphics[width=3in]{infection_rate_SIR_PDSIR}
-	\end{center}
+		\centering
+		\includegraphics[scale = 0.4]{infection_rate_SIR_PDSIR.jpg}
 		\caption{Infection dynamics of the SIR (dashed line) and PD-SIR (solid line) processes in a small population ($(S(0), I(0)) = (10,2)$) with $\beta=1$.
 The schedule of the PD-SIR is $t_{0:3}=(0, 0.2, 0.4, 0.6)$.
-		Vertical dotted lines indicate infection (red) and removal (blue) times. \label{fig:mu}}
+		Vertical dotted lines indicate infection (red) and removal (blue) times.}
+		\label{fig:mu}
 	\end{figure}
@@
 By construction, under the PD-SIR model the variable $S(t)$ follows a linear death process (LDP) during each interval with death rate $\mu_k = \beta I(t_{k-1})$.
-The LDP is a counting process possessing the so-called order statistics property \citep{Neuts.1971}; that is, given the number of events registered on some interval, the event times are distributed as the order statistics of independent and identically distributed (i.i.d.) random variables following, in the case of the LDP, a exponential distribution truncated to that interval.
+	The LDP is a counting process possessing the so-called order statistics property \cite{Neuts.1971}; that is, given the number of events registered on some interval, the event times are distributed as the order statistics of independent and identically distributed (i.i.d.) random variables following, in the case of the LDP, a exponential distribution truncated to that interval.
 	\begin{theorem}
@@
 	Appendix \ref{app:ldp} contains the proof of this theorem.
-Generating values from a truncated exponential distribution can be done extremely efficiently via the inverse cumulative distribution function method.
+Generating values from a truncated exponential distribution can be done extremely efficiently via the inverse cumulative distribution function method \cite{Devroye.2006}.
 
@@
 Algorithm \ref{alg:PD-SIR} provides a simple recursion to compute this variable efficiently.
-For $j \in \mathcal{I}_k$, following Theorem \ref{theo:ldp}, we sample the infection times $z^I_j \sim \textsf{TrunExp}(\mu_k; t_{k-1}, t_k)$, a truncated exponential distribution with rate $\mu_k$ bounded between $t_{k-1}$ and $t_k$.
+For $j \in \mathcal{I}_k$, following Theorem \ref{theo:ldp}, we sample the infection times $z^I_j \sim \textsf{TruncExp}(\mu_k; t_{k-1}, t_k)$, a truncated exponential distribution with rate $\mu_k$ bounded between $t_{k-1}$ and $t_k$.
 For the same indices $j$, we then generate the removal times $z^R_j$ from the removal dynamics of the SIR.
 To accomplish this, we sample the removal time of individual $j$ from the mixed distribution
-$$z^R_j|z^I_j \sim (1 - p_j) \delta_{\infty} + p_j \textsf{TrunExp}(\gamma; z^I_j, t_{end})$$
+	$$z^R_j|z^I_j \sim (1 - p_j) \delta_{\infty} + p_j \textsf{TruncExp}(\gamma; z^I_j, t_{end})$$
 	placing point mass $(1 - p_j)$ at $\infty$ and continuous mass on the interval $(z^I_j, t_{end}]$,
@@
 		q(\mathbf{z}|\theta) 
-	& = && \prod_{k=1}^K \prod_{j\in\mathcal{I}_k} \textsf{TrunExp}(z^I_j; \mu_{k}, t_{k-1}, t_{k}) \\
-	& && \times \prod_{i=1}^{n} \left( 1 - p_i \right)^{\mathbf{1}(z^R_i = \infty)} \left( p_i \textsf{TrunExp}(z^R_i; \gamma, z^I_i, t_{end}) \right)^{\mathbf{1}(z^R_i \le t_{end})}  \nonumber
+		& = && \prod_{k=1}^K \prod_{j\in\mathcal{I}_k} \textsf{TruncExp}(z^I_j; \mu_{k}, t_{k-1}, t_{k}) \\
+		& && \times \prod_{i=1}^{n} \left( 1 - p_i \right)^{\mathbf{1}(z^R_i = \infty)} \left( p_i \textsf{TruncExp}(z^R_i; \gamma, z^I_i, t_{end}) \right)^{\mathbf{1}(z^R_i \le t_{end})}
 	\end{alignat}
 	where %$\mu_k = \beta I(t_{k-1})$ and
-$$\textsf{TrunExp}(x; \mu, l, u) = \dfrac{\mu \exp\{-\mu x\}}{\exp\{-\mu l\} - \exp\{-\mu u\}}, \quad x \in (l, u)$$
+	$$\textsf{TruncExp}(x; \mu, l, u) = \dfrac{\mu \exp\{-\mu x\}}{\exp\{-\mu l\} - \exp\{-\mu u\}}, \quad x \in (l, u)$$
 	denotes the density of a truncated exponential distribution with parameters as notated previously.
@@
     			\STATE $p_j \leftarrow 1 - \exp\{-\gamma (t_{end} - 0)\}$
-		\STATE $z^R_j \sim (1 - p_j) \delta_\infty + p_j \textsf{TrunExp}(\gamma, 0, t_{end})$
+    			\STATE $z^R_j \sim (1 - p_j) \delta_\infty + p_j \textsf{TruncExp}(\gamma, 0, t_{end})$
 			\ENDFOR
@@
     			\STATE $\mu_k \leftarrow \beta I(t_{k-1})$
-		\STATE $X_{\mathcal{I}_k} \sim \textsf{TrunExp}(\mu_k, t_{k-1}, t_k)$ i.i.d.\
+    			\STATE $X_{\mathcal{I}_k} \sim \textsf{TruncExp}(\mu_k, t_{k-1}, t_k)$ i.i.d.\
 
@@
         			\STATE $p_j \leftarrow 1 - \exp\{-\gamma(t_{end} - z^I_j)\}$
-		\STATE $z^R_j \sim (1 - p_j) \delta_\infty + p_j \textsf{TrunExp}(\gamma, z^I_j, t_{end})$ (removal times)
+        			\STATE $z^R_j \sim (1 - p_j) \delta_\infty + p_j \textsf{TruncExp}(\gamma, z^I_j, t_{end})$ (removal times)
     			\ENDFOR
@@
 	Third, if all event times are updated ($\rho = 1$), then the proposed and current latent data are independent conditionally on the current values of the parameters: $\mathbf{Z}^{(k-1)} \perp \mathbf{Z}^* | \theta^{(k)}$.
-Our proposal scheme can therefore be said to be semi-independent, a characteristic that is crucial for proving the uniform ergodicity of the resulting Markov chain in Section \ref{sec:uni}.
+Our proposal scheme can therefore be said to be semi-independent, a characteristic that is crucial for proving the uniform ergodicity of the resulting Markov chain in Section \eqref{sec:uni}.
 This contrasts with existing DA-MCMC algorithms which only update a small fraction of the latent data per iteration and generate consecutive configurations of the latent data that are mostly identical.
@@
 	\begin{figure}
-	\begin{center}
+		\centering
 		\begin{subfigure}[b]{0.49\textwidth}
@@
 		\end{subfigure}
-	\end{center}
-	\caption{Compartment trajectories in a SIR (in colors, with $S$ in red, $I$ in green and $R$ in blue) and a PD-SIR process (in grey) with the same infection incidence data $I_{1:K}$.
-\label{fig:comparison}}
+		\caption{Trajectories of the compartments $S$, $I$ and $R$ in a SIR (in colors, with $S$ in red, $I$ in green and $R$ in blue) and PD-SIR processes (in grey) with the same infection incidence data $I_{1:K}$.}
+		\label{fig:comparison}
 	\end{figure}
-%\smallskip
+	\smallskip
 
@@
 	We now turn to analyze the convergence properties of the Markov chain $\left\{\left(\theta^{(m)}, \mathbf{z}^{(m)}\right)\right\}_m$ underlying the proposed MCMC algorithm.
-By construction, the joint posterior distribution $\pi(\theta, \mathbf{z}|\mathbf{Y})$ is invariant for the Gibbs kernel $P_\theta$ that updates $\theta$ and for the MH kernel $P_\mathbf{z}$ updating $\mathbf{z}$, and is therefore also invariant for the composite kernel $P=P_\theta P_\mathbf{z}$ of the MCMC algorithm \citep{Tierney.1994}.
+By construction, the joint posterior distribution $\pi(\theta, \mathbf{z}|\mathbf{Y})$ is invariant for the Gibbs kernel $P_\theta$ that updates $\theta$ and for the MH kernel $P_\mathbf{z}$ updating $\mathbf{z}$, and is therefore also invariant for the composite kernel $P=P_\theta P_\mathbf{z}$ of the MCMC algorithm \cite{Tierney.1994}.
 Moreover, $P$ is strictly positive, implying that the chain is aperiodic and $\pi$-irreducible.
@@
 	where $\Vert \mu \Vert_{TV} := \sup_A \mu(A)$ denotes the total variation norm of a signed measure $\mu$.
-\cite{Jones.2004b} shows that uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$. That is,
+	Jones \cite{Jones.2004b} shows that uniform ergodicity ensures the existence of a Central Limit Theorem for $\bar{g}_m$ whenever $E_\pi |g|^2 < \infty$. That is,
 	$$\sqrt{m}(\bar{g}_m - E_\pi g) \Rightarrow N(0, \sigma^2_g)$$
 	for any initial distribution and for some positive, finite $\sigma^2_g$.
-The constant $\sigma^2_g$ can be estimated via regeneration sampling, batch means or a spectral variance analysis (see \cite{Flegal.2010}).
+The constant $\sigma^2_g$ can be estimated via regeneration sampling, batch means or a spectral variance analysis \cite{Flegal.2010}.
 
@@
 	Finally, it is easy to see that, if $\rho < 1$, Equation \eqref{eq:uni} holds taking 
-$m=\lfloor \rho^{-1} \rfloor$ for some $\epsilon$ and $\nu$.
+	$m=\lfloor \rho^{-1} \rfloor$ and for some $\epsilon$ and $\nu$.
 
@@
 	\begin{figure}
-	\begin{center}
+		\centering
 		\begin{subfigure}[b]{0.32\textwidth}
@@
 		%\end{subfigure}
-	\end{center}
-	\caption{Performance of DA-MCMC in a medium-sized population for the parameter $\gamma$.
-		\label{fig:E1}}
+		\caption{Performance of DA-MCMC in a medium-sized population for the parameter $\gamma$.}
+		\label{fig:E1}
 	\end{figure}
@@
     \begin{table}
-	\caption{Empirical coverage of $90\%$ posterior credible intervals and summary statistics of posterior means across $2000$ independent runs in a medium-sized population, $n=1000$.
-The true values of the parameters are $(\beta, \gamma, R_0) = (0.0025, 1, 2.5)$.
-%The standard errors for the observed coverage rate are calculated as $\sqrt{\hat{p}(1-\hat{p})/2000}$, which results in a largest standard error of $6.8$e-3.}
-		\label{tab:coverage}}
-	\begin{center}
+    \centering
     \begin{tabular}{ C{2cm}| *{3}{C{3cm}}}
@@
     \end{tabular}
-	\end{center}
+    \caption{Empirical coverage of $90\%$ posterior credible intervals and summary statistics of posterior means across $2000$ independent runs in a medium-sized population, $n=1000$.
+The true values of the parameters are $(\beta, \gamma, R_0) = (0.0025, 1, 2.5)$.} %The standard errors for the observed coverage rate are calculated as $\sqrt{\hat{p}(1-\hat{p})/2000}$, which results in a largest standard error of $6.8$e-3.}
+    \label{tab:coverage}
     \end{table}
@@
 		\end{subfigure}
-	\end{center}
-	\caption{Impact of $\rho$ on the performance of the DA-MCMC in a population of size $250$ (solid), $500$ (dotted) and $1000$ (dashed).
-		\label{fig:E3}}
+		\caption{Impact of $\rho$ on the performance of the DA-MCMC in a population of size $250$ (solid), $500$ (dotted) and $1000$ (dashed).}
+		\label{fig:E3}
 	\end{figure}
@@
 	Finally, we showcase the efficacy of our joint proposal scheme through a comparison %the performance of the proposed DA-MCMC algorithm in which event times are jointly proposed 
-to a SSU DA-MCMC that is similar in spirit to those in \cite{Gibson.1998}, \cite{ONeill.1999} and \cite{Fintzi.2017}.
-We do not consider the particle filters of \cite{King.2015} nor the diffusion approximations of \cite{Fintzi.2020} since these methods are not suited to situations in which counts are observed without noise; see \cite{Ho.2018}.
+	to a SSU DA-MCMC that is similar in spirit to those in \cite{Gibson.1998, ONeill.1999, Fintzi.2017}.
+We do not consider particle filters \cite{King.2015} nor diffusion approximations \cite{Fintzi.2020} since these methods are not suited to situations in which do not provide a noisy distribution modeling counts; see \cite{Ho.2018}.
 	The SSU sampler that we consider here is a special case of our DA-MCMC in which $\rho = (S(0)+I(0))^{-1}$, so that the infection and removal times of a single individual are updated per iteration.
@@
 		\end{subfigure}
-	\end{center}
-	\caption{Traceplots and auto-correlation functions of the proposed DA-MCMC and the SSU DA-MCMC for $\beta$.
-		\label{fig:E6}}
+		\caption{Traceplots and auto-correlation functions of the proposed DA-MCMC and the SSU DA-MCMC for $\beta$.}
+		\label{fig:E6}
 	\end{figure}
@@
 	 \begin{table}
-	\caption{Effective sample size per second for the proposed DA-MCMC and the single site update (SSU) DA-MCMC.
-		\label{tab:E6}}
-	\begin{center}
+        \centering
         \begin{tabular}{ C{2cm}| *{2}{C{3cm}}}
@@
         \end{tabular}
-	\end{center}
+        \caption{Effective sample size per second for the proposed DA-MCMC and the single site update (SSU) DA-MCMC.}
+        \label{tab:E6}
     \end{table}
@@
 The virus, which has a fatality rate of $70\%$, was responsible for the death of almost $2000$ people in Guinea alone.
-\cite{Coltart.2017} traced back the origin of the outbreak to the Gu\'eck\'edou prefecture of Guinea at the end of November $2013$.
+	The outbreak is believed to have originated in the Gu\'eck\'edou prefecture of Guinea at the end of November $2013$ \cite{Coltart.2017}.
 Weekly infection incidence counts are available for each prefecture for the $73$ weeks between the end of December $2013$ and May $2015$.
@@
 	The Metropolis-Hastings step for the latent data proposals achieves a healthy $20.1\%$ acceptance rate, and the first 50,000 iterations of the Markov chain are discarded as a burn-in.
-Figure \ref{fig:ebola} shows the marginal posterior distributions $\gamma^{-1}$ which indicates that people remained infectious for around $9$ days on average, which is consistent with the existing literature.
+	Figure \ref{fig:ebola} shows the marginal posterior distributions $\gamma$, whose means is $0.109$.
+This indicates that people remained infectious for around $9$ days on average, which is consistent with the existing literature.
+
+	%As noted by numerous authors, conditionally on partially observed data, the parameters $\beta$ and $\gamma$ appear positively correlated.
+	%The posterior distribution of $R_0$ is unimodal, relatively symmetric, and centered around $1$.
+A basic reproduction number close to $1$ is consistent with an outbreak that lasted several months without infecting the majority of the population.
 
 	\begin{figure}
-	\begin{center}
+		\centering
 		\begin{subfigure}[b]{0.45\textwidth}
@@
 		\end{subfigure}
-	\end{center}
-	\caption{Posterior distribution and traceplot of the expected infection length ($\gamma^{-1}$) for the Gu\'eck\'edou prefecture.
-		\label{fig:ebola}}
+		\caption{Posterior distribution and traceplot of the expected infection length ($\gamma^{-1}$) for the Gu\'eck\'edou prefecture.}
+		\label{fig:ebola}
 	\end{figure}
@@
 %Existing attempts using Markov chain Monte Carlo either mix poorly or do not scale to populations of more than a few hundred individuals, while methods relying on forward simulation are limited to moderate-sized outbreaks and may suffer from degeneracy issues on a case-by-case basis depending on the amount of missing data.
-In contrast to methods that rely on approximate computation of simplifying assumptions necessitated by computational considerations such as \cite{McKinley.2018} and \cite{Fintzi.2020}, 
-our data-augmented algorithm enables exact and fast Bayesian inference, even for large outbreaks, and leverages a well-studied, transparent MCMC framework with  guarantees of uniform ergodicity.
+	In contrast to methods that rely on approximate computation of simplifying assumptions necessitated by computational considerations \cite{McKinley.2018,Fintzi.2020}, 
+	our data-augmented algorithm enables fast and exact Bayesian inference, even for large outbreaks, and leverages a well-studied, transparent MCMC framework with  guarantees of uniform ergodicity.
 
@@
 Moreover, the piece-wise constant infection rate of the PD-SIR is only updated $K$ times, as opposed to after each event in the SIR model.
-Second, generating a PD-SIR process that is constrained to be compatible with the observed incidence data can be done at no additional cost; in contrast, directly generating a SIR process compatible with the observed data would be prohibitively slow \citep{Hobolth.2009}.
+Second, generating a PD-SIR process that is constrained to be compatible with the observed incidence data can be done at no additional cost; in contrast, generating a SIR process compatible with the observed data would be prohibitively slow \cite{Hobolth.2009}.
 Third, the dynamics of the PD-SIR process closely resemble those of the SIR process: the removal dynamics are identical and, for short observation time intervals, the infection dynamics are also very similar in the two processes.
@@
 As a result, the Markov chain makes large jumps in the latent space and has very good mixing properties.
-In contrast, existing DA-MCMC algorithms \citep{Gibson.1998, ONeill.1999, Fintzi.2017} keep most of the latent space fixed across iterations  and their Markov chains and therefore mix much more slowly.
+In contrast, existing DA-MCMC algorithms keep most of the latent space fixed across iterations \cite{Gibson.1998, ONeill.1999, Fintzi.2017} and their Markov chains therefore mix much more slowly.
 
@@
 	The DA-MCMC algorithm proposed in this article is specific to 
-the stochastic SIR model.
+	the stochastic SIR model for which our algorithm is designed.
 However, one can argue that it is a simplistic representation of the spread of disease, relying on assumptions such as perfect reporting, exponentially-distributed infectious periods, homogeneously mixing population and a constant infection rate.
-Our data-augmentation framework and the idea of a faithful surrogate proposal can be extended to models of increasing realism where these assumptions are relaxed.
-It will be fruitful to consider extensions to epidemic models accounting for under-reporting \citep{Fintzi.2017,Morozova.2021}, non-Markovian dynamics \citep{Streftaris.2002}, non-homogeneous mixing \citep{Severo.1969,Lomeli.2021} as well as time-varying infection rates \citep{Kypraios.2018} in future work.
+Our data-augmentation framework and the idea of a faithful mechanistic proposal can be extended to models of increasing realism where these assumptions are relaxed.
+It will be fruitful to consider extensions to epidemic models accounting for under-reporting \cite{Fintzi.2017,morozova2021one}, non-Markovian dynamics \cite{Streftaris.2002}, non-homogeneous mixing \cite{Severo.1969,martinez2021statistical} as well as time-varying infection rates \cite{Kypraios.2018} in future work.
 We invite readers to consider applications of these ideas in other stochastic process models with complex latent spaces that may similarly be made navigable through designing tractable, mechanistic proposal processes.
@@
 
-Theorem \ref{theo:ldp} was first proved by \cite{Neuts.1971}. \cite{Ross.1996} provide a simpler proof which we now give.
+	Theorem \ref{theo:ldp} was first proved by \cite{Neuts.1971}. Ross \cite{Ross.1996} provides a simpler proof which we now give.
 
@@
 	Consider a linear pure death process with $n$ particles and individual death rate $\mu$.
-	Let $T_i$ be the time of the $i$th death.
-Then $W_1 = T_1 \sim \textsf{Exp}(n\mu)$ and $W_i = T_i - T_{i-1} \sim \textsf{Exp}((n-i)\mu)$ independently.
-Let $N$ be the number of deaths by time $t$.
-Then,
+	Let $T_i$ be the time of the $i$th death. Then $W_1 = T_1 \sim \Exp(n\mu)$ and $W_i = T_i - T_{i-1} \sim \Exp((n-i)\mu)$ independently. Let $N$ be the number of deaths by time $t$. Then, 
 	\begin{align*}
@@
 			\min\{Ga(x;a+A,b), Ga(x;a,b+B)\}, & x_a^* \wedge x_b^*\le x < x_{a+A}^* \vee x_{b+B}^*\\
-	\end{cases}.$$
+		\end{cases}$$
 	\end{lemma}
@@
 		\begin{table}[H]
-		\caption{Values of $(\alpha, \beta)$ that minimize $Ga(x;a+\alpha,b+\beta)$ for different values $x$.
-			\label{tab:ga9}}
-		\begin{center}
+			\centering
 			\begin{tabular}{l|c c c}
@@
 			\end{tabular}
-		\end{center}
+			\caption{Values of $(\alpha, \beta)$ that minimize $Ga(x;a+\alpha,b+\beta)$ for different values $x$.}
+			\label{tab:ga9}
 		\end{table}
@@
 		\end{subfigure}
-	\end{center}
-	\caption{Example of minorization of the density of a gamma distribution $Ga(2+\alpha, 0.5+\beta)$.
-		\label{fig:gam}}
+		\caption{Example of minorization of the density of a gamma distribution $Ga(2+\alpha, 0.5+\beta)$.}
+		\label{fig:gam}
 	\end{figure}
@@
 
-	First, we obtain a closed form minorization of $\pi(\beta_2| \mathbf{z}_1)$ using Proposition \ref{pro:ga1} as follows,
+		First, we can obtain a closed form minorization of $\pi(\beta_2| \mathbf{z}_1)$ using Proposition \ref{pro:ga1} as follows,
 		\begin{align*}
@@
 		since $n_I = \sum_k I_k$ is known and the sufficient statistic from $\mathbf{z}_1$, 
-	$\int_{0}^{t_{end}}S(t)I(t)dt$, is bounded between $0$ and $n (n_I+I_0)t_{end}$. Similarly, by Proposition \ref{pro:ga2}, we minorize $\pi(\gamma_2| \mathbf{z}_1)$ as follows,
+		$\int_{0}^{t_{end}}S(t)I(t)dt$, is bounded between $0$ and $n (n_I+I_0)t_{end}$. Similarly, by Proposition \ref{pro:ga2}, we can minorize $\pi(\gamma_2| \mathbf{z}_1)$ as follows,
 		\begin{align*}
@@
 		    \dfrac{q(\mathbf{z}_1|\theta_2)}{\pi(\theta_2, \mathbf{z}_1)} 
-		& = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \textsf{TrunExp}(z^I_j; \mu_k, t_{k-1}, t_k) \prod_{i=1}^n (1-p_i)^{\mathbf{1}\{z^R_i=\infty\}} \left(p_i \textsf{TrunExp}(z^R_i;\gamma, z^I_i, t_{end})\right)^{\mathbf{1}\{z^R_i \le t_{end}\}}}
+		    & = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \textsf{TruncExp}(z^I_j; \mu_k, t_{k-1}, t_k) \prod_{i=1}^n (1-p_i)^{\mathbf{1}\{z^R_i=\infty\}} \left(p_i \textsf{TruncExp}(z^R_i;\gamma, z^I_i, t_{end})\right)^{\mathbf{1}\{z^R_i \le t_{end}\}}}
         	{\beta^{n_I} \prod_{l\in \mathcal{I}} I(z^I_l) \gamma^{n_R} \exp\{-\beta \int_0^{t_{end}} S(t)I(t)dt -\gamma \int_0^{t_{end}} I(t)dt\}} \\
-		& = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \textsf{TrunExp}(z^I_j; \mu_k, t_{k-1}, t_k)}
+        	& = \dfrac{\prod_{k=1}^K \prod_{j\in \mathcal{I}_k} \textsf{TruncExp}(z^I_j; \mu_k, t_{k-1}, t_k)}
         	{\prod_{l\in \mathcal{I}} \beta I(z^I_l) \exp\{-\beta \int_0^{t_{end}} S(t)I(t)dt\}}\\
@@
         \begin{align*}
-		& \dfrac{\prod_{i=1}^n (1-p_i)^{\mathbf{1}\{z^R_i=\infty\}} (p_i \textsf{TrunExp}(z^R_i;\gamma, z^I_i, t_{end}))^{\mathbf{1}\{z^R_i \le t_{end}\}}}{\gamma^{n_R} \exp\{- \gamma \int_0^{t_end} I(t)dt\}} \\
+        	& \dfrac{\prod_{i=1}^n (1-p_i)^{\mathbf{1}\{z^R_i=\infty\}} (p_i \textsf{TruncExp}(z^R_i;\gamma, z^I_i, t_{end}))^{\mathbf{1}\{z^R_i \le t_{end}\}}}{\gamma^{n_R} \exp\{- \gamma \int_0^{t_end} I(t)dt\}} \\
         	& = \dfrac{\prod_{j \in \mathcal{R}^c} \exp\{-\gamma(t_{end}-z^I_j)\} \prod_{k \in \mathcal{R}} (1-\exp\{\gamma(t_{end}-z^I_k)\}) \dfrac{\gamma \exp\{-\gamma (z^R_k - z^I_k)\}}{1-\exp\{\gamma(t_{end}-z^I_k)\}} }{\gamma^{n_R} \exp\{ \gamma \sum_{i=1}^n \min\{z^R_i, t_{end}\} - z^I_i\}} \\
@@
 
-%\begin{figure}
-%\begin{center}
-%\includegraphics[width=3in]{fig1.pdf}
-%\end{center}
-%\caption{Consistency comparison in fitting surrogate model in the tidal
-%power example. \label{fig:first}}
-%\end{figure}
-%
-%\begin{table}
-%\caption{D-optimality values for design $X$ under five different scenarios.  \label{tab:tabone}}
-%\begin{center}
-%\begin{tabular}{rrrrr}
-%one & two & three & four & five\\\hline
-%1.23 & 3.45 & 5.00 & 1.21 & 3.41 \\
-%1.23 & 3.45 & 5.00 & 1.21 & 3.42 \\
-%1.23 & 3.45 & 5.00 & 1.21 & 3.43 \\
-%\end{tabular}
-%\end{center}
-%\end{table}
-
-%\bigskip
-%\begin{center}
-%{\large\bf SUPPLEMENTARY MATERIAL}
-%\end{center}
-%
-%\begin{description}
-%
-%\item[Title:] Brief description. (file type)
-%
-%\item[R-package for  MYNEW routine:] R-package OMYNEWO containing code to perform the diagnostic methods described in the article.
-The package also contains all datasets used as examples in the article.
-(GNU zipped tar file)
-%
-%\item[HIV data set:] Data set used in the illustration of MYNEW method in Section~ 3.2. (.txt file)
-%
-%\end{description}
-
-\bibliographystyle{Chicago}
+	% Gamma 2
+	%can be used to minorize a density of the form
+	% \begin{equation}
+	% 	\label{eq:ga}
+	% 	Ga(x;a+\alpha,b+\beta) = \frac{(b+\beta)^{a+\alpha}}{\Gamma(a+\alpha)}x^{a+A-1}\exp\{-x(b+B)\}, \quad 0\le\alpha\le A, 0\le\beta\le B.
+	% \end{equation}
+	% over $\alpha$ and $\beta$ for each value of $x$. We can show the following.
+	% \begin{lemma}	
+	% 	\label{pro:ga2}
+	% 	For a fix $x>0$, the density $Ga(x;a+\alpha,b+\beta)$ is minimized by $(\alpha, \beta) \in \{(A, 0), (0, B)\}$, the minimizing set of values depending on $x$. In particular,	
+	% 	$$\inf_{\begin{aligned}
+	% 			0\le \alpha\le A \\ 0\le \beta\le B
+	% 	\end{aligned}}Ga(x;a+\alpha,b+\beta) = 
+	% 	\begin{cases}
+	% 		Ga(x;a+A,b)                     , & x<x_a         \text{ or } x<x_{a+A}^* \vee x_b^*\\
+	% 		Ga(x;a,b+B)                     , & x_{a+A}^* < x \text{ or } x_a^* \wedge x_{b+B}^* < x\\
+	% 		\min\{Ga(x;a+A,b), Ga(x;a,b+B)\}, & x_a^* \wedge x_b^*\le x < x_{a+A}^* \vee x_{b+B}^*\\
+	% 	\end{cases}$$
+	% \end{lemma}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+	% Truncated Exponential
+%	Proof of Lemma \ref{pro:tru}
+
+%	\begin{proof}
+%		Remember that $\textsf{TruncExp}(u; \mu, 0, u) = \frac{\mu \exp\{-u\mu\}}{1-\exp\{-u\mu\}}$ is the density of an exponential distribution bounded above by $u$.
+%		\begin{align*}
+%			\frac{d}{d\mu}\textsf{TruncExp}(u; \mu, 0, u)
+%			& = \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \left[ 1-\mu u - \mu u \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \right]
+%		\end{align*}
+%		Now let $g(\mu) = 1-\mu u - \mu u \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right)$. By L'H\^{o}pital's rule, $g(0) = 0$. Moreover, for $\mu>0$, we have
+%		\begin{align*}
+%			g'(\mu)
+%			& = - \mu \left(1+\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \left[ 1-u\mu \left(\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \right]  \\
+%			& = - \mu \left(1+\frac{\exp\{-u\mu\}}{1-\exp\{-u\mu\}} \right) \left[ 1- \frac{u\mu}{\exp\{u\mu\}-1} \right] \\
+%			& \le 0
+%		\end{align*}
+%		where the inequality follows from $\frac{u\mu}{\exp\{u\mu\}-1}\le1$.
+%		This implies that $g(\mu) \le 0$ for $\mu>0$. We therefore have $\frac{d}{d\mu}\textsf{TruncExp}(u; \mu, 0, u) \le 0$ and the result of the proposition follows.
+%	\end{proof}
+
+	\bibliographystyle{plain}
 	\bibliography{bibliography}

### 2022-01-13 Manuscripts/jasa_test/JASA-template.tex
jasa minimal test

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%

\begin{document}

%\bibliographystyle{natbib}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if1\blind
{
  \title{\bf Title}
  \author{Author 1\thanks{
    The authors gratefully acknowledge \textit{please remember to list all relevant funding sources in the unblinded version}}\hspace{.2cm}\\
    Department of YYY, University of XXX\\
    and \\
    Author 2 \\
    Department of ZZZ, University of WWW}
  \maketitle
} \fi

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Title}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
The text of your abstract. 200 or fewer words.
\end{abstract}

\noindent%
{\it Keywords:}  3 to 6 keywords, that do not appear in the title
\vfill

\newpage
\spacingset{1.9} % DON'T change the spacing!
\section{Introduction}
\label{sec:intro}

Body of paper.  Margins in this document are roughly 0.75 inches all
around, letter size paper.

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig1.pdf}
\end{center}
\caption{Consistency comparison in fitting surrogate model in the tidal
power example. \label{fig:first}}
\end{figure}

\begin{table}
\caption{D-optimality values for design $X$ under five different scenarios.  \label{tab:tabone}}
\begin{center}
\begin{tabular}{rrrrr}
one & two & three & four & five\\\hline
1.23 & 3.45 & 5.00 & 1.21 & 3.41 \\
1.23 & 3.45 & 5.00 & 1.21 & 3.42 \\
1.23 & 3.45 & 5.00 & 1.21 & 3.43 \\
\end{tabular}
\end{center}
\end{table}

\begin{itemize}
\item Note that figures and tables (such as Figure~\ref{fig:first} and
Table~\ref{tab:tabone}) should appear in the paper, not at the end or
in separate files.
\item In the latex source, near the top of the file the command
\verb+\newcommand{\blind}{1}+ can be used to hide the authors and
acknowledgements, producing the required blinded version.
\item Remember that in the blind version, you should not identify authors
indirectly in the text.  That is, don't say ``In Smith et. al.  (2009) we
showed that ...''.  Instead, say ``Smith et. al. (2009) showed that ...''.
\item These points are only intended to remind you of some requirements.
Please refer to the instructions for authors
at \url{http://amstat.tandfonline.com/action/authorSubmission?journalCode=uasa20&page=instructions#.VFkk7fnF_0c}
\item For more about ASA\ style, please see \url{http://journals.taylorandfrancis.com/amstat/asa-style-guide/}
\item If you have supplementary material (e.g., software, data, technical
proofs), identify them in the section below.  In early stages of the
submission process, you may be unsure what to include as supplementary
material.  Don't worry---this is something that can be worked out at later stages.
\end{itemize}

\section{Methods}
\label{sec:meth}
Don't take any of these section titles seriously.  They're just for
illustration.

\section{Verifications}
\label{sec:verify}
This section will be just long enough to illustrate what a full page of
text looks like, for margins and spacing.

\addtolength{\textheight}{.5in}%

\cite{Campbell02}, \cite{Schubert13}, \cite{Chi81}

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
{\bf With this spacing we have 25 lines per page.}
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

%new margin nere

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

\addtolength{\textheight}{-.3in}%

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

\section{Conclusion}
\label{sec:conc}

\bigskip
\begin{center}
{\large\bf SUPPLEMENTARY MATERIAL}
\end{center}

\begin{description}

\item[Title:] Brief description. (file type)

\item[R-package for  MYNEW routine:] R-package MYNEW containing code to perform the diagnostic methods described in the article. The package also contains all datasets used as examples in the article. (GNU zipped tar file)

\item[HIV data set:] Data set used in the illustration of MYNEW method in Section~ 3.2. (.txt file)

\end{description}

\section{BibTeX}

We hope you've chosen to use BibTeX!\ If you have, please feel free to use the package natbib with any bibliography style you're comfortable with. The .bst file agsm has been included here for your convenience.

\bibliographystyle{Chicago}

\bibliography{Bibliography-MM-MC}
\end{document}
